### 2013-12-18 paper.tex
Backup of current progress.
Improved lperf, updated perfcounters for Haswell, and started outline
of paper.

\documentclass[a4paper,11pt,twocolumn,twoside]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
\usepackage{authblk}

\lstset{language=C, basicstyle=\ttfamily, columns=flexible}

\pgfplotsset{compat=1.8}

\author{Lars Kirkholt Melhus}
\author{Rune Erlend Jensen}
\affil{Norwegian University of Science and Technology}
\title{Sources of Measurement Bias on Modern CPUs}
\date{} % This is a timeless piece

\begin{document}

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
    Seemingly innocuous properties of the execution context, such as the contents of system environment variables, or variations in link ordering, can impact the performance of computer programs. 
    Variations in external properties like these can bias programs towards certain configurations. 
    These effects have been shown to be a significant issue in performance analysis, but unpredictable and difficult to deal with.

    Abstract This paper focuses on the underlying reasons for bias effects that can be experienced for example by changing environment variables, or using different link orders. 
    Through experimentation and careful measurements using performance counters, we identify several potential sources of bias on the 4th generation Intel Core i7 architecture Haswell. 
    Limitations imposed by the Loop Stream Detector is revealed, along with effects from 4K address aliasing. We show that bias is in fact not completely unpredictable, and discuss measures for avoiding it.
    \end{abstract}
    \bigskip

  \end{@twocolumnfalse}
]

\section{Introduction}
Optimizing for performance is an important topic in computer science, and high performance computing (HPC) in particular. 
A great amount of research and effort is put into designing better algorithms and compiler optimizations, in order to produce an “optimal” executable. 
However, performance of computer programs is not just a function of the collection of machine instructions executed. 
Various external properties, such as the operating system and user environment, can have have an impact on how fast programs execute. 
Characteristics of the operating system can decide things like the placement of code and data in memory, which in turn interacts with various hardware mechanisms. 
Corner cases in hardware introduces potential performance cliffs; If for example a memory access happens to cross a page boundary, the cost of an additional TLB miss can be significant. 
The set of conditions under which a program executes is called the execution context, which includes virtual memory layout in particular. 
Variations within contexts can lead to programs being biased towards certain configurations. 
Measuring performance of the same program on two machines with identical hardware can sometimes yield very different results, an effect known as measurement bias.

Previous work has studied parameters such as the size of Unix environment variables and program link order. 
Both are found to potentially have a significant effect on performance, even in standardized benchmarks. 
Unfortunately, the effects also appear to be unpredictable, and therefore difficult to deal with. 
This poses a problem for researchers and performance analysts, who will need to account for effects of measurement bias with more rigorous methodologies and statistical methods.

%\pgfplotstableread{resources/motivation.dat}{\motivationtable}
%\pgfplotstabletypeset[string type]{\motivationtable}

%\begin{figure}[t]
%  \caption{Ticz figure}
%  \begin{tikzpicture}
    %\draw circle (5);
%    \begin{axis}[
        %width = \columnwidth,
        %scale mode = stretch to fill,
%        font = \small,
        %scaled ticks = false,
%        xlabel=Characters added to environment,
        %ymin = 600000, ymax = 900000,
%        grid = major,
 %       cycle list name = exotic
 %%     ]
 %     \addplot[smooth] table[y = cycles:u] \motivationtable ;
%      \addlegendentry{cycles:u} ;
%    \end{axis}
%  \end{tikzpicture}
%\end{figure}

In this paper, we unveil some of the actual causes of bias that can be experienced by altering memory layout.
Both the size of environment variables and link ordering ultimately has an effect on memory layout of running processes. 
We will not look at things like cache efficiency, which often can be the reason for bad performance under certain memory contexts. 
Instead, we look at effects from less known optimizations in the out-of-order execution engine and instructions fetch pipeline. 
Because bias effects are intrinsically connected to various hardware features, this paper is focued on the Intel® Core™ i7 “Haswell” architecture specifically. 
However, the bias sources we present has also been shown to behave similarly on previous Intel architectures as well.
With a better understanding of intricate properties of the CPU, we show how one is able to not only predict, but also explicitly optimize for beneficial memory layouts.

In the paper appropriately named “Producing Wrong Data Without Doing Anything Obviously Wrong” by T. Mytkowicz et al. \cite{Mytkowicz:2009:WrongData}, the authors show how simple changes to the execution environment can have an impact on performance. 
One of their examples is a program where performance varies significantly with changes to the environment variables. 
Unix environment variables contain various information about the system, such as the name of the currently logged in user, home- and current directory path.
This means that the length of your name -- by extension you user login -- could in principle be the tipping point between good and bad performance.

Previous work focuses mostly on how to mitigate effects of bias, and avoid drawing the wrong conclusions based on misleading measurements \cite{Mytkowicz:2008:Easy, Mytkowicz:2009:WrongData}.
The goal of this paper is to show not only that bias exists, but also \emph{why} it occurs.

With more accurate models of how bias occur, we will not only be able to actively avoid in performance, but in some cases also gain a real speedup on average.

Modern microprocessors are extremely complex in design and functionality. 
Some features of recent Intel processors includes several layers of cache to camouflage slow memory, multiple prefetchers, speculative out-of-order execution and branch prediction, just to name a few.
Hardware features and optimizations interact with memory layout of program code and data in various ways.
We will unveil characteristics about two different architectural features in recent Intel CPUs, and show how they can bias performance towards certain memory contexts.

First we will look at an aliasing effect between memory addresses of loads and stores, causing false dependencies in the out-of-order execution pipeline. 
This effect can be triggered for example by changing stack position, and can explain bias from altering environment variables. 
Secondly, we will look at the Loop Stream Detector, which in previous work on measurement bias has been suggested as a possible explanation.
We therefore choose to study this particular optimization in detail, and show how measurement bias can occur from changing link order.

\section{Methodology}
When studying bias effects, carefully controlling all relevant properties of the environment is crucial. 
Previous work on measurement bias and observer effect describes a set of best practices for how the measurement infrastructure should be configured \cite{Mytkowicz:2009:WrongData, Mytkowicz:2008:OE&MB}. 
\begin{itemize}
  \item Address space layout randomization (ASLR) is disabled
  \item Hyper-threading is disabled
  \item System load is kept at a minimum
  \item Automatic CPU frequency scaling is disabled
\end{itemize}
Our hardware configuration consists of an Intel® Core™ i7-4770K @ 3.50 GHz and 32 GB DDR3 @ 1600 MHz, running a 64-bit installation of Ubuntu 13.10. GCC version 4.8.1 is used as compiler toolchain.

We use perf to acquire statistics and measurements.
Perf is a utility to interface with hardware performance counters, and available as part of the Linux kernel.
Measurements are aggregated over several runs to collect all of the about 200 performance counters provided on our architecture [Intel:2013:Volume3B]. 
Unless otherwise specified, a minimal environment is used for all benchmarks. 
To measure statistics with n bytes added to the environment, a dummy variable is set to \(0^{n}\) (repeated zero characters n times).

Cache conflicts can in many cases be a potential explanation for bias effects observed by changing memory layout.
Although cache is often a likely explanation of performance variations, bias caused by cache issues is well understood and \emph{not} the focus of this paper.
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias. 
The hit rates of micro-ops for each level of cache can be monitored by performance counters \cite{Intel:2012:OptimizationManual}.

\section{4K Address Aliasing}
An event known as “4K aliasing” can occur when the addresses of a store instruction followed by a load instruction differ by a multiple of 4 KiB.
On recent Intel architectures, only the last 12 address bits are used to determine if operations refer to the same address \cite{Intel:2012:OptimizationManual}.
This is a heuristic employed by the CPU to enable out of order execution of memory access operations.
Independent pairs of loads and stores where the last address bits differ can then be reordered and issued out of order.
Doing only a partial compare, \emph{false} dependencies can be detected when the addresses are equal $\bmod 4096$, as in the following example: 
\begin{lstlisting}[language={[x86masm]Assembler}]
mov %rax ($0x601010)
mov ($0x604010) %rbx
\end{lstlisting}
Situations like these causes the pipeline to stall and wait for the operations to finish in proper order.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\small{Evnt Mask \\ 0x07 0x01}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.
\end{description}
Performance implications of 4K aliasing is discussed to some extent in the optimization manual from Intel. 
There are even some concrete suggestions for how avoid aliasing cases to improve performance (Rule 8 and 9) \cite{Intel:2012:OptimizationManual}.
In the following sections, we look at how address aliasing can explain some instances of measurement bias.

\subsection{Bias from Environment Size}
Variations in performance with bias towards certain environment sizes has been observed by other authors.
Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack. 
Figure \ref{fig:stack} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory.

\pgfplotstableread{resources/stack-offset.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \small,
        xlabel=Bytes added to environment,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        legend pos = outer north east,
        cycle list name = exotic
      ]
      \draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

The stack is normally placed close to the upper address 0x7fff'ffffffff.\footnote{Modern processors do not actually use the full 64 bit space, only the low order 54 bits are used for addressing memory}
Environment variables are allocated before the stack, offsetting all call frames and addresses of stack allocated variables at run time.
There is still some level of stack alignment happening independent of environment offsets. For example, the compiler will often enforce alignment to some number of zero bits.
As long as this number is less than 12, environment size can impact addresses of automatic variables $\bmod 4096$.

Text and data sections contain program code and static data respectively.
These regions are mapped towards the low en of the address spectrum, and the virtual addresses of variables allocated here will not change with environment size.

Note that the initial position of stack will not depend on environment size this way when address randomization (ASLR) is enabled.
\begin{figure}
  \begin{tikzpicture}
    \draw [fill=white, thick] (0, 0) rectangle (2.6, 6) ;
    \draw [fill=white, thin] (0.1, 5.3) rectangle (2.5, 5.9) ;
    \draw [fill=white, thin] (0.1, 4.0) rectangle (2.5, 5.0) ;
    \draw [fill=white, thin] (0.1, 1.6) rectangle (2.5, 2.5) ;
    \draw [fill=white, thin] (0.1, 0.6) rectangle (2.5, 1.5) ;
    \draw [fill=white, thin] (0.1, 0.1) rectangle (2.5, 0.5) ;

    \node [below right] at (2.6, 6.0) {0x7fff'ffffffff} ;
    \node [above right] at (2.6, 0.5) {0x400000} ;
    \node [above right] at (2.6, 0.0) {0x0} ;

    \node [below right] at (0.1, 5.9) {environment} ;
    \node [below right] at (0.1, 5.0) {stack} ;
    \node [below right] at (0.1, 2.5) {data} ;
    \node [below right] at (0.1, 1.5) {text} ;

    \draw [<->, line width=1, color=black] (1.3, 5.3) -- (1.3, 5.0) ;

  \end{tikzpicture}
  \caption{Stack position is offset by size of environment variables}
  \label{fig:stack}
\end{figure}

Addresses of stack allocated variables can vary depending on environment size, while other segments of memory such as text and data remain constant over all environments.

To illustrate how this can cause bias, we revisit the example first presented in \cite{Mytkowicz:2009:WrongData}.
Their program contains four variables, two of which are stack allocated and i, j and k that are statically allocated.

\begin{lstlisting}[language=C]
static int i, j, k;
int main() {
  int g = 0, inc = 1;
  for(; g < 65536; g++) {
    i += inc;
    j += inc;
    k += inc; 
  }
  return 0;
}
\end{lstlisting}

Measuring all available performance counters for this program under different environment sizes.

\begin{small}
\begin{table}
  \caption{Properties of the LSD}
  \begin{tabular}{@{} l r @{}}
    \toprule
      UnHalted Core Cycles & 1 \\
      LD\_BLOCKS\_PARTIAL.\_ADDRESS\_ALIAS & 0.89 \\
      CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING & 0.97 \\
      RESOURCE\_STALLS.ANY & 0.59 \\
    \bottomrule
  \end{tabular}
\end{table}
\end{small}

\subsection{Heap Address Aliasing}
As an example of a “sliding window” program working on pairs of memory buffers, consider the function shown in Listing [alg:convolution]. 
It computes the convolution between an input array and a fixed kernel, writing the result to another array (endpoints skipped for simplicity).

\begin{lstlisting}[language=C]
static float k[5]={0.1,0.25,0.3,0.25,0.1};
void convolve(int size,float *in,float *out)
{
  int i, j;
  for (i=2;i<size-2; ++i) {
    out[i] = 0;
    for (j=0;j<5; ++j)
      out[i]+=in[i-2+j]*k[j];}
}
int main() {
  float *in=malloc(N*sizeof(float));
  float *out=malloc(N*sizeof(float));
  convolve(N, in, out);
  return 0;
}
\end{lstlisting}

\section{The Loop Stream Detector}
Typical software spends most of the time executing the same instructions repeatedly in loops, where the same instructions are fetched, decoded and executed many times. 
The Loop Stream Detector (LSD) is a front-end hardware optimization that is able to detect small software loops.
A cache of decoded instructions is kept and analyzed for cyclic chains in the instruction stream.
When a loop is detected, can be fetched directly from a queue of already decoded micro operations.
This enables some instruction decode logic in the processor front end to be temporarily disabled.
The LSD is a power saving optimization, but can also improve performance in cases where the decode pipeline is a bottleneck.

Utilization of the Loop Stream Detector can be measured by the following performance counter.
\begin{description}
  \item{LSD.UOPS}\marginpar{\small{Evnt Mask \\ 0xA8 0x01}} \hfill \\
  Counts the number of micro-operations delivered by the LSD.
\end{description}

The LSD has been suggested in previous work as a potential source of measurement bias [Mytkowicz:2009:WrongData].
The Loop Stream Detector is dependent on code layout, which can change with external factors such as variations in link ordering. 
Interacting with the alignment and position of instructions in memory, the LSD can cause performance differences for programs with otherwise \emph{identical} instruction sequences.

Some effort has been made to model properties of the LSD on previous Intel architectures, as an aid in low-level optimization. 
Hundt et. al. \cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
For loops to be recognized, they have to fit within a set of restrictions imposed by the hardware implementation that tries to detect them. 
For instance, because the LSD cache is of fixed size there will be a maximum number of instructions a loop can contain for it to be recognized.
The capabilities of the LSD has improved with newer micro-architectures.
Some important limitations for Haswell's LSD are listed below.
\begin{itemize}
  \item Up to 12 chunk fetches of 32 B instruction memory.\footnote{The optimization manual actually states 8 chunk fetches for HTT enabled, and 11 for HTT disabled. Our experiments with a minimal test case shows that the LSD is active for loops spanning up to 12 chunks -- regardless of hyper threading.}
  \item Up to 28 micro-operations for HTT enabled, 56 for HTT disabled.
  \item Can not contain CALL or RET instructions
  \item Stack operations must match, i.e push must be matched with a pop.
\end{itemize}

\subsection{Bias from Link Order}
Different link orders affect the relative position of code and static data within the final compiled binary file.
More importantly, \emph{alignment} of code can be affected by link order.
A loop that was aligned to 32 byte for one ordering might end up at 16 byte alignment in another. 
Alignment is important for determining the number of 32 byte chunks a loop is spanning.

\subsection{Other Triggers of Bias Effects}
The focus of previous work, and elaborated in this section, is measurement bias caused by altering link ordering.
The important factor is not the process of linking itself, but the effect it has on final code layout in the compiled binary.

\paragraph{Order of functions within a file}
As with link order, the order of functions within a source file usually determines their relative position in the text segment of the compiled binary. 
Given a program with functions foo and bar, listing foo before bar and vice versa generates two “different” programs with respect to memory layout. 
GCC prints functions to the text segment in the order they appear in the source code.

\paragraph{Length of external symbols}
Symbols such as function names are placed in a symbol table section in the ELF file.
Symbols that needs to be resolved dynamically are allocable, and mapped to virtual address space before the text segment. 
This means that names of external symbols resolved at run-time does affect code layout. 
Longer function names (or more external dependencies) will offset the text segment because the size of allocable segments related to dynamic linking changes. 
his includes the “dynsym”, “dynstr”, and “rela” sections in an executable made with GCC.

Some particularly devious “bugs” can occur from this. 
Consider a scenario where a program does not align properly to utilize the Loop Stream Detector. 
Assume a printf statement is added to a completely unrelated part of the code. 
The additional external symbol offsets instruction alignment such that the LSD can be used, giving a huge speedup.
Later removing the printf statement will reduce the size of external dependencies, and again result in worse performance.

Any interaction that affects code layout can potentially trigger performance cliffs caused by the Loop Stream Detector.

\section{Conclusions}

\onecolumn

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2013-12-20 paper.tex
Backup

@@
 \usepackage{authblk}
+
+\usetikzlibrary{patterns,shapes,positioning,calc}
 
@@
 
-%\pgfplotstableread{resources/motivation.dat}{\motivationtable}
-%\pgfplotstabletypeset[string type]{\motivationtable}
-
-%\begin{figure}[t]
-%  \caption{Ticz figure}
-%  \begin{tikzpicture}
-    %\draw circle (5);
-%    \begin{axis}[
-        %width = \columnwidth,
-        %scale mode = stretch to fill,
-%        font = \small,
-        %scaled ticks = false,
-%        xlabel=Characters added to environment,
-        %ymin = 600000, ymax = 900000,
-%        grid = major,
- %       cycle list name = exotic
- %%     ]
- %     \addplot[smooth] table[y = cycles:u] \motivationtable ;
-%      \addlegendentry{cycles:u} ;
-%    \end{axis}
-%  \end{tikzpicture}
-%\end{figure}
-
-In this paper, we unveil some of the actual causes of bias that can be experienced by altering memory layout.
 Both the size of environment variables and link ordering ultimately has an effect on memory layout of running processes. 
@@
       ]
-      \draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
+      %\draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
       \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
@@
 
-The stack is normally placed close to the upper address 0x7fff'ffffffff.\footnote{Modern processors do not actually use the full 64 bit space, only the low order 54 bits are used for addressing memory}
-Environment variables are allocated before the stack, offsetting all call frames and addresses of stack allocated variables at run time.
-There is still some level of stack alignment happening independent of environment offsets. For example, the compiler will often enforce alignment to some number of zero bits.
+The stack is normally placed close to the upper address 0x7fff'ffffffff.\footnote{Modern processors do not actually use the full 64 bit space, only the low order 47 bits are used for addressing memory}
+Environment variables and program arguments are allocated in the stack section, before the first call frame.
+Changing environment variables will therefore offset the addresses of all stack allocated data.
+There is still some level of stack alignment happening, independent of environment offsets. 
+The compiler can enforce alignment to some number of zero bits.
 As long as this number is less than 12, environment size can impact addresses of automatic variables $\bmod 4096$.
@@
 Note that the initial position of stack will not depend on environment size this way when address randomization (ASLR) is enabled.
+(TODO: Does it offset from page boundary?
+That would be good, same alias behavior :)
+
 \begin{figure}
-  \begin{tikzpicture}
-    \draw [fill=white, thick] (0, 0) rectangle (2.6, 6) ;
-    \draw [fill=white, thin] (0.1, 5.3) rectangle (2.5, 5.9) ;
-    \draw [fill=white, thin] (0.1, 4.0) rectangle (2.5, 5.0) ;
-    \draw [fill=white, thin] (0.1, 1.6) rectangle (2.5, 2.5) ;
-    \draw [fill=white, thin] (0.1, 0.6) rectangle (2.5, 1.5) ;
-    \draw [fill=white, thin] (0.1, 0.1) rectangle (2.5, 0.5) ;
+  \begin{tikzpicture}[font=\footnotesize]
 
-    \node [below right] at (2.6, 6.0) {0x7fff'ffffffff} ;
-    \node [above right] at (2.6, 0.5) {0x400000} ;
-    \node [above right] at (2.6, 0.0) {0x0} ;
+    % See page 453 in pgf manual
+    \node [
+      rectangle split, rectangle split parts=9, 
+      rectangle split part fill={white, lightgray, white, lightgray, white, lightgray, white},
+      draw, anchor=center, text width=2cm
+    ] (m)  
+      {
+        \nodepart{one}
+          stack
+        \nodepart{three}
+          mmap area
+        \nodepart{five}
+          heap
+        \nodepart{seven}
+          bss
+        \nodepart{eight}
+          data
+        \nodepart{nine}
+          text
+      };
 
-    \node [below right] at (0.1, 5.9) {environment} ;
-    \node [below right] at (0.1, 5.0) {stack} ;
-    \node [below right] at (0.1, 2.5) {data} ;
-    \node [below right] at (0.1, 1.5) {text} ;
+    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.six split west) 
+      node [black, midway, xshift=-1.5cm, text width=2.0cm] 
+        { Program code and static data } ;
 
-    \draw [<->, line width=1, color=black] (1.3, 5.3) -- (1.3, 5.0) ;
+    \node[below right] at (m.north east)
+      { \scriptsize{0x7fff'ffffffff} } ;
+
+    \node[above right] at (m.south east)
+      { \scriptsize{0x400000} } ;
 
   \end{tikzpicture}
-  \caption{Stack position is offset by size of environment variables}
+  \caption{Important sections of virtual memory on a 64 bit process}
   \label{fig:stack}
@@
 To illustrate how this can cause bias, we revisit the example first presented in \cite{Mytkowicz:2009:WrongData}.
-Their program contains four variables, two of which are stack allocated and i, j and k that are statically allocated.
+Their program contains five variables, two of which are stack allocated and i, j and k that are statically allocated.
 
@@
 
-\begin{small}
-\begin{table}
-  \caption{Properties of the LSD}
-  \begin{tabular}{@{} l r @{}}
-    \toprule
-      UnHalted Core Cycles & 1 \\
-      LD\_BLOCKS\_PARTIAL.\_ADDRESS\_ALIAS & 0.89 \\
-      CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING & 0.97 \\
-      RESOURCE\_STALLS.ANY & 0.59 \\
-    \bottomrule
-  \end{tabular}
-\end{table}
-\end{small}
+\subsection{Heap Address Aliasing}
+Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
+The previous section looked specifically at an example of collision between stack variables and static data.
+Another scenario to consider is collisions in heap allocated memory.
+In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
+Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
+A typical malloc implementation uses two different strategies for how to allocate memory.
+The \emph{heap} area in virtual memory is used for smaller allocations.
+Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
+For larger allocations, it is typically faster to call mmap instead of increasing the ``normal'' heap.
+The heap starts at a low address close to the data section and grows towards higher addresses. 
+Mmap allocates much closer to the stack.
+While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
+This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
+However, mmap has an interesting property in that allocations will always be page aligned.
+The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{Libc's version of malloc adds 16 bytes of medatada at the beginning, thus every mmap'ed address ends with 0x010}
 
-\subsection{Heap Address Aliasing}
-As an example of a “sliding window” program working on pairs of memory buffers, consider the function shown in Listing [alg:convolution]. 
-It computes the convolution between an input array and a fixed kernel, writing the result to another array (endpoints skipped for simplicity).
+Consider the function shown in Listing [alg:convolution].
+It computes the convolution between an input array and a fixed five-element kernel, writing the result to another array.
+For simplicity, endpoints are not handled.
 
@@
 static float k[5]={0.1,0.25,0.3,0.25,0.1};
-void convolve(int size,float *in,float *out)
+void conv(int n, float *in, float *out)
 {
   int i, j;
-  for (i=2;i<size-2; ++i) {
+  for (i = 2; i < n-2; ++i) {
     out[i] = 0;
-    for (j=0;j<5; ++j)
-      out[i]+=in[i-2+j]*k[j];}
+    for (j = 0; j < 5; ++j)
+      out[i] += in[i-2+j] * k[j];
   }
-int main() {
-  float *in=malloc(N*sizeof(float));
-  float *out=malloc(N*sizeof(float));
-  convolve(N, in, out);
-  return 0;
 }
 \end{lstlisting}
+
+\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
+\begin{figure}[t]
+  \caption{Offseting arrays}
+  \begin{tikzpicture}
+    \begin{axis}[
+        font = \small,
+        xlabel=Number of float (4B) offset,
+        cycle list name=black white
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convolutiontable ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convolutiontable ;
+      \addlegendentry{cycles:u} ;
+      \addlegendentry{r0107:u} ;
+    \end{axis}
+  \end{tikzpicture}
+\end{figure}
 

### 2013-12-21 paper.tex
Backup

@@
 \usepackage{authblk}
+\usepackage{caption} % For margin top bottom
 
@@
 \affil{Norwegian University of Science and Technology}
-\title{Sources of Measurement Bias on Modern CPUs}
+\title{Sources of Measurement Bias in Modern CPUs}
 \date{} % This is a timeless piece
@@
 In the paper appropriately named “Producing Wrong Data Without Doing Anything Obviously Wrong” by T.
-Mytkowicz et al.
-\cite{Mytkowicz:2009:WrongData}, the authors show how simple changes to the execution environment can have an impact on performance.
+Mytkowicz et al.~\cite{Mytkowicz:2009:WrongData}, the authors show how simple changes to the execution environment can have an impact on performance.
 One of their examples is a program where performance varies significantly with changes to the environment variables. 
@@
 
-Previous work focuses mostly on how to mitigate effects of bias, and avoid drawing the wrong conclusions based on misleading measurements \cite{Mytkowicz:2008:Easy, Mytkowicz:2009:WrongData}.
+Previous work focuses mostly on how to mitigate effects of bias, and avoid drawing the wrong conclusions based on misleading measurements~\cite{Mytkowicz:2008:Easy, Mytkowicz:2009:WrongData}.
 The goal of this paper is to show not only that bias exists, but also \emph{why} it occurs.
@@
 Cache related metrics are monitored in order to rule out cache as the underlying cause of bias. 
-The hit rates of micro-ops for each level of cache can be monitored by performance counters \cite{Intel:2012:OptimizationManual}.
+The hit rates of micro-ops for each level of cache can be monitored by performance counters~\cite{Intel:2012:OptimizationManual}.
 
@@
 Performance implications of 4K aliasing is discussed to some extent in the optimization manual from Intel. 
-There are even some concrete suggestions for how avoid aliasing cases to improve performance (Rule 8 and 9) \cite{Intel:2012:OptimizationManual}.
+There are even some concrete suggestions for how avoid aliasing cases to improve performance (Rule 8 and 9)~\cite{Intel:2012:OptimizationManual}.
 In the following sections, we look at how address aliasing can explain some instances of measurement bias.
@@
 Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack. 
-Figure \ref{fig:stack} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory.
+Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory.
 
@@
 \begin{figure*}[t]
-  \caption{Bias from environment size}
+  \caption{Bias from environment size for program in Listing~\ref{lst:loopkernel}.
+Measure of average 100 cycle counts for 512 different environments, spread over two 4K periods.
+Compiled with no optimization.}
+  \label{fig:envbias}
   \begin{tikzpicture}
@@
 
-Text and data sections contain program code and static data respectively.
-These regions are mapped towards the low en of the address spectrum, and the virtual addresses of variables allocated here will not change with environment size.
-
-Note that the initial position of stack will not depend on environment size this way when address randomization (ASLR) is enabled.
-(TODO: Does it offset from page boundary?
-That would be good, same alias behavior :)
-
-\begin{figure}
+\begin{figure}[float=h]
+  \caption{Important sections of virtual memory on a 64 bit process}
+  \label{fig:virtualmemory}
   \begin{tikzpicture}[font=\footnotesize]
@@
   \end{tikzpicture}
-  \caption{Important sections of virtual memory on a 64 bit process}
-  \label{fig:stack}
 \end{figure}
+
+Text and data sections contain program code and static data respectively.
+These regions are mapped towards the low en of the address spectrum, and the virtual addresses of variables allocated here will not change with environment size.
+
+Note that the initial position of stack will not depend on environment size this way when address randomization (ASLR) is enabled.
+(TODO: Does it offset from page boundary?
+That would be good, same alias behavior :)
 
@@
 
-To illustrate how this can cause bias, we revisit the example first presented in \cite{Mytkowicz:2009:WrongData}.
-Their program contains five variables, two of which are stack allocated and i, j and k that are statically allocated.
+To illustrate how this can cause bias, we revisit the example first presented in~\cite{Mytkowicz:2009:WrongData}, shown in Listing~\ref{lst:loopkernel}. 
+This example is interesting for several reasons. 
+The bias effects are significant and easily reproducible, while the example code is simple and easy to analyze.
+Still, no satisfactory explanation as to what causes bias was given in the original paper.
 
-\begin{lstlisting}[language=C]
+\begin{lstlisting}[float=h, language=C, caption={Microkernel succeptible to aliasing between static variables and automatic variables for certain environment sizes}, label={lst:loopkernel}, frame=lines]
 static int i, j, k;
@@
     int g = 0, inc = 1;
-  for(; g < 65536; g++) {
+    for (; g < 65536; g++) {
         i += inc;
@@
 
-Measuring all available performance counters for this program under different environment sizes.
+Their program contains five variables, two of which are stack allocated and i, j and k that are statically allocated.
+Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
+Sampling is done for every 16 byte increment of environment size, covering two 4K periods of addresses.
+A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
+The program is compiled with GCC using no optimization.
+Any optimization would likely disregard most of the function as redundant code, and reduce it to return zero immediately.
+
+There are clearly two worst cases, indicated by significant spikes around the 3100 and 7100 byte offsets.
+We analyzed these cases by sampling an extensive set of performance counters and correlating them with cycle counts.
+The results show a high number of resource stalls when the spikes occur, indicated by high values for performance counters RESOURCE\_STALLS.ANY, CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING, CYCLE\_ACTIVITY.CYCLE\_NO\_EXECUTE and RESOURCE\_STALLS.RS.
+We also find that LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS correlates almost perfectly with cycle count, which can explain the high number of resource stalls with false dependency from aliasing.
+
+A small amount of assembly code was used to calculate the addresses of each stack allocated variable and output them directly using system calls, without impacting run time addresses.
+We find that the spike in cycle count occurs when the address of g and inc alias with i and j, respectively.
+Resource stalls are generated because of false dependencies between the addresses of (g, i) and (inc, j), resulting in worse performance.
+Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment.
+The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.
 
@@
 Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
-For larger allocations, it is typically faster to call mmap instead of increasing the ``normal'' heap.
+For larger allocations, it is typically faster to use anonymous memory mapping by calling mmap, instead of growing the ``normal'' heap.
 The heap starts at a low address close to the data section and grows towards higher addresses. 
@@
 However, mmap has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{Libc's version of malloc adds 16 bytes of medatada at the beginning, thus every mmap'ed address ends with 0x010}
+The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, thus every mmap'ed address ends with 0x010}
 
-Consider the function shown in Listing [alg:convolution].
+Consider the function shown in Listing~\ref{lst:conv}.
 It computes the convolution between an input array and a fixed five-element kernel, writing the result to another array.
@@
 
-\begin{lstlisting}[language=C]
+\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
+\begin{figure}[t]
+  \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}.
+Offset 0 means equal 12 bit address suffix, which default behavior for mmap allocations.
+Array size $n=2^{20}$, compiled with -O3.
+}
+  \label{fig:heapalias}
+  \begin{tikzpicture}
+    \begin{axis}[
+        font = \small,
+        xlabel=Offset in \lstinline!sizeof(float)! bytes,
+        cycle list name=black white
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convolutiontable ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convolutiontable ;
+      \addlegendentry{cycles:u} ;
+      \addlegendentry{r0107:u} ;
+    \end{axis}
+  \end{tikzpicture}
+\end{figure}
+
+\begin{lstlisting}[float=h, language=C, caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}, label={lst:conv}, frame=lines]
 static float k[5]={0.1,0.25,0.3,0.25,0.1};
@@
 
-\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
-\begin{figure}[t]
-  \caption{Offseting arrays}
-  \begin{tikzpicture}
-    \begin{axis}[
-        font = \small,
-        xlabel=Number of float (4B) offset,
-        cycle list name=black white
-      ]
-      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convolutiontable ;
-      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convolutiontable ;
-      \addlegendentry{cycles:u} ;
-      \addlegendentry{r0107:u} ;
-    \end{axis}
-  \end{tikzpicture}
-\end{figure}
+Figure \ref{fig:heapalias} shows ...
+
+Using malloc to allocate input and output float arrays of size n, they will alias if mmap is used internally.
+For sufficiently large values of n, this will always be the case.
+Compiling with optimization O3 and n=$2^{20}$, a very large number of alias events and resource stalls are generated compared to cycle count.
 
@@
 Some effort has been made to model properties of the LSD on previous Intel architectures, as an aid in low-level optimization. 
-Hundt et. al. \cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
+Hundt et. al.~\cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
 For loops to be recognized, they have to fit within a set of restrictions imposed by the hardware implementation that tries to detect them. 
@@
 
-\onecolumn
+%\onecolumn
 

### 2014-01-01 paper.tex
Updated perf counters and content

@@
-\documentclass[a4paper,11pt,twocolumn,twoside]{article}
+\documentclass[a4paper,10pt,twocolumn,twoside]{article}
 \usepackage{fullpage}
@@
 \begin{description}
-  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\small{Evnt Mask \\ 0x07 0x01}} \hfill \\
+  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
   Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.
@@
 
-\pgfplotstableread{resources/stack-offset.dat}{\stackoffsettable}
+\pgfplotstableread{bin/stack-offset.dat}{\stackoffsettable}
 \begin{figure*}[t]
@@
     \node [
-      rectangle split, rectangle split parts=9, 
-      rectangle split part fill={white, lightgray, white, lightgray, white, lightgray, white},
+      rectangle split, rectangle split parts=10, 
+      rectangle split part fill={white, white, lightgray, white, lightgray, white, lightgray, white},
       draw, anchor=center, text width=2cm
@@
         \nodepart{one}
+          environment
+        \nodepart{two}
           stack
-        \nodepart{three}
+        \nodepart{four}
           mmap area
-        \nodepart{five}
+        \nodepart{six}
           heap
-        \nodepart{seven}
+        \nodepart{eight}
           bss
-        \nodepart{eight}
+        \nodepart{nine}
           data
-        \nodepart{nine}
+        \nodepart{ten}
           text
@@
 
-    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.six split west) 
+    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.seven split west) 
       node [black, midway, xshift=-1.5cm, text width=2.0cm] 
@@
 
-    \node[below right] at (m.north east)
+    \node[right] at (m.north east)
       { \scriptsize{0x7fff'ffffffff} } ;
 
-    \node[above right] at (m.south east)
+    \node[right] at (m.south east)
       { \scriptsize{0x400000} } ;
@@
 We analyzed these cases by sampling an extensive set of performance counters and correlating them with cycle counts.
-The results show a high number of resource stalls when the spikes occur, indicated by high values for performance counters RESOURCE\_STALLS.ANY, CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING, CYCLE\_ACTIVITY.CYCLE\_NO\_EXECUTE and RESOURCE\_STALLS.RS.
-We also find that LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS correlates almost perfectly with cycle count, which can explain the high number of resource stalls with false dependency from aliasing.
+The results show a high number of resource stalls when the spikes occur, indicated by high values for performance counters \small{RESOURCE\_STALLS.ANY}, \small{CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING}, \small{CYCLE\_ACTIVITY.CYCLE\_NO\_EXECUTE} and \small{RESOURCE\_STALLS.RS}.
+We also find that \small{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS} correlates almost perfectly with cycle count, which can explain the high number of resource stalls with false dependency from aliasing.
 
@@
 However, mmap has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, thus every mmap'ed address ends with 0x010}
+The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}
 
@@
   \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}.
-Offset 0 means equal 12 bit address suffix, which default behavior for mmap allocations.
+Offset 0 means equal 12 bit address suffix, which is default behavior for mmap allocations.
 Array size $n=2^{20}$, compiled with -O3.
@@
         font = \small,
-        xlabel=Offset in \lstinline!sizeof(float)! bytes,
+        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
         cycle list name=black white
@@
 static float k[5]={0.1,0.25,0.3,0.25,0.1};
-void conv(int n, float *in, float *out)
-{
+void conv(int n, float *in, float *out) {
     int i, j;
@@
 
-Figure \ref{fig:heapalias} shows ...
+Figure \ref{fig:heapalias} shows how this function behaves when given input with different alignment.
+The default behavior for large heap allocations is that both arrays alias.
+To force different alignments, we allocate some extra bytes at the end and offset the output pointer argument.
 
-Using malloc to allocate input and output float arrays of size n, they will alias if mmap is used internally.
-For sufficiently large values of n, this will always be the case.
-Compiling with optimization O3 and n=$2^{20}$, a very large number of alias events and resource stalls are generated compared to cycle count.
+\begin{lstlisting}
+output = malloc(n * (sizeof(float) + offset));
+conv(n, input, output + offset);
+\end{lstlisting}
+
+Listing \ref{lst:conv}, together with a small main method handling the allocation and offset, is compiled with optimization O3 and input size n=$2^{20}$.
+Figure \ref{fig:heapalias} plots performance counter statistics for cycle count and 4K alias events over increasing amount of offset.
+This is an example of how false dependencies from address aliasing can cause a significant performance hit.
+Because large heap allocations always alias, the processor thinks memory accesses to input[i] potentially conflicts with output[i], stalling the pipeline.
+By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\% (verify).
+This is a speedup on top of compiler optimization O3, which is not able to account for aliasing effects.
+
+It is perhaps unfair to assume the compiler can fix this when only given a generic optimization flag, as aliasing effects is (presumably) an Intel-specific issue.
+You might also wamt to utilize the C99 keyword restrict to explicitly tell the compiler that the buffers are disjoint in memory.
+
+\begin{lstlisting}[language=C]
+void conv( int n,
+            const float * restrict in,
+            const float * restrict out );
+\end{lstlisting}
+
+Even with the updated signature compiled with -O3 -march=native, there is still room for improvement by inserting manual offsets.
 
@@
 The LSD is a power saving optimization, but can also improve performance in cases where the decode pipeline is a bottleneck.
-
 Utilization of the Loop Stream Detector can be measured by the following performance counter.
 \begin{description}
-  \item{LSD.UOPS}\marginpar{\small{Evnt Mask \\ 0xA8 0x01}} \hfill \\
-  Counts the number of micro-operations delivered by the LSD.
+  \item{LSD.UOPS\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0xA8 & 0x01\end{tabular}}} \hfill \\
+  Counts the number of micro-operations delivered by the Loop Stream Detector.
 \end{description}
@@
 Some effort has been made to model properties of the LSD on previous Intel architectures, as an aid in low-level optimization. 
-Hundt et. al.~\cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
-For loops to be recognized, they have to fit within a set of restrictions imposed by the hardware implementation that tries to detect them. 
+Hundt et al.~\cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
+For loops to be recognized, they have to match a set of restrictions imposed by the hardware implementation that tries to detect them. 
 For instance, because the LSD cache is of fixed size there will be a maximum number of instructions a loop can contain for it to be recognized.
@@
 Alignment is important for determining the number of 32 byte chunks a loop is spanning.
+
+\begin{lstlisting}[language=C, escapeinside=``, caption={Example of bias from link order. Instruction alignment of the loop changes with ordering. }]
+`loop.c`
+#define B22(n) b += i*i*a + 1
+volatile int i = 42;
+
+int loop(int a) {
+    register int b = i == 42 ? 1 : 0;
+    do {
+        // 10 bytes mov(6), test(2), jne(2)
+        if (!i) B22(1); 
+        if (!i) B22(2);
+        if (!i) B22(3);
+        if (!i) B22(4);
+        if (!i) B22(5);
+        if (!i) B22(6);
+        if (!i) B22(7);
+        if (!i) B22(8);
+        if (!i) B22(9);
+        if (!i) B22(10);
+        if (!i) B22(11);
+        // 27 bytes for increment and test
+    } while (i++ < 0x12345678);
+    return b;
+}
+`foo.c`
+int foo(int n) {
+    int a = 1;
+    return a;
+}
+`main.c`
+#include <stdio.h>
+int main() {
+    int i = loop(10);
+    int f = foo(10);
+    printf("Loop: %d\n", i);
+    printf("Foo: %d\n", f);
+    return 0;
+}
+\end{lstlisting}
 
@@
 \section{Conclusions}
+Any change to virtual memory layout of data can potentially introduce bias effects from address aliasing. 
+We find that compilers, even when provided with architecture specific optimization flags, still generates code that suffers from aliasing. 
+Typical heap allocators return page-aligned memory by default, which is the worst case with respect to aliasing for many algorithms.
+Despite recommendations from Intel, we are not aware of any allocators that specifically addresses this issue.
+
+The Loop Stream Detector can provide significant speedup, but subtle changes to code layout sometimes prevents this optimization to be utilized. 
+Programmers rarely care, or even know, about intricate details such as the addresses and alignment of code. 
+We expect, and probably rightfully so, that sensible compilers and linkers will organize code somewhat optimally. 
+In cases where this is not true, weird effects that might be categorized as “bias” occurs. 
+For developers, it is useful to be aware of the fact that any change to instruction addresses can potentially have a huge impact on performance.
 

### 2014-01-28 paper.tex
Remove section on LSD, and start cleanup lplot

@@
 \usepackage{authblk}
-\usepackage{caption} % For margin top bottom
+\usepackage{caption}
 
@@
 
+\subsection{Analysis}
+
 A small amount of assembly code was used to calculate the addresses of each stack allocated variable and output them directly using system calls, without impacting run time addresses.
@@
 The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.
+
+\subsection{Avoid Aliasing -- A Proof of Concept}
+Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
+In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as address layout randomization. 
+Although one can not easily know if a collision is going to happen for a given environment, one can try to change the program to account for possible alias effects.
+The following strategy is a proof of concept of how alias-free code can be generated in this particular case.
+If the addresses do alias, branch to an alternative but semantically equivalent code path.
+Allocate a new set of variables to avoid aliasing.
+
+\begin{lstlisting}[float=h, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopkernel}, frame=lines]
+static int i, j, k;
+
+int main() {
+    int g = 0, inc = 1;
+
+    if ((((long)&inc) & 0xfff) == (((long)&i) & 0xfff) || 
+        (((long)&g) & 0xfff) == (((long)&i) & 0xfff))
+        return main();
+
+    for (; g < 65536*10; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+
+    return 0;
+}
+\end{lstlisting}
 
@@
 
-\begin{lstlisting}[float=h, language=C, caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}, label={lst:conv}, frame=lines]
+\begin{lstlisting}[float=t, language=C, caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}, label={lst:conv}, frame=lines]
 static float k[5]={0.1,0.25,0.3,0.25,0.1};
@@
 
-\section{The Loop Stream Detector}
-Typical software spends most of the time executing the same instructions repeatedly in loops, where the same instructions are fetched, decoded and executed many times. 
-The Loop Stream Detector (LSD) is a front-end hardware optimization that is able to detect small software loops.
-A cache of decoded instructions is kept and analyzed for cyclic chains in the instruction stream.
-When a loop is detected, can be fetched directly from a queue of already decoded micro operations.
-This enables some instruction decode logic in the processor front end to be temporarily disabled.
-The LSD is a power saving optimization, but can also improve performance in cases where the decode pipeline is a bottleneck.
-Utilization of the Loop Stream Detector can be measured by the following performance counter.
-\begin{description}
-  \item{LSD.UOPS\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0xA8 & 0x01\end{tabular}}} \hfill \\
-  Counts the number of micro-operations delivered by the Loop Stream Detector.
-\end{description}
-
-The LSD has been suggested in previous work as a potential source of measurement bias [Mytkowicz:2009:WrongData].
-The Loop Stream Detector is dependent on code layout, which can change with external factors such as variations in link ordering. 
-Interacting with the alignment and position of instructions in memory, the LSD can cause performance differences for programs with otherwise \emph{identical} instruction sequences.
-
-Some effort has been made to model properties of the LSD on previous Intel architectures, as an aid in low-level optimization. 
-Hundt et al.~\cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
-For loops to be recognized, they have to match a set of restrictions imposed by the hardware implementation that tries to detect them. 
-For instance, because the LSD cache is of fixed size there will be a maximum number of instructions a loop can contain for it to be recognized.
-The capabilities of the LSD has improved with newer micro-architectures.
-Some important limitations for Haswell's LSD are listed below.
-\begin{itemize}
-  \item Up to 12 chunk fetches of 32 B instruction memory.\footnote{The optimization manual actually states 8 chunk fetches for HTT enabled, and 11 for HTT disabled.
-Our experiments with a minimal test case shows that the LSD is active for loops spanning up to 12 chunks -- regardless of hyper threading.}
-  \item Up to 28 micro-operations for HTT enabled, 56 for HTT disabled.
-  \item Can not contain CALL or RET instructions
-  \item Stack operations must match, i.e push must be matched with a pop.
-\end{itemize}
-
-\subsection{Bias from Link Order}
-Different link orders affect the relative position of code and static data within the final compiled binary file.
-More importantly, \emph{alignment} of code can be affected by link order.
-A loop that was aligned to 32 byte for one ordering might end up at 16 byte alignment in another. 
-Alignment is important for determining the number of 32 byte chunks a loop is spanning.
-
-\begin{lstlisting}[language=C, escapeinside=``, caption={Example of bias from link order. Instruction alignment of the loop changes with ordering. }]
-`loop.c`
-#define B22(n) b += i*i*a + 1
-volatile int i = 42;
-
-int loop(int a) {
-    register int b = i == 42 ? 1 : 0;
-    do {
-        // 10 bytes mov(6), test(2), jne(2)
-        if (!i) B22(1); 
-        if (!i) B22(2);
-        if (!i) B22(3);
-        if (!i) B22(4);
-        if (!i) B22(5);
-        if (!i) B22(6);
-        if (!i) B22(7);
-        if (!i) B22(8);
-        if (!i) B22(9);
-        if (!i) B22(10);
-        if (!i) B22(11);
-        // 27 bytes for increment and test
-    } while (i++ < 0x12345678);
-    return b;
-}
-`foo.c`
-int foo(int n) {
-    int a = 1;
-    return a;
-}
-`main.c`
-#include <stdio.h>
-int main() {
-    int i = loop(10);
-    int f = foo(10);
-    printf("Loop: %d\n", i);
-    printf("Foo: %d\n", f);
-    return 0;
-}
-\end{lstlisting}
-
-\subsection{Other Triggers of Bias Effects}
-The focus of previous work, and elaborated in this section, is measurement bias caused by altering link ordering.
-The important factor is not the process of linking itself, but the effect it has on final code layout in the compiled binary.
-
-\paragraph{Order of functions within a file}
-As with link order, the order of functions within a source file usually determines their relative position in the text segment of the compiled binary. 
-Given a program with functions foo and bar, listing foo before bar and vice versa generates two “different” programs with respect to memory layout. 
-GCC prints functions to the text segment in the order they appear in the source code.
-
-\paragraph{Length of external symbols}
-Symbols such as function names are placed in a symbol table section in the ELF file.
-Symbols that needs to be resolved dynamically are allocable, and mapped to virtual address space before the text segment. 
-This means that names of external symbols resolved at run-time does affect code layout. 
-Longer function names (or more external dependencies) will offset the text segment because the size of allocable segments related to dynamic linking changes. 
-his includes the “dynsym”, “dynstr”, and “rela” sections in an executable made with GCC.
-
-Some particularly devious “bugs” can occur from this. 
-Consider a scenario where a program does not align properly to utilize the Loop Stream Detector. 
-Assume a printf statement is added to a completely unrelated part of the code. 
-The additional external symbol offsets instruction alignment such that the LSD can be used, giving a huge speedup.
-Later removing the printf statement will reduce the size of external dependencies, and again result in worse performance.
-
-Any interaction that affects code layout can potentially trigger performance cliffs caused by the Loop Stream Detector.
-
 \section{Conclusions}
@@
 Despite recommendations from Intel, we are not aware of any allocators that specifically addresses this issue.
-
-The Loop Stream Detector can provide significant speedup, but subtle changes to code layout sometimes prevents this optimization to be utilized. 
-Programmers rarely care, or even know, about intricate details such as the addresses and alignment of code. 
-We expect, and probably rightfully so, that sensible compilers and linkers will organize code somewhat optimally. 
-In cases where this is not true, weird effects that might be categorized as “bias” occurs. 
-For developers, it is useful to be aware of the fact that any change to instruction addresses can potentially have a huge impact on performance.
 

### 2014-02-01 code/perf-env/printenv.c
Program and script to determine env size running perf-stat

@@ -1,0 +1,39 @@
+/// Program that prints its own environment, illustrates how running perf-stat
+/// adds some environment variables.
+///
+/// $ perf stat env -i ./printenv
+///     <empty>
+///
+/// $ env -i perf stat ./printenv
+///	    PWD=/home/lars/bias-paper/code/perf-env
+///     SHLVL=0
+///     PERF_BUILDID_DIR=.debug
+///     PATH=/usr/libexec/perf-core:/usr/lib/linux-tools/3.11.0-15-generic:/usr/local/bin:/usr/bin:/bin
+///
+/// $ perf stat ./printenv
+///	    PWD=/home/lars/bias-paper/code/perf-env
+///     SHLVL=0
+///     PERF_BUILDID_DIR=.debug
+///     PATH=/usr/libexec/perf-core:/usr/lib/linux-tools/3.11.0-15-generic:/usr/local/bin:/usr/bin:/bin
+///     (+ whatever was already there)
+///
+#include <stdio.h>
+#include <string.h>
+
+int main() {
+	char file[64], command[128];
+	sprintf(file, "/proc/%d/environ", getpid());
+
+	strcpy(command, "cat ");
+	strcat(command, file);
+	system(command);
+
+	putchar('\n');
+
+	strcat(command, " | wc -c");
+	system(command);
+
+	printf("stack: %p\n", (void*)file);
+
+	return 0;
+}

### 2014-02-01 code/perf-env/printenv.py

@@ -1,0 +1,19 @@
+#!/usr/bin/env python 
+
+import subprocess
+
+# Generating plots with lplot are done using python and subprocess. This script
+# is used to invoke printenv, to show how subprocess.open with an empty
+# environment can possibly affect addresses compared to executing from shell.
+# In conclusion, perf stat adds about 170 bytes, including PWD which will
+# obviously vary between setups. Launching through subprocess.Popen does not
+# affect environment or variable addresses. 
+
+# $ ./printenv.py
+# PWD=/home/lars/bias-paper/code/perf-env
+# SHLVL=0
+# PERF_BUILDID_DIR=.debug
+# PATH=/usr/libexec/perf-core:/usr/lib/linux-tools/3.11.0-15-generic:/usr/local/bin:/usr/bin:/bin
+
+p = subprocess.Popen("perf stat -e cycles:u ./printenv", env={}, shell=True)
+p.wait()

### 2014-02-01 code/env-alias/loop.c
Add code files for loop kernel analysis

@@ -1,0 +1,20 @@
+/**
+ * Copy of microkernel from the paper
+ * "Producing Wrong Data Without Doing Anything Obviously Wrong!"
+ * 
+ * Interesting counters
+ * cycles:u,r0107:u,r003c:u,r013c:u,r025c:u,r01a2:u,r02a3:u
+ *
+ */
+
+static int i, j, k;
+
+int main() {
+    int g = 0, inc = 1;
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    return 0;
+}

### 2014-02-01 code/env-alias/recursion-fix.c

@@ -1,0 +1,28 @@
+/// Analyzing run-time addresses of automatic variables g and inc revealed
+/// that worst case performance happens when i alias with inc.
+static int i, j, k;
+
+int main() {
+    int g = 0, inc = 1;
+
+    // Offset by another call frame in case of alias. Note that which 
+    // variables alias is compiler-dependent. For a different static allocation
+    // of i, j, k, this case could be impossible due to alignment guarantees,
+    // but instead there would be another pair of aliasing variables. For
+    // the same case in thesis "Analyzing Contextual Bias of Program
+    // Execution on Modern CPUs", we found that inc alias with j.
+    // Plot twist: The order inc and g is allocated on stack depends on the check
+    // performed here, causing this approach to fail when only checking one
+    // pair for collision => the other pair will collide instead.
+    if ((((long)&inc) & 0xfff) == (((long)&i) & 0xfff) || 
+        (((long)&g) & 0xfff) == (((long)&i) & 0xfff))
+        return main();
+
+    for (; g < 65536*10; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    
+    return 0;
+}

### 2014-02-02 paper.tex
Add analysis stuff on loop

@@
 
-\lstset{language=C, basicstyle=\ttfamily, columns=flexible}
+\lstset{language=C, basicstyle=\ttfamily\small, columns=flexible}
 
@@
 
-\subsection{Bias from Environment Size}
+\section{Bias from Environment Size}
 Variations in performance with bias towards certain environment sizes has been observed by other authors.
@@
 \subsection{Analysis}
+To be able to explain what is happening in this case we need to know the addresses of all variables at run time.
+For statically allocated data, memory allocation is decided at compile time.
+The location of variables i, j and k can be found by looking at the symbol table in the ELF executable.
+Below is an excerpt of \lstinline$readelf -s$.
+\begin{lstlisting}[float=h, basicstyle=\small]
+37: 000000000060103c    4  (...)  25 i
+38: 0000000000601040    4  (...)  25 j
+39: 0000000000601044    4  (...)  25 k
+\end{lstlisting}
 
 A small amount of assembly code was used to calculate the addresses of each stack allocated variable and output them directly using system calls, without impacting run time addresses.
-We find that the spike in cycle count occurs when the address of g and inc alias with i and j, respectively.
-Resource stalls are generated because of false dependencies between the addresses of (g, i) and (inc, j), resulting in worse performance.
+We find that the spike in cycle count occurs when the address of inc alias with i, where the first spike happens when inc is at 0x7fffffffe03c.
+Because the stack is aligned to 16 byte or 4 words, there are a couple of different scenarios that could have been observed here.
+Static area is fixed and covers 12 contiguous bytes (3 words) \lstinline$[x|x| |x]$.
+Automatic variables occupies 8 contiguous bytes on stack (2 words) [ | |x|x]. 
+The space of possible addresses for (inc, g) in this case is (0x7fffffffe03c, 0x7fffffffe038) - n*4096. 
+In this scenario, g will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of i, j or k.
+An even worse case with respect to the number of alias events is if allocation of static data allows collisions with both stack allocated variables.
+This can be achieved by reserving 8 extra bytes to offset i, j into the 0x8, 0xc slots. 
+While this will give significantly more alias counts, it has little effect on the number of cycles executed.
+
 Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment.
+Resource stalls are generated because of false dependencies between stack and static data, resulting in worse performance.
 The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.
@@
 \begin{lstlisting}[float=h, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopkernel}, frame=lines]
+#define ALIAS(a, b) \
+    ((((long)&a)&0xfff)==(((long)&b)&0xfff))
 static int i, j, k;
-
 int main() {
     int g = 0, inc = 1;
-
-    if ((((long)&inc) & 0xfff) == (((long)&i) & 0xfff) || 
-        (((long)&g) & 0xfff) == (((long)&i) & 0xfff))
+    if (ALIAS(inc, i) || ALIAS(g, i))
         return main();
-
     for (; g < 65536*10; g++) {
@@
     }
-
     return 0;
@@
 
-\subsection{Heap Address Aliasing}
+\section{Bias from Heap Allocations}
 Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
@@
 
+\section{Aliasing in BLAS}
+Another test case, show numbers for ATLAS OR default blas in Ubuntu, or maybe both. Copy thesis more or less I guess.
+
 \section{Conclusions}

### 2014-02-02 code/env-alias/loop.c

@@ -1,12 +1,3 @@
-/**
- * Copy of microkernel from the paper
- * "Producing Wrong Data Without Doing Anything Obviously Wrong!"
- * 
- * Interesting counters
- * cycles:u,r0107:u,r003c:u,r013c:u,r025c:u,r01a2:u,r02a3:u
- *
- */
-
 static int i, j, k;
 
 int main() {

### 2014-02-02 code/env-alias/recursion-fix.c

@@ -1,10 +1,13 @@
-/// Analyzing run-time addresses of automatic variables g and inc revealed
-/// that worst case performance happens when i alias with inc.
+#define ALIAS(a, b) \
+    ((((long)&a)&0xfff)==(((long)&b)&0xfff))
+
 static int i, j, k;
 
 int main() {
     int g = 0, inc = 1;
 
+    // Analyzing run-time addresses of automatic variables g and inc revealed
+    // that worst case performance happens when i alias with inc.
     // Offset by another call frame in case of alias. Note that which 
     // variables alias is compiler-dependent. For a different static allocation
     // of i, j, k, this case could be impossible due to alignment guarantees,
@@ -14,8 +17,7 @@ int main() {
     // Plot twist: The order inc and g is allocated on stack depends on the check
     // performed here, causing this approach to fail when only checking one
     // pair for collision => the other pair will collide instead.
-    if ((((long)&inc) & 0xfff) == (((long)&i) & 0xfff) || 
-        (((long)&g) & 0xfff) == (((long)&i) & 0xfff))
+    if (ALIAS(inc, i) || ALIAS(g, i))
         return main();
 
     for (; g < 65536*10; g++) {
@@ -23,6 +25,6 @@ int main() {
         j += inc;
         k += inc;
     }
-    
+
     return 0;
 }

### 2014-02-11 code/blas/gemv.c
Add BLAS test code

@@ -1,0 +1,50 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <cblas.h>
+
+// A is M x N -> rows x cols.
+#define M 8192
+#define N 8192
+
+// Matrix-vector multiplication, y = alpha*Ax + beta*y
+// As a relic from Fortran, BLAS prefers col-major order. 
+// When CblasNoTrans is set, an additional copy operation of A is done.
+int main(int argc, char **argv)
+{
+    int offset1 = argc > 1 ? atoi(argv[1]) : 0;
+    int offset2 = argc > 2 ? atoi(argv[2]) : 0;
+    int iters   = argc > 3 ? atoi(argv[3]) : 1;
+
+    const double alpha = 1.0, beta = 0.0;
+
+    if (offset1) malloc(sizeof(char) * offset1);
+
+    double *A = malloc(sizeof(double) * M * N);
+    for (int i = 0; i < M * N; ++i)
+        A[i] = (double) rand();
+
+    double *x = malloc(sizeof(double) * N);
+    for (int i = 0; i < N; ++i)
+        x[i] = (double) rand();
+
+    if (offset2) malloc(sizeof(char) * offset2);
+
+    double *y = malloc(sizeof(double) * M);
+
+    printf("A:%p b:%p c:%p\n", A, x, y);
+
+    for (int i = 0; i < iters; ++i)
+    cblas_dgemv(CblasColMajor, 
+        CblasNoTrans, // Transpose A
+        M, N,         // rows, cols
+        alpha, 
+        A, M,         // lda (offset to next column with same row index)
+        x, 1, 
+        beta, 
+        y, 1
+    );
+
+    free(A), free(x), free(y);
+
+    return 0;
+}

### 2014-02-11 paper.tex
Start BLAS section

@@
 
+Need to observe run time addresses without introducing any observer effects.
 A small amount of assembly code was used to calculate the addresses of each stack allocated variable and output them directly using system calls, without impacting run time addresses.
@@
 An even worse case with respect to the number of alias events is if allocation of static data allows collisions with both stack allocated variables.
-This can be achieved by reserving 8 extra bytes to offset i, j into the 0x8, 0xc slots. 
+This can be achieved by reserving an extra 8 bytes to offset i, j into the 0x8, 0xc slots. 
 While this will give significantly more alias counts, it has little effect on the number of cycles executed.
@@
 
-\subsection{Avoid Aliasing -- A Proof of Concept}
+\subsection{Offset to Avoid Aliasing}
 Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
@@
 
-\section{Aliasing in BLAS}
-Another test case, show numbers for ATLAS OR default blas in Ubuntu, or maybe both. Copy thesis more or less I guess.
+\subsection{Bias from Heap Allocators}
+Addresses of dynamic memory at run time is governed by the particular heap allocator used.
+Like the contents of environment varialbes, the exact operation of heap allocation can typically not be known ahead of time.
+For the code examples we present in C, the dynamic linker will resolve malloc to whichever version of libc is available on the system.
+In other words, it is part of the environment, and a potential source of measurement bias.
+(table, todo)
+
+\section{Performance impacts in BLAS}
+
+Basic Linear Algebra Subroutines (BLAS), is the de facto standard API for high performance linear algebra routines.
+The functionality is divided into three categories:
+
+\begin{description}
+  \item{Level 1} Scalar and vector operations, such as dot product and vector addition.
+  \item{Level 2} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
+  \item{Level 3} Matrix-matrix operations, including the widely applied gemm routine for general matrix-matrix multiplication. 
+\end{description}
+
+Many highly optimized implementations of BLAS exists, ATLAS being a widely used and open source alternative.
+One of the key features of ATLAS is that it uses automatic tuning to optimize for cache efficiency [Whaley:2000:ATLAS].
+
+We use the currently latest version 3.10.1 of ATLAS, built as a shared library from source.
+The automatic tuning happens during the build process.
+A series of test programs are run to determine cache edges and other properties of the hardware, which in turn affects the resulting binary.
+
+Carefully monitoring cache is particularly important in this case study, as BLAS performance heavily relies on cache efficiency.
+Relevant performance statistics for cache hit ratios will be considered to rule out cache as the cause of any bias effects.
+
+The idea is to reproduce similar scenarios to what we saw in Section ?, where aliasing was caused by linear accesses to pairs of aligned heap allocated data.
+Functions operating on vectors, from BLAS Level 1 or 2, seems most likely to have potential for similar characteristics.
+Using heap allocated memory, address aliasing is an artifact of the allocator used.
+With different versions or configurations of memory allocators, any performance impact from aliasing can qualify as measurement bias.
+In the following sections, we will study bias from address aliasing in the Level 2 function gemv specifically, which computes matrix-vector multiplication.
+
+\iffalse
+
+Consider matrix-vector multiplication of the form $\boldsymbol{y}=A\boldsymbol{x}$. Let A be of size M\times N, where M is the number of rows.
+
+\left[\begin{array}{ccccc}
+a_{0,0} & a_{0,1} &  &  & a_{0,N}\\
+a_{1,0}\\
+ &  &  & \ddots\\
+a_{M,0} &  &  &  & a_{M,N}
+\end{array}\right]\left[\begin{array}{c}
+x_{0}\\
+x_{1}\\
+\vdots\\
+\\
+x_{N}
+\end{array}\right]=\left[\begin{array}{c}
+y_{0}\\
+y_{1}\\
+\vdots\\
+y_{M}
+\end{array}\right]
+
+The corresponding BLAS function is the level 2 gemv routine, computing the more general matrix-vector product given as
+
+$\boldsymbol{y}=\alpha\text{op}\left(A\right)\boldsymbol{x}+\beta\boldsymbol{y}$
+
+Here, \alpha and \beta are constants, and \text{op}\left(A\right) is an optional transpose or complex conjugate of the matrix.
+A typical invocation of this routine is shown in Listing [?].
+We set \alpha=1, \beta=0 and \text{op}\left(A\right)=A to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
+Note that A is declared as CblasColMajor, meaning we impose a column major ordering of the data.
+The prefix indicates data type, in this case d for double precision.
+
+\fi
 

### 2014-02-11 analysis/env-alias/printf-analysis.c
Rename code dir and add lots of garbage

@@ -1,0 +1,31 @@
+#include <stdio.h>
+
+static int i, j, k;
+
+static char* str = "%p, %p\n";
+
+void output(void* a, void* b);
+
+int main() {
+    int g = 0, inc = 1;
+
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    output(&g, &inc);
+/*
+    asm __volatile__ (
+    	"mov $1 %0\n"
+    	""
+    	:
+    	: "r"(g), "r"(inc)
+    	: );
+*/
+    return 0;
+}
+
+void output(void* a, void* b) {
+    printf(str, a, b);
+}

### 2014-02-11 analysis/heap-alias/convolution-optimized.c

@@ -1,0 +1,43 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+// For automated testing, set -DOFFSET
+#ifndef OFFSET
+    #define OFFSET 0x00
+#endif
+
+// Hoard:    export LD_PRELOAD=~/Allocators/Hoard/libhoard.so
+// tcmalloc: libgoogle-perftools-dev, -ltcmalloc
+// cc -O3 convolution.c; cc -O3 convolution.c -S; perf stat -e cycles:u,r0107:u,r01a2:u,instructions:u -r 10 ./a.out
+// Best result with -O3 -std=c99 -march=native and no offset
+// -- Then we can save 250k alias and 300k RS with +0x30. Exact same instruction count (but sometimes slightly higher?)
+
+#define N 0x100000
+static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
+
+void convolve(int size, const float * restrict input, float * restrict output)
+{
+    int i, j;
+    for (i = 2; i < size - 2; ++i)
+    {
+        output[i] = 0;
+        for (j = 0; j < 5; ++j)
+            output[i] += input[i-2+j] * kernel[j];
+    }
+}
+
+int main(int argc, char **argv)
+{
+    // Needed for automated test script
+    int offset = argc > 1 ? atoi(argv[1]) : 0, i = 0;
+
+    float *input  = malloc( N          * sizeof(float));
+    float *result = malloc((N + 0x100) * sizeof(float));
+
+    //for (i = 0; i < 2; i++)
+    convolve(N, input, (result + offset));
+
+    printf("Offset: %d : (%p, %p)\n", offset, input, result);
+
+    return 0;
+}

### 2014-02-11 analysis/heap-alias/convolution.c

@@ -1,0 +1,31 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+#define N 0x100000
+static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
+
+void convolve(int size, float *input, float *output)
+{
+    int i, j;
+    for (i = 2; i < size - 2; ++i)
+    {
+        output[i] = 0;
+        for (j = 0; j < 5; ++j)
+            output[i] += input[i-2+j] * kernel[j];
+    }
+}
+
+int main(int argc, char **argv)
+{
+    // Needed for automated test script
+    int offset = argc > 1 ? atoi(argv[1]) : 0;
+
+    float *input  = malloc( N          * sizeof(float));
+    float *result = malloc((N + 0x100) * sizeof(float));
+
+    convolve(N, input, (result + offset));
+
+    printf("Offset: %d : (%p, %p)\n", offset, input, result);
+
+    return 0;
+}

### 2014-02-11 analysis/heap-alias/print-address.c

@@ -1,0 +1,43 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+/** Try different malloc implementations and check what addresses are returned
+ */
+#define PRINT(v) printf(#v "\t : %p\n", v);
+#define TEST(n) { \
+		float *a = malloc(n * sizeof(float)); \
+		float *b = malloc(n * sizeof(float)); \
+		printf(" %-10s : (%p, %p)\n", #n, a, b); \
+		free(a), free(b); \
+	}
+
+int main(void)
+{
+	TEST(0x01);
+	TEST(0x10);
+	TEST(0x20);
+	TEST(0xf0);
+	TEST(0x100);
+	TEST(0x123);
+	TEST(0xf00);
+	TEST(0xfed);
+	TEST(0xfff);
+	TEST(0x1000);
+	TEST(0x1234);
+	TEST(0xf000);
+	TEST(0xfedc);
+	TEST(0xffff);
+	TEST(0x10000);
+	TEST(0x12345);
+	TEST(0xf0000);
+	TEST(0xfedcb);
+	TEST(0xfffff);
+	TEST(0x100000);
+	TEST(0x123456);
+	TEST(0xf00000);
+	TEST(0xfedcba);
+	TEST(0xffffff);
+	TEST(0x1000000);
+
+	return 0;
+}

### 2014-02-12 paper.tex
Cleanup and small changes

@@
 
-\subsection{Analysis}
+\subsection{Finding Collisions}
 To be able to explain what is happening in this case we need to know the addresses of all variables at run time.
@@
 
-\section{Bias from Heap Allocations}
+\section{Bias from Heap Allocators}
 Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
@@
 
-\subsection{Bias from Heap Allocators}
+\subsection{Behavior of other Allocators}
 Addresses of dynamic memory at run time is governed by the particular heap allocator used.
@@
 
-We use the currently latest version 3.10.1 of ATLAS, built as a shared library from source.
+We use the currently latest stable version 3.10.1 of ATLAS, built as a shared library from source.
 The automatic tuning happens during the build process.

### 2014-02-16 analysis/blas/gemv.c
Results for large 8k test run

@@ -2,49 +2,54 @@
 #include <stdio.h>
 #include <cblas.h>
 
-// A is M x N -> rows x cols.
+// A is M rows x N cols
 #define M 8192
 #define N 8192
 
-// Matrix-vector multiplication, y = alpha*Ax + beta*y
-// As a relic from Fortran, BLAS prefers col-major order. 
-// When CblasNoTrans is set, an additional copy operation of A is done.
-int main(int argc, char **argv)
-{
+int main(int argc, char **argv) {
     int offset1 = argc > 1 ? atoi(argv[1]) : 0;
     int offset2 = argc > 2 ? atoi(argv[2]) : 0;
     int iters   = argc > 3 ? atoi(argv[3]) : 1;
 
+    // Matrix-vector multiplication, y = alpha*Ax + beta*y
     const double alpha = 1.0, beta = 0.0;
+    double *A, *x, *y;
 
-    if (offset1) malloc(sizeof(char) * offset1);
+    // Ensure random numbers are deterministic. Fill matrix A and input vector
+    // x with non-zero values for a more realistic test scenario. However,
+    // the cycle count is significantly reduced by using zero input, reinforcing
+    // aliasing effects.
+    srand(0);
 
-    double *A = malloc(sizeof(double) * M * N);
-    for (int i = 0; i < M * N; ++i)
-        A[i] = (double) rand();
+    // The matrix is big and will be mmap allocated.
+    A = malloc(sizeof(double) * M * N);
+    for (int i = 0; i < M * N; A[i++] = (double) rand())
+        ;
 
-    double *x = malloc(sizeof(double) * N);
-    for (int i = 0; i < N; ++i)
-        x[i] = (double) rand();
+    // Offset to manipulate normal heap addresses.
+    if (offset1) malloc(offset1);
 
-    if (offset2) malloc(sizeof(char) * offset2);
+    // x and y are small and willb e allocated on normal heap.
+    x = malloc(sizeof(double) * N);
+    for (int i = 0; i < N; x[i++] = (double) rand())
+        ;
 
-    double *y = malloc(sizeof(double) * M);
+    // Offset to separate x and y addresses.
+    if (offset2) malloc(offset2);
 
-    printf("A:%p b:%p c:%p\n", A, x, y);
+    y = malloc(sizeof(double) * M);
 
     for (int i = 0; i < iters; ++i)
-    cblas_dgemv(CblasColMajor, 
-        CblasNoTrans, // Transpose A
-        M, N,         // rows, cols
[…]

### 2014-02-19 paper.tex
Add more blas content
Most of it is copypaste from masters. Table containing results from new
benchmark.

@@
 \usepackage{caption}
+\usepackage{mathtools}
 
@@
 
-\section{Performance impacts in BLAS}
+\section{Performance Impact in Numeric Applications}
 
@@
 
-\iffalse
+Consider matrix-vector multiplication of the form $\boldsymbol{y} = A\boldsymbol{x} + \boldsymbol{b}$ . Let A be of size $M \times N$, where M is the number of rows.
 
-Consider matrix-vector multiplication of the form $\boldsymbol{y}=A\boldsymbol{x}$. Let A be of size M\times N, where M is the number of rows.
-
+$
 \left[\begin{array}{ccccc}
@@
 \end{array}\right]
+$
 
-The corresponding BLAS function is the level 2 gemv routine, computing the more general matrix-vector product given as
+The corresponding BLAS function is the level 2 gemv routine, computing the more general matrix-vector product given as $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$
+Here, $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix. 
+We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
 
-$\boldsymbol{y}=\alpha\text{op}\left(A\right)\boldsymbol{x}+\beta\boldsymbol{y}$
+To better be able to assess the effects of aliasing, the call to gemv needs to be isolated. 
+We make two benchmarks, one with iteration count $K = 1$ and another with $K = 101$.
+An approximation of the cost for a single invocation of dgemv can be expressed as 
+$$
+t_{\text{estimate}}=\frac{t_{K=101}-t_{K=1}}{100}
+$$
+where t represents some metric, such as the number of cycles.
+Subtracting the $K = 1$ run removes the constant overhead from the $K = 101$ run.
+Dividing by $100$ averages the values from the remaining iterations.
+Other iteration counts could have been used as well.
 
-Here, \alpha and \beta are constants, and \text{op}\left(A\right) is an optional transpose or complex conjugate of the matrix.
-A typical invocation of this routine is shown in Listing [?].
-We set \alpha=1, \beta=0 and \text{op}\left(A\right)=A to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
-Note that A is declared as CblasColMajor, meaning we impose a column major ordering of the data.
-The prefix indicates data type, in this case d for double precision.
+\begin{table*}[t]
+  \caption{Estimated cost of invoking cblas\_dgemv with a matrix size $8192 \times 8192$}
+  \label{tab:atlas8k}
+  \small
+  \centering
+  \begin{tabular}{l r r r}
+    \toprule
+      & (0x010, 0x010, 0x020) & (0x010, 0x010, 0x0b0) & (0x010, 0x140, 0x150) \\
+    \midrule
+    CPU\_CLK\_UNHALTED.THREAD\_P & 162,926,895 & 152,151,263 & 144,583,160 \\
+    LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS & \textbf{68,408,790} & \textbf{5,154,981} & \textbf{2,919} \\
 
-\fi
+    MEM\_UOPS\_RETIRED.ALL.LOADS & 100,701,777 & 100,702,138 & 100,701,752 \\
+    MEM\_LOAD\_UOPS\_RETIRED.HIT\_LFB & 31,628,539 & 32,495,287 & 29,853,850 \\
+    MEM\_LOAD\_UOPS\_RETIRED.L1\_HIT & 68,309,447 & 65,625,881 & 69,387,368 \\
+    MEM\_LOAD\_UOPS\_RETIRED.L2\_HIT & 711,973 & 1,540,956 & 864,720 \\
+    MEM\_LOAD\_UOPS\_RETIRED.L3\_HIT & 154,345 & 651,985 & 363,331 \\
+    MEM\_LOAD\_UOPS\_RETIRED.L3\_MISS & 106,313 & 350,449 & 295,235 \\
+    MEM\_UOPS\_RETIRED.ALL.STORES & 33,562,692 & 33,562,692 & 33,562,692 \\
+
+    LOAD\_HIT\_PRE.HW\_PF & 42,914,191 & 47,315,935 & 59,211,320 \\
+    \bottomrule
+  \end{tabular}
+\end{table*}
+
+Notes: 
+ - More BR\_INST.COND.TAKEN taken in case (a), but number of retired instructions are the same. Suggests replays of instructions.
+ - Inverse HW prefetch numbers from masters
+
+Estimated performance counter statistics for each of the three heap address configurations are shown in Table [?tab:gemv-estimate].
+In addition to cycle count and alias events, a number of relevant metrics related to cache activity are also included.
+Notice that almost all load micro-ops are served by either the line fill buffer or L1 cache in all cases.
+Only a small amount of loads come from L2, and almost none from L3.
+The hit rate for L1 actually decreases somewhat (considering LFB as well) with better execution time.
+This could be explained by less time for prefetchers to feed the L1 cache with data.
+The hardware prefetch counter indicates more hits in cases a and b.
+Again, we find that cache efficiency does not explain the performance cliffs we observe.
+
+Our results show that address aliasing between matrix and vector heap buffers can significantly impact performance of dgemv in ATLAS.
+Variations in heap addresses alone can give a speedup of more than 30 \%.
+
+\subsection{Dealing with Aliasing}
+For the particular case we investigated, a good heuristic is to align heap segments “far apart” within the 4 KiB area of 12 bit suffixes.
+More specifically, it appears that address suffixes of A and $\boldsymbol{y}$ are the most important to separate.
+
+As discussed in section [x], aliasing cases like these can be accounted for in software using padding techniques.
+A possible run time solution to adjust heap addresses can be realized as follows:
+
+\begin{itemize}
+  \item Allocate some extra space for one of the vectors when calling malloc, for instance sizeof(double) * (M + 0x100) for $\boldsymbol{y}$.
+
+  \item Check the returned pointers for potential alias, i.e. the difference between \&A and \&y. Offset using pointer arithmetic into the array with extra padding at the end, i.e. y += 0x100. 
+\end{itemize}
+
+Another option is to explicitly account for “worst cases” in the implementation of routines that are vulnerable to aliasing.
+Addresses can be explicitly checked for potential conflicts in cblas\_dgemv, and if possible branch to code that will not suffer from aliasing.
 

### 2014-02-23 analysis/allocators/address.c
Update test and add initial results

@@ -1,43 +1,0 @@
-#include <stdlib.h>
-#include <stdio.h>
-
-/** Try different malloc implementations and check what addresses are returned
- */
-#define PRINT(v) printf(#v "\t : %p\n", v);
-#define TEST(n) { \
-		float *a = malloc(n * sizeof(float)); \
-		float *b = malloc(n * sizeof(float)); \
-		printf(" %-10s : (%p, %p)\n", #n, a, b); \
-		free(a), free(b); \
-	}
-
-int main(void)
-{
-	TEST(0x01);
-	TEST(0x10);
-	TEST(0x20);
-	TEST(0xf0);
-	TEST(0x100);
-	TEST(0x123);
-	TEST(0xf00);
-	TEST(0xfed);
-	TEST(0xfff);
-	TEST(0x1000);
-	TEST(0x1234);
-	TEST(0xf000);
-	TEST(0xfedc);
-	TEST(0xffff);
-	TEST(0x10000);
-	TEST(0x12345);
-	TEST(0xf0000);
-	TEST(0xfedcb);
-	TEST(0xfffff);
-	TEST(0x100000);
-	TEST(0x123456);
-	TEST(0xf00000);
-	TEST(0xfedcba);
-	TEST(0xffffff);
-	TEST(0x1000000);
-
-	return 0;
-}

### 2014-02-23 analysis/allocators/heapalloc.c

@@ -1,0 +1,26 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+#define alloc_pair(n) { \
+        float *a = malloc(n); \
+        float *b = malloc(n); \
+        printf("%d,%p,%p\n", n, a, b); \
+    }
+
+int main(void) {
+    alloc_pair(64)
+    alloc_pair(1024)
+    alloc_pair(3001)
+    alloc_pair(4096)
+
+    alloc_pair(5111)
+    alloc_pair(8192)
+    alloc_pair(20000)
+    alloc_pair(100000)
+    alloc_pair(1 << 20) // 1 048 576
+    alloc_pair(1 << 22)
+    alloc_pair(1 << 24)
+    alloc_pair(1 << 28)
+    alloc_pair(1 << 30) // 1 GiB
+    return 0;
+}

### 2014-02-23 analysis/heap-alias/convolution-optimized.c
Remove unrelated comment

@@ -6,8 +6,6 @@
     #define OFFSET 0x00
 #endif
 
-// Hoard:    export LD_PRELOAD=~/Allocators/Hoard/libhoard.so
-// tcmalloc: libgoogle-perftools-dev, -ltcmalloc
 // cc -O3 convolution.c; cc -O3 convolution.c -S; perf stat -e cycles:u,r0107:u,r01a2:u,instructions:u -r 10 ./a.out
 // Best result with -O3 -std=c99 -march=native and no offset
 // -- Then we can save 250k alias and 300k RS with +0x30. Exact same instruction count (but sometimes slightly higher?)

### 2014-02-24 paper.tex
Refactor abstract and introduction

@@
     \begin{abstract}
-    Seemingly innocuous properties of the execution context, such as the contents of system environment variables, or variations in link ordering, can impact the performance of computer programs. 
-    Variations in external properties like these can bias programs towards certain configurations. 
-    These effects have been shown to be a significant issue in performance analysis, but unpredictable and difficult to deal with.
+    Seemingly innocuous properties of the execution context, such as the contents of system environment variables, can impact the performance of computer programs.
+    Factors impacting performance that are not directly related to the subject being measured can introduce bias towards certain configurations.
+    Measurement bias has been shown to be a significant issue in performance analysis, appearing to be unpredictable and difficult to deal with.
 
-    Abstract This paper focuses on the underlying reasons for bias effects that can be experienced for example by changing environment variables, or using different link orders. 
-    Through experimentation and careful measurements using performance counters, we identify several potential sources of bias on the 4th generation Intel Core i7 architecture Haswell. 
-    Limitations imposed by the Loop Stream Detector is revealed, along with effects from 4K address aliasing.
-We show that bias is in fact not completely unpredictable, and discuss measures for avoiding it.
+    This paper focuses on the underlying reasons for bias effects that can be experienced for example by changing environment variables or using different dynamic allocation libraries.
+    Each alteration of the environment we look at has the effect of changing run time addresses of data or code.
+    Through experimentation and careful measurements using hardware performance counters, we identify that address aliasing on the low 12 bits between reads and writes can cause stalls in the execution pipeline.
+    Depending on initial memory layout at program start, this can be a common source of bias.
+    For some applications, performance penalties from unfavorable memory address allocations can be as high as double the cycle count.
+    In a case study of auto-tuned BLAS implementation ATLAS, we estimate a potential 30\% performance variation in one matrix-vector multiplication routine, depending on the location in memory of each buffer.
+    Note that these effects are not a result of cache issues, but a heuristic in the out of order engine found in recent Intel microarchitectures.
+
+    With an understanding of 4K aliasing, we show that bias is in fact not completely unpredictable, and discuss measures for avoiding it.
     \end{abstract}
@@
 Optimizing for performance is an important topic in computer science, and high performance computing (HPC) in particular. 
-A great amount of research and effort is put into designing better algorithms and compiler optimizations, in order to produce an “optimal” executable. 
-However, performance of computer programs is not just a function of the collection of machine instructions executed. 
+A great amount of research and effort is put into designing better algorithms and compiler optimizations in order to produce an ``optimal'' executable. 
+However, performance of computer programs is not just a function of the sequence of machine instructions executed. 
 Various external properties, such as the operating system and user environment, can have have an impact on how fast programs execute. 
-Characteristics of the operating system can decide things like the placement of code and data in memory, which in turn interacts with various hardware mechanisms. 
-Corner cases in hardware introduces potential performance cliffs; If for example a memory access happens to cross a page boundary, the cost of an additional TLB miss can be significant. 
-The set of conditions under which a program executes is called the execution context, which includes virtual memory layout in particular. 
+Characteristics of the operating system can affect things like the placement of code and data in memory, which in turn interacts with various hardware mechanisms at run time.
+Corner cases in hardware introduces potential performance cliffs; If for example a memory access happens to cross a page boundary, the cost of an additional TLB miss could be significant. 
+Modern microprocessors are extremely complex in design and functionality; Some features of recent Intel processors includes several layers of cache to camouflage slow memory, multiple prefetchers, speculative out-of-order execution and branch prediction, just to name a few.
+Hardware features and optimizations interact with memory layout of program code and data in various ways, and in certain worst cases one can observe a performance degradation.
+We will not look at things like cache efficiency, which often can be the reason for bad performance under certain memory contexts. 
+Instead, we look at effects from less known optimizations in the out-of-order execution engine and instructions fetch pipeline. 
+We will unveil characteristics about one specific microarchitectural feature known as 4K aliasing, showing how false dependencies in the out-of-order execution pipeline can bias performance towards certain memory contexts.
+This effect can be triggered for example by changing stack position, and can explain bias from altering environment variables.
+With more accurate models of how bias occur, we will not only be able to actively avoid in performance, but in some cases also gain a real speedup on average.
+Because bias effects are intrinsically connected to various hardware features, this paper focuses on the 4th generation Intel® Core™ i7 ``Haswell'' architecture specifically. 
+However, the bias sources we present has also been shown to behave similarly on previous Intel architectures as well.
+
+\subsection{Background}
+The set of external, and a-priori unknown, conditions under which a program executes is called the execution context. 
 Variations within contexts can lead to programs being biased towards certain configurations. 
@@
 
-Previous work has studied parameters such as the size of Unix environment variables and program link order. 
+Previous work has studied parameters such as the size of Unix environment variables and program link order \cite{Mytkowicz:2008:OE&MB}. 
 Both are found to potentially have a significant effect on performance, even in standardized benchmarks. 
-Unfortunately, the effects also appear to be unpredictable, and therefore difficult to deal with. 
+Unix environment variables contain various information about the system, such as the name of the currently logged in user, home- and current directory path.
+In the paper appropriately named ``Producing Wrong Data Without Doing Anything Obviously Wrong'' by T.
+Mytkowicz et al.~\cite{Mytkowicz:2009:WrongData}, the authors show an examples of a program where performance varies significantly by changing merely the size of the environment.
+This means that the length of your name -- by extension you user login -- could in principle be the tipping point between good and bad performance.
+Unfortunately, the effects appear to be unpredictable, and therefore difficult to deal with. 
 This poses a problem for researchers and performance analysts, who will need to account for effects of measurement bias with more rigorous methodologies and statistical methods.
 
-Both the size of environment variables and link ordering ultimately has an effect on memory layout of running processes. 
-We will not look at things like cache efficiency, which often can be the reason for bad performance under certain memory contexts. 
-Instead, we look at effects from less known optimizations in the out-of-order execution engine and instructions fetch pipeline. 
-Because bias effects are intrinsically connected to various hardware features, this paper is focued on the Intel® Core™ i7 “Haswell” architecture specifically. 
-However, the bias sources we present has also been shown to behave similarly on previous Intel architectures as well.
-With a better understanding of intricate properties of the CPU, we show how one is able to not only predict, but also explicitly optimize for beneficial memory layouts.
-
-In the paper appropriately named “Producing Wrong Data Without Doing Anything Obviously Wrong” by T.
-Mytkowicz et al.~\cite{Mytkowicz:2009:WrongData}, the authors show how simple changes to the execution environment can have an impact on performance.
-One of their examples is a program where performance varies significantly with changes to the environment variables. 
-Unix environment variables contain various information about the system, such as the name of the currently logged in user, home- and current directory path.
-This means that the length of your name -- by extension you user login -- could in principle be the tipping point between good and bad performance.
-
-Previous work focuses mostly on how to mitigate effects of bias, and avoid drawing the wrong conclusions based on misleading measurements~\cite{Mytkowicz:2008:Easy, Mytkowicz:2009:WrongData}.
+In previous work, the focus has mostly been on how to mitigate effects of bias, and avoid drawing the wrong conclusions based on misleading measurements~\cite{Mytkowicz:2008:Easy, Mytkowicz:2009:WrongData}.
 The goal of this paper is to show not only that bias exists, but also \emph{why} it occurs.
 
-With more accurate models of how bias occur, we will not only be able to actively avoid in performance, but in some cases also gain a real speedup on average.
-
-Modern microprocessors are extremely complex in design and functionality. 
-Some features of recent Intel processors includes several layers of cache to camouflage slow memory, multiple prefetchers, speculative out-of-order execution and branch prediction, just to name a few.
-Hardware features and optimizations interact with memory layout of program code and data in various ways.
-We will unveil characteristics about two different architectural features in recent Intel CPUs, and show how they can bias performance towards certain memory contexts.
-
-First we will look at an aliasing effect between memory addresses of loads and stores, causing false dependencies in the out-of-order execution pipeline. 
-This effect can be triggered for example by changing stack position, and can explain bias from altering environment variables. 
-Secondly, we will look at the Loop Stream Detector, which in previous work on measurement bias has been suggested as a possible explanation.
-We therefore choose to study this particular optimization in detail, and show how measurement bias can occur from changing link order.
-
-\section{Methodology}
+\subsection{Methodology}
 When studying bias effects, carefully controlling all relevant properties of the environment is crucial. 
@@
 \end{itemize}
-Our hardware configuration consists of an Intel® Core™ i7-4770K @ 3.50 GHz and 32 GB DDR3 @ 1600 MHz, running a 64-bit installation of Ubuntu 13.10. GCC version 4.8.1 is used as compiler toolchain.
+Experimental hardware configuration consists of an Intel® Core™ i7-4770K @ 3.50 GHz, 32 GiB DDR3 memory running at 1600 MHz, on a 64-bit installation of Ubuntu 13.10.
+GCC version 4.8.1 is used as compiler toolchain.
 
-We use perf to acquire statistics and measurements.
-Perf is a utility to interface with hardware performance counters, and available as part of the Linux kernel.
-Measurements are aggregated over several runs to collect all of the about 200 performance counters provided on our architecture [Intel:2013:Volume3B]. 
+We use perf to acquire hardware performance counter statistics.
+Measurements are aggregated over several runs to collect an extensice sample of about 200 performance counters and combinations provided on our architecture~\cite{Intel:2013:Volume3B}. 
 Unless otherwise specified, a minimal environment is used for all benchmarks. 
-To measure statistics with n bytes added to the environment, a dummy variable is set to \(0^{n}\) (repeated zero characters n times).
+To measure statistics with n bytes added to the environment, a dummy variable is set to \(0^{n}\) (repeated zero characters n times).\footnote{Because perf-stat adds a few variables, the environment will never be completely empty.}
 
@@
 Although cache is often a likely explanation of performance variations, bias caused by cache issues is well understood and \emph{not} the focus of this paper.
-Cache related metrics are monitored in order to rule out cache as the underlying cause of bias. 
-The hit rates of micro-ops for each level of cache can be monitored by performance counters~\cite{Intel:2012:OptimizationManual}.
+Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
 

### 2014-02-28 paper.tex
Rewrite and structure abstract

@@
     \maketitle
-    \begin{abstract}
-    Seemingly innocuous properties of the execution context, such as the contents of system environment variables, can impact the performance of computer programs.
-    Factors impacting performance that are not directly related to the subject being measured can introduce bias towards certain configurations.
-    Measurement bias has been shown to be a significant issue in performance analysis, appearing to be unpredictable and difficult to deal with.
+    \begin{abstract} % Should be 150 -- 200 words max
+    % Motivation
+    Performance analysis in computer science is prone to measurement bias, where seemingly innocious properties of the execution context, such as the size of system environment variables, can impact performance in non-obvious ways.
+    Measurement bias has been shown to be a significant, but unpredictable and difficult to deal with.
+    In order to get reliable data for potentially biased programs, it might be necessary to average experiments over multiple environment configurations.
 
-    This paper focuses on the underlying reasons for bias effects that can be experienced for example by changing environment variables or using different dynamic allocation libraries.
-    Each alteration of the environment we look at has the effect of changing run time addresses of data or code.
-    Through experimentation and careful measurements using hardware performance counters, we identify that address aliasing on the low 12 bits between reads and writes can cause stalls in the execution pipeline.
-    Depending on initial memory layout at program start, this can be a common source of bias.
+    % Problem statement
+    Our goal is to identify some of the underlying mechanisms in the Intel ``Haswell'' processor architecture that causes measurement bias, and see how that can be accounted for in an attempt at optimization.
+
+    % Approach
+    Starting with the high level known causes of bias from previous work, we analyze a minimal C-program that exhibits bias from environment size.
+    In order to understand why performance varies, we use hardware performance counters to obtain detailed statistics for different environments.
+
+    % Results
+    We find that an effect known as 4K aliasing, causing resource stalls in the CPU execution pipeline, can be a common source of bias from external factors that alters run time addresses of program data.
     For some applications, performance penalties from unfavorable memory address allocations can be as high as double the cycle count.
-    In a case study of auto-tuned BLAS implementation ATLAS, we estimate a potential 30\% performance variation in one matrix-vector multiplication routine, depending on the location in memory of each buffer.
-    Note that these effects are not a result of cache issues, but a heuristic in the out of order engine found in recent Intel microarchitectures.
+    For ATLAS, an auto-tuned BLAS implementation, we estimate a potential 30\% performance variation in one matrix-vector multiplication routine, depending on the location in memory of each buffer.
 
-    With an understanding of 4K aliasing, we show that bias is in fact not completely unpredictable, and discuss measures for avoiding it.
+    % Conclusions
+    With an understanding of 4K aliasing, we show that bias is in fact not completely unpredictable, and discuss measures for avoiding bad performance because of it.
     \end{abstract}

### 2014-02-28 paper.tex
Refactoring

@@
 
-Previous work has studied parameters such as the size of Unix environment variables and program link order \cite{Mytkowicz:2008:OE&MB}. 
+Previous work has studied parameters such as the size of system environment variables and program link order \cite{Mytkowicz:2008:OE&MB}. 
 Both are found to potentially have a significant effect on performance, even in standardized benchmarks. 
-Unix environment variables contain various information about the system, such as the name of the currently logged in user, home- and current directory path.
+Environment variables contain various information, such as the name of the currently logged in user, home- and current directory path.
 In the paper appropriately named ``Producing Wrong Data Without Doing Anything Obviously Wrong'' by T.
@@
 \begin{itemize}
-  \item Address space layout randomization (ASLR) is disabled
-  \item Hyper-threading is disabled
-  \item System load is kept at a minimum
-  \item Automatic CPU frequency scaling is disabled
+  \item Address space layout randomization (ASLR) is disabled.
+  \item Hyper-threading is disabled.
+  \item System load is kept at a minimum.
+  \item Automatic CPU frequency scaling is disabled.
 \end{itemize}
@@
 Unless otherwise specified, a minimal environment is used for all benchmarks. 
-To measure statistics with n bytes added to the environment, a dummy variable is set to \(0^{n}\) (repeated zero characters n times).\footnote{Because perf-stat adds a few variables, the environment will never be completely empty.}
+To measure statistics with n bytes added to the environment, a dummy variable is set to \(0^{n}\) (repeated zero characters n times).\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
 
@@
 \end{lstlisting}
-Situations like these causes the pipeline to stall and wait for the operations to finish in proper order.
+Situations like these can cause the pipeline to stall and wait for the operations to finish in proper order.
 The number of times this happens can be counted by the following performance counter:
@@
 \end{description}
-Performance implications of 4K aliasing is discussed to some extent in the optimization manual from Intel. 
+Performance implications of 4K aliasing is discussed to some extent in the Intel optimization manual. 
 There are even some concrete suggestions for how avoid aliasing cases to improve performance (Rule 8 and 9)~\cite{Intel:2012:OptimizationManual}.
@@
 \section{Bias from Environment Size}
-Variations in performance with bias towards certain environment sizes has been observed by other authors.
-Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack. 
+Unless the program explicitly accesses environment variables, it is typically not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack.
 Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory.

### 2014-04-17 paper.tex
Add heap allocation table
Notes on why make results in different mmap addresses included in separate file.

@@
 \usepackage{mathtools}
+\usepackage{underscore}
 
@@
 \date{} % This is a timeless piece
+
+% Pretty printing of performance counter names
+\newcommand{\perfctr}[1] {
+  \small{\uppercase{#1}}
+}
 
@@
 
-\section{Bias from Heap Allocators}
+\section{Bias from Heap Allocation}
 Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
@@
 
+\subsection{Aligned Sequential Access}
 Consider the function shown in Listing~\ref{lst:conv}.
@@
 
-\begin{lstlisting}[float=t, language=C, caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}, label={lst:conv}, frame=lines]
+\begin{lstlisting}[float=h, language=C, caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}, label={lst:conv}, frame=lines]
 static float k[5]={0.1,0.25,0.3,0.25,0.1};
@@
 Addresses of dynamic memory at run time is governed by the particular heap allocator used.
-Like the contents of environment varialbes, the exact operation of heap allocation can typically not be known ahead of time.
+Like the contents of environment variables, the exact operation of heap allocation can typically not be known ahead of time.
 For the code examples we present in C, the dynamic linker will resolve malloc to whichever version of libc is available on the system.
 In other words, it is part of the environment, and a potential source of measurement bias.
-(table, todo)
 
-\section{Performance Impact in Numeric Applications}
+\begin{table*}[t]
+  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers}
+  \label{tab:mallocompare}
+  \centering
+  \begin{tabular}{l r r r r}
+    \toprule
+      & ptmalloc & tcmalloc & jemalloc & Hoard \\
+    \midrule
+    64 B       & 0x602010 & 0xe0e000 & 0x2aaaabc0e040 & 0x2aaaaab20070 \\
+               & 0x602060 & 0xe0e080 & 0x2aaaabc0e080 & 0x2aaaaab200b0 \\
+    \midrule
+    1024 B     & 0x6020b0 & 0xe12000 & 0x2aaaabc0f400 & 0x2aaaaab30070 \\
+               & 0x6024c0 & 0xe12800 & 0x2aaaabc0f800 & 0x2aaaaab30470 \\
+    \midrule
+    5120 B     & 0x6028d0 & 0xe16000 & 0x2aaaabc1f000 & 0x2aaaaab40070 \\
+               & 0x603ce0 & 0xe18800 & 0x2aaaabc21000 & 0x2aaaaab42070 \\
+    \midrule
+    $2^{20}$ B & 0x2aaaaaaf3010 & 0xe64000 & 0x2aaaabc86000 & 0x2aaaabf40070 \\
+               & 0x2aaaab098010 & 0xf64000 & 0x2aaaabd86000 & 0x2aaaac060070 \\
+    \bottomrule
+  \end{tabular}
+\end{table*}
 
+Using heap allocated memory, address aliasing is an artifact of the allocator used.
+With different versions or configurations of memory allocators, any performance impact from aliasing can qualify as measurement bias.
+
+\section{Optimizing Alignment in BLAS}
+
+To illustrate how real world applications can be impacted by aliasing, we look at low level numerical applications.
+In this section we show that even highly optimized numerical libraries suffer performance degradation from aliasing.
 Basic Linear Algebra Subroutines (BLAS), is the de facto standard API for high performance linear algebra routines.
@@
 \begin{description}
-  \item{Level 1} Scalar and vector operations, such as dot product and vector addition.
-  \item{Level 2} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
-  \item{Level 3} Matrix-matrix operations, including the widely applied gemm routine for general matrix-matrix multiplication. 
+  \item{Level~1} Scalar and vector operations, such as dot product and vector addition.
+  \item{Level~2} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
+  \item{Level~3} Matrix-matrix operations, including the widely applied gemm routine for general matrix-matrix multiplication. 
 \end{description}
 
+Functions operating on vectors, from BLAS Level 1 or 2, intuitively appears the most likely to have potential for aliasing.
 Many highly optimized implementations of BLAS exists, ATLAS being a widely used and open source alternative.
@@
 We use the currently latest stable version 3.10.1 of ATLAS, built as a shared library from source.
-The automatic tuning happens during the build process.
-A series of test programs are run to determine cache edges and other properties of the hardware, which in turn affects the resulting binary.
+The automatic tuning happens during the build process; A series of test programs are run to determine cache edges and other properties of the hardware, which in turn affects the resulting binary.
 
-Carefully monitoring cache is particularly important in this case study, as BLAS performance heavily relies on cache efficiency.
-Relevant performance statistics for cache hit ratios will be considered to rule out cache as the cause of any bias effects.
+Carefully monitoring cache metrics is particularly important in this case study, as BLAS performance heavily relies on cache efficiency.
+Relevant performance statistics for cache hit ratios was considered to rule out cache as the cause of any bias effects.
+In addition, we disable hardware prefetching in BIOS to further minimize cache impacts. 
+Note however that the same conclusions can be drawn independently of these measures.
 
-The idea is to reproduce similar scenarios to what we saw in Section ?, where aliasing was caused by linear accesses to pairs of aligned heap allocated data.
-Functions operating on vectors, from BLAS Level 1 or 2, seems most likely to have potential for similar characteristics.
-Using heap allocated memory, address aliasing is an artifact of the allocator used.
-With different versions or configurations of memory allocators, any performance impact from aliasing can qualify as measurement bias.
-In the following sections, we will study bias from address aliasing in the Level 2 function gemv specifically, which computes matrix-vector multiplication.
+\subsection{Aliasing in cblas\_daxpy}
+Found aliasing here as well. Two spikes, not sure why. Might want to test other BLAS libraries. Include or leave out? Too much handwaving and too little analysis maybe.
 
-Consider matrix-vector multiplication of the form $\boldsymbol{y} = A\boldsymbol{x} + \boldsymbol{b}$ . Let A be of size $M \times N$, where M is the number of rows.
+\subsection{Matrix vector multiplication}
+
+Consider matrix-vector multiplication of the form $\boldsymbol{y} = A\boldsymbol{x} + \boldsymbol{b}$.
+Let A be of size $M \times N$, where M is the number of rows.
 
@@
 
-The corresponding BLAS function is the level 2 gemv routine, computing the more general matrix-vector product given as $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$
+General matrix vector multiplication is implemented by the level 2 gemv routine, computing $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$
 Here, $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix. 
 We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
+[There are three memory buffers involved, and one might suspect that implementations are succeptible to aliasing between of $\boldsymbol{A}$, $\boldsymbol{x}$ or $\boldsymbol{y}$.]
+For the following discussion we look at a particular configuration where $A$ is of size $64\times64$.
 
-To better be able to assess the effects of aliasing, the call to gemv needs to be isolated. 
-We make two benchmarks, one with iteration count $K = 1$ and another with $K = 101$.
-An approximation of the cost for a single invocation of dgemv can be expressed as 
+In order to accurately measure any effects of aliasing, we need to isolate the actual function call from heap allocation and initialization code.
+This is done by running the program twice for each memory configuration, one where cblas\_dgemv is executed $K = 101$ times in a tight loop.
+An approximation of the cost for a single function invocation can be expressed as 
 $$
@@
 where t represents some metric, such as the number of cycles.
-Subtracting the $K = 1$ run removes the constant overhead from the $K = 101$ run.
-Dividing by $100$ averages the values from the remaining iterations.
+Subtracting the $K = 1$ run removes the constant overhead from the $K = 101$ run, and dividing by $100$ averages the values from the remaining iterations.
 Other iteration counts could have been used as well.
+
+Figure \ref{fig:heatmap} shows estimated number of alias events for every relevant memory configuration possible for cblas\_dgemv with matrix size 64 $\times$ 64. 
+While matrix A is kept fixed at address suffix 0x000, every reasonable variation of (\&x, \&y) is measured.
+Each increment represents an address offset of 16 bytes, which is the same granularity one would get from the default heap allocator.
+In a 4K address space, there are thus $4096/16 = 256$ possible address suffixes for $\boldsymbol{x}$ and $\boldsymbol{y}$ independently, resulting in a grid of $256^2$ values.
+
+\begin{figure}[h]
+  \caption{Estimated aliasing cost of invoking cblas\_dgemv with a matrix size $64 \times 64$ for different memory layouts.
+Axis show address offsets times 0x10 for vectors $\boldsymbol{x}$ and $\boldsymbol{y}$.
+Matrix $A$ has fixed to address suffix 0x000.}
+  \label{fig:heatmap}
+  \includegraphics[width=\columnwidth]{resources/heatmap.eps}
+\end{figure}
+
+It appears that aliasing is mostly dependent on the location of $\boldsymbol{y}$, in this case with 8 distinct worst case regions.
+This is consistent with false dependencies between $\boldsymbol{y}$ and specific rows in A. 
+The first bad region is where $\&y ~= 0$, matching the first row of A.
 
@@
     \midrule
-    CPU\_CLK\_UNHALTED.THREAD\_P & 162,926,895 & 152,151,263 & 144,583,160 \\
-    LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS & \textbf{68,408,790} & \textbf{5,154,981} & \textbf{2,919} \\
+    \perfctr{cpu_clk_unhalted.thread_p} & 162,926,895 & 152,151,263 & 144,583,160 \\
+    \perfctr{ld_blocs_partial.address_alias} & \textbf{68,408,790} & \textbf{5,154,981} & \textbf{2,919} \\
 
-    MEM\_UOPS\_RETIRED.ALL.LOADS & 100,701,777 & 100,702,138 & 100,701,752 \\
-    MEM\_LOAD\_UOPS\_RETIRED.HIT\_LFB & 31,628,539 & 32,495,287 & 29,853,850 \\
-    MEM\_LOAD\_UOPS\_RETIRED.L1\_HIT & 68,309,447 & 65,625,881 & 69,387,368 \\
-    MEM\_LOAD\_UOPS\_RETIRED.L2\_HIT & 711,973 & 1,540,956 & 864,720 \\
-    MEM\_LOAD\_UOPS\_RETIRED.L3\_HIT & 154,345 & 651,985 & 363,331 \\
-    MEM\_LOAD\_UOPS\_RETIRED.L3\_MISS & 106,313 & 350,449 & 295,235 \\
-    MEM\_UOPS\_RETIRED.ALL.STORES & 33,562,692 & 33,562,692 & 33,562,692 \\
+    \perfctr{mem_uops_retired.all.loads} & 100,701,777 & 100,702,138 & 100,701,752 \\
+    \perfctr{mem_load_uops_retired.hit_lfb} & 31,628,539 & 32,495,287 & 29,853,850 \\
+    \perfctr{mem_load_uops_retired.l1_hit} & 68,309,447 & 65,625,881 & 69,387,368 \\
+    \perfctr{mem_load_uops_retired.l2_hit} & 711,973 & 1,540,956 & 864,720 \\
+    \perfctr{mem_load_uops_retired.l3_hit} & 154,345 & 651,985 & 363,331 \\
+    \perfctr{mem_load_uops_retired.l3_miss} & 106,313 & 350,449 & 295,235 \\
+    \perfctr{mem_load_uops_retired.all_stores} & 33,562,692 & 33,562,692 & 33,562,692 \\
 
-    LOAD\_HIT\_PRE.HW\_PF & 42,914,191 & 47,315,935 & 59,211,320 \\
+    \perfctr{load_hit_pre.hw_pf} & 42,914,191 & 47,315,935 & 59,211,320 \\
     \bottomrule
@@
  - Inverse HW prefetch numbers from masters
+
+\paragraph{Prefetching}
+Measurements repeated with prefetching disabled, giving similar results.
 

### 2014-04-17 analysis/allocators/heapalloc.c

@@ -2,25 +2,16 @@
 #include <stdio.h>
 
 #define alloc_pair(n) { \
-        float *a = malloc(n); \
-        float *b = malloc(n); \
+        char *a = malloc(n); \
+        char *b = malloc(n); \
         printf("%d,%p,%p\n", n, a, b); \
     }
 
-int main(void) {
+int main(int argc, char* argv[])
+{
     alloc_pair(64)
     alloc_pair(1024)
-    alloc_pair(3001)
-    alloc_pair(4096)
-
-    alloc_pair(5111)
-    alloc_pair(8192)
-    alloc_pair(20000)
-    alloc_pair(100000)
+    alloc_pair(5120)
     alloc_pair(1 << 20) // 1 048 576
-    alloc_pair(1 << 22)
-    alloc_pair(1 << 24)
-    alloc_pair(1 << 28)
-    alloc_pair(1 << 30) // 1 GiB
     return 0;
 }

### 2014-04-20 paper.tex
Add section on allocators, and improve conclusion

@@
 \usepackage{underscore}
+\usepackage{url}
 
@@
 \affil{Norwegian University of Science and Technology}
-\title{Sources of Measurement Bias in Modern CPUs}
+\title{Measurement Bias from Address Aliasing}
 \date{} % This is a timeless piece
@@
 
-\subsection{Offset to Avoid Aliasing}
+\subsection{Avoiding Aliasing}
 Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
@@
 It is perhaps unfair to assume the compiler can fix this when only given a generic optimization flag, as aliasing effects is (presumably) an Intel-specific issue.
-You might also wamt to utilize the C99 keyword restrict to explicitly tell the compiler that the buffers are disjoint in memory.
+You might also want to utilize the C99 keyword restrict to explicitly tell the compiler that the buffers are disjoint in memory.
 
@@
 \subsection{Behavior of other Allocators}
-Addresses of dynamic memory at run time is governed by the particular heap allocator used.
-Like the contents of environment variables, the exact operation of heap allocation can typically not be known ahead of time.
-For the code examples we present in C, the dynamic linker will resolve malloc to whichever version of libc is available on the system.
-In other words, it is part of the environment, and a potential source of measurement bias.
+As with the contents of environment variables, the exact behavior of dynamic memory allocation can typically not be known ahead of time.
+Heap allocation routines such as malloc and free are usually dynamically linked and resolved at run-time, which again represents a source of potential bias.
+For the code examples presented earlier, the dynamic linker will resolve malloc to some version of glibc, which is the default on our system.
+Variations between different versions of libc, or local configurations on parameters such as M\_MMAP\_THRESHOLD, could trigger aliasing cases.
 
-\begin{table*}[t]
+In addition to glibc's ptmalloc, we look at the following alternatives.
+\begin{itemize}
+ \item Thread-Caching Malloc (ptmalloc) is developed by Google. It aims to be time and space efficient, and reduces lock contention for multithreaded applications. \cite{TCMalloc}
+ \item jemalloc was originally developed for FreeBSD. It aims to be a scalable general purpose allocator in single or multithreaded environments. \cite{JEMalloc}
+ \item Hoard is another project that focuses extensively on improving performance for multithreaded programs. \cite{Berger:2000:Hoard}
+\end{itemize}
+
+\begin{table}[t]
   \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers}
@@
   \centering
-  \begin{tabular}{l r r r r}
+  \small
+  \begin{tabular}{l r r}
     \toprule
-      & ptmalloc & tcmalloc & jemalloc & Hoard \\
+                      & 5120 B          & $2^{20}$ B \\
     \midrule
-    64 B       & 0x602010 & 0xe0e000 & 0x2aaaabc0e040 & 0x2aaaaab20070 \\
-               & 0x602060 & 0xe0e080 & 0x2aaaabc0e080 & 0x2aaaaab200b0 \\
+    glibc             & 0x6028d0        & 0x2aaaaaaf3010 \\
+    (ptmalloc)        & 0x603ce0        & 0x2aaaab098010 \\
     \midrule
-    1024 B     & 0x6020b0 & 0xe12000 & 0x2aaaabc0f400 & 0x2aaaaab30070 \\
-               & 0x6024c0 & 0xe12800 & 0x2aaaabc0f800 & 0x2aaaaab30470 \\
+    tcmalloc          & 0xe16000        & 0xe64000 \\
+                      & 0xe18800        & 0xf64000 \\
     \midrule
-    5120 B     & 0x6028d0 & 0xe16000 & 0x2aaaabc1f000 & 0x2aaaaab40070 \\
-               & 0x603ce0 & 0xe18800 & 0x2aaaabc21000 & 0x2aaaaab42070 \\
+    jemalloc          & 0x2aaaabc1f000  & 0x2aaaabc86000 \\
+                      & 0x2aaaabc21000  & 0x2aaaabd86000 \\
     \midrule
-    $2^{20}$ B & 0x2aaaaaaf3010 & 0xe64000 & 0x2aaaabc86000 & 0x2aaaabf40070 \\
-               & 0x2aaaab098010 & 0xf64000 & 0x2aaaabd86000 & 0x2aaaac060070 \\
+    Hoard             & 0x2aaaaab40070  & 0x2aaaabf40070 \\
+                      & 0x2aaaaab42070  & 0x2aaaac060070 \\
     \bottomrule
   \end{tabular}
-\end{table*}
+\end{table}
 
+Table \ref{tab:mallocompare} illustrates some differences that can be observed by swithing heap allocator. \footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
+The addresses of two equally sized char buffers allocated with malloc are observed for size parameters 5120 and $2^{20}$ bytes. 
+We do not care about performance, only the actual memory addresses returned and whether they alias or not.
+
+For smaller allocations, we see that glibc and tcmalloc uses the normal heap area -- returning numerically low addresses.
+Interestingly, jemalloc and Hoard appears to never use brk, but always allocate to mmaped areas even for smaller sizes.
+We see one example where one allocator yields aliasing buffers while another does not. 
+Allocating $2 \times 5120 B$ returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
+Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.
+
+Most malloc alternatives focuses heavily on performance in multithreaded environments.
+Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
+To our knowledge, there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
+This dispite explicit recommendation from Intel that special purpose allocators should care about aliasing for optimization. [cite]
+
+The purpose of this is to illustrate that allocators is yet another source of bias.
 Using heap allocated memory, address aliasing is an artifact of the allocator used.
@@
 \section{Conclusions}
-Any change to virtual memory layout of data can potentially introduce bias effects from address aliasing. 
+We have studied how address aliasing can affect program performance under different memory layouts. 
+This effect is caused by the way speculative and out-of-order memory operations are handled by the CPU, only considering the last 12 address bits to resolve conflicts load and store operations.
+
+We have shown how aliasing can be the root cause of measurement bias from variations in environment size. 
+Analyzing the example provided in \ref{Mytkowicz:2009:WrongData}, we determined that collisions between stack variables and static data caused aliasing for certain stack positions.
+Variations in environment size changes virtual addresses of stack allocated data, and measurement bias observed by this can in some cases be attributed to address aliasing.
+
+In general, any change to virtual memory layout of data can potentially introduce bias effects from address aliasing. 
+We present a code example with extreme sensitivity of data alignment, with more than 50 \% performance variation between different memory layouts.
 We find that compilers, even when provided with architecture specific optimization flags, still generates code that suffers from aliasing. 
@@
 Despite recommendations from Intel, we are not aware of any allocators that specifically addresses this issue. 
+We show how manual padding of variables, and alternative alias-free code paths, can be used to avoid aliasing at run time.
+This means that an understanding of 4K address aliasing is sometimes needed to achieve optimal performance.
 

### 2014-04-20 analysis/blas/axpy.c
Update blas analysis files

@@ -1,0 +1,38 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <cblas.h>
+#include <sys/mman.h>
+
+// void cblas_daxpy (const int N, const double alpha, const double *X, const int incX, double *Y, const int incY);
+
+#define N (1 << 10)
+#define T float
+#define axpy cblas_saxpy
+
+#define ALIGNED_ALLOC(n) \
+    ((char*) mmap(NULL, n, PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0))
+
+int main(int argc, char **argv) {
+    int n  = argc > 1 ? atoi(argv[1]) : 1;
+    int dx = argc > 2 ? atoi(argv[2]) : 0;
+    int i;
+
+    // y = alpha x + y
+    const T alpha = 1.0;
+
+    // Make y page aligned, offset x from page alignment
+    T *y = (T*) ALIGNED_ALLOC(N*sizeof(T));
+    T *x = (T*) (ALIGNED_ALLOC(N*sizeof(T) + 4096)) + dx;
+
+    srand(0);
+
+    // Initialize separately to not introduce additional alias
+    for (i = 0; i < N; ++i) y[i] = (T) rand();
+    for (i = 0; i < N; ++i) x[i] = (T) rand();
+
+    for (i = 0; i < n; ++i)
+        axpy(N, alpha, x, 1, y, 1);
+
+    //printf("x:%p y:%p\n", x, y);
+    return 0;
+}

### 2014-04-20 analysis/blas/gemv-grid.c

@@ -4,8 +4,8 @@
 #include <sys/mman.h>
 
 // M rows x N cols
-#define M 256
-#define N 256
+#define M 8192
+#define N 8192
 
 #define ALIGNED_ALLOC(n) \
     ((char*) mmap(NULL, n, PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0))
@@ -22,36 +22,34 @@ int main(int argc, char **argv) {
     double *A, *x, *y;
 
     // Ensure page alignment as baseline. Keep matrix A at offset 0x000 even though
-    // for libc allocator it will by default always be 0x010. Have not observed any
-    // side effects of this.
-
-    // Easiest way to achieve this is probably mmap, so will just use that
+    // for libc allocator it will by default always be 0x010. Easiest way to 
+    // achieve this is probably mmap, so might as well use that.
     A = (double*) ALIGNED_ALLOC(M*N*sizeof(double));
     x = (double*) (ALIGNED_ALLOC(N*sizeof(double)+4096) + dx);
     y = (double*) (ALIGNED_ALLOC(M*sizeof(double)+4096) + dy);
 
-    // TODO: Check if use of aligned_alloc is actually sane, not mmaped?
-    //A = (double*) (aligned_alloc(4096, (M*N*sizeof(double)+4096) & -4096) + 0);
-    // Allocate an extra page for possible offset of x and y vectors
-    //x = (double*) (aligned_alloc(4096, (N*sizeof(double)+4096*2) & -4096) + dx);
-    //y = (double*) (aligned_alloc(4096, (M*sizeof(double)+4096*2) & -4096) + dy);
+    // Aliasing effects are slightly more visible for sparse (or zero) matrices, 
+    // resulting in fewer operations overall, while the number of alias events is 
+    // constant. Also keeps the example simple.
+    for (i = 0; i < N; ++i) {
+        A[i*N + i] = 1; 
+        x[i] = i;
+    }
 
-    srand(0);
-
-    for (i = 0; i < M*N; ++i) A[i] = (double) rand();
-    for (i = 0; i < N; ++i) x[i] = (double) rand();
-
+    // Repeat invocation to filter out constant overhead from allocation and
+    // initialization.
     for (i = 0; i < n; ++i)
-        cblas_dgemv(CblasColMajor, 
-            CblasNoTrans, // Transpose A
-            M, N, // rows, cols
+        cblas_dgemv(
+            CblasColMajor, 
[…]

### 2014-04-22 paper.tex
Remove BLAS case study
Trim abstract

@@
     \begin{abstract} % Should be 150 -- 200 words max
+
     % Motivation
-    Performance analysis in computer science is prone to measurement bias, where seemingly innocious properties of the execution context, such as the size of system environment variables, can impact performance in non-obvious ways.
-    Measurement bias has been shown to be a significant, but unpredictable and difficult to deal with.
-    In order to get reliable data for potentially biased programs, it might be necessary to average experiments over multiple environment configurations.
+    % Seemingly innocious properties of the execution context can cause subtle interactions with the low level hardware boundaries.
+    % Because of the vast complexity of modern hardware, these effects are found to be unpredictable and difficult to deal with.
+    % Programs are said to be \emph{biased} towards certain machines and configurations in various ways, .
+
+    % Interaction with cache boundaries, branch prediction and other intricate details of the underlying hardware can make accurate performance measurements difficult.
 
     % Problem statement
-    Our goal is to identify some of the underlying mechanisms in the Intel ``Haswell'' processor architecture that causes measurement bias, and see how that can be accounted for in an attempt at optimization.
+    In this paper we show how a lesser known optimization in the out-of-order execution engine can be a common cause of measurement bias on moderns Intel microarchitectures.
 
     % Approach
-    Starting with the high level known causes of bias from previous work, we analyze a minimal C-program that exhibits bias from environment size.
-    In order to understand why performance varies, we use hardware performance counters to obtain detailed statistics for different environments.
+    Our discovery stems from an analysis of a small program that was presented as a curiosity in previous work.
+The program shows clear bias in that for certain sizes of system environment variables, the performance is significantly worse than average.
+    Using hardware performance counters we identify the source of bias to be resource stalls from falsely detected dependencies in the out-of-order execution pipeline, reported as 4K address aliasing.
 
     % Results
-    We find that an effect known as 4K aliasing, causing resource stalls in the CPU execution pipeline, can be a common source of bias from external factors that alters run time addresses of program data.
-    For some applications, performance penalties from unfavorable memory address allocations can be as high as double the cycle count.
-    For ATLAS, an auto-tuned BLAS implementation, we estimate a potential 30\% performance variation in one matrix-vector multiplication routine, depending on the location in memory of each buffer.
+    We show how address aliasing can cause measurement bias for two particular external properties; environment variable size and dynamically linked heap allocation libraries.
+    Using a constructed example, we find that performance penalties from unfavorable memory address allocations can be as high as double the cycle count.
+    Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give worst case performance by default for large allocations.
 
     % Conclusions
-    With an understanding of 4K aliasing, we show that bias is in fact not completely unpredictable, and discuss measures for avoiding bad performance because of it.
+    % In principle anything that can alter run time addresses of program data can potentially result in bias from aliasing.
+    % We show that bias caused by address aliasing is in fact not completely unpredictable, and should be taken into account by optimizing 
     \end{abstract}
@@
 
+\iffalse
+
 \section{Optimizing Alignment in BLAS}
@@
 
+\fi
+
 \section{Conclusions}
-We have studied how address aliasing can affect program performance under different memory layouts. 
-This effect is caused by the way speculative and out-of-order memory operations are handled by the CPU, only considering the last 12 address bits to resolve conflicts load and store operations.
+In this paper we have looked at how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
+The effect is caused by the way speculative and out-of-order memory operations are handled by the CPU, only considering the last 12 address bits to resolve conflicts load and store operations.
 
-We have shown how aliasing can be the root cause of measurement bias from variations in environment size. 
-Analyzing the example provided in \ref{Mytkowicz:2009:WrongData}, we determined that collisions between stack variables and static data caused aliasing for certain stack positions.
-Variations in environment size changes virtual addresses of stack allocated data, and measurement bias observed by this can in some cases be attributed to address aliasing.
+Analyzing the example originally presented by Mytkowicz et al.
+\cite{Mytkowicz:2009:WrongData}, we determined that collisions between stack variables and static data resulted in aliasing for certain stack positions.
+In this case, the external trigger of bias is variations in environment size, which in turn offset the virtual addresses of stack allocated variables.
 
@@
 We present a code example with extreme sensitivity of data alignment, with more than 50 \% performance variation between different memory layouts.
+
 We find that compilers, even when provided with architecture specific optimization flags, still generates code that suffers from aliasing. 

### 2014-04-27 paper.tex
Scrap old introduction and start over

@@
     \begin{abstract} % Should be 150 -- 200 words max
+    % Motivation -- Problem statement -- Approach -- Results -- Conclusions
 
-    % Motivation
-    % Seemingly innocious properties of the execution context can cause subtle interactions with the low level hardware boundaries.
-    % Because of the vast complexity of modern hardware, these effects are found to be unpredictable and difficult to deal with.
-    % Programs are said to be \emph{biased} towards certain machines and configurations in various ways, .
+    In this paper we show how an optimization in the out-of-order execution engine on modern Intel microarchtectures can affect performance depending on run-time memory layout.
 
-    % Interaction with cache boundaries, branch prediction and other intricate details of the underlying hardware can make accurate performance measurements difficult.
+    We analyse a small program that was presented as a curiosity in previous work on \emph{measurement bias};
+    Code that for unexplained reasons performs significantly worse than average for certain configurations of system environment variables.
 
-    % Problem statement
-    In this paper we show how a lesser known optimization in the out-of-order execution engine can be a common cause of measurement bias on moderns Intel microarchitectures.
+    Using hardware performance counters, we identify a strong correlation with resource stalls, ultimately caused by address aliasing events.
+    The CPU uses an optimistic strategy when issuing operations out of order.
+    Only the last 12 address bits are compared to determine if there was a conflict, sometimes resulting in performance penalties from false dependencies.
 
-    % Approach
-    Our discovery stems from an analysis of a small program that was presented as a curiosity in previous work.
-The program shows clear bias in that for certain sizes of system environment variables, the performance is significantly worse than average.
-    Using hardware performance counters we identify the source of bias to be resource stalls from falsely detected dependencies in the out-of-order execution pipeline, reported as 4K address aliasing.
-
-    % Results
-    We show how address aliasing can cause measurement bias for two particular external properties; environment variable size and dynamically linked heap allocation libraries.
-    Using a constructed example, we find that performance penalties from unfavorable memory address allocations can be as high as double the cycle count.
+    We show how address aliasing can cause measurement bias for two particular external properties; environment variable size, and dynamically linked heap allocation libraries.
+    Performance penalties from unfavorable memory address allocations can be more than double the cycle count, even for real world applications.
     Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give worst case performance by default for large allocations.
 
-    % Conclusions
-    % In principle anything that can alter run time addresses of program data can potentially result in bias from aliasing.
-    % We show that bias caused by address aliasing is in fact not completely unpredictable, and should be taken into account by optimizing 
     \end{abstract}
@@
 \section{Introduction}
+\iffalse
+- Measurement bias in systems research
+  - What is measurement bias
+    - fundamentally anything external
+    - Room temperature
+  - Studied before
+    - External properties that affect memory layout
+      - link order (not really)
+      - env size
+  - Has been shown to be significant and commonplace
+    - But difficult to deal with
+    - Need to apply statistical methods to experiments
+      - randomization of experimental setups
+- Want to look at flipside: understand what why bias occurs
+  - Use for optimization
+  - Find interesting cases
+  - Known causes: cache alignment, ignore.
+- Analysis: 
+  - Use known example that exhibit bias
+  - Find lesser known source in address aliasing
+- Results:
+  - Applicable general knowledge about Intel architecture
+  - Useful for optimization
+  - Libraries and compilers fail to account for it in reality.
+\fi
+
+How fast does an algorithm run?
+Is program A faster than program B on my machine?
+These types of questions are, perhaps surprisingly, not always trivial to answer.
+The usual approach to quantify performance is to simply measure some metric, such as the wall time or number of cycles executed.
+However, performance is not just a function of the collection of instructions in an executable, but also dependent on various \emph{external} factors.
+In other fields of science, accounting for sampling bias is well understood.
+Trying to determine the political opinion of a country just by sampling a narrow geographical area will probably not be very accurate.
+The same concept applies for software; measuring a metric once is analogous to sampling only a tiny geographical area in the set of possible system configurations.
+
+A number of external variables has the potential to affect system performance.
+In particular, anything that has the potential to offset memory addresses of code or data at run-time can be significant.
+Memory accesses interacts with low level hardware buffers, which can heavily favor certain memory layouts over others.
+Consequentially, any environmental factor that can alter memory layout has a potential performance impact.
+Failing to account for these effects lead to \emph{measurement bias}.
+
+Measurement bias has been shown to be commonplace and significant in real applications.
+Previous work looks at environmental factors such as the size of Unix environment variables, which affects memory by offseting the initial position of stack.
+Because of the enormous complexety of modern hardware, it is often impossible to predict exactly how seemingly innocious things such as environment variables ultimately affect program execution.
+In fact, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
+\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}-
+
+Effects of bias has been shown to be commonplace and significant, yet unpredictable and difficult to deal with.
+Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods.
+Our approach is instead to try to understand some of the underlying reasons to why bias can occur. 
+Usually, the root explanation will be that different memory layouts interact with various hardware buffers.
+Certain mechanisms such as cache contention or branch misprediction are already well understood.
+In this paper, we show how a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline can be the cause of measurement bias.
+Analyzing a program that exhibits strong bias from environment size, we find a correlation with resource stalls from \emph{address aliasing}.
+The CPU uses a conservative heuristic for determining whether memory accesses can be retired out of order, comparing only the last 12 address bits.
+False dependencies can incur performance penalties, which can explain many common cases of measurement bias.
+
+In addition to environment variables, we look at dynamic heap allocation libraries as an external bias trigger.
+Heap allocation turns out to be interesting in that most libraries tend to produce worst case behavior by default with respect to aliasing.
+We show that the cost of address aliasing can be significant, and usually beyond the reach of optimizing compilers to take into account.
+However, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
+In fact, for some applications it might make sense for programmers to actively take aliasing into account for architecture specific optimization, as neither compilers nor heap allocators seem to handle it well.
+
+\subsection{Methodology}
+\iffalse
+ - Experimental setup
+  - Intel haswell architecture
+    - Restrict to a single cpu to limit scope
+ - Hardware performance counters
+   - MSR, architecture specific
+   - Closest thing to the CPU
+   - Use perf-stat
+ - Avoid observer effect
+ - Best practices for controlled experimental setup
+   - No ASLR, what it is
+   - No hyperthreading etc.
+\fi
+
+\iffalse
 Optimizing for performance is an important topic in computer science, and high performance computing (HPC) in particular. 
@@
 Characteristics of the operating system can affect things like the placement of code and data in memory, which in turn interacts with various hardware mechanisms at run time.
-Corner cases in hardware introduces potential performance cliffs; If for example a memory access happens to cross a page boundary, the cost of an additional TLB miss could be significant. 
+Corner cases in hardware introduces potential \emph{performance cliffs}; If for example a memory access happens to cross a page boundary, the cost of an additional TLB miss could be significant.
+
 Modern microprocessors are extremely complex in design and functionality; Some features of recent Intel processors includes several layers of cache to camouflage slow memory, multiple prefetchers, speculative out-of-order execution and branch prediction, just to name a few.
 Hardware features and optimizations interact with memory layout of program code and data in various ways, and in certain worst cases one can observe a performance degradation.
+
 We will not look at things like cache efficiency, which often can be the reason for bad performance under certain memory contexts. 
@@
 \subsection{Background}
-The set of external, and a-priori unknown, conditions under which a program executes is called the execution context. 
-Variations within contexts can lead to programs being biased towards certain configurations. 
-Measuring performance of the same program on two machines with identical hardware can sometimes yield very different results, an effect known as measurement bias.
+The set of external, and a-priori unknown, conditions under which a program executes is called the \emph{execution context}. 
+Performance variations within contexts lead to programs being biased towards certain configurations. 
+The same program under different contexts can sometimes yield very different results, an effect known as \emph{measurement bias}.
+
+A related concept of \emph{observer effect} also applies to computer systems.
+Any instrumentation or observation probes has the potential to alter the observed behavior.
 
@@
 Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
+\fi
 
@@
 Doing only a partial compare, \emph{false} dependencies can be detected when the addresses are equal $\bmod 4096$, as in the following example: 
-\begin{lstlisting}[language={[x86masm]Assembler}]
-mov %rax ($0x601010)
-mov ($0x604010) %rbx
+\begin{lstlisting}
+store (0x601010)
+load (0x604010)
 \end{lstlisting}

### 2014-04-29 paper.tex
Rewrite methodology section

@@
 However, performance is not just a function of the collection of instructions in an executable, but also dependent on various \emph{external} factors.
+
 In other fields of science, accounting for sampling bias is well understood.
@@
 
-A number of external variables has the potential to affect system performance.
-In particular, anything that has the potential to offset memory addresses of code or data at run-time can be significant.
+There are many external variables that has the potential to affect system performance.
+In particular, anything which can offset memory addresses of code or data at run-time can be significant.
 Memory accesses interacts with low level hardware buffers, which can heavily favor certain memory layouts over others.
@@
 In fact, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
-\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}-
+\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}
 
@@
 Analyzing a program that exhibits strong bias from environment size, we find a correlation with resource stalls from \emph{address aliasing}.
-The CPU uses a conservative heuristic for determining whether memory accesses can be retired out of order, comparing only the last 12 address bits.
-False dependencies can incur performance penalties, which can explain many common cases of measurement bias.
+In short, the CPU uses a conservative heuristic for determining whether memory accesses can be retired out of order, comparing only the last 12 virtual address bits.
+False dependencies incur performance penalties, which can explain many common cases of measurement bias.
 
 In addition to environment variables, we look at dynamic heap allocation libraries as an external bias trigger.
-Heap allocation turns out to be interesting in that most libraries tend to produce worst case behavior by default with respect to aliasing.
-We show that the cost of address aliasing can be significant, and usually beyond the reach of optimizing compilers to take into account.
-However, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
-In fact, for some applications it might make sense for programmers to actively take aliasing into account for architecture specific optimization, as neither compilers nor heap allocators seem to handle it well.
+Heap allocation turns out to be particularly interesting for the following reason: most libraries tend to produce worst case behavior by default with respect to aliasing.
+We show that the cost of address aliasing can be significant, exemplified with a simple program yielding more than 50~\% speedup based on heap address alignment alone.
+Fortunately, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
+For some applications it might even make sense for programmers to actively take aliasing into account for architecture specific optimization, as neither compilers nor heap allocators seem to handle it well.
 
@@
 \iffalse
- - Experimental setup
-  - Intel haswell architecture
-    - Restrict to a single cpu to limit scope
+ - Goal: Produce empirical data showing what causes bias
+ - Two pitfalls:
+  - Observer effect
+  - Measurement bias
  - Hardware performance counters
@@
    - Use perf-stat
- - Avoid observer effect
  - Best practices for controlled experimental setup
@@
    - No hyperthreading etc.
+ - Experimental setup
+  - Intel haswell architecture
+    - Restrict to a single cpu to limit scope
+ - Control cache metrics
 \fi
+In order to understand bias behavior it is important to be able to do precise measurements.
+Besides measurement bias in itself, one must also consider \emph{observer effect}.
+Any kind of instrumentation added to the code, such as a counter variable for each invocation of a function, has the potential to produce misleading results.
+Instead, we use instrumentation provided directly by the hardware; in the form of performance counters. 
+Modern processors have a set of model specific registers (MSRs) that can be configured to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
+Recent Intel architectures have several hundred different available events, which can give a very detailed view of what happens inside the CPU.
+
+Performance counters are supported in the Linux kernel, and via a tool called \texttt{perf}. 
+We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the reference manual \cite{Intel:2013:Volume3B}.
+A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
+Metrics related to cache, such as hit rates for different levels, are closely monitored in order to exclude cache as the underlying cause of bias.
+We attempt to isolate interesting events by computing correlation to cycle count, measuring all counters over a series of execution contexts.
+
+Results are averaged over multiple runs, and all benchmarks are done on a machine under minimal load.
+Our setup consists of an Intel® Core™ i7-4770K CPU @ 3.50 GHz, 32 GiB DDR3 1600 Mhz memory, running a 64-bit installation of Ubuntu 13.10.
+Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus only on the Intel ``Haswell'' microarchitecture specifically.
+Code samples are compiled using the GCC toolchain, version 4.8.1.
+
+We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe.
+Most importantly, this means keeping the memory address space under control.
+For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as Address Space Layout Randomization.\cite{Shackham:2004:ASLR} % Should maybe have more generic reference
+By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
+Bias from environment size is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.
+\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
+
+Finally, we disable Hyper-threading, which reduces the possibility for resource contention in the CPU between threads.
 

### 2014-05-01 paper.tex
Some inline notes on 4k aliasing

@@
 \section{Introduction}
-\iffalse
-- Measurement bias in systems research
-  - What is measurement bias
-    - fundamentally anything external
-    - Room temperature
-  - Studied before
-    - External properties that affect memory layout
-      - link order (not really)
-      - env size
-  - Has been shown to be significant and commonplace
-    - But difficult to deal with
-    - Need to apply statistical methods to experiments
-      - randomization of experimental setups
-- Want to look at flipside: understand what why bias occurs
-  - Use for optimization
-  - Find interesting cases
-  - Known causes: cache alignment, ignore.
-- Analysis: 
-  - Use known example that exhibit bias
-  - Find lesser known source in address aliasing
-- Results:
-  - Applicable general knowledge about Intel architecture
-  - Useful for optimization
-  - Libraries and compilers fail to account for it in reality.
-\fi
+% - Measurement bias in systems research
+%   - What is measurement bias
+%     - fundamentally anything external
+%     - Room temperature
+%   - Studied before
+%     - External properties that affect memory layout
+%       - link order (not really)
+%       - env size
+%   - Has been shown to be significant and commonplace
+%     - But difficult to deal with
+%     - Need to apply statistical methods to experiments
+%       - randomization of experimental setups
+% - Want to look at flipside: understand what why bias occurs
+%   - Use for optimization
+%   - Find interesting cases
+%   - Known causes: cache alignment, ignore.
+% - Analysis: 
+%   - Use known example that exhibit bias
+%   - Find lesser known source in address aliasing
+% - Results:
+%   - Applicable general knowledge about Intel architecture
+%   - Useful for optimization
+%   - Libraries and compilers fail to account for it in reality.
 
@@
 \subsection{Methodology}
-\iffalse
- - Goal: Produce empirical data showing what causes bias
- - Two pitfalls:
-  - Observer effect
-  - Measurement bias
- - Hardware performance counters
-   - MSR, architecture specific
-   - Closest thing to the CPU
-   - Use perf-stat
- - Best practices for controlled experimental setup
-   - No ASLR, what it is
-   - No hyperthreading etc.
- - Experimental setup
-  - Intel haswell architecture
-    - Restrict to a single cpu to limit scope
- - Control cache metrics
-\fi
+% - Goal: Produce empirical data showing what causes bias
+% - Two pitfalls:
+%  - Observer effect
+%  - Measurement bias
+% - Hardware performance counters
+%   - MSR, architecture specific
+%   - Closest thing to the CPU
+%   - Use perf-stat
+%   - Control cache metrics
+% - Best practices for controlled experimental setup
+%   - No ASLR, what it is
+%   - No hyperthreading, disable frequency scaling.
+% - Experimental setup
+%  - Intel haswell architecture
+%    - Restrict to a single cpu to limit scope
+
 In order to understand bias behavior it is important to be able to do precise measurements.
 Besides measurement bias in itself, one must also consider \emph{observer effect}.
-Any kind of instrumentation added to the code, such as a counter variable for each invocation of a function, has the potential to produce misleading results.
+Any kind of instrumentation added to the code, such as a counter variable for each invocation of a function, has the potential to produce misleading results~\cite{Mytkowicz:2008:OE&MB}.
 Instead, we use instrumentation provided directly by the hardware; in the form of performance counters. 
@@
 
-Performance counters are supported in the Linux kernel, and via a tool called \texttt{perf}. 
+Performance counters are supported in the Linux kernel, and via a tool called perf. 
 We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the reference manual \cite{Intel:2013:Volume3B}.
 A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
+Interesting events are isolated by computing correlation to cycle count, measuring all counters over a series of execution contexts.
+Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
 Metrics related to cache, such as hit rates for different levels, are closely monitored in order to exclude cache as the underlying cause of bias.
-We attempt to isolate interesting events by computing correlation to cycle count, measuring all counters over a series of execution contexts.
 
@@
 By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
-Bias from environment size is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.
-\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
+Bias from environment size is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
 
+Frequency scaling is disabled, keeping the processor's clock speed fixed.
 Finally, we disable Hyper-threading, which reduces the possibility for resource contention in the CPU between threads.
@@
 \section{4K Address Aliasing}
-An event known as “4K aliasing” can occur when the addresses of a store instruction followed by a load instruction differ by a multiple of 4 KiB.
+% Discussed in Optimization manual (pp. 144, 403, 564, 583)
+% - Said to mean many things
+%   - "False store forwarding" -- Only checks last 12 bits. Does it attempt to forward?)
+%   - Loads that have partial address match with preceding stores, causing the load to be reissued. 
+%     Estimate 5 cycle cost
+%   - "False dependencies in MOB due to partial compare on address"
+%   - Very clear separate event in VTune Amplifier documentation online 
+% - Detected by LOAD_BLOCK.OVERLAP_STORE on previous architectures (pp. 599)
+% - Present event code and description
+% Optimization guidelines
+% - Has some cost associated with it
+% - Intel suggests measures for mitigating cost
+%   - Change offsets between input and output buffers
+% Potentially bias inducing
+% - Anything that affects memory layout
+%   - Stack variable offset from env size
+%   - Heap addresses from dynamic allocation libraries
+
+% User/Source Coding Rule 8. (H impact, ML generality) Consider using a special memory allocation
+% library with address offset capability to avoid aliasing. .................................................. 3-55
+% User/Source Coding Rule 9. (M impact, M generality) When padding variable declarations to avoid
+% aliasing, the greatest benefit comes from avoiding aliasing on second-level cache lines, suggesting
+% an offset of 128 bytes or more ................................................................................... 3-55
+
+% There seems to be two separate issues here
+% 1: store-to-load forwarding speculative, then discover that the addresses are not the same. (pp 500)
+% 2: store-load issuing the load first speculatively. WAR conflict in MOB, reissue load. VTune event "4K alias"
+%
+% MOB: "Accepts and processes all memory operations". pp 62
+%      Allows issue load before preceding store, speculatively
+
+An event known as “4K aliasing” can occur when the addresses of a store instruction followed by a load instruction differ by a multiple of 4 KiB, or in other words; when the last 12 address bits are the same.
+
 On recent Intel architectures, only the last 12 address bits are used to determine if operations refer to the same address \cite{Intel:2012:OptimizationManual}.
@@
 Independent pairs of loads and stores where the last address bits differ can then be reordered and issued out of order.
+But of course this can only happen if the memory accesses are known to not conflict -- probably determined by MOB as 12 bit address compare.
+If correct, this means that 4K aliasing implies lost opportunity for out-of-order execution.
+It also appears that this mechanism is used to determine if store forwarding should be performed, 4K aliasing implies that store forwarding can be performed.
+In optimization manual states that 4k aliasing causes the load to be reissued, period.
+Again, that has to mean the load was attempted and something went wrong.
+It could read data from a store buffer, when it really should have read from L1. 
+So not really sure this has anything to do with reordering, but rather a failed assumption that the load can read the MOB value from the store. 
+The dependency is later discovered by the MOB to be wrong before retirement, and the load has to be reissued.
+
 Doing only a partial compare, \emph{false} dependencies can be detected when the addresses are equal $\bmod 4096$, as in the following example: 
@@
       \addlegendentry{cycles:u} ;
-      \addlegendentry{r0107:u} ;
+      %\addlegendentry{r0107:u} ;
     \end{axis}

### 2014-05-02 paper.tex
=Minor change to abstract only.

@@
 
-    In this paper we show how an optimization in the out-of-order execution engine on modern Intel microarchtectures can affect performance depending on run-time memory layout.
+	% Bad motivation start:
+    % In this paper we show how an optimization in the out-of-order execution engine on modern Intel microarchtectures can affect performance depending on run-time memory layout. 
+    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchtectures. 
+    The magnitude of this performance bias has been shown to be significant, and can be as high as a factor two.
 

### 2014-05-02 paper.tex
=Comments and changes to introduction. Fixed spelling some places.

@@
 
-    We analyse a small program that was presented as a curiosity in previous work on \emph{measurement bias};
+    We analyze a small program that was presented as a curiosity in previous work on \emph{measurement bias};
     Code that for unexplained reasons performs significantly worse than average for certain configurations of system environment variables.
@@
 How fast does an algorithm run?
-Is program A faster than program B on my machine?
+Is program A faster than program B on \emph{my} machine?
 These types of questions are, perhaps surprisingly, not always trivial to answer.
@@
 
-Measurement bias has been shown to be commonplace and significant in real applications.
-Previous work looks at environmental factors such as the size of Unix environment variables, which affects memory by offseting the initial position of stack.
-Because of the enormous complexety of modern hardware, it is often impossible to predict exactly how seemingly innocious things such as environment variables ultimately affect program execution.
-In fact, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
+Measurement bias has been shown to be commonplace and significant in real applications [move cite here?].
+Previous work looks at environmental factors such as the size of Unix environment variables, which affects memory by offsetting the initial position of stack.
+Because of the [enormous] complexity of modern hardware, it is often impossible to predict exactly how seemingly innocuous things such as environment variables ultimately affect program execution.
+[In fact], the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts [not selling our case here].
 \cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}
 
-Effects of bias has been shown to be commonplace and significant, yet unpredictable and difficult to deal with.
+Effects of bias has been shown to be commonplace and significant, yet unpredictable and difficult to deal with [cite/ref].
 Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods.
@@
 Certain mechanisms such as cache contention or branch misprediction are already well understood.
-In this paper, we show how a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline can be the cause of measurement bias.
+[We collect 200+ events using HW counters, and search for correlations.]
+[In this paper,] we show how a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline can [be the] cause of measurement bias.
 Analyzing a program that exhibits strong bias from environment size, we find a correlation with resource stalls from \emph{address aliasing}.
-In short, the CPU uses a conservative heuristic for determining whether memory accesses can be retired out of order, comparing only the last 12 virtual address bits.
+In short, the CPU uses a [conservative] heuristic for determining whether memory accesses can be retired out of order, comparing only the last 12 virtual address bits.
 False dependencies incur performance penalties, which can explain many common cases of measurement bias.
 
-In addition to environment variables, we look at dynamic heap allocation libraries as an external bias trigger.
-Heap allocation turns out to be particularly interesting for the following reason: most libraries tend to produce worst case behavior by default with respect to aliasing.
+In addition to environment variables, we look at dynamic heap allocation libraries as an external bias trigger [amplifier?].
+We find that [Heap allocation turns out to be particularly interesting for the following reason:] most libraries tend to produce worst case behavior by default with respect to aliasing.
 We show that the cost of address aliasing can be significant, exemplified with a simple program yielding more than 50~\% speedup based on heap address alignment alone.
+[This next part might be added to the abstract in some form:] 
 Fortunately, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
@@
 Results are averaged over multiple runs, and all benchmarks are done on a machine under minimal load.
-Our setup consists of an Intel® Core™ i7-4770K CPU @ 3.50 GHz, 32 GiB DDR3 1600 Mhz memory, running a 64-bit installation of Ubuntu 13.10.
+Our setup consists of an Intel® Core™ i7-4770K CPU @ 3.50 GHz, 32 GiB DDR3 1600 MHz memory, running a 64-bit installation of Ubuntu 13.10.
 Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus only on the Intel ``Haswell'' microarchitecture specifically.
@@
 We use perf to acquire hardware performance counter statistics.
-Measurements are aggregated over several runs to collect an extensice sample of about 200 performance counters and combinations provided on our architecture~\cite{Intel:2013:Volume3B}. 
+Measurements are aggregated over several runs to collect an extensive sample of about 200 performance counters and combinations provided on our architecture~\cite{Intel:2013:Volume3B}. 
 Unless otherwise specified, a minimal environment is used for all benchmarks. 
@@
 
-Table \ref{tab:mallocompare} illustrates some differences that can be observed by swithing heap allocator. \footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
+Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator. \footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
 The addresses of two equally sized char buffers allocated with malloc are observed for size parameters 5120 and $2^{20}$ bytes. 
@@
 \subsection{Aliasing in cblas\_daxpy}
-Found aliasing here as well. Two spikes, not sure why. Might want to test other BLAS libraries. Include or leave out? Too much handwaving and too little analysis maybe.
+Found aliasing here as well.
+Two spikes, not sure why.
+Might want to test other BLAS libraries.
+Include or leave out?
+Too much handwaving and too little analysis maybe.
+Depends on the target paper length as well?
 

### 2014-05-02 paper.tex
=Comments and changes to Methodology.

@@
 
-Performance counters are supported in the Linux kernel, and via a tool called perf. 
-We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the reference manual \cite{Intel:2013:Volume3B}.
-A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
-Interesting events are isolated by computing correlation to cycle count, measuring all counters over a series of execution contexts.
-Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
-Metrics related to cache, such as hit rates for different levels, are closely monitored in order to exclude cache as the underlying cause of bias.
+Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
+We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
+A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture. [Might add "and environmental sizes".]
+Interesting events are identified [isolated] by computing correlation to cycle count, measuring all counters over a series of execution contexts.
+[Both lines overlap:]
+Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of [micro-ops] for each level of cache~\cite{Intel:2012:OptimizationManual}.
+[Metrics related to cache, such as hit rates for different levels, are closely monitored in order to exclude cache as the underlying cause of bias.]
 
-Results are averaged over multiple runs, and all benchmarks are done on a machine under minimal load.
+Results are averaged [not median?] over multiple runs, and all benchmarks are done on a machine under minimal load.
 Our setup consists of an Intel® Core™ i7-4770K CPU @ 3.50 GHz, 32 GiB DDR3 1600 MHz memory, running a 64-bit installation of Ubuntu 13.10.
 Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus only on the Intel ``Haswell'' microarchitecture specifically.
-Code samples are compiled using the GCC toolchain, version 4.8.1.
+Code samples are compiled using the GCC toolchain, version 4.8.1 [Perhaps the full name? 4.4.6-11ubuntu2].
 
+[Add a link to a paper that lists good practice?: ] 
 We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe.
@@
 For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as Address Space Layout Randomization.\cite{Shackham:2004:ASLR} % Should maybe have more generic reference
-By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
+By disabling \emph{ASLR}, we are able to execute the same program multiple times with identical virtual address spaces.
 Bias from environment size is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
 
-Frequency scaling is disabled, keeping the processor's clock speed fixed.
-Finally, we disable Hyper-threading, which reduces the possibility for resource contention in the CPU between threads.
+\emph{Frequency scaling} is disabled, keeping the processor's clock speed fixed.
+Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention [in the CPU] between threads.
 
@@
 Optimizing for performance is an important topic in computer science, and high performance computing (HPC) in particular. 
-A great amount of research and effort is put into designing better algorithms and compiler optimizations in order to produce an ``optimal'' executable. 
+A great amount of research and effort is put into designing better algorithms and compiler optimizations in order to produce an ``optimal'' executable]. 
 However, performance of computer programs is not just a function of the sequence of machine instructions executed. 

### 2014-05-02 paper.tex
Update introduction and methodology
Fixes based on feedback

@@
 
-	% Bad motivation start:
-    % In this paper we show how an optimization in the out-of-order execution engine on modern Intel microarchtectures can affect performance depending on run-time memory layout. 
     In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchtectures. 
@@
 
-Measurement bias has been shown to be commonplace and significant in real applications [move cite here?].
+% This is basically former background section
+Measurement bias has been shown to be commonplace and significant in real applications \cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
 Previous work looks at environmental factors such as the size of Unix environment variables, which affects memory by offsetting the initial position of stack.
-Because of the [enormous] complexity of modern hardware, it is often impossible to predict exactly how seemingly innocuous things such as environment variables ultimately affect program execution.
-[In fact], the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts [not selling our case here].
-\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}
+Because of the complexity of modern hardware, it is often impossible to predict exactly how seemingly innocuous things such as environment variables ultimately affect program execution.
+While effects of bias has been shown to be commonplace and significant, they are unpredictable and difficult to deal with.
+Without an accurate understanding of the underlying memory interactions, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
+Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods.
+% Cite is basically the same as above, all three are about the same -- difficult to split.
 
-Effects of bias has been shown to be commonplace and significant, yet unpredictable and difficult to deal with [cite/ref].
-Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods.
-Our approach is instead to try to understand some of the underlying reasons to why bias can occur.
+Our approach is instead to try to understand some of the underlying reasons to why bias can occur, in an attempt to utilize this knowlege for potential optimization.
+% Could have a sentence with cite of Blind Optimization.
 Usually, the root explanation will be that different memory layouts interact with various hardware buffers.
-Certain mechanisms such as cache contention or branch misprediction are already well understood.
-[We collect 200+ events using HW counters, and search for correlations.]
-[In this paper,] we show how a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline can [be the] cause of measurement bias.
-Analyzing a program that exhibits strong bias from environment size, we find a correlation with resource stalls from \emph{address aliasing}.
-In short, the CPU uses a [conservative] heuristic for determining whether memory accesses can be retired out of order, comparing only the last 12 virtual address bits.
-False dependencies incur performance penalties, which can explain many common cases of measurement bias.
+Certain mechanisms such as cache contention or branch misprediction are already well understood, thus we are looking for the more esoteric processor features.
+Analyzing a program that exhibits strong bias from environment size, collecting about 200 events using hardware performance counters, we find strong correlation with resource stalls from \emph{address aliasing}.
+Aliasing is an artifact of a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline.
+In short, the CPU uses a heuristic for determining whether memory loads are dependent on previous writes, comparing only the last 12 virtual address bits.
+False dependencies incur performance penalties, and we show how this can explain many common cases of measurement bias.
 
-In addition to environment variables, we look at dynamic heap allocation libraries as an external bias trigger [amplifier?].
-We find that [Heap allocation turns out to be particularly interesting for the following reason:] most libraries tend to produce worst case behavior by default with respect to aliasing.
-We show that the cost of address aliasing can be significant, exemplified with a simple program yielding more than 50~\% speedup based on heap address alignment alone.
-[This next part might be added to the abstract in some form:] 
+In addition to environment variables, we look at dynamic heap allocation libraries as an external factor that can trigger (or amplify) bias resulting from address aliasing.
+We find that most libraries tend to produce worst case behavior by default with respect to aliasing, caused by favoring page alignment for large allocations.
+The cost of address aliasing can be significant, exemplified with a simple program yielding more than 50~\% speedup based on heap address alignment alone.
+% [This next part might be added to the abstract in some form:] 
 Fortunately, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
@@
 We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
-A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture. [Might add "and environmental sizes".]
-Interesting events are identified [isolated] by computing correlation to cycle count, measuring all counters over a series of execution contexts.
-[Both lines overlap:]
-Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of [micro-ops] for each level of cache~\cite{Intel:2012:OptimizationManual}.
-[Metrics related to cache, such as hit rates for different levels, are closely monitored in order to exclude cache as the underlying cause of bias.]
+A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
+The same script simulates changes to environment size, which is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
+Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
+Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
+Results are averaged over multiple runs to reduce potential random error, supplying the \texttt{-r} option to perf-stat. % Not median, good question. man perf-stat says average + std.dev.
 
-Results are averaged [not median?] over multiple runs, and all benchmarks are done on a machine under minimal load.
-Our setup consists of an Intel® Core™ i7-4770K CPU @ 3.50 GHz, 32 GiB DDR3 1600 MHz memory, running a 64-bit installation of Ubuntu 13.10.
-Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus only on the Intel ``Haswell'' microarchitecture specifically.
-Code samples are compiled using the GCC toolchain, version 4.8.1 [Perhaps the full name? 4.4.6-11ubuntu2].
+Our setup consists of an Intel\textregistered{} Core\texttrademark{} i7-4770K CPU @ 3.50 GHz, running 64-bit Ubuntu 13.10.
+Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus only on this processor and the Intel ``Haswell'' microarchitecture specifically.
+Code samples are compiled using the GCC toolchain, version 4.8.1-10ubuntu9.
 
-[Add a link to a paper that lists good practice?: ] 
-We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe.
+We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe, and in general follow best practices on gathering data \cite{Mytkowicz:2009:WrongData}.
 Most importantly, this means keeping the memory address space under control.
@@
 By disabling \emph{ASLR}, we are able to execute the same program multiple times with identical virtual address spaces.
-Bias from environment size is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
-
-\emph{Frequency scaling} is disabled, keeping the processor's clock speed fixed.
-Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention [in the CPU] between threads.
+All benchmarks are done on a machine under minimal load, and \emph{Frequency scaling} is disabled to keep the processor's clock speed fixed.
+Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads.
 

### 2014-05-02 paper.tex
Remove old commented out intro sections

@@
 Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads.
-
-\iffalse
-Optimizing for performance is an important topic in computer science, and high performance computing (HPC) in particular. 
-A great amount of research and effort is put into designing better algorithms and compiler optimizations in order to produce an ``optimal'' executable]. 
-However, performance of computer programs is not just a function of the sequence of machine instructions executed. 
-Various external properties, such as the operating system and user environment, can have have an impact on how fast programs execute. 
-Characteristics of the operating system can affect things like the placement of code and data in memory, which in turn interacts with various hardware mechanisms at run time.
-Corner cases in hardware introduces potential \emph{performance cliffs}; If for example a memory access happens to cross a page boundary, the cost of an additional TLB miss could be significant.
-
-Modern microprocessors are extremely complex in design and functionality; Some features of recent Intel processors includes several layers of cache to camouflage slow memory, multiple prefetchers, speculative out-of-order execution and branch prediction, just to name a few.
-Hardware features and optimizations interact with memory layout of program code and data in various ways, and in certain worst cases one can observe a performance degradation.
-
-We will not look at things like cache efficiency, which often can be the reason for bad performance under certain memory contexts. 
-Instead, we look at effects from less known optimizations in the out-of-order execution engine and instructions fetch pipeline. 
-We will unveil characteristics about one specific microarchitectural feature known as 4K aliasing, showing how false dependencies in the out-of-order execution pipeline can bias performance towards certain memory contexts.
-This effect can be triggered for example by changing stack position, and can explain bias from altering environment variables.
-With more accurate models of how bias occur, we will not only be able to actively avoid in performance, but in some cases also gain a real speedup on average.
-Because bias effects are intrinsically connected to various hardware features, this paper focuses on the 4th generation Intel® Core™ i7 ``Haswell'' architecture specifically. 
-However, the bias sources we present has also been shown to behave similarly on previous Intel architectures as well.
-
-\subsection{Background}
-The set of external, and a-priori unknown, conditions under which a program executes is called the \emph{execution context}. 
-Performance variations within contexts lead to programs being biased towards certain configurations. 
-The same program under different contexts can sometimes yield very different results, an effect known as \emph{measurement bias}.
-
-A related concept of \emph{observer effect} also applies to computer systems.
-Any instrumentation or observation probes has the potential to alter the observed behavior.
-
-Previous work has studied parameters such as the size of system environment variables and program link order \cite{Mytkowicz:2008:OE&MB}. 
-Both are found to potentially have a significant effect on performance, even in standardized benchmarks. 
-Environment variables contain various information, such as the name of the currently logged in user, home- and current directory path.
-In the paper appropriately named ``Producing Wrong Data Without Doing Anything Obviously Wrong'' by T.
-Mytkowicz et al.~\cite{Mytkowicz:2009:WrongData}, the authors show an examples of a program where performance varies significantly by changing merely the size of the environment.
-This means that the length of your name -- by extension you user login -- could in principle be the tipping point between good and bad performance.
-Unfortunately, the effects appear to be unpredictable, and therefore difficult to deal with. 
-This poses a problem for researchers and performance analysts, who will need to account for effects of measurement bias with more rigorous methodologies and statistical methods.
-
-In previous work, the focus has mostly been on how to mitigate effects of bias, and avoid drawing the wrong conclusions based on misleading measurements~\cite{Mytkowicz:2008:Easy, Mytkowicz:2009:WrongData}.
-The goal of this paper is to show not only that bias exists, but also \emph{why} it occurs.
-
-\subsection{Methodology}
-When studying bias effects, carefully controlling all relevant properties of the environment is crucial. 
-Previous work on measurement bias and observer effect describes a set of best practices for how the measurement infrastructure should be configured \cite{Mytkowicz:2009:WrongData, Mytkowicz:2008:OE&MB}.
-\begin{itemize}
-  \item Address space layout randomization (ASLR) is disabled.
-  \item Hyper-threading is disabled.
-  \item System load is kept at a minimum.
-  \item Automatic CPU frequency scaling is disabled.
-\end{itemize}
-Experimental hardware configuration consists of an Intel® Core™ i7-4770K @ 3.50 GHz, 32 GiB DDR3 memory running at 1600 MHz, on a 64-bit installation of Ubuntu 13.10.
-GCC version 4.8.1 is used as compiler toolchain.
-
-We use perf to acquire hardware performance counter statistics.
-Measurements are aggregated over several runs to collect an extensive sample of about 200 performance counters and combinations provided on our architecture~\cite{Intel:2013:Volume3B}. 
-Unless otherwise specified, a minimal environment is used for all benchmarks. 
-To measure statistics with n bytes added to the environment, a dummy variable is set to \(0^{n}\) (repeated zero characters n times).\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
-
-Cache conflicts can in many cases be a potential explanation for bias effects observed by changing memory layout.
-Although cache is often a likely explanation of performance variations, bias caused by cache issues is well understood and \emph{not} the focus of this paper.
-Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
-\fi
 

### 2014-05-02 paper.tex
Small fixes in methodology
Rephrasing and reordering.

@@
 % This is basically former background section
-Measurement bias has been shown to be commonplace and significant in real applications \cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
+Measurement bias has been shown to be commonplace and significant in real applications~\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
 Previous work looks at environmental factors such as the size of Unix environment variables, which affects memory by offsetting the initial position of stack.
@@
 
-Our setup consists of an Intel\textregistered{} Core\texttrademark{} i7-4770K CPU @ 3.50 GHz, running 64-bit Ubuntu 13.10.
-Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus only on this processor and the Intel ``Haswell'' microarchitecture specifically.
+Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus on the Intel ``Haswell'' microarchitecture specifically.
+Our setup consists of a 4th generation Intel\textsuperscript{\textregistered} Core\textsuperscript{\texttrademark} i7-4770K processor, running 64-bit Ubuntu 13.10.
 Code samples are compiled using the GCC toolchain, version 4.8.1-10ubuntu9.
 
-We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe, and in general follow best practices on gathering data \cite{Mytkowicz:2009:WrongData}.
+We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe, and in general follow best practices on gathering data~\cite{Mytkowicz:2009:WrongData}.
 Most importantly, this means keeping the memory address space under control.
-For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as Address Space Layout Randomization.\cite{Shackham:2004:ASLR} % Should maybe have more generic reference
-By disabling \emph{ASLR}, we are able to execute the same program multiple times with identical virtual address spaces.
-All benchmarks are done on a machine under minimal load, and \emph{Frequency scaling} is disabled to keep the processor's clock speed fixed.
+For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Shackham:2004:ASLR}.
+% Should maybe have more generic reference
+By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
+All benchmarks are done on a machine under minimal load, and \emph{frequency scaling} is disabled to keep the CPU's clock speed fixed.
 Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads.

### 2014-05-03 paper.tex
Slight rewrite of 4k alias section

@@
 
-\section{4K Address Aliasing}
+% === Random notes section on 4K alias ===
+% There seems to be two separate issues here
+% 1: store-to-load forwarding speculative, then discover that the addresses are not the same. (pp 500)
+% 2: store-load issuing the load first speculatively. WAR conflict in MOB, reissue load. VTune event "4K alias"
+%
+% MOB: "Accepts and processes all memory operations". pp 62
+%      Allows issue load before preceding store, speculatively
+
 % Discussed in Optimization manual (pp. 144, 403, 564, 583)
 % - Said to mean many things
-%   - "False store forwarding" -- Only checks last 12 bits. Does it attempt to forward?)
+%   - "False store forwarding" -- Only checks last 12 bits. Does it attempt to forward?) (pp. 564)
 %   - Loads that have partial address match with preceding stores, causing the load to be reissued. 
@@
 % library with address offset capability to avoid aliasing. .................................................. 3-55
+%   > Examples mention avoiding L1 and L2 cache conflicts.
 % User/Source Coding Rule 9. (M impact, M generality) When padding variable declarations to avoid
@@
 
-% There seems to be two separate issues here
-% 1: store-to-load forwarding speculative, then discover that the addresses are not the same. (pp 500)
-% 2: store-load issuing the load first speculatively. WAR conflict in MOB, reissue load. VTune event "4K alias"
-%
-% MOB: "Accepts and processes all memory operations". pp 62
-%      Allows issue load before preceding store, speculatively
+% Facts
+% * Loads can be issued before stores, speculating the store is not going to be to a conflicting address. (pp. 62)
+% * All memory operations go through the memory order buffer. (pp. 63)
+% * Memory disambiguation has to do with prediction. Predicted ok load from L1. (pp. 65)
+% * Example of speculative execution of loads, with mem. disambiguation (pp 131)
+% * Rule 56, 8 & 9. Aliasing case wrt Store forwarding. Claims 4K causes stall (not reissue) (pp. 141)
+% * Statement: 4K aliasing causes stall = included in LOAD_BLOCK.OVERLAP_STORE
+% * For Sandy Bridge, description of LD_BLOCKS_PARTIAL.ADDRESS_ALIAS states reissue. (pp. 583)
 
-An event known as “4K aliasing” can occur when the addresses of a store instruction followed by a load instruction differ by a multiple of 4 KiB, or in other words; when the last 12 address bits are the same.
+\section{4K Address Aliasing}
+% Mechanisms for increased throughput and parallelism of memory operations
+% In particular loads that might depend on earlier stores
+% - Worst case have to wait for store to retire
+% - Supports /speculative/ issue based on prediction on whether load conflicts with previous store.
+%   - Later verified (memory disambiguation)
+% - If refer to same address, take value from sotre buffers via /store forwarding/. 
+%   - Don't have to wait for store to finish
+% An event known as 4K aliasing occurs when the load and store addresses match in the lower 12 bits
+% - CPU detects /false/ dependencies between memory operations.
+% - Counted by event ...
+% - Sometimes refered to as /false store forwarding/. (pp. 564)
+% Optimization guidelines
+%   - Avoid store and non-dependent load with 4K between (rule 56)
+%   - Special memory allocator (rule 8) -- for 64 KiB alias
+%   - Padding variables (rule 9) -- For 64 KiB alias
+% - All classified as "High" impact.
+% - We demonstrate how this effect causes bias for memory allocator and store/load single
+% - How programmers can adjust layout manually
+Modern processors are \emph{superscalar}, and achieves parallelism by issuing multiple instructions simultaneously and out of order.
+One of the issues that can limit throughput is dependencies between a load and previous stores.
+Typical software consists of about 38~\% memory operations, with about two thirds of them being loads~\cite{Intel:2006:InsideICM:SmartMemoryAccess}.
+More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
+Loads can therefore be issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store.
+If the load and store locations are the same, the value can be \emph{forwarded} from the store before it retires.
 
-On recent Intel architectures, only the last 12 address bits are used to determine if operations refer to the same address \cite{Intel:2012:OptimizationManual}.
-This is a heuristic employed by the CPU to enable out of order execution of memory access operations.
-Independent pairs of loads and stores where the last address bits differ can then be reordered and issued out of order.
-But of course this can only happen if the memory accesses are known to not conflict -- probably determined by MOB as 12 bit address compare.
-If correct, this means that 4K aliasing implies lost opportunity for out-of-order execution.
-It also appears that this mechanism is used to determine if store forwarding should be performed, 4K aliasing implies that store forwarding can be performed.
-In optimization manual states that 4k aliasing causes the load to be reissued, period.
-Again, that has to mean the load was attempted and something went wrong.
-It could read data from a store buffer, when it really should have read from L1. 
-So not really sure this has anything to do with reordering, but rather a failed assumption that the load can read the MOB value from the store. 
-The dependency is later discovered by the MOB to be wrong before retirement, and the load has to be reissued.
-
-Doing only a partial compare, \emph{false} dependencies can be detected when the addresses are equal $\bmod 4096$, as in the following example: 
-\begin{lstlisting}
-store (0x601010)
-load (0x604010)
-\end{lstlisting}
-Situations like these can cause the pipeline to stall and wait for the operations to finish in proper order.
+While optimizations such as these are good on the average case, there are corner cases. 
+In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
+A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12 bit address suffix 0x020 is the same in both. 
+Dispite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
 The number of times this happens can be counted by the following performance counter:
@@
   \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
-  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.
+  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
 \end{description}
-Performance implications of 4K aliasing is discussed to some extent in the Intel optimization manual. 
-There are even some concrete suggestions for how avoid aliasing cases to improve performance (Rule 8 and 9)~\cite{Intel:2012:OptimizationManual}.
-In the following sections, we look at how address aliasing can explain some instances of measurement bias.
+
+As address aliasing depends on the memory addresses of loads and stores, environmental factors that affects memory has the potential to induce aliasing conditions.
+In the following sections, we show how performance penalties from address aliasing can be the root cause of measurement bias.
 
@@
 
-\begin{lstlisting}[float=h, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopkernel}, frame=lines]
+\begin{lstlisting}[float=h, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
 #define ALIAS(a, b) \

### 2014-05-04 paper.tex
Started working on env bias section

@@
 \usepackage{url}
+\usepackage{numprint}
 
@@
 \newcommand{\perfctr}[1] {
-  \small{\uppercase{#1}}
+  \scriptsize{\uppercase{#1}}
 }
@@
 For some applications it might even make sense for programmers to actively take aliasing into account for architecture specific optimization, as neither compilers nor heap allocators seem to handle it well.
+
+% Include preview of aliasing program early to provike interest. 
+% Also lightly referenced in introduction/background
+\begin{lstlisting}[float=t, language=C, caption={Micro-kernel succeptible to aliasing between static and automatic variables depending on environment sizes}, label={lst:loopkernel}, frame=lines]
+static int i, j, k;
+int main() {
+    int g = 0, inc = 1;
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc; 
+    }
+    return 0;
+}
+\end{lstlisting}
 
@@
 In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
-A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12 bit address suffix 0x020 is the same in both. 
+A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12 bit address suffix of 0x020 is the same in both. 
 Dispite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
@@
 \section{Bias from Environment Size}
-Unless the program explicitly accesses environment variables, it is typically not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack.
+% General intro on why environment triggers aliasing
+% - Image of memory organization, this can be generic.
+%   - Each area affected by randomization differently
+% - Default 16 bit stack alignment guarantee.
+% Analysis section
+% - Present data on correlation
+% - Plot graph of 2 x 4K areas. 
+% - Addresses of static variables from symbol table
+% - Addresses of automatic variables from asm output
+% - Outline different scenarios that can happen for aliasing with 1 or 2 variables
+% - Conclusion: was bias from false dependencies
+% Avoid aliasing for env size
+% - Show altered program
+% - Proof of concept, in theory possible for compilers to generate such code.
+
+Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack.
+Environment variables and program arguments are allocated in the stack section of virtual memory, before the first call frame.
+Changing environment variables will therefore offset the addresses of stack, and consequently all stack allocated variables.
+
+% Generic intro to memory layout (with focus on stack/env interaction)
 Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory. 
+The stack is normally placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64 bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables.
+There is still some level of stack alignment happening, independent of environment offsets. 
+Normally, the stack is aligned to 16 bytes, which will be enforced by the compiler. % [ref]
+Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
+Any one of these contexts could trigger aliasing issues between stack and some other area of memory remaining unaffected by environment size.
+This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data.
 
-\pgfplotstableread{bin/stack-offset.dat}{\stackoffsettable}
-\begin{figure*}[t]
-  \caption{Bias from environment size for program in Listing~\ref{lst:loopkernel}.
-Measure of average 100 cycle counts for 512 different environments, spread over two 4K periods.
-Compiled with no optimization.}
-  \label{fig:envbias}
-  \begin{tikzpicture}
-    \begin{axis}[
-        width = \textwidth,
-        height = 6cm,
-        font = \small,
-        xlabel=Bytes added to environment,
-        domain = 0:8192,
-        xtick = {0,1024,...,8192},
-        xmin = 0,
-        xmax = 8192,
-        legend pos = outer north east,
-        cycle list name = exotic
-      ]
-      %\draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
-      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
-      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
-      \addlegendentry{cycles:u} ;
-      %\addlegendentry{r0107:u} ;
-    \end{axis}
-  \end{tikzpicture}
-\end{figure*}
+Note that there is no clear relationship between environment size and stack location with address randomization enabled.
+However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurences of measurement bias indeed random.
+% In addition to affecting stack, ASLR also randomizes location of dynamically linked files (memory mapped), and initial position of heap.
+% Moving to the lower end of the address space, we find the text, data and bss sections, which contains program code and statically allocated data.
+% The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.
 
-The stack is normally placed close to the upper address 0x7fff'ffffffff.\footnote{Modern processors do not actually use the full 64 bit space, only the low order 47 bits are used for addressing memory}
-Environment variables and program arguments are allocated in the stack section, before the first call frame.
-Changing environment variables will therefore offset the addresses of all stack allocated data.
-There is still some level of stack alignment happening, independent of environment offsets. 
-The compiler can enforce alignment to some number of zero bits.
-As long as this number is less than 12, environment size can impact addresses of automatic variables $\bmod 4096$.
-
-\begin{figure}[float=h]
-  \caption{Important sections of virtual memory on a 64 bit process}
+\begin{figure}[float=t]
+  \caption{Organization of virtual memory for a 64 bit process.}
   \label{fig:virtualmemory}
@@
 
-Text and data sections contain program code and static data respectively.
-These regions are mapped towards the low en of the address spectrum, and the virtual addresses of variables allocated here will not change with environment size.
+\subsection{Analysis of micro-kernel}
+To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by T.
+Mytkowicz et.
+al.~\cite{Mytkowicz:2009:WrongData}, reproduced here in Listing~\ref{lst:loopkernel}.
+This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straight forward to analyze.
+Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
 
-Note that the initial position of stack will not depend on environment size this way when address randomization (ASLR) is enabled.
-(TODO: Does it offset from page boundary?
-That would be good, same alias behavior :)
+% Large plot of cycle count for two 4K stack sections.
+\pgfplotstableread{bin/stack-offset.dat}{\stackoffsettable}
+\begin{figure*}[t]
+  \caption{Bias from environment size for micro-kernel in Listing~\ref{lst:loopkernel}.
+Measured average of 100 cycle count samples for 512 different environments.
+Spikes show aliasing case, occurring once for each 4K period}
+  \label{fig:envbias}
+  \begin{tikzpicture}
+    \begin{axis}[
+        width = \textwidth,
+        height = 6cm,
+        font = \small,
+        xlabel=Bytes added to environment,
+        ylabel=Cycles,
+        domain = 0:8192,
+        xtick = {0,1024,...,8192},
+        xmin = 0,
+        xmax = 8192,
+        %legend pos = outer north east,
+        cycle list name = exotic
+      ]
+      %\draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
+      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
+      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
+      %\addlegendentry{cycles:u} ;
+      %\addlegendentry{r0107:u} ;
 
-Addresses of stack allocated variables can vary depending on environment size, while other segments of memory such as text and data remain constant over all environments.
+      % Could annotate area with correlation analysis, (spike - 50*16), + 100*16
+      % This is not correct atm, uses 100 samples with byte increment. 
+      %\draw [|<->|, gray] (axis cs:2392,720000)--(axis cs:3992,720000) node[right] {Analysis};
 
-To illustrate how this can cause bias, we revisit the example first presented in~\cite{Mytkowicz:2009:WrongData}, shown in Listing~\ref{lst:loopkernel}. 
-This example is interesting for several reasons. 
-The bias effects are significant and easily reproducible, while the example code is simple and easy to analyze.
-Still, no satisfactory explanation as to what causes bias was given in the original paper.
+    \end{axis}
+  \end{tikzpicture}
+\end{figure*}
 
-\begin{lstlisting}[float=h, language=C, caption={Microkernel succeptible to aliasing between static variables and automatic variables for certain environment sizes}, label={lst:loopkernel}, frame=lines]
-static int i, j, k;
-int main() {
-    int g = 0, inc = 1;
-    for (; g < 65536; g++) {
-        i += inc;
-        j += inc;
-        k += inc; 
-    }
-    return 0;
-}
-\end{lstlisting}
+Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
+We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
+A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
+The program is compiled with \small{GCC} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
 
-Their program contains five variables, two of which are stack allocated and i, j and k that are statically allocated.
-Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
-Sampling is done for every 16 byte increment of environment size, covering two 4K periods of addresses.
-A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
-The program is compiled with GCC using no optimization.
-Any optimization would likely disregard most of the function as redundant code, and reduce it to return zero immediately.
+There are clearly two worst cases, indicated by significant spikes at the 3192 and 7288 byte offsets.
+We sampled an extensive set of performance counters around the first spike, incrementing environment size by 1 instead of 16 to capture the aliasing case multiple times.
+A total of 100 environment offsets were measured, ranging from 3142 to 3242 bytes added.
+Events with the highest (absolute value) correlation to cycle count is listed in Table \ref{tab:loopcorrelation}, values close to $1$ or $-1$ indicate strong correlation.
 
-There are clearly two worst cases, indicated by significant spikes around the 3100 and 7100 byte offsets.
-We analyzed these cases by sampling an extensive set of performance counters and correlating them with cycle counts.
+% This table is a bit dubius, are correlation values really that interesting?
+\begin{table}[h]
+  \caption{Events with significant correlation to cycle count.}
+  \label{tab:loopcorrelation}
+  \npdecimalsign{.}
+  \nprounddigits{3}
+  \centering
+  \scriptsize
+  \begin{tabular}{|l|n{2}{3}|}
+    \toprule
+      LD_BLOCKS_PARTIAL.ADDRESS_ALIAS & 0.998961259046 \\ \midrule
+      CYCLE_ACTIVITY.CYCLES_LDM_PENDING & 0.998618840476 \\ \midrule
+      RESOURCE_STALLS.ANY & 0.998582438059 \\ \midrule
+      CPL_CYCLES.RING123 & 0.997305209986 \\ \midrule
+      UOPS_EXECUTED_PORT.PORT_2 & 0.995814514248 \\ \midrule
+      CYCLE_ACTIVITY.STALLS_L2_PENDING & 0.841261809569 \\ \midrule
+      UOPS_EXECUTED_PORT.PORT_3 & -0.9780404694 \\ \midrule
+      UOPS_EXECUTED_PORT.PORT_7 & -0.991034639943 \\ \midrule
+      UOPS_EXECUTED_PORT.PORT_4 & -0.995703596298 \\ \midrule
+      UOPS_EXECUTED_PORT.PORT_6 & -0.998681490923 \\ \midrule
+      UOPS_EXECUTED_PORT.PORT_0 & -0.998757079731 \\ \midrule
+      UOPS_EXECUTED_PORT.PORT_1 & -0.99886851364 \\ \midrule
+      UOPS_EXECUTED_PORT.PORT_5 & -0.998956858377 \\ \midrule
+      RESOURCE_STALLS.RS & -0.998963674679 \\
+    \bottomrule
+  \end{tabular}
+  \npnoround
+\end{table}
+
+% Todo: continue here
 The results show a high number of resource stalls when the spikes occur, indicated by high values for performance counters \small{RESOURCE\_STALLS.ANY}, \small{CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING}, \small{CYCLE\_ACTIVITY.CYCLE\_NO\_EXECUTE} and \small{RESOURCE\_STALLS.RS}.
@@
 
-\subsection{Finding Collisions}
 To be able to explain what is happening in this case we need to know the addresses of all variables at run time.
+The program contains five variables, two of which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} that are statically allocated.
 For statically allocated data, memory allocation is decided at compile time.
-The location of variables i, j and k can be found by looking at the symbol table in the ELF executable.
-Below is an excerpt of \lstinline$readelf -s$.
+We can find the location of variables i, j and k by looking at the symbol table in the ELF executable, below is an excerpt of \lstinline$readelf -s$.
 \begin{lstlisting}[float=h, basicstyle=\small]

### 2014-05-04 paper.tex
Work on micro-kernel analysis section
Changed table to contain worst/best case numbers instead
of just correlation number.

@@
 \usepackage{url}
-\usepackage{numprint}
 
@@
 \newcommand{\perfctr}[1] {
-  \scriptsize{\uppercase{#1}}
+  {\lowercase{#1}}
 }
@@
 A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
-The program is compiled with \small{GCC} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
+The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
 
@@
 A total of 100 environment offsets were measured, ranging from 3142 to 3242 bytes added.
-Events with the highest (absolute value) correlation to cycle count is listed in Table \ref{tab:loopcorrelation}, values close to $1$ or $-1$ indicate strong correlation.
+From this we extracted counters with clear correlation to cycle count.
+Table \ref{tab:loopcorrelation} shows numerical counter values for the average case of 3176 bytes added to the environment, compared to worst case of 3192 bytes.
+% Events with the highest (absolute value) correlation to cycle count is listed in Table \ref{tab:loopcorrelation}, values close to $1$ or $-1$ indicate strong correlation.
 
-% This table is a bit dubius, are correlation values really that interesting?
 \begin{table}[h]
@@
   \label{tab:loopcorrelation}
-  \npdecimalsign{.}
-  \nprounddigits{3}
   \centering
-  \scriptsize
-  \begin{tabular}{|l|n{2}{3}|}
+  \small
+  \newcolumntype{H}{>{\lrbox0}c<{\endlrbox}@{}}
+  \renewcommand{\tabcolsep}{2pt}
+  \begin{tabular}{lHrr}
     \toprule
-      LD_BLOCKS_PARTIAL.ADDRESS_ALIAS & 0.998961259046 \\ \midrule
-      CYCLE_ACTIVITY.CYCLES_LDM_PENDING & 0.998618840476 \\ \midrule
-      RESOURCE_STALLS.ANY & 0.998582438059 \\ \midrule
-      CPL_CYCLES.RING123 & 0.997305209986 \\ \midrule
-      UOPS_EXECUTED_PORT.PORT_2 & 0.995814514248 \\ \midrule
-      CYCLE_ACTIVITY.STALLS_L2_PENDING & 0.841261809569 \\ \midrule
-      UOPS_EXECUTED_PORT.PORT_3 & -0.9780404694 \\ \midrule
-      UOPS_EXECUTED_PORT.PORT_7 & -0.991034639943 \\ \midrule
-      UOPS_EXECUTED_PORT.PORT_4 & -0.995703596298 \\ \midrule
-      UOPS_EXECUTED_PORT.PORT_6 & -0.998681490923 \\ \midrule
-      UOPS_EXECUTED_PORT.PORT_0 & -0.998757079731 \\ \midrule
-      UOPS_EXECUTED_PORT.PORT_1 & -0.99886851364 \\ \midrule
-      UOPS_EXECUTED_PORT.PORT_5 & -0.998956858377 \\ \midrule
-      RESOURCE_STALLS.RS & -0.998963674679 \\
+       &  & 3176 & 3192 \\
+    \midrule
+      \perfctr{CPU_CLK_UNHALTED.THREAD_P} & cycles:u & 698,101 & 829,103 \\
+      \perfctr{CPU_CLK_THREAD_UNHALTED.REF_XCLK} & bus-cycles:u & 19,698 & 23,632 \\ 
+      \perfctr{LD_BLOCKS_PARTIAL.ADDRESS_ALIAS} & r0107:u & \bf{121} & \bf{327,908} \\ 
+      \perfctr{CYCLE_ACTIVITY.CYCLES_LDM_PENDING} & r02a3:u & 1,183,945 & 1,449,964 \\ 
+      \perfctr{RESOURCE_STALLS.ANY} & r01a2:u & 276,648 & 408,071 \\ 
+      \perfctr{CPL_CYCLES.RING123} & r025c:u & 690,895 & 832,806 \\ 
+      \perfctr{UOPS_EXECUTED_PORT.PORT_2} & r04a1:u & 336,035 & 345,766 \\ 
+      \perfctr{CYCLE_ACTIVITY.STALLS_L2_PENDING} & r05a3:u & 492,163 & 512,649 \\ 
+      %L2_RQSTS.RFO_MISS & r2224:u & 123 & 127 \\ 
+      %OFFCORE_REQUESTS.DEMAND_RFO & r04b0:u & 129 & 130 \\
+      %L2_RQSTS.ALL_RFO & re224:u & 245 & 245 \\ 
+      %OFFCORE_REQUESTS_OUTSTANDING.DEMAND_RFO & r0460:u & 120154 & 1067959 \\ % Completely random numbers
+      %L2_TRANS.RFO & r02f0:u & 249 & 250 \\ 
+      %DTLB_STORE_MISSES.STLB_HIT_4K & r2049:u & 294 & 294 \\ 
+      %L2_TRANS.L2_WB & r40f0:u & 402 & 395 \\
+
+      %L2_RQSTS.ALL_DEMAND_DATA_RD & re124:u & 1097 & 1091 \\ 
+      %MEM_LOAD_UOPS_RETIRED.L2_MISS & r10d1:u & 565 & 562 \\ 
+      %L2_RQSTS.RFO_HIT & r4224:u & 69 & 67 \\ 
+      %LSD.UOPS & r01a8:u & 1059131 & 1058564 \\
+
+      % Less extreme, but all ports change
+      \perfctr{UOPS_EXECUTED_PORT.PORT_3} & r08a1:u & 358,324 & 355,080 \\ 
+      \perfctr{UOPS_EXECUTED_PORT.PORT_7} & r80a1:u & 137,862 & 131,882 \\ 
+      \perfctr{UOPS_EXECUTED_PORT.PORT_4} & r10a1:u & 280,908 & 276,586 \\
+
+      \perfctr{UOPS_EXECUTED_PORT.PORT_6} & r40a1:u & 234,672 & 180,675 \\ 
+      \perfctr{UOPS_EXECUTED_PORT.PORT_0} & r01a1:u & 239,356 & 106,916 \\ 
+      \perfctr{UOPS_EXECUTED_PORT.PORT_1} & r02a1:u & 247,572 & 99,270 \\ 
+      \perfctr{UOPS_EXECUTED_PORT.PORT_5} & r20a1:u & 250,450 & 122,051 \\ 
+      \perfctr{RESOURCE_STALLS.RS} & r04a2:u & 272,833 & 135,900 \\
+
     \bottomrule
+
+    % \toprule
+    %   CPU_CLK_THREAD_UNHALTED.REF_XCLK & 0.99946409726 \\ \midrule
+    %   LD_BLOCKS_PARTIAL.ADDRESS_ALIAS & 0.998961259046 \\ \midrule
+    %   CYCLE_ACTIVITY.CYCLES_LDM_PENDING & 0.998618840476 \\ \midrule
+    %   RESOURCE_STALLS.ANY & 0.998582438059 \\ \midrule
+    %   CPL_CYCLES.RING123 & 0.997305209986 \\ \midrule
+    %   UOPS_EXECUTED_PORT.PORT_2 & 0.995814514248 \\ \midrule
+    %   CYCLE_ACTIVITY.STALLS_L2_PENDING & 0.841261809569 \\ \midrule
+    %   UOPS_EXECUTED_PORT.PORT_3 & -0.9780404694 \\ \midrule
+    %   UOPS_EXECUTED_PORT.PORT_7 & -0.991034639943 \\ \midrule
+    %   UOPS_EXECUTED_PORT.PORT_4 & -0.995703596298 \\ \midrule
+    %   UOPS_EXECUTED_PORT.PORT_6 & -0.998681490923 \\ \midrule
+    %   UOPS_EXECUTED_PORT.PORT_0 & -0.998757079731 \\ \midrule
+    %   UOPS_EXECUTED_PORT.PORT_1 & -0.99886851364 \\ \midrule
+    %   UOPS_EXECUTED_PORT.PORT_5 & -0.998956858377 \\ \midrule
+    %   RESOURCE_STALLS.RS & -0.998963674679 \\
+    % \bottomrule
   \end{tabular}
-  \npnoround
 \end{table}
 
-% Todo: continue here
-The results show a high number of resource stalls when the spikes occur, indicated by high values for performance counters \small{RESOURCE\_STALLS.ANY}, \small{CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING}, \small{CYCLE\_ACTIVITY.CYCLE\_NO\_EXECUTE} and \small{RESOURCE\_STALLS.RS}.
-We also find that \small{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS} correlates almost perfectly with cycle count, which can explain the high number of resource stalls with false dependency from aliasing.
+Notice the most extreme change from average case to worst case is clearly the number of alias events.
+If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
+The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
+Some of the other counters are less interesting; for example, bus- and cpl cycles, which will naturally vary with the total cycle count.
 
-To be able to explain what is happening in this case we need to know the addresses of all variables at run time.
-The program contains five variables, two of which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} that are statically allocated.
-For statically allocated data, memory allocation is decided at compile time.
-We can find the location of variables i, j and k by looking at the symbol table in the ELF executable, below is an excerpt of \lstinline$readelf -s$.
-\begin{lstlisting}[float=h, basicstyle=\small]
-37: 000000000060103c    4  (...)  25 i
-38: 0000000000601040    4  (...)  25 j
-39: 0000000000601044    4  (...)  25 k
-\end{lstlisting}
+% Need input here, does this make sense?
+On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270'000 to about 135'000.
+The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Intel:2013:Volume3B}.
+Fewer stalls in the aliasing case could indicate less contention on the reservation station.
+This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port events, as all but one of the in total 8 execution ports seem to have fever operations going in parallel.
+Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stays in the execution pipeline for a shorter time.
+Our assumption is that this is a by-product of stalls generated from aliasing.
 
-Need to observe run time addresses without introducing any observer effects.
-A small amount of assembly code was used to calculate the addresses of each stack allocated variable and output them directly using system calls, without impacting run time addresses.
-We find that the spike in cycle count occurs when the address of inc alias with i, where the first spike happens when inc is at 0x7fffffffe03c.
-Because the stack is aligned to 16 byte or 4 words, there are a couple of different scenarios that could have been observed here.
-Static area is fixed and covers 12 contiguous bytes (3 words) \lstinline$[x|x| |x]$.
-Automatic variables occupies 8 contiguous bytes on stack (2 words) [ | |x|x]. 
-The space of possible addresses for (inc, g) in this case is (0x7fffffffe03c, 0x7fffffffe038) - n*4096. 
-In this scenario, g will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of i, j or k.
-An even worse case with respect to the number of alias events is if allocation of static data allows collisions with both stack allocated variables.
-This can be achieved by reserving an extra 8 bytes to offset i, j into the 0x8, 0xc slots. 
+% More deep dive section (maybe separate headline...)
+These numbers clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
+For that we need to know the addresses of all variables at run time.
+The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
+Virtual memory layout of static data is decided at compile time, and we can find the location of these variables by looking at the symbol table in the ELF executable.
+We find the addresses to be \texttt{\&i} = 0x60103c, \texttt{\&j} = 0x601040, and \texttt{\&k} = 0x601044.\footnote{ELF symbol tables can be read using \texttt{readelf -s}}
+
+Observing addresses of stack allocated data at run time is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
+A small amount of assembly code was used to calculate the addresses of \texttt{g} and \texttt{inc} and outputing to stdout direcly using syscall.
+The instrumented program has the exact same bias to environment size, making no change to the stack allocation instructions and not affecting the addresses of static variables.
+We find that the spike in cycle count occurs precicely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe038 and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
+
+Because the stack is aligned to 16 bytes, or 4 words, there are a couple of different scenarios that could have been observed here.
+Static variables are fixed and covers 12 contiguous bytes (3 words), in our case the addresses end in 0xc, 0x0 and 0x4, leaving the 0x8 slot free.
+Automatic variables occupies 8 contiguous bytes on stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots..
+In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.
+A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots.
 While this will give significantly more alias counts, it has little effect on the number of cycles executed.
 
-Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment.
-Resource stalls are generated because of false dependencies between stack and static data, resulting in worse performance.
+In conclusion, we identified that address aliasing is the root cause of measurement bias from environment size for this program.
+Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between stack and static data.
 The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.
@@
 In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as address layout randomization. 
-Although one can not easily know if a collision is going to happen for a given environment, one can try to change the program to account for possible alias effects.
-The following strategy is a proof of concept of how alias-free code can be generated in this particular case.
-If the addresses do alias, branch to an alternative but semantically equivalent code path.
-Allocate a new set of variables to avoid aliasing.
+Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
+Shown in Listing \ref{lst:loopfixed} is a proof of concept of how alias-free code could be generated in this particular case.
+If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
+Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.
 
-\begin{lstlisting}[float=h, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
+\begin{lstlisting}[float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
 #define ALIAS(a, b) \
@@
 \end{lstlisting}
+
+While this particular solution is clearly not something one would want to do in practice, it at least shows that compilers and programmers \emph{could} take measures to account for aliasing.
+This becomes much more enticing when the performance penalties are more significant, which we find is often the case for aliasing heap allocations.
 

### 2014-05-06 paper.tex
Outline for heap alias section

@@
 \section{Bias from Heap Allocation}
-Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
+% Intro
+% - Bias from aliasing caused by accesses to any part of memory
+% - Dynamic memory usually from dynamically linked libs = bias
+% - More interesting: default behavior provokes aliasing 
+% Heap allocation alias by default
+% - Normal heap with brk/sbrk syscalls
+%   - Manage heap and increase on demand
+% - Different depending on size (and other parameters)
+%   - Use mmap for performance, usually large allocations
+%   - Posix function, anonymous memory mapping
+%   - Page aligned
+% - Normal addr. vs mmap addr.
+% - Makes no difference ++ everything is conceptually the same heap, but alias by default!
+% Aligned sequential access
+% - Intel warning, copy routine. 
+% - Trouble with aligned read/write buffers, creating constant conflicts in sliding window.
+% - More contrived example; convolution
+%   - Simple code disregarding endpoints
+%   - Manipulate start addresses by offseting pointers
+%   - Compiled with gcc O3, shows bad performance (graph) + discussion
+% - Table with lots of data points, showing actual numbers for each offset
+% Architectural optimization
+% - Modified example with restrict and -march=i7
+% - Other compilers: Clang (pretty easy), Icc (already used preview version I think...)
+% Bahavior of other Allocators
+% - Intro each allocator
+% - Present table
+% - Conclude; not good...
+
+Address aliasing can be caused by conflicting pairs of load and store operations to any part of memory.
 The previous section looked specifically at an example of collision between stack variables and static data.
@@
 Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
+
+\subsection{\texttt{malloc} Alias by Default}
 A typical malloc implementation uses two different strategies for how to allocate memory.
@@
 
+\subsection{Architectural Optimization}
 It is perhaps unfair to assume the compiler can fix this when only given a generic optimization flag, as aliasing effects is (presumably) an Intel-specific issue.

### 2014-05-06 analysis/heap-alias/conv.c
Update measurement infrastructure for heap bias case

@@ -1,0 +1,12 @@
+static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
+
+void convolve(int size, float *input, float *output)
+{
+    int i, j;
+    for (i = 2; i < size - 2; ++i)
+    {
+        output[i] = 0;
+        for (j = 0; j < 5; ++j)
+            output[i] += input[i-2+j] * kernel[j];
+    }
+}

### 2014-05-06 analysis/heap-alias/estimate.py

@@ -1,0 +1,30 @@
+#!/usr/bin/env python 
+import sys, argparse, numpy
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--longrun', type=argparse.FileType('r'))
+    parser.add_argument('--shortrun', type=argparse.FileType('r'))
+    parser.add_argument('-o', '--outfile', type=argparse.FileType('w'), default=sys.stdout)
+    parser.add_argument('-r', '--delta', type=int, default=1)
+    args = parser.parse_args()
+
+    # Propagate header
+    args.shortrun.readline()
+    args.outfile.write(args.longrun.readline())
+
+    # Split into ([name, mnemonic], [values]), discard correlation
+    # TODO: Check if correlation is there... BUGS CURRENTLY!
+    def splitline(line):
+        return (line.split(',')[:2], map(int, line.split(',')[3:]))
+
+    # Calculate estimate counter value based on two series
+    def diffline(onerun, manyruns, howmany):
+        return map(lambda (one, many): (many - one) // howmany, zip(onerun, manyruns))
+
+    # Read files in parallel and output estimate
+    for x, y in zip(args.shortrun, args.longrun):
+        header, onecount = splitline(x)
+        _, manycounts = splitline(y)
+        args.outfile.write(','.join(header + map(str, diffline(onecount, manycounts, args.delta))))
+        args.outfile.write('\n')

### 2014-05-06 analysis/heap-alias/main.c

@@ -2,18 +2,6 @@
 #include <stdio.h>
 
 #define N 0x100000
-// static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
-
-// void convolve(int size, float *input, float *output)
-// {
-//     int i, j;
-//     for (i = 2; i < size - 2; ++i)
-//     {
-//         output[i] = 0;
-//         for (j = 0; j < 5; ++j)
-//             output[i] += input[i-2+j] * kernel[j];
-//     }
-// }
 
 void convolve(int, float *, float *);
 

### 2014-05-07 util/correlate.py
Split scripts into simpler components

@@ -1,0 +1,37 @@
+#!/usr/bin/env python 
+import sys, argparse, numpy
+from scipy import stats
+
+# Calculate linear correlation of lperf formatted performance data. Argument
+# matches a mnemonic from input file, typically cycles:u.
+# From [name, mnemonic, 0, 1, 2, 3, ...]
+# To   [name, mnemonic, correlation, 0, 1, 2, 3, ...]
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--input', type=argparse.FileType('r'), default=sys.stdin)
+    parser.add_argument('-o', '--output', type=argparse.FileType('w'), default=sys.stdout)
+    parser.add_argument('event')
+    args = parser.parse_args()
+
+    # Consume and update header
+    header = args.input.readline().split(",")
+    header.insert(2, "Correlation")
+    args.output.write(",".join(header))
+
+    def splitline(line):
+        return (line.split(',')[:2], map(int, line.split(',')[2:]))
+
+    # Read all lines as [([name, mnemonic], [counts])]
+    events = map(splitline, args.input)
+
+    # Extract counts for event to calculate correlation with
+    reference = filter(lambda e: e[0][1] == args.event, events)[0][1]
+
+    # Calculate correlation for all events
+    numpy.seterr(invalid='ignore')
+    correlations = map(lambda (x): stats.pearsonr(reference, x)[0], map(lambda (x, y): y, events))
+
+    # Output rest
+    for ((names, counts), correlation) in zip(events, correlations):
+        args.output.write(",".join(map(str, names + [correlation] + counts)))
+        args.output.write('\n')

### 2014-05-07 util/estimate.py

@@ -1,6 +1,11 @@
 #!/usr/bin/env python 
 import sys, argparse, numpy
 
+# Calculate the estimated cost of a program when subtracting constant overhead.
+# Input is two performance measurement sets, t_few and t_many. Estimation is
+# t_estimate = (t_few - t_many) / (many - few). 
+# Input can optionally contain correlation [name, mnemonic, correlation, 0, 1, 2, 3, ...]
+# Output removes correlation
 if __name__ == '__main__':
     parser = argparse.ArgumentParser()
     parser.add_argument('--longrun', type=argparse.FileType('r'))
@@ -9,14 +14,14 @@ if __name__ == '__main__':
     parser.add_argument('-r', '--delta', type=int, default=1)
     args = parser.parse_args()
 
-    # Propagate header
+    # Propagate header, discarding correlation
     args.shortrun.readline()
-    args.outfile.write(args.longrun.readline())
+    args.outfile.write(args.longrun.readline().replace("Correlation,",""))
 
-    # Split into ([name, mnemonic], [values]), discard correlation
-    # TODO: Check if correlation is there... BUGS CURRENTLY!
+    # Split into ([name, mnemonic], [values])
     def splitline(line):
-        return (line.split(',')[:2], map(int, line.split(',')[3:]))
+        tokens = line.replace("Correlation,","").split(",")
+        return (tokens[:2], map(int, tokens[2:]))
 
     # Calculate estimate counter value based on two series
     def diffline(onerun, manyruns, howmany):

### 2014-05-07 util/lperf.py

@@ -228,8 +228,8 @@ def parseargs():
     parser.add_argument('-e', '--events', type=perfctrs, default="all", help="Comma separated list of event code or mnemonic, ex. cycles:u,r0107:u")
     parser.add_argument('-n', '--iterations', type=int, default=1)
     parser.add_argument('-r', '--repeat', type=int, default=1, help="Average results over multiple runs, using perf-stat -r")
-    parser.add_argument('-o', '--outfile', type=argparse.FileType('w'), default=sys.stdout)
-    parser.add_argument('--correlate', type=perfctr, default=None)
+    parser.add_argument('-o', '--output', type=argparse.FileType('w'), default=sys.stdout)
+    #parser.add_argument('--correlate', type=perfctr, default=None)
     parser.add_argument('--env-offset', type=int, default=0, help="Number of bytes initially added to environment")
     parser.add_argument('--env-increment', type=int, default=1, help="Number of characters added to environment each iteration")
     parser.add_argument('--enumerate', default=False, action='store_true', help="Iteration number as program argument")
@@ -247,7 +247,7 @@ def parseargs():
 # counters each invocation of perf because of register limitations.
 def benchmark(args):
     counters = args.events
-    data = { counter: {'count': [0] * args.iterations, 'correlation': 0 } for counter in args.events }
+    data = { counter: {'count': [0] * args.iterations } for counter in args.events }
 
     for x in range(args.iterations):
         argument = "" if not args.enumerate else str(x)
@@ -268,10 +268,10 @@ def benchmark(args):
 
             process.wait()
     
-    if args.correlate != None:
-        numpy.seterr(invalid='ignore')
-        for event in args.events:
-            data[event]['correlation'], _ = stats.pearsonr( data[args.correlate]['count'], data[event]['count'] )
+    # if args.correlate != None:
+    #     numpy.seterr(invalid='ignore')
+    #     for event in args.events:
+    #         data[event]['correlation'], _ = stats.pearsonr( data[args.correlate]['count'], data[event]['count'] )
 
     return data
 
@@ -279,13 +279,13 @@ if __name__ == '__main__':
     args = parseargs()
     data = benchmark(args)
 
-    args.outfile.write("Performance counter,Mnemonic,")
-    if args.correlate != None and args.iterations > 1:
-        args.outfile.write("Correlation,")
-    args.outfile.write(",".join(map(str, range(args.iterations))) + "\n")
+    args.output.write("Performance counter,Mnemonic,")
[…]

### 2014-05-07 util/pgfpconv.py
Update pgf converter

@@ -1,22 +1,28 @@
 #!/usr/bin/env python
 import sys
 
+# Convert output of lperf to a format that is easier to parse for the LaTeX 
+# pgfplots package.
 #
-# Script to convert output of lperf to a format that is easier to parse for
-# the LaTeX pgfplots package. 
+# From: [name, mnemonic {,correlation}?, 0, 1, 2, ...]
+#
+# To:     n     cycles:u    r0107:u   ...
+#         0        50          8
+#         1         6          1
+#
 # Usage: lperf <args> | pgfconv > output.dat
-
+#
 if __name__ == '__main__':
-	data = [[]]
-	for line in sys.stdin:
-		if line[0] == '#':
-			sys.stdout.write(line)
-		else:
-			row = line.strip().split(',')
-			header, numbers = row[1], row[3:]
-			data.append([header] + numbers)
+    data = [[]]
+    header = sys.stdin.readline()
+    startidx = header.split(',').index('0')
+    
+    for line in sys.stdin:
+        row = line.strip().split(',')
+        event, numbers = row[1], row[startidx:]
+        data.append([event] + numbers)
 
-	n = len(data[1])
-	data[0] = ["n"] + map(str, range(0, n - 1))
-	for i in range(n):
-		print '\t'.join(data[k][i] for k in range(len(data)))
+    n = len(data[1])
+    data[0] = ["n"] + map(str, range(0, n - 1))
+    for i in range(n):
+        print '\t'.join(data[k][i] for k in range(len(data)))

### 2014-05-07 paper.tex
Add conv table

@@
 \usepackage{url}
+\usepackage{rotating}
 
@@
 
-\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
+% Plot of cycles vs alias for estimated cost of O3 convolution
+\pgfplotstableread{bin/conv-all-estimate.dat}{\convolutiontable}
 \begin{figure}[t]
@@
 This is a speedup on top of compiler optimization O3, which is not able to account for aliasing effects.
+
+% Table with detailed performance data, estimated for O3
+\pgfplotstabletranspose[
+  col sep=comma,
+  columns={n, cycles:u,r0107:u,r02a3:u,r01a2:u,r05a3:u,r04a2:u,r20a1:u,r0860:u,r0160:u,r2049:u,r6049:u},
+  colnames from=n % Automapped to new 'colnames' column
+  ]\loadedtable{bin/conv-all-estimate.dat}
+
+\begin{sidewaystable*}[p]
+  \caption{Most interesting performance counters for O3 convolution code, estimated cost accounting for constant overhead of heap allocation in main. (see /resources for full table)}
+  \label{tab:convstats}
+  \pgfkeys{/pgf/number format/.cd,int trunc} % Output whole numbers
+  \pgfplotstabletypeset[
+    font=\normalsize,
+    columns={[index]0, [index]1, [index]2, [index]3, [index]4, [index]5, [index]11, [index]21}, % Select interesting offsets
+    columns/colnames/.style={string type,column name={}, column type={l}},
+    columns/colnames/.append style={string replace={cycles:u}{Cycles}}, % Pretty print counter mnemonics
+    columns/colnames/.append style={string replace={r0107:u}{Alias}},
+    columns/colnames/.append style={string replace={r02a3:u}{cycle_activity.cycles_ldm_pending}},
+    columns/colnames/.append style={string replace={r01a2:u}{resource_stalls.any}},
+    columns/colnames/.append style={string replace={r05a3:u}{cycle_activity.stalls_l2_pending}},
+    columns/colnames/.append style={string replace={r04a2:u}{resource_stalls.rs}},
+    columns/colnames/.append style={string replace={r20a1:u}{uops_executed_port.port_5}},
+    columns/colnames/.append style={string replace={r0860:u}{offcore_requests_outstanding.all_data_rd}},
+    columns/colnames/.append style={string replace={r0160:u}{offcore_requests_outstanding.demand_data_rd}},
+    columns/colnames/.append style={string replace={r2049:u}{dtlb_store_misses.stlb_hit_4k}},
+    columns/colnames/.append style={string replace={r6049:u}{dtlb_store_misses.stlb_hit}},
+    every head row/.style={before row=\toprule, after row=\midrule},
+    every last row/.style={after row=\bottomrule}
+    ]{\loadedtable}
+\end{sidewaystable*}
 

### 2014-05-08 analysis/heap-alias/conv-restrict.c
Cleanup in heap alias experiments
Remove old files, and add some new compiler optimizations
Graph of 4k range show expected flat performance.

@@ -1,0 +1,12 @@
+static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
+
+void convolve(int size, const float * restrict input, float * restrict output)
+{
+    int i, j;
+    for (i = 2; i < size - 2; ++i)
+    {
+        output[i] = 0;
+        for (j = 0; j < 5; ++j)
+            output[i] += input[i-2+j] * kernel[j];
+    }
+}

### 2014-05-08 analysis/heap-alias/convolution-optimized.c

@@ -1,41 +1,0 @@
-#include <stdlib.h>
-#include <stdio.h>
-
-// For automated testing, set -DOFFSET
-#ifndef OFFSET
-    #define OFFSET 0x00
-#endif
-
-// cc -O3 convolution.c; cc -O3 convolution.c -S; perf stat -e cycles:u,r0107:u,r01a2:u,instructions:u -r 10 ./a.out
-// Best result with -O3 -std=c99 -march=native and no offset
-// -- Then we can save 250k alias and 300k RS with +0x30. Exact same instruction count (but sometimes slightly higher?)
-
-#define N 0x100000
-static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
-
-void convolve(int size, const float * restrict input, float * restrict output)
-{
-    int i, j;
-    for (i = 2; i < size - 2; ++i)
-    {
-        output[i] = 0;
-        for (j = 0; j < 5; ++j)
-            output[i] += input[i-2+j] * kernel[j];
-    }
-}
-
-int main(int argc, char **argv)
-{
-    // Needed for automated test script
-    int offset = argc > 1 ? atoi(argv[1]) : 0, i = 0;
-
-    float *input  = malloc( N          * sizeof(float));
-    float *result = malloc((N + 0x100) * sizeof(float));
-
-    //for (i = 0; i < 2; i++)
-    convolve(N, input, (result + offset));
-
-    printf("Offset: %d : (%p, %p)\n", offset, input, result);
-
-    return 0;
-}

### 2014-05-08 analysis/heap-alias/convolution.c

@@ -1,44 +1,0 @@
-#include <stdlib.h>
-#include <stdio.h>
-
-#define N 0x100000
-// static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
-
-// void convolve(int size, float *input, float *output)
-// {
-//     int i, j;
-//     for (i = 2; i < size - 2; ++i)
-//     {
-//         output[i] = 0;
-//         for (j = 0; j < 5; ++j)
-//             output[i] += input[i-2+j] * kernel[j];
-//     }
-// }
-
-void convolve(int, float *, float *);
-
-int main(int argc, char **argv)
-{
-    // Needed for automated test script
-    int repeat = argc > 1 ? atoi(argv[1]) : 0; // Repeat function invocation
-    int offset = argc > 2 ? atoi(argv[2]) : 0; // Offset between buffers
-    int i;
-
-    float *input  = malloc( N          * sizeof(float));
-    float *output = malloc((N + 0x100) * sizeof(float));
-
-    // To make a realistic test, initialize the otherwise all-zero input array
-    for (srand(0), i = 0; i < N; ++i) {
-        input[i] = (float)rand();
-    }
-
-    // Repeat a variable number of times, to be able to subtract the 
-    // constant overhead from allocating and initializing input arrays.
-    for (i = 0; i < repeat; ++i) {
-        convolve(N, input, (output + offset));
-    }
-
-    //printf("Offset: %d : (%p, %p), repeat %d\n", offset, input, output, repeat);
-
-    return 0;
-}

### 2014-05-08 analysis/heap-alias/main.c

@@ -13,7 +13,7 @@ int main(int argc, char **argv)
     int i;
 
     float *input  = malloc( N          * sizeof(float));
-    float *output = malloc((N + 0x100) * sizeof(float));
+    float *output = malloc((N + 0x400) * sizeof(float)); // Room for 4k offset
 
     // To make a realistic test, initialize the otherwise all-zero input array
     for (srand(0), i = 0; i < N; ++i) {

### 2014-05-08 util/lplot.py
Fix plot for missing correlation

@@ -61,8 +61,9 @@ def plot(args):
     args.input.readline()
     for line in args.input:
         e = map(lambda s : s.strip(), line.split(','))
+        start = 3 if "Correlation" in e else 2
         if args.events == ['all'] or e[1].lower() in args.events:
-            numbers = map(float, e[3:])
+            numbers = map(float, e[start:])
             plt.plot( \
                 [args.stride*v for v in range(len(numbers))], \
                 numbers, \

### 2014-05-08 util/lplot.py
Fix plotting take 2

@@ -58,10 +58,10 @@ def plot(args):
         locator = MultipleLocator(args.ticks)
         plt.gca().xaxis.set_major_locator(locator)
 
-    args.input.readline()
+    header = args.input.readline().split(",")
+    start = 3 if "Correlation" in header else 2
     for line in args.input:
         e = map(lambda s : s.strip(), line.split(','))
-        start = 3 if "Correlation" in e else 2
         if args.events == ['all'] or e[1].lower() in args.events:
             numbers = map(float, e[start:])
             plt.plot( \

### 2014-05-08 paper.tex
Update heap measurements in paper

--- 2014-05-06 paper.tex
@@
 % - Modified example with restrict and -march=i7
-% - Other compilers: Clang (pretty easy), Icc (already used preview version I think...)
 % Bahavior of other Allocators
@@
 % - Present table
-% - Conclude; not good...
+% Conclude
+% - not good...
+% - Not tested other compilers, not the purpose of this paper. 
+% - How did we find this example? Intelligent design, literally first guess
 
@@
 Another scenario to consider is collisions in heap allocated memory.
+Heap allocators are typically dynamically linked, and therefore part of the execution context.
 In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
+
+\subsection{Alias by Default with \texttt{malloc}}
 Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
+A typical malloc implementation uses two different strategies for how to allocate memory, depeding on the size of the request.
+The \emph{heap} area in virtual memory is used for smaller allocations.
+Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk system calls.
+For larger allocations, instead of growing the regular heap it is typically faster to use anonymous memory mapping by calling the Posix function \texttt{mmap}.
 
-\subsection{\texttt{malloc} Alias by Default}
-A typical malloc implementation uses two different strategies for how to allocate memory.
-The \emph{heap} area in virtual memory is used for smaller allocations.
-Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
-For larger allocations, it is typically faster to use anonymous memory mapping by calling mmap, instead of growing the ``normal'' heap.
-The heap starts at a low address close to the data section and grows towards higher addresses. 
-Mmap allocates much closer to the stack.
+The heap starts at a relatively low address close to the data section, and grows upwards.
+Memory mapped chunks from mmap are placed much closer to the stack.
 While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
@@
 
-\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
+% Plot of cycles vs alias for estimated cost of O3 convolution
+\pgfplotstableread{bin/default-o3.estimate.dat}{\convolutiontable}
 \begin{figure}[t]
@@
 Because large heap allocations always alias, the processor thinks memory accesses to input[i] potentially conflicts with output[i], stalling the pipeline.
-By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\% (verify).
+By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\%.
 This is a speedup on top of compiler optimization O3, which is not able to account for aliasing effects.
+
+% Table with detailed performance data, estimated for O3
+\pgfplotstabletranspose[
+  col sep=comma,
+  columns={n, cycles:u,r0107:u,r02a3:u,r01a2:u,r05a3:u,r04a2:u,r20a1:u,r0860:u,r0160:u,r2049:u,r6049:u},
+  colnames from=n % Automapped to new 'colnames' column
+  ]\loadedtable{bin/default-o3.estimate.dat}
+
+\begin{sidewaystable*}[p]
+  \caption{Most interesting performance counters for O3 convolution code, estimated cost accounting for constant overhead of heap allocation in main. (see /resources for full table)}
+  \label{tab:convstats}
+  \pgfkeys{/pgf/number format/.cd,int trunc} % Output whole numbers
+  \pgfplotstabletypeset[
+    font=\normalsize,
+    columns={[index]0, [index]1, [index]2, [index]3, [index]4, [index]5, [index]11, [index]21, [index]31}, % Select interesting offsets
+    columns/colnames/.style={string type,column name={}, column type={l}},
+    columns/colnames/.append style={string replace={cycles:u}{Cycles}}, % Pretty print counter mnemonics
+    columns/colnames/.append style={string replace={r0107:u}{Alias}},
+    columns/colnames/.append style={string replace={r02a3:u}{cycle_activity.cycles_ldm_pending}},
+    columns/colnames/.append style={string replace={r01a2:u}{resource_stalls.any}},
+    columns/colnames/.append style={string replace={r05a3:u}{cycle_activity.stalls_l2_pending}},
+    columns/colnames/.append style={string replace={r04a2:u}{resource_stalls.rs}},
+    columns/colnames/.append style={string replace={r20a1:u}{uops_executed_port.port_5}},
+    columns/colnames/.append style={string replace={r0860:u}{offcore_requests_outstanding.all_data_rd}},
+    columns/colnames/.append style={string replace={r0160:u}{offcore_requests_outstanding.demand_data_rd}},
+    columns/colnames/.append style={string replace={r2049:u}{dtlb_store_misses.stlb_hit_4k}},
+    columns/colnames/.append style={string replace={r6049:u}{dtlb_store_misses.stlb_hit}},
+    every head row/.style={before row=\toprule, after row=\midrule},
+    every last row/.style={after row=\bottomrule}
+    ]{\loadedtable}
+\end{sidewaystable*}
 

### 2014-05-11 util/median.py
New utility scripts, median and select

@@ -1,0 +1,24 @@
+#!/usr/bin/env python 
+import sys, argparse, numpy
+
+# Calculate median value of each counter
+#
+# From [name, mnemonic, [correlation], 0, 1, 2, 3, ...]
+# To   [name, mnemonic, median]
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--input', type=argparse.FileType('r'), default=sys.stdin)
+    parser.add_argument('-o', '--output', type=argparse.FileType('w'), default=sys.stdout)
+    args = parser.parse_args()
+
+    # Consume and update header
+    header = args.input.readline().split(",")
+    start = header.index("0")
+    args.output.write(",".join(header[:2] + ["Median\n"]))
+
+    def transform(line):
+        return line.split(',')[:2] + [int(numpy.median(map(int, line.split(',')[start:])))]
+
+    # Read all lines as [[name, mnemonic, median]]
+    for row in map(transform, args.input):
+        args.output.write(",".join(map(str, row)) + "\n")

### 2014-05-11 util/select.py

@@ -1,0 +1,18 @@
+#!/usr/bin/env python 
+import sys, argparse, numpy
+
+# Filter rows matching perf counter mnemonic
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+    parser.add_argument("--input", type=argparse.FileType("r"), default=sys.stdin)
+    parser.add_argument("-o", "--output", type=argparse.FileType("w"), default=sys.stdout)
+    parser.add_argument("-e", "--events")
+    args = parser.parse_args()
+
+    events = args.events.split(",")
+
+    args.output.write(args.input.readline())
+
+    for row in args.input:
+        if row.split(",")[1] in events:
+            args.output.write(row)

### 2014-05-11 util/lperf.py
Remove commented code

@@ -229,7 +229,6 @@ def parseargs():
     parser.add_argument('-n', '--iterations', type=int, default=1)
     parser.add_argument('-r', '--repeat', type=int, default=1, help="Average results over multiple runs, using perf-stat -r")
     parser.add_argument('-o', '--output', type=argparse.FileType('w'), default=sys.stdout)
-    #parser.add_argument('--correlate', type=perfctr, default=None)
     parser.add_argument('--env-offset', type=int, default=0, help="Number of bytes initially added to environment")
     parser.add_argument('--env-increment', type=int, default=1, help="Number of characters added to environment each iteration")
     parser.add_argument('--enumerate', default=False, action='store_true', help="Iteration number as program argument")
@@ -243,16 +242,16 @@ def parseargs():
 
     return args
 
-# Measure all counters under x different environments. Sample at most 4
-# counters each invocation of perf because of register limitations.
 def benchmark(args):
     counters = args.events
     data = { counter: {'count': [0] * args.iterations } for counter in args.events }
 
+    # Measure all counters under n different environments
     for x in range(args.iterations):
         argument = "" if not args.enumerate else str(x)
         env = {'X': '0' * (args.env_offset + x*args.env_increment)}
 
+        # Sample at most 4 events at a time because of register limitations
         for i in range(0, len(counters), 4):
             current = counters[i:i + 4]
 
@@ -267,11 +266,6 @@ def benchmark(args):
                 sys.stderr.write(line)
 
             process.wait()
-    
-    # if args.correlate != None:
-    #     numpy.seterr(invalid='ignore')
-    #     for event in args.events:
-    #         data[event]['correlation'], _ = stats.pearsonr( data[args.correlate]['count'], data[event]['count'] )
 
     return data
 
@@ -280,12 +274,8 @@ if __name__ == '__main__':
     data = benchmark(args)
 
     args.output.write("Performance counter,Mnemonic,")
-    # if args.correlate != None and args.iterations > 1:
-    #     args.output.write("Correlation,")
     args.output.write(",".join(map(str, range(args.iterations))) + "\n")
 
     for event in args.events:
         row = [event.name, event.mnemonic] + data[event]['count']
-        # if args.correlate != None and args.iterations > 1:
-        #     row.insert(2, data[event]['correlation'])
         args.output.write(",".join(map(str, row)) + "\n")

### 2014-05-11 analysis/env-alias/printf-analysis.c
Improve loop kernel analysis
Fix bug with offsets being wrong (not multiple of 16), and use
median and select on a master set of data.

@@ -1,31 +1,0 @@
-#include <stdio.h>
-
-static int i, j, k;
-
-static char* str = "%p, %p\n";
-
-void output(void* a, void* b);
-
-int main() {
-    int g = 0, inc = 1;
-
-    for (; g < 65536; g++) {
-        i += inc;
-        j += inc;
-        k += inc;
-    }
-    output(&g, &inc);
-/*
-    asm __volatile__ (
-    	"mov $1 %0\n"
-    	""
-    	:
-    	: "r"(g), "r"(inc)
-    	: );
-*/
-    return 0;
-}
-
-void output(void* a, void* b) {
-    printf(str, a, b);
-}

### 2014-05-12 paper.tex
Replace data behind microkernel plot and table
Link directly to analysis/../results, with minor post-processing
from parent makefile.

@@
 % Large plot of cycle count for two 4K stack sections.
-\pgfplotstableread{bin/stack-offset.dat}{\stackoffsettable}
+\pgfplotstableread{bin/micro-kernel-cycles.dat}{\stackoffsettable}
 \begin{figure*}[t]
@@
         xmax = 8192,
-        %legend pos = outer north east,
         cycle list name = exotic
       ]
-      %\draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
       \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
-      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
-      %\addlegendentry{cycles:u} ;
-      %\addlegendentry{r0107:u} ;
-
-      % Could annotate area with correlation analysis, (spike - 50*16), + 100*16
-      % This is not correct atm, uses 100 samples with byte increment. 
-      %\draw [|<->|, gray] (axis cs:2392,720000)--(axis cs:3992,720000) node[right] {Analysis};
-
     \end{axis}
@@
 
-There are clearly two worst cases, indicated by significant spikes at the 3192 and 7288 byte offsets.
-We sampled an extensive set of performance counters around the first spike, incrementing environment size by 1 instead of 16 to capture the aliasing case multiple times.
-A total of 100 environment offsets were measured, ranging from 3142 to 3242 bytes added.
-From this we extracted counters with clear correlation to cycle count.
-Table \ref{tab:loopcorrelation} shows numerical counter values for the average case of 3176 bytes added to the environment, compared to worst case of 3192 bytes.
-% Events with the highest (absolute value) correlation to cycle count is listed in Table \ref{tab:loopcorrelation}, values close to $1$ or $-1$ indicate strong correlation.
+There are clearly two worst cases, indicated by significant spikes at the 3200 and 7296 byte offsets.
+We sampled an extensive set of performance counters in addition to cycle count for each execution context.
+By looking at the \emph{median} value of each counter compared to the extreme cases, we narrowed down the performance events that behaved similarly to cycle count.
+Table \ref{tab:loopcorrelation} shows numerical counter values for median over all environments, compared to worst cases of 3200 and 7296 bytes added.
 
+% Detailed comparison between median and worst cases.
 \begin{table}[h]
@@
   \centering
-  \small
-  \newcolumntype{H}{>{\lrbox0}c<{\endlrbox}@{}}
-  \renewcommand{\tabcolsep}{2pt}
-  \begin{tabular}{lHrr}
-    \toprule
-       &  & 3176 & 3192 \\
-    \midrule
-      \perfctr{CPU_CLK_UNHALTED.THREAD_P} & cycles:u & 698,101 & 829,103 \\
-      \perfctr{CPU_CLK_THREAD_UNHALTED.REF_XCLK} & bus-cycles:u & 19,698 & 23,632 \\ 
-      \perfctr{LD_BLOCKS_PARTIAL.ADDRESS_ALIAS} & r0107:u & \bf{121} & \bf{327,908} \\ 
-      \perfctr{CYCLE_ACTIVITY.CYCLES_LDM_PENDING} & r02a3:u & 1,183,945 & 1,449,964 \\ 
-      \perfctr{RESOURCE_STALLS.ANY} & r01a2:u & 276,648 & 408,071 \\ 
-      \perfctr{CPL_CYCLES.RING123} & r025c:u & 690,895 & 832,806 \\ 
-      \perfctr{UOPS_EXECUTED_PORT.PORT_2} & r04a1:u & 336,035 & 345,766 \\ 
-      \perfctr{CYCLE_ACTIVITY.STALLS_L2_PENDING} & r05a3:u & 492,163 & 512,649 \\ 
-      %L2_RQSTS.RFO_MISS & r2224:u & 123 & 127 \\ 
-      %OFFCORE_REQUESTS.DEMAND_RFO & r04b0:u & 129 & 130 \\
-      %L2_RQSTS.ALL_RFO & re224:u & 245 & 245 \\ 
-      %OFFCORE_REQUESTS_OUTSTANDING.DEMAND_RFO & r0460:u & 120154 & 1067959 \\ % Completely random numbers
-      %L2_TRANS.RFO & r02f0:u & 249 & 250 \\ 
-      %DTLB_STORE_MISSES.STLB_HIT_4K & r2049:u & 294 & 294 \\ 
-      %L2_TRANS.L2_WB & r40f0:u & 402 & 395 \\
 
-      %L2_RQSTS.ALL_DEMAND_DATA_RD & re124:u & 1097 & 1091 \\ 
-      %MEM_LOAD_UOPS_RETIRED.L2_MISS & r10d1:u & 565 & 562 \\ 
-      %L2_RQSTS.RFO_HIT & r4224:u & 69 & 67 \\ 
-      %LSD.UOPS & r01a8:u & 1059131 & 1058564 \\
+  \renewcommand{\tabcolsep}{1pt} % Less padding between cells
+  \pgfkeys{/pgf/number format/.cd,int trunc} % Output whole numbers
 
-      % Less extreme, but all ports change
-      \perfctr{UOPS_EXECUTED_PORT.PORT_3} & r08a1:u & 358,324 & 355,080 \\ 
-      \perfctr{UOPS_EXECUTED_PORT.PORT_7} & r80a1:u & 137,862 & 131,882 \\ 
-      \perfctr{UOPS_EXECUTED_PORT.PORT_4} & r10a1:u & 280,908 & 276,586 \\
-
-      \perfctr{UOPS_EXECUTED_PORT.PORT_6} & r40a1:u & 234,672 & 180,675 \\ 
-      \perfctr{UOPS_EXECUTED_PORT.PORT_0} & r01a1:u & 239,356 & 106,916 \\ 
-      \perfctr{UOPS_EXECUTED_PORT.PORT_1} & r02a1:u & 247,572 & 99,270 \\ 
-      \perfctr{UOPS_EXECUTED_PORT.PORT_5} & r20a1:u & 250,450 & 122,051 \\ 
-      \perfctr{RESOURCE_STALLS.RS} & r04a2:u & 272,833 & 135,900 \\
-
-    \bottomrule
-
-    % \toprule
-    %   CPU_CLK_THREAD_UNHALTED.REF_XCLK & 0.99946409726 \\ \midrule
-    %   LD_BLOCKS_PARTIAL.ADDRESS_ALIAS & 0.998961259046 \\ \midrule
-    %   CYCLE_ACTIVITY.CYCLES_LDM_PENDING & 0.998618840476 \\ \midrule
-    %   RESOURCE_STALLS.ANY & 0.998582438059 \\ \midrule
-    %   CPL_CYCLES.RING123 & 0.997305209986 \\ \midrule
-    %   UOPS_EXECUTED_PORT.PORT_2 & 0.995814514248 \\ \midrule
-    %   CYCLE_ACTIVITY.STALLS_L2_PENDING & 0.841261809569 \\ \midrule
-    %   UOPS_EXECUTED_PORT.PORT_3 & -0.9780404694 \\ \midrule
-    %   UOPS_EXECUTED_PORT.PORT_7 & -0.991034639943 \\ \midrule
-    %   UOPS_EXECUTED_PORT.PORT_4 & -0.995703596298 \\ \midrule
-    %   UOPS_EXECUTED_PORT.PORT_6 & -0.998681490923 \\ \midrule
-    %   UOPS_EXECUTED_PORT.PORT_0 & -0.998757079731 \\ \midrule
-    %   UOPS_EXECUTED_PORT.PORT_1 & -0.99886851364 \\ \midrule
-    %   UOPS_EXECUTED_PORT.PORT_5 & -0.998956858377 \\ \midrule
-    %   RESOURCE_STALLS.RS & -0.998963674679 \\
-    % \bottomrule
-  \end{tabular}
+  \pgfplotstabletypeset[
+    font=\scriptsize,
+    col sep=comma,
+    columns={Performance counter, Median, [index]3, [index]4},
+    column type=r,
+    columns/Performance counter/.style={
+      string type, 
+      column type=l,
+      postproc cell content/.append code={
+        \pgfkeysalso{@cell content=\perfctr{##1}}
+      }
+    },
+    every head row/.style={
+      output empty row,
+      before row={\toprule
+        Performance counter & Median & Spike 1 & Spike 2 \\
+        },
+      after row=\midrule,
+    },
+    every last row/.style={after row=\bottomrule}
+  ]{bin/micro-kernel-comparison.csv}
 \end{table}
@@
 \begin{sidewaystable*}[p]
-  \caption{Most interesting performance counters for O3 convolution code, estimated cost accounting for constant overhead of heap allocation in main. (see /resources for full table)}
+  \caption{Most interesting performance counters for O3 convolution code, estimated cost accounting for constant overhead of heap allocation in main.}
   \label{tab:convstats}

### 2014-05-12 util/pgfpconv.py

@@ -4,25 +4,24 @@ import sys
 # Convert output of lperf to a format that is easier to parse for the LaTeX 
 # pgfplots package.
 #
-# From: [name, mnemonic {,correlation}?, 0, 1, 2, ...]
+# From: [name, mnemonic {,correlation}?, a, b, c, ...]
 #
 # To:     n     cycles:u    r0107:u   ...
-#         0        50          8
-#         1         6          1
+#         a        50          8
+#         b         6          1
 #
 # Usage: lperf <args> | pgfconv > output.dat
 #
 if __name__ == '__main__':
-    data = [[]]
-    header = sys.stdin.readline()
-    startidx = header.split(',').index('0')
+    header = sys.stdin.readline().strip().split(",")
+    startidx = 3 if "Correlation" in header else 2
+    data = [["n"] + header[startidx:]]
     
     for line in sys.stdin:
         row = line.strip().split(',')
         event, numbers = row[1], row[startidx:]
         data.append([event] + numbers)
 
-    n = len(data[1])
-    data[0] = ["n"] + map(str, range(0, n - 1))
+    n = len(data[0])
     for i in range(n):
         print '\t'.join(data[k][i] for k in range(len(data)))

### 2014-05-13 analysis/env-alias/recursion-fix.c
Remove random multiply by 10

@@ -20,7 +20,7 @@ int main() {
     if (ALIAS(inc, i) || ALIAS(g, i))
         return main();
 
-    for (; g < 65536*10; g++) {
+    for (; g < 65536; g++) {
         i += inc;
         j += inc;
         k += inc;

### 2014-05-13 paper.tex
Add assembly listing inline

@@
 
-Notice the most extreme change from average case to worst case is clearly the number of alias events.
+Notice the most extreme change from median to worst case is clearly the number of alias events.
 If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
@@
 Some of the other counters are less interesting; for example, bus- and cpl cycles, which will naturally vary with the total cycle count.
+These counters such as these are obviously not indicative of any causal relationship, and are ommitted.
 
@@
 Fewer stalls in the aliasing case could indicate less contention on the reservation station.
-This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port events, as all but one of the in total 8 execution ports seem to have fever operations going in parallel.
+This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports seem to have fewer operations going in parallel.
 Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stays in the execution pipeline for a shorter time.
@@
 We find that the spike in cycle count occurs precicely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe038 and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
+
+% Refer to relevant section of the generated assembly file (in gas syntax)
+\lstinputlisting[
+  linerange={22-38}, 
+  numbers=left,
+  firstnumber=22
+  ]{bin/micro-kernel-annotated.s}
 
@@
 
+% See analysis for results showing this does not alias
 \begin{lstlisting}[float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]

### 2014-05-13 analysis/heap-alias/conv.c
Add O2 targets and set const input buffer default

@@ -1,6 +1,6 @@
 static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
 
-void convolve(int size, float *input, float *output)
+void convolve(int size, const float *input, float *output)
 {
     int i, j;
     for (i = 2; i < size - 2; ++i)

### 2014-05-13 paper.tex
Improved handling of raw data for convolution example
More sane way of importing and drawing table/graph

@@
   \centering
-
   \renewcommand{\tabcolsep}{1pt} % Less padding between cells
-  \pgfkeys{/pgf/number format/.cd,int trunc} % Output whole numbers
-
   \pgfplotstabletypeset[
     font=\scriptsize,
+    int detect, % Output whole numbers for counter values
     col sep=comma,
@@
 % - More interesting: default behavior provokes aliasing 
+% - Intel warning, copy routine. 
+% - Trouble with aligned read/write buffers, creating constant conflicts in sliding window.
 % Heap allocation alias by default
@@
 % - Makes no difference ++ everything is conceptually the same heap, but alias by default!
-% Aligned sequential access
-% - Intel warning, copy routine. 
-% - Trouble with aligned read/write buffers, creating constant conflicts in sliding window.
+% Convolution kernel Analysis
 % - More contrived example; convolution
@@
 % Architectural optimization
-% - Modified example with restrict and -march=i7
+% - Modified example with restrict and -march=native
 % Bahavior of other Allocators
@@
 
-\subsection{Aligned Sequential Access}
+\subsection{Analysis of Convolution kernel}
 Consider the function shown in Listing~\ref{lst:conv}.
@@
 % Plot of cycles vs alias for estimated cost of O3 convolution
-\pgfplotstableread{bin/default-o3.estimate.dat}{\convolutiontable}
+\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convolutiontable}
 \begin{figure}[t]
@@
 
-\begin{lstlisting}[float=t, language=C, caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}, label={lst:conv}, frame=lines]
-static float k[5]={0.1,0.25,0.3,0.25,0.1};
-void conv(int n, float *in, float *out) {
-    int i, j;
-    for (i = 2; i < n-2; ++i) {
-      out[i] = 0;
-      for (j = 0; j < 5; ++j)
-          out[i] += in[i-2+j] * k[j];
-    }
-}
-\end{lstlisting}
+\lstinputlisting[
+  float=t, 
+  language=C, 
+  caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.},
+  label={lst:conv},
+  frame=lines
+]{bin/convolution-kernel.c}
 
@@
 
-% Table with detailed performance data, estimated for O3
-\pgfplotstabletranspose[
-  col sep=comma,
-  columns={n, cycles:u,r0107:u,r02a3:u,r01a2:u,r05a3:u,r04a2:u,r20a1:u,r0860:u,r0160:u,r2049:u,r6049:u},
-  colnames from=n % Automapped to new 'colnames' column
-  ]\loadedtable{bin/default-o3.estimate.dat}
-
-\begin{sidewaystable*}[p]
+% Table with detailed performance data, estimated for O3 (sidewaystable)
+\begin{table*}[t]
   \caption{Most interesting performance counters for O3 convolution code, estimated cost accounting for constant overhead of heap allocation in main.}
   \label{tab:convstats}
-  \pgfkeys{/pgf/number format/.cd,int trunc} % Output whole numbers
+
+  \renewcommand{\tabcolsep}{3pt} % Less padding between cells
   \pgfplotstabletypeset[
-    font=\normalsize,
-    columns={[index]0, [index]1, [index]2, [index]3, [index]4, [index]5, [index]11, [index]21, [index]31}, % Select interesting offsets
-    columns/colnames/.style={string type,column name={}, column type={l}},
-    columns/colnames/.append style={string replace={cycles:u}{Cycles}}, % Pretty print counter mnemonics
-    columns/colnames/.append style={string replace={r0107:u}{Alias}},
-    columns/colnames/.append style={string replace={r02a3:u}{cycle_activity.cycles_ldm_pending}},
-    columns/colnames/.append style={string replace={r01a2:u}{resource_stalls.any}},
-    columns/colnames/.append style={string replace={r05a3:u}{cycle_activity.stalls_l2_pending}},
-    columns/colnames/.append style={string replace={r04a2:u}{resource_stalls.rs}},
-    columns/colnames/.append style={string replace={r20a1:u}{uops_executed_port.port_5}},
-    columns/colnames/.append style={string replace={r0860:u}{offcore_requests_outstanding.all_data_rd}},
-    columns/colnames/.append style={string replace={r0160:u}{offcore_requests_outstanding.demand_data_rd}},
-    columns/colnames/.append style={string replace={r2049:u}{dtlb_store_misses.stlb_hit_4k}},
-    columns/colnames/.append style={string replace={r6049:u}{dtlb_store_misses.stlb_hit}},
-    every head row/.style={before row=\toprule, after row=\midrule},
+    font=\footnotesize,
+    int detect, % Output whole numbers for counter values
+    col sep=comma,
+    columns={Performance counter, Correlation, 0, 2, 4, 8, 16, 24},
+    column type=r,
+    columns/Performance counter/.style={
+      string type, 
+      column type=l,
+      postproc cell content/.append code={
+        \pgfkeysalso{@cell content=\perfctr{##1}}
+      }
+    },
+    columns/Correlation/.style={
+      fixed,
+      fixed zerofill,
+      precision=2
+    },
+    every head row/.style={
+      output empty row,
+      before row={\toprule
+        Performance counter & $r$ & 0 & 2 & 4 & 8 & 16 & 24 \\
+        },
+      after row=\midrule,
+    },
     every last row/.style={after row=\bottomrule}
-    ]{\loadedtable}
-\end{sidewaystable*}
+  ]{bin/conv-default-o3.estimate.csv}
+\end{table*}
 

### 2014-05-14 paper.tex
Start rewrite intro on heap alias

@@
 % More deep dive section (maybe separate headline...)
+\subsection{The Offending Variables}
 These numbers clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
@@
 % - Modified example with restrict and -march=native
+% - Vectorization -- probably less loads that causes reduction mostly
 % Bahavior of other Allocators
@@
 Address aliasing can be caused by conflicting pairs of load and store operations to any part of memory.
-The previous section looked specifically at an example of collision between stack variables and static data.
-Another scenario to consider is collisions in heap allocated memory.
-Heap allocators are typically dynamically linked, and therefore part of the execution context.
-In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
+In the previous sections we saw collisions between static data and stack, observing bias from external conditions that affected addresses of automatic variables.
+Most dynamic memory is allocated on the \emph{heap}, which is managed by an \emph{allocator}.
+A heap allocator is responsible for managing dynamic memory, and ultimately assigns the actual addresses of heap allocated variables at run time.
+Heap allocation routines such as \texttt{malloc} and \texttt{free} are typically dynamically linked, for example as part of glibc, or some other library.
+The particular library used therefore constitutes an important part of the execution context, as linking to a different library, or a library with some alternative configuration, can impact heap addresses at run time.
 
-\subsection{Alias by Default with \texttt{malloc}}
-Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
-A typical malloc implementation uses two different strategies for how to allocate memory, depeding on the size of the request.
-The \emph{heap} area in virtual memory is used for smaller allocations.
-Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk system calls.
-For larger allocations, instead of growing the regular heap it is typically faster to use anonymous memory mapping by calling the Posix function \texttt{mmap}.
+\subsection{\texttt{malloc} will Alias by Default}
+Acquiring dynamic memory at run time is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
+A typical malloc implementation uses different strategies for how to allocate memory depending on the size of the request.
+\begin{itemize}
+  \item The ``regular'' heap area in virtual memory is used for smaller allocations. 
+        Programs are initially given a relatively small heap, which the allocator can increase by calling \texttt{brk} or \texttt{sbrk} system calls.
+        The heap section starts at a relatively low address (close to the data section), and grows upwards.
+  \item For larger allocations, instead of growing the regular heap it is typically faster to use anonymous memory mapping by calling the Posix function \texttt{mmap}.
+      Memory mapped chunks are typically placed towards the upper end of the virtual address space, closer to the stack.
+\end{itemize}
+Addresses for smaller requests that are placed in the regular heap can look something like 0x16e30a0 or 0x1723020, while pointers returned by \texttt{mmap} can for example be 0x7f0318a8f010 or 0x7f03105d2010.
+This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
+However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
+The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}
 
-The heap starts at a relatively low address close to the data section, and grows upwards.
-Memory mapped chunks from mmap are placed much closer to the stack.
-While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
-This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
-However, mmap has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}
+\subsection{Aligned Sequential Access}
+Many functions operate on pairs of buffers in a ``sliding window'' fashion; reading from one and writing the result to another in some loop construct.
+This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointers are aliased and continuously generating false conflicts.
+This effect is mentioned in the Intel optimization manual, exemplified by an aliasing copy routine.
 
-\subsection{Analysis of Convolution kernel}
-Consider the function shown in Listing~\ref{lst:conv}.
-It computes the convolution between an input array and a fixed five-element kernel, writing the result to another array.
-For simplicity, endpoints are not handled.
+\lstinputlisting[
+  float=t, 
+  language=C, 
+  caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.},
+  label={lst:conv},
+  frame=lines
+]{bin/convolution-kernel.c}
+
+The function from Listing \ref{lst:conv} is a hand crafted naive implementation of linear correlation, containing lots of interleaved loads and stores while iterating over two independent memory buffers (for simplicity, endpoints are not handled).
+We show that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12 bits.
+For heap allocated memory, this depends entirely on what addresses the heap allocator provides.
+
+To force different alignments, we allocate some extra bytes at the end and offset the output pointer argument.
+
+\begin{lstlisting}
+output = malloc(n * (sizeof(float) + offset));
+conv(n, input, output + offset);
+\end{lstlisting}
+
+Figure \ref{fig:heapalias} shows how this function behaves when given input with different alignment.
+The default behavior for large heap allocations is that both arrays alias.
 
@@
 \end{figure}
-
-\lstinputlisting[
-  float=t, 
-  language=C, 
-  caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.},
-  label={lst:conv},
-  frame=lines
-]{bin/convolution-kernel.c}
-
-Figure \ref{fig:heapalias} shows how this function behaves when given input with different alignment.
-The default behavior for large heap allocations is that both arrays alias.
-To force different alignments, we allocate some extra bytes at the end and offset the output pointer argument.
-
-\begin{lstlisting}
-output = malloc(n * (sizeof(float) + offset));
-conv(n, input, output + offset);
-\end{lstlisting}
 

### 2014-05-15 paper.tex
More work on heap alias section
Add O2 together with O3 in the discussion, more clear to present
data for O2 in a table form.

@@
 % Detailed comparison between median and worst cases.
-\begin{table}[h]
+\begin{table}[t]
   \caption{Events with significant correlation to cycle count.}
@@
 Some of the other counters are less interesting; for example, bus- and cpl cycles, which will naturally vary with the total cycle count.
-These counters such as these are obviously not indicative of any causal relationship, and are ommitted.
+These counters are obviously not indicative of any causal relationship, and are ommitted.
 
@@
 Observing addresses of stack allocated data at run time is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
-A small amount of assembly code was used to calculate the addresses of \texttt{g} and \texttt{inc} and outputing to stdout direcly using syscall.
-The instrumented program has the exact same bias to environment size, making no change to the stack allocation instructions and not affecting the addresses of static variables.
+A small amount of assembly code was added to calculate the addresses of \texttt{g} and \texttt{inc}, outputing to stdout direcly using the \texttt{syscall} instruction.
+Making no change to the stack allocation instructions and not affecting the addresses of static variables, the instrumented program have the exact same bias to environment size, free from observer effects.
 We find that the spike in cycle count occurs precicely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe038 and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
+The assembly output of gcc for the inner loop is shown below, indicating the conflicting load and store instructions.
 
@@
 Acquiring dynamic memory at run time is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
-A typical malloc implementation uses different strategies for how to allocate memory depending on the size of the request.
+A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
 \begin{itemize}
@@
   Programs are initially given a relatively small heap, which the allocator can increase by calling \texttt{brk} or \texttt{sbrk} system calls.
-        The heap section starts at a relatively low address (close to the data section), and grows upwards.
-  \item For larger allocations, instead of growing the regular heap it is typically faster to use anonymous memory mapping by calling the Posix function \texttt{mmap}.
-      Memory mapped chunks are typically placed towards the upper end of the virtual address space, closer to the stack.
+
+  \item For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
+  One drawback of this is that it requires the memory to be zero initialized, thus whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
 \end{itemize}
-Addresses for smaller requests that are placed in the regular heap can look something like 0x16e30a0 or 0x1723020, while pointers returned by \texttt{mmap} can for example be 0x7f0318a8f010 or 0x7f03105d2010.
+The heap section starts at a relatively low address (close to the data section), growing upwards.
+Memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack.
+Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the actual pointer values returned.
+While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger -- for example 0x7f0318a8f010 or 0x7f03105d2010.
 This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
@@
 The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}
+This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.
 
 \subsection{Aligned Sequential Access}
-Many functions operate on pairs of buffers in a ``sliding window'' fashion; reading from one and writing the result to another in some loop construct.
-This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointers are aliased and continuously generating false conflicts.
-This effect is mentioned in the Intel optimization manual, exemplified by an aliasing copy routine.
+Many functions operate in a ``sliding window'' fashion; reading from one buffer and writing the result to another in some loop construction.
+This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointers are aliased, and continuously generating false conflicts.
+The function from Listing \ref{lst:conv} is an example of this.
+It is a naive implementation of linear correlation, containing lots of interleaved loads and stores while iterating over two independent memory buffers (for simplicity, endpoints are not handled).
 
@@
 
-The function from Listing \ref{lst:conv} is a hand crafted naive implementation of linear correlation, containing lots of interleaved loads and stores while iterating over two independent memory buffers (for simplicity, endpoints are not handled).
-We show that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12 bits.
-For heap allocated memory, this depends entirely on what addresses the heap allocator provides.
+The performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12 bits.
+By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
+To analyse performance for different addresses input and output buffers, we manually insert padding to offset one of the memory mapped pointers.
+This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset the input arguments when calling \texttt{convolution}.
+Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
+Allocating and managing these buffers at startup takes a non-neglegible amount of work.
+This constant over head can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
+\begin{lstlisting}
+    for (i = 0; i < k; ++i)
+        convolution(n, input, output + offset);
+\end{lstlisting}
+An estimate of the actual cost of one invocation is calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
 
-To force different alignments, we allocate some extra bytes at the end and offset the output pointer argument.
+\[
+t_{\text{estimate}} = \frac{t_{k} - t_{1}}{k - 1}
+\]
 
-\begin{lstlisting}
-output = malloc(n * (sizeof(float) + offset));
-conv(n, input, output + offset);
-\end{lstlisting}
+Our results are from using $k=11$, averaging over 10 invocations.
 
-Figure \ref{fig:heapalias} shows how this function behaves when given input with different alignment.
-The default behavior for large heap allocations is that both arrays alias.
-
-% Plot of cycles vs alias for estimated cost of O3 convolution
-\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convolutiontable}
-\begin{figure}[t]
+\pgfplotstableread{bin/conv-default-o2.estimate.dat}{\convtabletwo}
+\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
+\begin{figure*}[t]
   \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}.
-Offset 0 means equal 12 bit address suffix, which is default behavior for mmap allocations.
-Array size $n=2^{20}$, compiled with -O3.
-}
-  \label{fig:heapalias}
+Offset 0 means equal 12 bit address suffix, which is default behavior for \texttt{mmap} allocations.
+Input size $n=2^{20}$.}
+  \label{fig:conv-default}
   \begin{tikzpicture}
     \begin{axis}[
+        title=\texttt{cc -O2},
         font = \small,
@@
       ]
-      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convolutiontable ;
-      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convolutiontable ;
-      \addlegendentry{cycles:u} ;
-      \addlegendentry{r0107:u} ;
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwo ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwo ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
     \end{axis}
   \end{tikzpicture}
-\end{figure}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=\texttt{cc -O3},
+        font = \small,
+        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
+        cycle list name=black white
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethree ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethree ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+\end{figure*}
 
-Listing \ref{lst:conv}, together with a small main method handling the allocation and offset, is compiled with optimization O3 and input size n=$2^{20}$.
-Figure \ref{fig:heapalias} plots performance counter statistics for cycle count and 4K alias events over increasing amount of offset.
-This is an example of how false dependencies from address aliasing can cause a significant performance hit.
-Because large heap allocations always alias, the processor thinks memory accesses to input[i] potentially conflicts with output[i], stalling the pipeline.
-By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\%.
-This is a speedup on top of compiler optimization O3, which is not able to account for aliasing effects.
-
-% Table with detailed performance data, estimated for O3 (sidewaystable)
+% Table with detailed performance data, use estimated for O2 because it does not need as many columns (!)
 \begin{table*}[t]
-  \caption{Most interesting performance counters for O3 convolution code, estimated cost accounting for constant overhead of heap allocation in main.}
+  \caption{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}
   \label{tab:convstats}
 
-  \renewcommand{\tabcolsep}{3pt} % Less padding between cells
+  %\renewcommand{\tabcolsep}{3pt} % Less padding between cells
   \pgfplotstabletypeset[
-    font=\footnotesize,
+    font=\small,
     int detect, % Output whole numbers for counter values
     col sep=comma,
-    columns={Performance counter, Correlation, 0, 2, 4, 8, 16, 24},
+    columns={Performance counter, Correlation, 0, 2, 4, 8},
     column type=r,
@@
       before row={\toprule
-        Performance counter & $r$ & 0 & 2 & 4 & 8 & 16 & 24 \\
+        Performance counter & $r$ & 0 & 2 & 4 & 8 \\
         },
@@
     every last row/.style={after row=\bottomrule}
-  ]{bin/conv-default-o3.estimate.csv}
+  ]{bin/conv-default-o2.estimate.csv}
 \end{table*}
+
+Figure~\ref{fig:conv-default} shows how the convolution kernel behaves for increasing offsets modulo 4096 between heap addresses.
+The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
+The figures clearly indicate a relationship between address aliasing events and cycles executed.
+The numbers on the x axis represent the amount of offset, measures in number of \texttt{float} values.
+Note that an offset of zero is the default behavior for this program when using \texttt{malloc}, where both arrays will have addresses ending in 0x010 using the default allocator.
+Both for optimization level 2 and 3, the default alignment is close to worst case performance. 
+We only show the first 32 data points, but the lines are flat all the way through if extended to cover the full width of possible offsets within a 4K segment.
+
+% todo from here ...
+% - O2 easy, clear correlation. Look at numbers.
+% - O3 still the same. But more chaotic -- very much vectorized by gcc.
+
+Listing \ref{lst:conv}, together with a small main method handling the allocation and offset, is compiled with optimization O3 and input size n=$2^{20}$.
+Figure \ref{fig:conv-o2} plots performance counter statistics for cycle count and 4K alias events over increasing amount of offset.
+This is an example of how false dependencies from address aliasing can cause a significant performance hit.
+Because large heap allocations always alias, the processor thinks memory accesses to input[i] potentially conflicts with output[i], stalling the pipeline.
+By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\%.
+This is a speedup on top of compiler optimization O3, which is not able to account for aliasing effects.
 
@@
 For the code examples presented earlier, the dynamic linker will resolve malloc to some version of glibc, which is the default on our system.
-Variations between different versions of libc, or local configurations on parameters such as M\_MMAP\_THRESHOLD, could trigger aliasing cases.
+Variations between different versions of libc, or local configurations on parameters such as thresholds for when to use \texttt{mmap} over \texttt{brk}, could trigger aliasing cases.
 

### 2014-05-16 paper.tex
=Updates based on the feedback from Jørn.

@@
 
-    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchtectures. 
+    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchitectures. 
     The magnitude of this performance bias has been shown to be significant, and can be as high as a factor two.
 
-    We analyze a small program that was presented as a curiosity in previous work on \emph{measurement bias};
+    We analyze a small program that was presented as a curiosity in [others] previous work on \emph{measurement bias};
     Code that for unexplained reasons performs significantly worse than average for certain configurations of system environment variables.
@@
 These types of questions are, perhaps surprisingly, not always trivial to answer.
-The usual approach to quantify performance is to simply measure some metric, such as the wall time or number of cycles executed.
+The usual approach to quantify performance is to simply use some metric, such as the wall time or number of cycles executed.
 However, performance is not just a function of the collection of instructions in an executable, but also dependent on various \emph{external} factors.
@@
 Trying to determine the political opinion of a country just by sampling a narrow geographical area will probably not be very accurate.
-The same concept applies for software; measuring a metric once is analogous to sampling only a tiny geographical area in the set of possible system configurations.
+The same concept applies for software; measuring once is analogous to sampling only a tiny geographical area in the set of possible system configurations.
 
 There are many external variables that has the potential to affect system performance.
-In particular, anything which can offset memory addresses of code or data at run-time can be significant.
+In particular, anything which can offset memory addresses of code or data at runtime can be significant.
 Memory accesses interacts with low level hardware buffers, which can heavily favor certain memory layouts over others.
@@
 
-Our approach is instead to try to understand some of the underlying reasons to why bias can occur, in an attempt to utilize this knowlege for potential optimization.
+Our approach is instead to try to understand some of the underlying reasons to why bias can occur, in an attempt to utilize this knowledge for potential optimization.
 % Could have a sentence with cite of Blind Optimization.
-Usually, the root explanation will be that different memory layouts interact with various hardware buffers.
+Usually, the explanation will be that different memory layouts interact with various hardware buffers.
 Certain mechanisms such as cache contention or branch misprediction are already well understood, thus we are looking for the more esoteric processor features.
@@
 We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
-A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
+A small Python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
 The same script simulates changes to environment size, which is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
@@
 In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
-A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12 bit address suffix of 0x020 is the same in both. 
-Dispite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
+A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
+Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
 The number of times this happens can be counted by the following performance counter:
@@
 % Generic intro to memory layout (with focus on stack/env interaction)
-Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory. 
-The stack is normally placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64 bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables.
+Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
+The stack is normally placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables.
 There is still some level of stack alignment happening, independent of environment offsets. 
@@
 
-Note that there is no clear relationship between environment size and stack location with address randomization enabled.
-However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurences of measurement bias indeed random.
+Note that there is no clear relationship between environment size and stack location with ASLR enabled.
+However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
 % In addition to affecting stack, ASLR also randomizes location of dynamically linked files (memory mapped), and initial position of heap.
@@
 \begin{figure}[float=t]
-  \caption{Organization of virtual memory for a 64 bit process.}
+  \caption{Organization of virtual memory for a 64-bit process.}
   \label{fig:virtualmemory}
@@
 
-\subsection{Analysis of micro-kernel}
+\subsection{Microkernel Analysis}
 To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by T.
@@
 Some of the other counters are less interesting; for example, bus- and cpl cycles, which will naturally vary with the total cycle count.
-These counters are obviously not indicative of any causal relationship, and are ommitted.
+These counters are obviously not indicative of any causal relationship, and are omitted.
 
 % Need input here, does this make sense?
-On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270'000 to about 135'000.
+On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270,000 to about 135,000.
 The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Intel:2013:Volume3B}.
@@
 These numbers clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
-For that we need to know the addresses of all variables at run time.
+For that we need to know the addresses of all variables at runtime.
 The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
@@
 
-Observing addresses of stack allocated data at run time is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
-A small amount of assembly code was added to calculate the addresses of \texttt{g} and \texttt{inc}, outputing to stdout direcly using the \texttt{syscall} instruction.
+Observing addresses of stack allocated data at runtime is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
+A small amount of assembly code was added to calculate the addresses of \texttt{g} and \texttt{inc}, outputting to stdout directly using the \texttt{syscall} instruction.
 Making no change to the stack allocation instructions and not affecting the addresses of static variables, the instrumented program have the exact same bias to environment size, free from observer effects.
-We find that the spike in cycle count occurs precicely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe038 and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
+We find that the spike in cycle count occurs precisely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe{\underline{038} and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
 The assembly output of gcc for the inner loop is shown below, indicating the conflicting load and store instructions.
@@
 Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
-In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as address layout randomization. 
+In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR. 
 Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
@@
 Most dynamic memory is allocated on the \emph{heap}, which is managed by an \emph{allocator}.
-A heap allocator is responsible for managing dynamic memory, and ultimately assigns the actual addresses of heap allocated variables at run time.
+A heap allocator is responsible for managing dynamic memory, and ultimately assigns the actual addresses of heap allocated variables at runtime.
 Heap allocation routines such as \texttt{malloc} and \texttt{free} are typically dynamically linked, for example as part of glibc, or some other library.
-The particular library used therefore constitutes an important part of the execution context, as linking to a different library, or a library with some alternative configuration, can impact heap addresses at run time.
+The particular library used therefore constitutes an important part of the execution context, as linking to a different library, or a library with some alternative configuration, can impact heap addresses at runtime.
 
 \subsection{\texttt{malloc} will Alias by Default}
-Acquiring dynamic memory at run time is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
+Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
 A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
@@
 However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}
+The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every mmap'ed address ends with 0x010}
 This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.
@@
 
-The performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12 bits.
+The performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.
 By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
@@
   \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}.
-Offset 0 means equal 12 bit address suffix, which is default behavior for \texttt{mmap} allocations.
+Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
 Input size $n=2^{20}$.}
@@
 
-Listing \ref{lst:conv}, together with a small main method handling the allocation and offset, is compiled with optimization O3 and input size n=$2^{20}$.
+Listing \ref{lst:conv}, together with a small main function handling the allocation and offset, is compiled with optimization O3 and input size n=$2^{20}$.
 Figure \ref{fig:conv-o2} plots performance counter statistics for cycle count and 4K alias events over increasing amount of offset.
@@
 By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\%.
-This is a speedup on top of compiler optimization O3, which is not able to account for aliasing effects.
+This is a speedup on top of compiler optimization level O3, which is not able to account for aliasing effects.
 
@@
 It is perhaps unfair to assume the compiler can fix this when only given a generic optimization flag, as aliasing effects is (presumably) an Intel-specific issue.
-You might also want to utilize the C99 keyword restrict to explicitly tell the compiler that the buffers are disjoint in memory.
+Another option is to utilize the C99 keyword \emph{restrict} to explicitly tell the compiler that the buffers are disjoint in memory.
 
@@
 As with the contents of environment variables, the exact behavior of dynamic memory allocation can typically not be known ahead of time.
-Heap allocation routines such as malloc and free are usually dynamically linked and resolved at run-time, which again represents a source of potential bias.
+Heap allocation routines such as malloc and free are usually dynamically linked and resolved at runtime, which again represents a source of potential bias.
 For the code examples presented earlier, the dynamic linker will resolve malloc to some version of glibc, which is the default on our system.
@@
 \begin{itemize}
- \item Thread-Caching Malloc (ptmalloc) is developed by Google. It aims to be time and space efficient, and reduces lock contention for multithreaded applications. \cite{TCMalloc}
+ \item Thread-Caching Malloc (tcmalloc) is developed by Google. It aims to be time and space efficient, and reduces lock contention for multithreaded applications. \cite{TCMalloc}
  \item jemalloc was originally developed for FreeBSD. It aims to be a scalable general purpose allocator in single or multithreaded environments. \cite{JEMalloc}
@@
 
-Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator. \footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
+Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
 The addresses of two equally sized char buffers allocated with malloc are observed for size parameters 5120 and $2^{20}$ bytes. 
@@
 For smaller allocations, we see that glibc and tcmalloc uses the normal heap area -- returning numerically low addresses.
-Interestingly, jemalloc and Hoard appears to never use brk, but always allocate to mmaped areas even for smaller sizes.
+Interestingly, jemalloc and Hoard appears to never use brk, but always allocate to memory mapped areas even for smaller sizes.
 We see one example where one allocator yields aliasing buffers while another does not. 
-Allocating $2 \times 5120 B$ returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
+Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
 Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.
@@
 To our knowledge, there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
-This dispite explicit recommendation from Intel that special purpose allocators should care about aliasing for optimization. [cite]
+This despite explicit recommendation from Intel that special purpose allocators should care about aliasing for optimization. [cite]
 
@@
 \subsection{Dealing with Aliasing}
-For the particular case we investigated, a good heuristic is to align heap segments “far apart” within the 4 KiB area of 12 bit suffixes.
+For the particular case we investigated, a good heuristic is to align heap segments “far apart” within the 4 KiB area of 12-bit suffixes.
 More specifically, it appears that address suffixes of A and $\boldsymbol{y}$ are the most important to separate.
@@
 As discussed in section [x], aliasing cases like these can be accounted for in software using padding techniques.
-A possible run time solution to adjust heap addresses can be realized as follows:
+A possible runtime solution to adjust heap addresses can be realized as follows:
 
@@
 
-Analyzing the example originally presented by Mytkowicz et al.
+By analyzing an example originally presented by Mytkowicz et al.
 \cite{Mytkowicz:2009:WrongData}, we determined that collisions between stack variables and static data resulted in aliasing for certain stack positions.
@@
 Despite recommendations from Intel, we are not aware of any allocators that specifically addresses this issue. 
-We show how manual padding of variables, and alternative alias-free code paths, can be used to avoid aliasing at run time.
+We show how manual padding of variables, and alternative alias-free code paths, can be used to avoid aliasing at runtime.
 This means that an understanding of 4K address aliasing is sometimes needed to achieve optimal performance.

### 2014-05-16 paper.tex
More work on heap alias

@@
 
-    We analyze a small program that was presented as a curiosity in [others] previous work on \emph{measurement bias};
+    We analyze a small program that was presented as a curiosity in related previous work on \emph{measurement bias};
     Code that for unexplained reasons performs significantly worse than average for certain configurations of system environment variables.
@@
         return main();
-    for (; g < 65536*10; g++) {
+    for (; g < 65536; g++) {
         i += inc;
@@
   \item For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
-  One drawback of this is that it requires the memory to be zero initialized, thus whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
+  One drawback of this is that it requires the memory to be zero initialized, and whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
 \end{itemize}
@@
 However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every mmap'ed address ends with 0x010}
+The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address end with 0x010.}
 This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.
@@
 \subsection{Aligned Sequential Access}
-Many functions operate in a ``sliding window'' fashion; reading from one buffer and writing the result to another in some loop construction.
-This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointers are aliased, and continuously generating false conflicts.
-The function from Listing \ref{lst:conv} is an example of this.
-It is a naive implementation of linear correlation, containing lots of interleaved loads and stores while iterating over two independent memory buffers (for simplicity, endpoints are not handled).
+Many functions operate in a ``sliding window'' fashion; reading and writing to different buffers in some loop construction.
+This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
+As an example of this, consider a naive implementation of linear correlation shown in Listing \ref{lst:conv} (for simplicity, endpoints are not handled).
+The function containins lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
+We will see that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.
 
@@
 
-The performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.
+% Use mmap directly instead maybe?
 By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
-To analyse performance for different addresses input and output buffers, we manually insert padding to offset one of the memory mapped pointers.
+By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
+To analyse performance for different addresses of input and output buffers, we manually insert padding to offset one of the memory mapped pointers.
 This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset the input arguments when calling \texttt{convolution}.
@@
 Allocating and managing these buffers at startup takes a non-neglegible amount of work.
-This constant over head can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
+The overhead can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
 \begin{lstlisting}
@@
 \end{lstlisting}
-An estimate of the actual cost of one invocation is calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
+An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
 
@@
 
-Our results are from using $k=11$, averaging over 10 invocations.
+Our results are from using $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
+For every run, performance counter measurements is also averaged over 10 samples, using the repeat mechanism of \texttt{perf-stat}.
 
@@
 \begin{figure*}[t]
-  \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}.
+  \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Listing~\ref{lst:conv}.
 Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
@@
 
-Figure~\ref{fig:conv-default} shows how the convolution kernel behaves for increasing offsets modulo 4096 between heap addresses.
+Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
 The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
-The figures clearly indicate a relationship between address aliasing events and cycles executed.
-The numbers on the x axis represent the amount of offset, measures in number of \texttt{float} values.
-Note that an offset of zero is the default behavior for this program when using \texttt{malloc}, where both arrays will have addresses ending in 0x010 using the default allocator.
+Numbers on the x axis represent the amount of offset, measured in number of \texttt{sizeof(float)} bytes.
+An offset of zero is the default behavior for this program when using \texttt{malloc} and moderately large inputs.
 Both for optimization level 2 and 3, the default alignment is close to worst case performance.
-We only show the first 32 data points, but the lines are flat all the way through if extended to cover the full width of possible offsets within a 4K segment.
+The differences in cycles executed is also very significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3.
+This phenomenon is only observed for address offsets close to zero, thus we only show the first 32 data points.
+If extended to cover the full width of 1024 possible offsets within a 4K segment, graph shows that the performance is uniform everywhere else.
 
-% todo from here ...
-% - O2 easy, clear correlation. Look at numbers.
-% - O3 still the same. But more chaotic -- very much vectorized by gcc.
+More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we have selected a subset that seem to correlate well with the change to cycle count.
+The numbers are for the level 2 optimization case, which was chosen because the behavior was less chaotic and thus more easily represented by only a few numbers.
+However, we find the performance data looks fairly similar between the two data sets, with a couple of events that seem to stand out.
+\begin{itemize}
+    \item A high number of resource stalls for the default alignment, which is reduced substantially together with increasing offsets. 
+    \item A high number of cycles with memory loads pending, indicating that the pipeline is stalled waiting for load operations to resolve.
+    \item Changes to micro-ops executed for certain ports.
+    \item Offcore requests. % todo!
+\end{itemize} 
+Interestingly, it looks like small address offsets incur a massive increase in operations issued on port 0 for the O2 case. 
+On Haswell, this port handles various ALU operations, and branching together with port 6.~\cite[Figure 2.1]{OptimizationManual}
+As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branch micro-ops are re-issued.
 
-Listing \ref{lst:conv}, together with a small main function handling the allocation and offset, is compiled with optimization O3 and input size n=$2^{20}$.
-Figure \ref{fig:conv-o2} plots performance counter statistics for cycle count and 4K alias events over increasing amount of offset.
-This is an example of how false dependencies from address aliasing can cause a significant performance hit.
-Because large heap allocations always alias, the processor thinks memory accesses to input[i] potentially conflicts with output[i], stalling the pipeline.
-By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\%.
-This is a speedup on top of compiler optimization level O3, which is not able to account for aliasing effects.
+It is also worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
+The L1 hit rate remains stable across all offsets, and only a neglible number of memory loads actually misses L1.
+
+Overall it seems reasonable to conclude that the resource stalling is causing delayed execution, ultimately generated by false dependencies from address aliasing.
+Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
+
+We do not attempt to pinpoint exactly which assembly instructions might alias, but on a high level we can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
+By manually adjusting the alignment of one of the input buffers, cycle count can be reduced by as much as 50\%.
+This is a speedup on top of aggressive compiler optimization, which is not able to account for aliasing effects.
 

### 2014-05-16 analysis/heap-alias/conv.c

@@ -1,9 +1,9 @@
 static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
 
-void convolve(int size, const float *input, float *output)
+void convolve(int n, const float *input, float *output)
 {
     int i, j;
-    for (i = 2; i < size - 2; ++i)
+    for (i = 2; i < n - 2; ++i)
     {
         output[i] = 0;
         for (j = 0; j < 5; ++j)

### 2014-05-17 paper.tex
Remove BLAS section

@@
 
-\iffalse
-
-\section{Optimizing Alignment in BLAS}
-
-To illustrate how real world applications can be impacted by aliasing, we look at low level numerical applications.
-In this section we show that even highly optimized numerical libraries suffer performance degradation from aliasing.
-Basic Linear Algebra Subroutines (BLAS), is the de facto standard API for high performance linear algebra routines.
-The functionality is divided into three categories:
-
-\begin{description}
-  \item{Level~1} Scalar and vector operations, such as dot product and vector addition.
-  \item{Level~2} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
-  \item{Level~3} Matrix-matrix operations, including the widely applied gemm routine for general matrix-matrix multiplication. 
-\end{description}
-
-Functions operating on vectors, from BLAS Level 1 or 2, intuitively appears the most likely to have potential for aliasing.
-Many highly optimized implementations of BLAS exists, ATLAS being a widely used and open source alternative.
-One of the key features of ATLAS is that it uses automatic tuning to optimize for cache efficiency [Whaley:2000:ATLAS].
-
-We use the currently latest stable version 3.10.1 of ATLAS, built as a shared library from source.
-The automatic tuning happens during the build process; A series of test programs are run to determine cache edges and other properties of the hardware, which in turn affects the resulting binary.
-
-Carefully monitoring cache metrics is particularly important in this case study, as BLAS performance heavily relies on cache efficiency.
-Relevant performance statistics for cache hit ratios was considered to rule out cache as the cause of any bias effects.
-In addition, we disable hardware prefetching in BIOS to further minimize cache impacts. 
-Note however that the same conclusions can be drawn independently of these measures.
-
-\subsection{Aliasing in cblas\_daxpy}
-Found aliasing here as well.
-Two spikes, not sure why.
-Might want to test other BLAS libraries.
-Include or leave out?
-Too much handwaving and too little analysis maybe.
-Depends on the target paper length as well?
-
-\subsection{Matrix vector multiplication}
-
-Consider matrix-vector multiplication of the form $\boldsymbol{y} = A\boldsymbol{x} + \boldsymbol{b}$.
-Let A be of size $M \times N$, where M is the number of rows.
-
-$
-\left[\begin{array}{ccccc}
-a_{0,0} & a_{0,1} &  &  & a_{0,N}\\
-a_{1,0}\\
- &  &  & \ddots\\
-a_{M,0} &  &  &  & a_{M,N}
-\end{array}\right]\left[\begin{array}{c}
-x_{0}\\
-x_{1}\\
-\vdots\\
-\\
-x_{N}
-\end{array}\right]=\left[\begin{array}{c}
-y_{0}\\
-y_{1}\\
-\vdots\\
-y_{M}
-\end{array}\right]
-$
-
-General matrix vector multiplication is implemented by the level 2 gemv routine, computing $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$
-Here, $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix. 
-We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
-[There are three memory buffers involved, and one might suspect that implementations are succeptible to aliasing between of $\boldsymbol{A}$, $\boldsymbol{x}$ or $\boldsymbol{y}$.]
-For the following discussion we look at a particular configuration where $A$ is of size $64\times64$.
-
-In order to accurately measure any effects of aliasing, we need to isolate the actual function call from heap allocation and initialization code.
-This is done by running the program twice for each memory configuration, one where cblas\_dgemv is executed $K = 101$ times in a tight loop.
-An approximation of the cost for a single function invocation can be expressed as 
-$$
-t_{\text{estimate}}=\frac{t_{K=101}-t_{K=1}}{100}
-$$
-where t represents some metric, such as the number of cycles.
-Subtracting the $K = 1$ run removes the constant overhead from the $K = 101$ run, and dividing by $100$ averages the values from the remaining iterations.
-Other iteration counts could have been used as well.
-
-Figure \ref{fig:heatmap} shows estimated number of alias events for every relevant memory configuration possible for cblas\_dgemv with matrix size 64 $\times$ 64. 
-While matrix A is kept fixed at address suffix 0x000, every reasonable variation of (\&x, \&y) is measured.
-Each increment represents an address offset of 16 bytes, which is the same granularity one would get from the default heap allocator.
-In a 4K address space, there are thus $4096/16 = 256$ possible address suffixes for $\boldsymbol{x}$ and $\boldsymbol{y}$ independently, resulting in a grid of $256^2$ values.
-
-\begin{figure}[h]
-  \caption{Estimated aliasing cost of invoking cblas\_dgemv with a matrix size $64 \times 64$ for different memory layouts.
-Axis show address offsets times 0x10 for vectors $\boldsymbol{x}$ and $\boldsymbol{y}$.
-Matrix $A$ has fixed to address suffix 0x000.}
-  \label{fig:heatmap}
-  \includegraphics[width=\columnwidth]{resources/heatmap.eps}
-\end{figure}
-
-It appears that aliasing is mostly dependent on the location of $\boldsymbol{y}$, in this case with 8 distinct worst case regions.
-This is consistent with false dependencies between $\boldsymbol{y}$ and specific rows in A. 
-The first bad region is where $\&y ~= 0$, matching the first row of A.
-
-\begin{table*}[t]
-  \caption{Estimated cost of invoking cblas\_dgemv with a matrix size $8192 \times 8192$}
-  \label{tab:atlas8k}
-  \small
-  \centering
-  \begin{tabular}{l r r r}
-    \toprule
-      & (0x010, 0x010, 0x020) & (0x010, 0x010, 0x0b0) & (0x010, 0x140, 0x150) \\
-    \midrule
-    \perfctr{cpu_clk_unhalted.thread_p} & 162,926,895 & 152,151,263 & 144,583,160 \\
-    \perfctr{ld_blocs_partial.address_alias} & \textbf{68,408,790} & \textbf{5,154,981} & \textbf{2,919} \\
-
-    \perfctr{mem_uops_retired.all.loads} & 100,701,777 & 100,702,138 & 100,701,752 \\
-    \perfctr{mem_load_uops_retired.hit_lfb} & 31,628,539 & 32,495,287 & 29,853,850 \\
-    \perfctr{mem_load_uops_retired.l1_hit} & 68,309,447 & 65,625,881 & 69,387,368 \\
-    \perfctr{mem_load_uops_retired.l2_hit} & 711,973 & 1,540,956 & 864,720 \\
-    \perfctr{mem_load_uops_retired.l3_hit} & 154,345 & 651,985 & 363,331 \\
-    \perfctr{mem_load_uops_retired.l3_miss} & 106,313 & 350,449 & 295,235 \\
-    \perfctr{mem_load_uops_retired.all_stores} & 33,562,692 & 33,562,692 & 33,562,692 \\
-
-    \perfctr{load_hit_pre.hw_pf} & 42,914,191 & 47,315,935 & 59,211,320 \\
-    \bottomrule
-  \end{tabular}
-\end{table*}
-
-Notes: 
- - More BR\_INST.COND.TAKEN taken in case (a), but number of retired instructions are the same. Suggests replays of instructions.
- - Inverse HW prefetch numbers from masters
-
-\paragraph{Prefetching}
-Measurements repeated with prefetching disabled, giving similar results.
-
-Estimated performance counter statistics for each of the three heap address configurations are shown in Table [?tab:gemv-estimate].
-In addition to cycle count and alias events, a number of relevant metrics related to cache activity are also included.
-Notice that almost all load micro-ops are served by either the line fill buffer or L1 cache in all cases.
-Only a small amount of loads come from L2, and almost none from L3.
-The hit rate for L1 actually decreases somewhat (considering LFB as well) with better execution time.
-This could be explained by less time for prefetchers to feed the L1 cache with data.
-The hardware prefetch counter indicates more hits in cases a and b.
-Again, we find that cache efficiency does not explain the performance cliffs we observe.
-
-Our results show that address aliasing between matrix and vector heap buffers can significantly impact performance of dgemv in ATLAS.
-Variations in heap addresses alone can give a speedup of more than 30 \%.
-
-\subsection{Dealing with Aliasing}
-For the particular case we investigated, a good heuristic is to align heap segments “far apart” within the 4 KiB area of 12-bit suffixes.
-More specifically, it appears that address suffixes of A and $\boldsymbol{y}$ are the most important to separate.
-
-As discussed in section [x], aliasing cases like these can be accounted for in software using padding techniques.
-A possible runtime solution to adjust heap addresses can be realized as follows:
-
-\begin{itemize}
-  \item Allocate some extra space for one of the vectors when calling malloc, for instance sizeof(double) * (M + 0x100) for $\boldsymbol{y}$.
-
-  \item Check the returned pointers for potential alias, i.e. the difference between \&A and \&y. Offset using pointer arithmetic into the array with extra padding at the end, i.e. y += 0x100. 
-\end{itemize}
-
-Another option is to explicitly account for “worst cases” in the implementation of routines that are vulnerable to aliasing.
-Addresses can be explicitly checked for potential conflicts in cblas\_dgemv, and if possible branch to code that will not suffer from aliasing.
-
-\fi
-
 \section{Conclusions}

### 2014-05-17 paper.tex
Some work on improving allocator comparison

@@
 
-\begin{table}[t]
-  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers}
+% \begin{table}[t]
+%   \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers}
+%   \label{tab:mallocompare}
+%   \centering
+%   \small
+%   \begin{tabular}{l r r}
+%     \toprule
+%                       & 5120 B          & $2^{20}$ B \\
+%     \midrule
+%     glibc             & 0x6028d0        & 0x2aaaaaaf3010 \\
+%     (ptmalloc)        & 0x603ce0        & 0x2aaaab098010 \\
+%     \midrule
+%     tcmalloc          & 0xe16000        & 0xe64000 \\
+%                       & 0xe18800        & 0xf64000 \\
+%     \midrule
+%     jemalloc          & 0x2aaaabc1f000  & 0x2aaaabc86000 \\
+%                       & 0x2aaaabc21000  & 0x2aaaabd86000 \\
+%     \midrule
+%     Hoard             & 0x2aaaaab40070  & 0x2aaaabf40070 \\
+%                       & 0x2aaaaab42070  & 0x2aaaac060070 \\
+%     \bottomrule
+%   \end{tabular}
+% \end{table}
+
+\begin{table*}[t]
+  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}
   \label{tab:mallocompare}
-  \centering
-  \small
-  \begin{tabular}{l r r}
-    \toprule
-                      & 5120 B          & $2^{20}$ B \\
-    \midrule
-    glibc             & 0x6028d0        & 0x2aaaaaaf3010 \\
-    (ptmalloc)        & 0x603ce0        & 0x2aaaab098010 \\
-    \midrule
-    tcmalloc          & 0xe16000        & 0xe64000 \\
-                      & 0xe18800        & 0xf64000 \\
-    \midrule
-    jemalloc          & 0x2aaaabc1f000  & 0x2aaaabc86000 \\
-                      & 0x2aaaabc21000  & 0x2aaaabd86000 \\
-    \midrule
-    Hoard             & 0x2aaaaab40070  & 0x2aaaabf40070 \\
-                      & 0x2aaaaab42070  & 0x2aaaac060070 \\
-    \bottomrule
-  \end{tabular}
-\end{table}
+
+  %\renewcommand{\tabcolsep}{3pt} % Less padding between cells
+  \pgfplotstabletypeset[
+    font=\small,
+    col sep=comma,
+    string type,
+    columns={Allocation, 64, 1024, 5120, 1048576},
+    column type=r,
+    columns/.style={
+        width=1cm
+    },
+    columns/Allocation/.style={
+      string type, 
+      column type=l
+    },
+    every head row/.style={
+      output empty row,
+      before row={\toprule
+         & 64 & 1024 & 5120 & 1048576 \\
+        },
+      after row=\midrule,
+    },
+    every last row/.style={after row=\bottomrule}
+  ]{bin/malloc-comparison.csv}
+\end{table*}
 

### 2014-05-17 analysis/allocators/heapalloc.c

@@ -1,17 +1,15 @@
 #include <stdlib.h>
 #include <stdio.h>
 
-#define alloc_pair(n) { \
-        char *a = malloc(n); \
-        char *b = malloc(n); \
-        printf("%d,%p,%p\n", n, a, b); \
-    }
-
 int main(int argc, char* argv[])
 {
-    alloc_pair(64)
-    alloc_pair(1024)
-    alloc_pair(5120)
-    alloc_pair(1 << 20) // 1 048 576
+	if (argc < 2)
+		return 1;
+	int size = atoi(argv[1]);
+
+	char *a = malloc(size);	
+	char *b = malloc(size);
+
+	printf("%p %p", a, b);
     return 0;
 }

### 2014-05-17 paper.tex
Improve allocator results
Fullwidth table and updated makefile for producing results

@@
   \label{tab:mallocompare}
-
-  %\renewcommand{\tabcolsep}{3pt} % Less padding between cells
   \pgfplotstabletypeset[
@@
     string type,
-    columns={Allocation, 64, 1024, 5120, 1048576},
+    columns={Allocation, 64, 1024, 65536, 1048576, 4194304},
     column type=r,
-    columns/.style={
-        width=1cm
-    },
     columns/Allocation/.style={
@@
       before row={\toprule
-         & 64 & 1024 & 5120 & 1048576 \\
+         & 64 B & 1,024 B & 65,536 B & 1,048,576 B & 4,194,304 B \\
         },
@@
     },
+    every odd row/.style={after row=\midrule},
     every last row/.style={after row=\bottomrule}

### 2014-05-17 analysis/allocators/heapalloc.c

@@ -10,6 +10,6 @@ int main(int argc, char* argv[])
 	char *a = malloc(size);	
 	char *b = malloc(size);
 
-	printf("%p %p", a, b);
+	printf("%p\n%p\n", a, b);
     return 0;
 }

### 2014-05-17 paper.tex
Update to 14.04
Plus minor adjustments in paper

@@
 % Also lightly referenced in introduction/background
-\begin{lstlisting}[float=t, language=C, caption={Micro-kernel succeptible to aliasing between static and automatic variables depending on environment sizes}, label={lst:loopkernel}, frame=lines]
+\begin{lstlisting}[float=t, language=C, caption={Micro-kernel presented by Mytkowitz et al~\cite{Mytkowicz:2009:WrongData}, succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
 static int i, j, k;
@@
 Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus on the Intel ``Haswell'' microarchitecture specifically.
-Our setup consists of a 4th generation Intel\textsuperscript{\textregistered} Core\textsuperscript{\texttrademark} i7-4770K processor, running 64-bit Ubuntu 13.10.
-Code samples are compiled using the GCC toolchain, version 4.8.1-10ubuntu9.
+Our setup consists of a 4th generation Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K processor, running 64-bit Ubuntu 14.04 LTS on kernel version 3.13.0-24-generic.
+Code samples are compiled using the {\small GCC} toolchain, version 4.8.2-19ubuntu1.
 
@@
 
-% \begin{table}[t]
-%   \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers}
-%   \label{tab:mallocompare}
-%   \centering
-%   \small
-%   \begin{tabular}{l r r}
-%     \toprule
-%                       & 5120 B          & $2^{20}$ B \\
-%     \midrule
-%     glibc             & 0x6028d0        & 0x2aaaaaaf3010 \\
-%     (ptmalloc)        & 0x603ce0        & 0x2aaaab098010 \\
-%     \midrule
-%     tcmalloc          & 0xe16000        & 0xe64000 \\
-%                       & 0xe18800        & 0xf64000 \\
-%     \midrule
-%     jemalloc          & 0x2aaaabc1f000  & 0x2aaaabc86000 \\
-%                       & 0x2aaaabc21000  & 0x2aaaabd86000 \\
-%     \midrule
-%     Hoard             & 0x2aaaaab40070  & 0x2aaaabf40070 \\
-%                       & 0x2aaaaab42070  & 0x2aaaac060070 \\
-%     \bottomrule
-%   \end{tabular}
-% \end{table}
-
 \begin{table*}[t]

### 2014-05-18 paper.tex
New mitigate alias section at the end
Avoid focus on compilers, not the point of the paper.

@@
   \label{tab:convstats}
-
-  %\renewcommand{\tabcolsep}{3pt} % Less padding between cells
   \pgfplotstabletypeset[
@@
 
-\subsection{Architectural Optimization}
-It is perhaps unfair to assume the compiler can fix this when only given a generic optimization flag, as aliasing effects is (presumably) an Intel-specific issue.
-Another option is to utilize the C99 keyword \emph{restrict} to explicitly tell the compiler that the buffers are disjoint in memory.
-
-\begin{lstlisting}[language=C]
-void conv( int n,
-            const float * restrict in,
-            const float * restrict out );
-\end{lstlisting}
-
-Even with the updated signature compiled with -O3 -march=native, there is still room for improvement by inserting manual offsets.
-
 \subsection{Behavior of other Allocators}
@@
 Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
-To our knowledge, there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
-This despite explicit recommendation from Intel that special purpose allocators should care about aliasing for optimization. [cite]
 
@@
 With different versions or configurations of memory allocators, any performance impact from aliasing can qualify as measurement bias.
+
+\subsection{Ways to Deal with Address Aliasing}
+Measurement bias from address aliasing can be significant in real applications, with potentially large performance penalties.
+However, with an understanding of the underlying mechanism that causes bias, the effects can to some extent be predicted and accounted for in software.
+The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
+In our convolution kernel example, there are several ways to limit the extent of aliasing.
+
+\paragraph{Mark buffers with \texttt{restrict}}
+In our convolution kernel implementation, the compiler has to account for the fact that input and output pointers might alias, or that the buffers partially overlap.
+This limits the extent generated code can keep data in registers without updating the values from memory, as a write to one buffer potentially could invalidate a cached value from the other.
+The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer does not alias with any other, allowing for more efficient code generation with fewer memory accesses.
+\begin{lstlisting}[language=C]
+void conv( int n,
+            const float * restrict input,
+            float * restrict output );
+\end{lstlisting}
+With this updated function prototype, most of the address aliasing events are gone for the default alignment.
+
+\paragraph{Use a special purpose allocator}
+We found that heap allocators are prone to generate pairwise aliasing buffers, the worst case for functions such as the convolution example.
+A potential solution could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
+This goes somewhat agains conventional wisdom that more alignment is better.
+The Intel optimization manual actually mentiones this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
+However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
+
+\paragraph{Manually adjust memory}
+In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.
+This can be achieved by exploiting the fact the \texttt{mmap} is in fact guaranteed to be placed at a page boundary, and use that directly instead of \texttt{malloc}.
+The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.
+\begin{lstlisting}[breaklines=true]
+    mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
+\end{lstlisting}
+The same pointer difference must of course by subtracted before unmapping the memory.
 
@@
 
-%\onecolumn
-
 \bibliographystyle{plain}

### 2014-05-18 analysis/heap-alias/conv-restrict.c

@@ -1,12 +1,12 @@
-static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
+static float k[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
 
-void convolve(int size, const float * restrict input, float * restrict output)
+void conv(int n, const float * restrict input, float * restrict output)
 {
     int i, j;
-    for (i = 2; i < size - 2; ++i)
+    for (i = 2; i < n - 2; ++i)
     {
         output[i] = 0;
         for (j = 0; j < 5; ++j)
-            output[i] += input[i-2+j] * kernel[j];
+            output[i] += input[i-2+j] * k[j];
     }
 }

### 2014-05-18 analysis/heap-alias/conv.c

@@ -1,12 +1,12 @@
-static float kernel[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
+static float k[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
 
-void convolve(int n, const float *input, float *output)
+void conv(int n, const float *input, float *output)
 {
     int i, j;
     for (i = 2; i < n - 2; ++i)
     {
         output[i] = 0;
         for (j = 0; j < 5; ++j)
-            output[i] += input[i-2+j] * kernel[j];
+            output[i] += input[i-2+j] * k[j];
     }
 }

### 2014-05-18 analysis/heap-alias/main.c

@@ -3,7 +3,7 @@
 
 #define N 0x100000
 
-void convolve(int, float *, float *);
+void conv(int, float *, float *);
 
 int main(int argc, char **argv)
 {
@@ -23,7 +23,7 @@ int main(int argc, char **argv)
     // Repeat a variable number of times, to be able to subtract the 
     // constant overhead from allocating and initializing input arrays.
     for (i = 0; i < repeat; ++i) {
-        convolve(N, input, (output + offset));
+        conv(N, input, (output + offset));
     }
 
     //printf("Offset: %d : (%p, %p), repeat %d\n", offset, input, output, repeat);

### 2014-05-18 paper.tex
Small fixes

@@
 As with the contents of environment variables, the exact behavior of dynamic memory allocation can typically not be known ahead of time.
-Heap allocation routines such as malloc and free are usually dynamically linked and resolved at runtime, which again represents a source of potential bias.
-For the code examples presented earlier, the dynamic linker will resolve malloc to some version of glibc, which is the default on our system.
+Heap allocation routines such as \texttt{malloc} and \texttt{free} are usually dynamically linked and resolved at runtime, which again represents a potential source of bias.
+For the code examples presented earlier, the dynamic linker will resolve \texttt{malloc} to some version of glibc, which is the default on our system.
 Variations between different versions of libc, or local configurations on parameters such as thresholds for when to use \texttt{mmap} over \texttt{brk}, could trigger aliasing cases.
-
-In addition to glibc's ptmalloc, we look at the following alternatives.
-\begin{itemize}
- \item Thread-Caching Malloc (tcmalloc) is developed by Google. It aims to be time and space efficient, and reduces lock contention for multithreaded applications. \cite{TCMalloc}
- \item jemalloc was originally developed for FreeBSD. It aims to be a scalable general purpose allocator in single or multithreaded environments. \cite{JEMalloc}
- \item Hoard is another project that focuses extensively on improving performance for multithreaded programs. \cite{Berger:2000:Hoard}
-\end{itemize}
 
@@
     string type,
-    columns={Allocation, 64, 1024, 65536, 1048576, 4194304},
+    columns={Allocation, 64, 5120, 65536, 1048576, 4194304},
     column type=r,
@@
       before row={\toprule
-         & 64 B & 1,024 B & 65,536 B & 1,048,576 B & 4,194,304 B \\
+         & 64 B & 5,120 B & 65,536 B & 1,048,576 B & 4,194,304 B \\
         },
@@
 Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
-The addresses of two equally sized char buffers allocated with malloc are observed for size parameters 5120 and $2^{20}$ bytes. 
-We do not care about performance, only the actual memory addresses returned and whether they alias or not.
+In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
+\begin{itemize}
+  \item Thread-Caching Malloc (tcmalloc) from Google. It claims to be time- and space efficient, and reduces lock contention for multithreaded applications~\cite{TCMalloc}.
+  \item jemalloc was originally developed for FreeBSD. It aims to be a scalable general purpose allocator in single or multithreaded environments~\cite{JEMalloc}.
+  \item Hoard is another project that focuses extensively on improving performance for multithreaded programs~\cite{Berger:2000:Hoard}.
+\end{itemize}
+The addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} are observed for different size parameters. 
+We do not care about performance, only whether the actual memory addresses returned alias or not.
 
 For smaller allocations, we see that glibc and tcmalloc uses the normal heap area -- returning numerically low addresses.
-Interestingly, jemalloc and Hoard appears to never use brk, but always allocate to memory mapped areas even for smaller sizes.
+Interestingly, jemalloc and Hoard appears to never use the normal heap, but allocate to memory mapped areas even for smaller sizes.
 We see one example where one allocator yields aliasing buffers while another does not.
@@
 
-Most malloc alternatives focuses heavily on performance in multithreaded environments.
-Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
-
-The purpose of this is to illustrate that allocators is yet another source of bias.
+%The purpose of this is to illustrate that allocators is yet another source of bias.
 Using heap allocated memory, address aliasing is an artifact of the allocator used.
 With different versions or configurations of memory allocators, any performance impact from aliasing can qualify as measurement bias.
+
+% Most malloc alternatives focuses heavily on performance in multithreaded environments.
+% Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
 
@@
 The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
-In our convolution kernel example, there are several ways to limit the extent of aliasing.
 
@@
 The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer does not alias with any other, allowing for more efficient code generation with fewer memory accesses.
-\begin{lstlisting}[language=C]
-void conv( int n,
-            const float * restrict input,
-            float * restrict output );
+\begin{lstlisting}[breaklines=true]
+    void conv(int n, const float * restrict input, float * restrict output);
 \end{lstlisting}
-With this updated function prototype, most of the address aliasing events are gone for the default alignment.
+With this updated function prototype, the number of alias events is reduced by about 10 million on optimization level O2 for the default alignment, with a corresponding improvement in cycle count.
+% File backing this in results
 
@@
 In this paper we have looked at how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
-The effect is caused by the way speculative and out-of-order memory operations are handled by the CPU, only considering the last 12 address bits to resolve conflicts load and store operations.
+The effect is caused by the way speculative and out-of-order memory operations are handled by the CPU, only considering the last 12 address bits to resolve conflicts between load and store operations.
 
@@
 In general, any change to virtual memory layout of data can potentially introduce bias effects from address aliasing. 
-We present a code example with extreme sensitivity of data alignment, with more than 50 \% performance variation between different memory layouts.
+We present a code example with extreme sensitivity to heap address alignment, with more than 50 \% performance variation between different memory layouts.
 
-We find that compilers, even when provided with architecture specific optimization flags, still generates code that suffers from aliasing. 
-Typical heap allocators return page-aligned memory by default, which is the worst case with respect to aliasing for many algorithms.
+We find that typical heap allocators return page-aligned memory by default, which is the worst case with respect to aliasing for many algorithms.
 Despite recommendations from Intel, we are not aware of any allocators that specifically addresses this issue. 

### 2014-05-18 paper.tex
Add acknowledgements
Update microkernel analysis
Same results offset 16 byte :)

@@
 \usepackage{url}
-\usepackage{rotating}
+\usepackage{lmodern}
 
@@
   \caption{Bias from environment size for micro-kernel in Listing~\ref{lst:loopkernel}.
-Measured average of 100 cycle count samples for 512 different environments.
+Measured average of 10 cycle count samples for 512 different environments.
 Spikes show aliasing case, occurring once for each 4K period}
@@
 
-There are clearly two worst cases, indicated by significant spikes at the 3200 and 7296 byte offsets.
+There are clearly two worst cases, indicated by significant spikes at the 3184 and 7280 byte offsets.
 We sampled an extensive set of performance counters in addition to cycle count for each execution context.
 By looking at the \emph{median} value of each counter compared to the extreme cases, we narrowed down the performance events that behaved similarly to cycle count.
-Table \ref{tab:loopcorrelation} shows numerical counter values for median over all environments, compared to worst cases of 3200 and 7296 bytes added.
+Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases of 3184 and 7280 bytes, compared to the median over all environments.
 
@@
 The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
-Some of the other counters are less interesting; for example, bus- and cpl cycles, which will naturally vary with the total cycle count.
-These counters are obviously not indicative of any causal relationship, and are omitted.
+% Some of the other counters are less interesting; for example bus cycles, which will naturally vary with the total cycle count.
+% Performance events that are obviously not indicative of any causal relationship are omitted.
 
@@
 Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stays in the execution pipeline for a shorter time.
-Our assumption is that this is a by-product of stalls generated from aliasing.
+Our assumption is that this is a by-product of stalls generated from aliasing, causing micro-operations to be reissued.
 
@@
 \subsection{The Offending Variables}
-These numbers clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
-For that we need to know the addresses of all variables at runtime.
+Our performance counter data clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
+For that we need to know the addresses of each variable at runtime.
 The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
@@
 \section{Bias from Heap Allocation}
-% Intro
-% - Bias from aliasing caused by accesses to any part of memory
-% - Dynamic memory usually from dynamically linked libs = bias
-% - More interesting: default behavior provokes aliasing 
-% - Intel warning, copy routine. 
-% - Trouble with aligned read/write buffers, creating constant conflicts in sliding window.
-% Heap allocation alias by default
-% - Normal heap with brk/sbrk syscalls
-%   - Manage heap and increase on demand
-% - Different depending on size (and other parameters)
-%   - Use mmap for performance, usually large allocations
-%   - Posix function, anonymous memory mapping
-%   - Page aligned
-% - Normal addr. vs mmap addr.
-% - Makes no difference ++ everything is conceptually the same heap, but alias by default!
-% Convolution kernel Analysis
-% - More contrived example; convolution
-%   - Simple code disregarding endpoints
-%   - Manipulate start addresses by offseting pointers
-%   - Compiled with gcc O3, shows bad performance (graph) + discussion
-% - Table with lots of data points, showing actual numbers for each offset
-% Architectural optimization
-% - Modified example with restrict and -march=native
-% - Vectorization -- probably less loads that causes reduction mostly
-% Bahavior of other Allocators
-% - Intro each allocator
-% - Present table
-% Conclude
-% - not good...
-% - Not tested other compilers, not the purpose of this paper. 
-% - How did we find this example? Intelligent design, literally first guess
-
 Address aliasing can be caused by conflicting pairs of load and store operations to any part of memory.
@@
 We present a code example with extreme sensitivity to heap address alignment, with more than 50 \% performance variation between different memory layouts.
-
 We find that typical heap allocators return page-aligned memory by default, which is the worst case with respect to aliasing for many algorithms.
@@
 
+\section{Acknowledgements}
+This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurment bias on the Intel Ivy Bridge architecture~\cite{MasterThesis}.
+
 \bibliographystyle{plain}

### 2014-05-18 paper.tex
Improve abstract
And other small fixes...

@@
 
-    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchitectures. 
-    The magnitude of this performance bias has been shown to be significant, and can be as high as a factor two.
+    Seemingly innocious properties of the environment, such as the contents of system environment variables, can affect the performance of computer programs.
+    This effect is known as \emph{measurement bias}, and has been shown to be significant, but unpredictable and difficult to deal with.
 
-    We analyze a small program that was presented as a curiosity in related previous work on \emph{measurement bias};
-    Code that for unexplained reasons performs significantly worse than average for certain configurations of system environment variables.
+    In this paper we identify one of the underlying mechanics resulting in measurement bias on modern Intel microarchitectures.
+    Using hardware performance counters, we identify that bias is ultimately caused by \emph{address aliasing}.
+    When issuing load operations, the CPU is only comparing the last 12 address bits to previous stores, sometimes resulting in performance penalties from \emph{false} dependencies.
 
-    Using hardware performance counters, we identify a strong correlation with resource stalls, ultimately caused by address aliasing events.
-    The CPU uses an optimistic strategy when issuing operations out of order.
-    Only the last 12 address bits are compared to determine if there was a conflict, sometimes resulting in performance penalties from false dependencies.
-
-    We show how address aliasing can cause measurement bias for two particular external properties; environment variable size, and dynamically linked heap allocation libraries.
-    Performance penalties from unfavorable memory address allocations can be more than double the cycle count, even for real world applications.
-    Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give worst case performance by default for large allocations.
-
+    % In addition to environment variable size, we show how address aliasing can cause measurement bias from variations in dynamically linked heap allocation libraries.
+    We find that performance penalties from unfavorable memory address allocations can be more than double the cycle count, even for real world applications.
+    Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give \emph{worst case} performance by default for large allocations.
+    An understanding of address aliasing is necessary in order to achieve optimal performance, and we outline how application developers can utilize this knowledge.
     \end{abstract}
@@
 \section{Introduction}
-% - Measurement bias in systems research
-%   - What is measurement bias
-%     - fundamentally anything external
-%     - Room temperature
-%   - Studied before
-%     - External properties that affect memory layout
-%       - link order (not really)
-%       - env size
-%   - Has been shown to be significant and commonplace
-%     - But difficult to deal with
-%     - Need to apply statistical methods to experiments
-%       - randomization of experimental setups
-% - Want to look at flipside: understand what why bias occurs
-%   - Use for optimization
-%   - Find interesting cases
-%   - Known causes: cache alignment, ignore.
-% - Analysis: 
-%   - Use known example that exhibit bias
-%   - Find lesser known source in address aliasing
-% - Results:
-%   - Applicable general knowledge about Intel architecture
-%   - Useful for optimization
-%   - Libraries and compilers fail to account for it in reality.
-
 How fast does an algorithm run?
@@
 
-Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
-We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
-A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
-The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
-
-There are clearly two worst cases, indicated by significant spikes at the 3184 and 7280 byte offsets.
-We sampled an extensive set of performance counters in addition to cycle count for each execution context.
-By looking at the \emph{median} value of each counter compared to the extreme cases, we narrowed down the performance events that behaved similarly to cycle count.
-Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases of 3184 and 7280 bytes, compared to the median over all environments.
-
 % Detailed comparison between median and worst cases.
@@
   \label{tab:loopcorrelation}
-  \centering
+  %\centering
   \renewcommand{\tabcolsep}{1pt} % Less padding between cells
@@
 \end{table}
+
+Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
+We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
+A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
+The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
+
+There are clearly two worst cases, indicated by significant spikes at the 3184 and 7280 byte offsets.
+We sampled an extensive set of performance counters in addition to cycle count for each execution context.
+By looking at the \emph{median} value of each counter compared to the extreme cases, we narrowed down the performance events that behaved similarly to cycle count.
+Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases of 3184 and 7280 bytes, compared to the median over all environments.
 

### 2014-05-19 paper.tex
Lots of smaller fixes

@@
     Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give \emph{worst case} performance by default for large allocations.
-    An understanding of address aliasing is necessary in order to achieve optimal performance, and we outline how application developers can utilize this knowledge.
+    An understanding of address aliasing is necessary in order to achieve optimal performance, and we outline how developers can utilize this knowledge.
     \end{abstract}
@@
 These types of questions are, perhaps surprisingly, not always trivial to answer.
-The usual approach to quantify performance is to simply use some metric, such as the wall time or number of cycles executed.
-However, performance is not just a function of the collection of instructions in an executable, but also dependent on various \emph{external} factors.
+The usual approach to quantify performance is to simply measure some metric, such as the wall time or number of cycles executed.
+However, performance is not just a function of the instructions in an executable, but also dependent on various \emph{external} factors.
 
-In other fields of science, accounting for sampling bias is well understood.
+In other fields of science, accounting for bias is well understood.
 Trying to determine the political opinion of a country just by sampling a narrow geographical area will probably not be very accurate.
-The same concept applies for software; measuring once is analogous to sampling only a tiny geographical area in the set of possible system configurations.
+The same concept applies to software; measuring performance for only one setup is analogous to sampling only a tiny geographical area in the set of possible system configurations.
 
@@
 In particular, anything which can offset memory addresses of code or data at runtime can be significant.
-Memory accesses interacts with low level hardware buffers, which can heavily favor certain memory layouts over others.
+Memory accesses interact with low level hardware buffers, which can heavily favor certain memory layouts over others~\cite{Hundt:2011:MAO}.
 Consequentially, any environmental factor that can alter memory layout has a potential performance impact.
@@
 % This is basically former background section
-Measurement bias has been shown to be commonplace and significant in real applications~\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
-Previous work looks at environmental factors such as the size of Unix environment variables, which affects memory by offsetting the initial position of stack.
-Because of the complexity of modern hardware, it is often impossible to predict exactly how seemingly innocuous things such as environment variables ultimately affect program execution.
-While effects of bias has been shown to be commonplace and significant, they are unpredictable and difficult to deal with.
+Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
+Previous work looks at environmental factors such as the size of Unix environment variables, which can affect memory by offsetting the initial position of stack.
+Because of the complexity of modern hardware, it is often impossible to predict exactly how things such as environment variables ultimately affect program execution.
 Without an accurate understanding of the underlying memory interactions, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
@@
 % Could have a sentence with cite of Blind Optimization.
-Usually, the explanation will be that different memory layouts interact with various hardware buffers.
-Certain mechanisms such as cache contention or branch misprediction are already well understood, thus we are looking for the more esoteric processor features.
-Analyzing a program that exhibits strong bias from environment size, collecting about 200 events using hardware performance counters, we find strong correlation with resource stalls from \emph{address aliasing}.
+Usually, the explanation will be that different memory layouts interact with various hardware buffers, hitting ``performance'' cliffs for certain worst or best cases.
+Certain mechanisms such as cache conflicts or branch misprediction are already well understood, thus we are looking for the more esoteric processor features.
+
+Our starting point is a small program that was first presented by Mytkowicz et.
+al, exhibiting significant bias from environment size without any clear answer as to why in the original paper~\cite{Mytkowicz:2009:WrongData}.
+Collecting about 200 events using hardware performance counters, we find strong correlation with resource stalls from \emph{address aliasing}.
 Aliasing is an artifact of a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline.
-In short, the CPU uses a heuristic for determining whether memory loads are dependent on previous writes, comparing only the last 12 virtual address bits.
+In short, the CPU uses a heuristic for determining whether memory loads are dependent on previous stores, comparing only the last 12 virtual address bits.
 False dependencies incur performance penalties, and we show how this can explain many common cases of measurement bias.
@@
 We find that most libraries tend to produce worst case behavior by default with respect to aliasing, caused by favoring page alignment for large allocations.
-The cost of address aliasing can be significant, exemplified with a simple program yielding more than 50~\% speedup based on heap address alignment alone.
-% [This next part might be added to the abstract in some form:] 
+The cost of address aliasing can be significant, exemplified by a simple program with more than $2x$ speedup based on heap address alignment alone.
 Fortunately, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
@@
 
-% Include preview of aliasing program early to provike interest. 
-% Also lightly referenced in introduction/background
-\begin{lstlisting}[float=t, language=C, caption={Micro-kernel presented by Mytkowitz et al~\cite{Mytkowicz:2009:WrongData}, succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
+% Include preview of aliasing program early to provike interest, also lightly referenced in introduction/background
+\begin{lstlisting}[float=t, language=C, caption={Micro-kernel presented by Mytkowicz et al~\cite{Mytkowicz:2009:WrongData}, succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
 static int i, j, k;
@@
 \subsection{Methodology}
-% - Goal: Produce empirical data showing what causes bias
-% - Two pitfalls:
-%  - Observer effect
-%  - Measurement bias
-% - Hardware performance counters
-%   - MSR, architecture specific
-%   - Closest thing to the CPU
-%   - Use perf-stat
-%   - Control cache metrics
-% - Best practices for controlled experimental setup
-%   - No ASLR, what it is
-%   - No hyperthreading, disable frequency scaling.
-% - Experimental setup
-%  - Intel haswell architecture
-%    - Restrict to a single cpu to limit scope
-
-In order to understand bias behavior it is important to be able to do precise measurements.
+In order to understand biased behavior it is important to be able to do precise measurements.
 Besides measurement bias in itself, one must also consider \emph{observer effect}.
 Any kind of instrumentation added to the code, such as a counter variable for each invocation of a function, has the potential to produce misleading results~\cite{Mytkowicz:2008:OE&MB}.
-Instead, we use instrumentation provided directly by the hardware; in the form of performance counters. 
+Instead, we use instrumentation provided directly by the hardware, in the form of performance counters. 
 Modern processors have a set of model specific registers (MSRs) that can be configured to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
-Recent Intel architectures have several hundred different available events, which can give a very detailed view of what happens inside the CPU.
+Recent Intel architectures have several hundred available events, which can give a very detailed view of what happens inside the CPU.
 
@@
 \section{Bias from Environment Size}
-% General intro on why environment triggers aliasing
-% - Image of memory organization, this can be generic.
-%   - Each area affected by randomization differently
-% - Default 16 bit stack alignment guarantee.
-% Analysis section
-% - Present data on correlation
-% - Plot graph of 2 x 4K areas. 
-% - Addresses of static variables from symbol table
-% - Addresses of automatic variables from asm output
-% - Outline different scenarios that can happen for aliasing with 1 or 2 variables
-% - Conclusion: was bias from false dependencies
-% Avoid aliasing for env size
-% - Show altered program
-% - Proof of concept, in theory possible for compilers to generate such code.
-
 Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack.
@@
 al.~\cite{Mytkowicz:2009:WrongData}, reproduced here in Listing~\ref{lst:loopkernel}.
-This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straight forward to analyze.
+This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straightforward to analyze.
 Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
@@
 
-% Need input here, does this make sense?
 On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270,000 to about 135,000.
@@
 
-% More deep dive section (maybe separate headline...)
 \subsection{The Offending Variables}
@@
 We find that the spike in cycle count occurs precisely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe{\underline{038} and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
-The assembly output of gcc for the inner loop is shown below, indicating the conflicting load and store instructions.
+The assembly output of {\small GCC} for the inner loop is shown below, indicating the relevant load and store instructions.
+There is only one store to variable \texttt{i}, which will alias with loads of \texttt{inc}.
 
@@
 
-Because the stack is aligned to 16 bytes, or 4 words, there are a couple of different scenarios that could have been observed here.
-Static variables are fixed and covers 12 contiguous bytes (3 words), in our case the addresses end in 0xc, 0x0 and 0x4, leaving the 0x8 slot free.
-Automatic variables occupies 8 contiguous bytes on stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots..
+Because the stack is aligned to 16 bytes, or 4 \emph{words}, there are a couple of different scenarios that could have been observed here.
+Static variables are fixed and covers 12 contiguous bytes (3 words), in our case the addresses end in 0x0, 0x4 and 0xc, leaving the 0x8 slot free.
+Automatic variables occupies 8 contiguous bytes on stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
 In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.
 A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots.
-While this will give significantly more alias counts, it has little effect on the number of cycles executed.
+While this will give significantly more alias counts, it has little effect on the total number of cycles executed.
 
@@
 #define ALIAS(a, b) \
-    ((((long)&a)&0xfff)==(((long)&b)&0xfff))
+    (((long)&a) & 0xfff == ((long)&b) & 0xfff)
 static int i, j, k;
@@
 By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
-To analyse performance for different addresses of input and output buffers, we manually insert padding to offset one of the memory mapped pointers.
-This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset the input arguments when calling \texttt{convolution}.
+To analyse performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
+This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the arguments before calling the function.
 Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
@@
 
-Our results are from using $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
-For every run, performance counter measurements is also averaged over 10 samples, using the repeat mechanism of \texttt{perf-stat}.
+Our results are with $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
+For every run, performance counter measurements is also averaged over 10 samples, using the repeat mechanism of perf.
 
@@
 Both for optimization level 2 and 3, the default alignment is close to worst case performance.
-The differences in cycles executed is also very significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3.
+The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
 This phenomenon is only observed for address offsets close to zero, thus we only show the first 32 data points.
-If extended to cover the full width of 1024 possible offsets within a 4K segment, graph shows that the performance is uniform everywhere else.
+If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
 
@@
 The numbers are for the level 2 optimization case, which was chosen because the behavior was less chaotic and thus more easily represented by only a few numbers.
-However, we find the performance data looks fairly similar between the two data sets, with a couple of events that seem to stand out.
+However, the performance data looks fairly similar between the two data sets, with a couple of events that seem to stand out.
 \begin{itemize}
@@
     \item A high number of cycles with memory loads pending, indicating that the pipeline is stalled waiting for load operations to resolve.
-    \item Changes to micro-ops executed for certain ports.
-    \item Offcore requests. % todo!
+    \item Changes to the number of micro-ops executed for certain ports.
+    \item Offcore requests. (++) % todo!
 \end{itemize}
@@
 On Haswell, this port handles various ALU operations, and branching together with port 6.~\cite[Figure 2.1]{OptimizationManual}
-As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branch micro-ops are re-issued.
+As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branch micro-ops are being re-issued.
 
@@
 We do not attempt to pinpoint exactly which assembly instructions might alias, but on a high level we can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
-By manually adjusting the alignment of one of the input buffers, cycle count can be reduced by as much as 50\%.
-This is a speedup on top of aggressive compiler optimization, which is not able to account for aliasing effects.
+By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
+This is a speedup on top of already aggressive compiler optimization.
 

### 2014-05-20 paper.tex
Change to ACM TACO formatting
Some adjustments to figures, tables and legends to accomodate
layout.

@@
-\documentclass[a4paper,10pt,twocolumn,twoside]{article}
-\usepackage{fullpage}
+% Typeset for Transactions on Architecture and Code Optimization
+\documentclass[prodmode,acmtaco]{acmsmall}
+%\usepackage{fullpage}
 \usepackage{graphicx}
@@
 \usepackage{booktabs}
-\usepackage{tabularx}
+%\usepackage{tabularx}
 \usepackage{tikz}
 \usepackage{listings}
-\usepackage{authblk}
-\usepackage{caption}
+%\usepackage{authblk} % acm error, \affil
+\usepackage{caption} % 
 \usepackage{mathtools}
@@
 \usepackage{url}
-\usepackage{lmodern}
+\usepackage{lmodern} % Much clearer text
 
@@
 
-\author{Lars Kirkholt Melhus}
-\author{Rune Erlend Jensen}
-\affil{Norwegian University of Science and Technology}
-\title{Measurement Bias from Address Aliasing}
-\date{} % This is a timeless piece
-
-% Pretty printing of performance counter names
 \newcommand{\perfctr}[1] {
@@
 
-\begin{document}
+%\acmVolume{V}
+%\acmNumber{N}
+%\acmArticle{A}
+%\acmYear{YYYY}
+%\acmMonth{0}
 
-\twocolumn[
-  \begin{@twocolumnfalse}
-    \maketitle
+% Page headers
+\markboth{L. K. Melhus and R. E. Jensen}{Measurement Bias from Address Aliasing}
+
+\title{Measurement Bias from Address Aliasing}
+\author{
+LARS KIRKHOLT MELHUS and RUNE ERLEND JENSEN
+  \affil{Norwegian University of Science and Technology}}
+
+% Motivation -- Problem statement -- Approach -- Results -- Conclusions
 \begin{abstract} % Should be 150 -- 200 words max
-    % Motivation -- Problem statement -- Approach -- Results -- Conclusions
-
 Seemingly innocious properties of the environment, such as the contents of system environment variables, can affect the performance of computer programs.
@@
 \end{abstract}
-    \bigskip
 
-  \end{@twocolumnfalse}
-]
+%\category{}{}{}
+\terms{Measurement bias, Optimization}
+\keywords{4K address aliasing, Intel ``Haswell'' architecture, heap allocators}
+\acmformat{Lars K. Melhus and Rune E. Jensen, 2014. Measurement Bias from Address Aliasing}
+
+\begin{document}
+
+\begin{bottomstuff}
+
+\end{bottomstuff}
+
+\maketitle
 
@@
 % This is basically former background section
-Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
+Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~%\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
 Previous work looks at environmental factors such as the size of Unix environment variables, which can affect memory by offsetting the initial position of stack.
@@
 % Include preview of aliasing program early to provike interest, also lightly referenced in introduction/background
-\begin{lstlisting}[float=t, language=C, caption={Micro-kernel presented by Mytkowicz et al~\cite{Mytkowicz:2009:WrongData}, succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
+% Cite does not work in acm style
+% Micro-kernel presented by Mytkowicz et al~\cite{Mytkowicz:2009:WrongData}, 
+\begin{lstlisting}[float=t, language=C, caption={Micro-kernel succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
 static int i, j, k;
@@
 \begin{description}
-  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
+  \item LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
   Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
@@
 
-\begin{figure}[float=t]
+\begin{figure}[h]
+  \centering
   \caption{Organization of virtual memory for a 64-bit process.}
@@
         height = 6cm,
-        font = \small,
+        font = \footnotesize,
         xlabel=Bytes added to environment,
@@
 % Detailed comparison between median and worst cases.
-\begin{table}[t]
-  \caption{Events with significant correlation to cycle count.}
+\begin{table}[b]
+  \centering
+  \caption{Events with significant correlation to cycle count}
   \label{tab:loopcorrelation}
-  %\centering
-  \renewcommand{\tabcolsep}{1pt} % Less padding between cells
+  %\renewcommand{\tabcolsep}{1pt} % Less padding between cells
+  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
   \pgfplotstabletypeset[
-    font=\scriptsize,
+    font=\footnotesize,
     int detect, % Output whole numbers for counter values
@@
 % See analysis for results showing this does not alias
-\begin{lstlisting}[float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
+\begin{lstlisting}[float=b, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
 #define ALIAS(a, b) \
@@
 \begin{figure*}[t]
+  \centering
   \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Listing~\ref{lst:conv}.
@@
         title=\texttt{cc -O2},
-        font = \small,
+        font=\footnotesize,
         xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
-        cycle list name=black white
+        cycle list name=black white,
+        width=\textwidth/1.9, % Make it fit to text width side by side
+        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
       ]
@@
         title=\texttt{cc -O3},
-        font = \small,
+        font=\footnotesize,
         xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
-        cycle list name=black white
+        cycle list name=black white,
+        width=\textwidth/1.9,
+        skip coords between index={20}{32}
       ]
@@
 \begin{table*}[t]
+  \centering
   \caption{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}
   \label{tab:convstats}
+  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
   \pgfplotstabletypeset[
-    font=\small,
+    font=\footnotesize,
     int detect, % Output whole numbers for counter values
@@
 \begin{table*}[t]
+  \centering
   \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}
   \label{tab:mallocompare}
+  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
   \pgfplotstabletypeset[
-    font=\small,
+    font=\footnotesize,
     col sep=comma,

### 2014-05-21 paper.tex
Fixes in preamble, minor abstract update

@@
 \documentclass[prodmode,acmtaco]{acmsmall}
-%\usepackage{fullpage}
 \usepackage{graphicx}
@@
 \usepackage{booktabs}
-%\usepackage{tabularx}
 \usepackage{tikz}
 \usepackage{listings}
-%\usepackage{authblk} % acm error, \affil
-\usepackage{caption} % 
+\usepackage{caption}
 \usepackage{mathtools}
@@
 
-In this paper we identify one of the underlying mechanics resulting in measurement bias on modern Intel microarchitectures.
+In this article we identify one of the underlying mechanics resulting in measurement bias on modern Intel microarchitectures.
 Using hardware performance counters, we identify that bias is ultimately caused by \emph{address aliasing}.
@@
 
-% In addition to environment variable size, we show how address aliasing can cause measurement bias from variations in dynamically linked heap allocation libraries.
-We find that performance penalties from unfavorable memory address allocations can be more than double the cycle count, even for real world applications.
-Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give \emph{worst case} performance by default for large allocations.
-An understanding of address aliasing is necessary in order to achieve optimal performance, and we outline how developers can utilize this knowledge.
+In addition to environment variable size, we show how bias can be triggered from variations in dynamically linked \emph{heap allocation} libraries.
+The impact of unfavorable memory address allocations can be significant, and we demonstrate up to $2x$ speedup for a real world application.
+Furthermore, the allocation pattern of common libraries such as glibc tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
+An understanding of address aliasing is necessary in order to achieve optimal performance, and we outline how to programmatically optimize for it.
 \end{abstract}
 
-%\category{}{}{}
-\terms{Measurement bias, Optimization}
-\keywords{4K address aliasing, Intel ``Haswell'' architecture, heap allocators}
-\acmformat{Lars K. Melhus and Rune E. Jensen, 2014. Measurement Bias from Address Aliasing}
+% Choose from classification tree http://www.acm.org/about/class/ccs98-html
+\category{C.4}{Performance of Systems}{Measurement techniques; Performance attributes}
+
+% Terms come from a fixed list of predefined terms
+\terms{Measurement, Performance}
+
+% Free to choose keywords, but must be in alphabetical order
+\keywords{4K address aliasing, heap allocators, measurement bias, memory disambiguation}
+
+% Minimum is author, year and title
+\acmformat{Lars K. Melhus and Rune E. Jensen. 2014. Measurement Bias from Address Aliasing}
 
@@
 
+% Acknowledgement of support, author(s) address(es)
 \begin{bottomstuff}
-
 \end{bottomstuff}
@@
 \section{Conclusions}
-In this paper we have looked at how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
+In this article we have looked at how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
 The effect is caused by the way speculative and out-of-order memory operations are handled by the CPU, only considering the last 12 address bits to resolve conflicts between load and store operations.

### 2014-05-21 paper.tex
Alias cleanup
Remove comments, and mention disambiguation

@@
 
-% Include preview of aliasing program early to provike interest, also lightly referenced in introduction/background
-% Cite does not work in acm style
-% Micro-kernel presented by Mytkowicz et al~\cite{Mytkowicz:2009:WrongData}, 
-\begin{lstlisting}[float=t, language=C, caption={Micro-kernel succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
-static int i, j, k;
-int main() {
-    int g = 0, inc = 1;
-    for (; g < 65536; g++) {
-        i += inc;
-        j += inc;
-        k += inc; 
-    }
-    return 0;
-}
-\end{lstlisting}
-
 \subsection{Methodology}
@@
 
-% === Random notes section on 4K alias ===
-% There seems to be two separate issues here
-% 1: store-to-load forwarding speculative, then discover that the addresses are not the same. (pp 500)
-% 2: store-load issuing the load first speculatively. WAR conflict in MOB, reissue load. VTune event "4K alias"
-%
-% MOB: "Accepts and processes all memory operations". pp 62
-%      Allows issue load before preceding store, speculatively
-
-% Discussed in Optimization manual (pp. 144, 403, 564, 583)
-% - Said to mean many things
-%   - "False store forwarding" -- Only checks last 12 bits. Does it attempt to forward?) (pp. 564)
-%   - Loads that have partial address match with preceding stores, causing the load to be reissued. 
-%     Estimate 5 cycle cost
-%   - "False dependencies in MOB due to partial compare on address"
-%   - Very clear separate event in VTune Amplifier documentation online 
-% - Detected by LOAD_BLOCK.OVERLAP_STORE on previous architectures (pp. 599)
-% - Present event code and description
-% Optimization guidelines
-% - Has some cost associated with it
-% - Intel suggests measures for mitigating cost
-%   - Change offsets between input and output buffers
-% Potentially bias inducing
-% - Anything that affects memory layout
-%   - Stack variable offset from env size
-%   - Heap addresses from dynamic allocation libraries
-
-% User/Source Coding Rule 8. (H impact, ML generality) Consider using a special memory allocation
-% library with address offset capability to avoid aliasing. .................................................. 3-55
-%   > Examples mention avoiding L1 and L2 cache conflicts.
-% User/Source Coding Rule 9. (M impact, M generality) When padding variable declarations to avoid
-% aliasing, the greatest benefit comes from avoiding aliasing on second-level cache lines, suggesting
-% an offset of 128 bytes or more ................................................................................... 3-55
-
-% Facts
-% * Loads can be issued before stores, speculating the store is not going to be to a conflicting address. (pp. 62)
-% * All memory operations go through the memory order buffer. (pp. 63)
-% * Memory disambiguation has to do with prediction. Predicted ok load from L1. (pp. 65)
-% * Example of speculative execution of loads, with mem. disambiguation (pp 131)
-% * Rule 56, 8 & 9. Aliasing case wrt Store forwarding. Claims 4K causes stall (not reissue) (pp. 141)
-% * Statement: 4K aliasing causes stall = included in LOAD_BLOCK.OVERLAP_STORE
-% * For Sandy Bridge, description of LD_BLOCKS_PARTIAL.ADDRESS_ALIAS states reissue. (pp. 583)
-
 \section{4K Address Aliasing}
-% Mechanisms for increased throughput and parallelism of memory operations
-% In particular loads that might depend on earlier stores
-% - Worst case have to wait for store to retire
-% - Supports /speculative/ issue based on prediction on whether load conflicts with previous store.
-%   - Later verified (memory disambiguation)
-% - If refer to same address, take value from sotre buffers via /store forwarding/. 
-%   - Don't have to wait for store to finish
-% An event known as 4K aliasing occurs when the load and store addresses match in the lower 12 bits
-% - CPU detects /false/ dependencies between memory operations.
-% - Counted by event ...
-% - Sometimes refered to as /false store forwarding/. (pp. 564)
-% Optimization guidelines
-%   - Avoid store and non-dependent load with 4K between (rule 56)
-%   - Special memory allocator (rule 8) -- for 64 KiB alias
-%   - Padding variables (rule 9) -- For 64 KiB alias
-% - All classified as "High" impact.
-% - We demonstrate how this effect causes bias for memory allocator and store/load single
-% - How programmers can adjust layout manually
 Modern processors are \emph{superscalar}, and achieves parallelism by issuing multiple instructions simultaneously and out of order.
 One of the issues that can limit throughput is dependencies between a load and previous stores.
-Typical software consists of about 38~\% memory operations, with about two thirds of them being loads~\cite{Intel:2006:InsideICM:SmartMemoryAccess}.
+According to Intel, typical software consists of about 38~\% memory access instructions, with about two thirds of them being loads.
+To increase parallelism, modern Intel architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}. 
 More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
-Loads can therefore be issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store.
-If the load and store locations are the same, the value can be \emph{forwarded} from the store before it retires.
+Loads are therefore issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
+The prediction is later verified, replaying any instructions that was wrongly assumed to have no dependencies.
+Similarly, if the load and store locations are the same, the value can be \emph{forwarded} from the store before it retires.
 
@@
 \begin{description}
-  \item LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
+  \item[LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS]
   Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
@@
 \end{figure}
+
+% Cite does not work in acm style
+% Micro-kernel presented by Mytkowicz et al~\cite{Mytkowicz:2009:WrongData}, 
+\begin{lstlisting}[float=h, language=C, caption={Micro-kernel succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
+static int i, j, k;
+int main() {
+    int g = 0, inc = 1;
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc; 
+    }
+    return 0;
+}
+\end{lstlisting}
 

### 2014-05-21 paper.tex
Prune references

@@
 % This is basically former background section
-Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~%\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
+Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~% todo: \cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
 Previous work looks at environmental factors such as the size of Unix environment variables, which can affect memory by offsetting the initial position of stack.
@@
 Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
-We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
+We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Volume3B}.
 A small Python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
@@
 Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
-Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
+Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
 Results are averaged over multiple runs to reduce potential random error, supplying the \texttt{-r} option to perf-stat. % Not median, good question. man perf-stat says average + std.dev.
@@
   \item[LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS]
-  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
+  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{OptimizationManual} % todo: where
 \end{description}
@@
 On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270,000 to about 135,000.
-The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Intel:2013:Volume3B}.
+The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
 Fewer stalls in the aliasing case could indicate less contention on the reservation station.

### 2014-05-24 paper.tex
Add acm bibtex style
Rewrite introduction

@@
 \section{Introduction}
-How fast does an algorithm run?
-Is program A faster than program B on \emph{my} machine?
-These types of questions are, perhaps surprisingly, not always trivial to answer.
-The usual approach to quantify performance is to simply measure some metric, such as the wall time or number of cycles executed.
-However, performance is not just a function of the instructions in an executable, but also dependent on various \emph{external} factors.
+Accurately measuring the performance of computer programs is important in order to evaluate new algorithms or compare different implementations.
+This process is complicated by \emph{measurment bias}, which can be a challenge for performance analysist and systems researchers.
+Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
+The reason this happens is because certain variables has the potential to offset memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
+Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to things like system environment variables ultimately affect program execution.
 
-In other fields of science, accounting for bias is well understood.
-Trying to determine the political opinion of a country just by sampling a narrow geographical area will probably not be very accurate.
-The same concept applies to software; measuring performance for only one setup is analogous to sampling only a tiny geographical area in the set of possible system configurations.
+To alliviate the impact of measurement bias in performance analysis, researchers have proposed to use causal analysis techniques and randomization of execution contexts~\cite{Mytkowicz:2008:OE&MB}.
+Others treat the existence bias as a potential optimization problem, using search over variant spaces to find optimal environments~\cite{Knights:2009:BlindOpt}.
+Our approach is analyze isolated programs in detail using \emph{performance counters}, in an attempt to identify which underlying hardware mechanisms are causing bias.
 
-There are many external variables that has the potential to affect system performance.
-In particular, anything which can offset memory addresses of code or data at runtime can be significant.
-Memory accesses interact with low level hardware buffers, which can heavily favor certain memory layouts over others~\cite{Hundt:2011:MAO}.
-Consequentially, any environmental factor that can alter memory layout has a potential performance impact.
-Failing to account for these effects lead to \emph{measurement bias}.
+In this paper we show how some instances of measurement bias can be explained by \emph{address aliasing}, an artifact of a presumably Intel specific optimization on the out-of-order execution pipeline.
+The CPU uses a heuristic for determining whether loads are dependent on previous stores, comparing only the last 12 virtual address bits.
+False dependencies can happen for aliased memory accesses, incurring performance penalties.
+We look at how address aliasing can trigger bias from two external factors; size of system environment variables and configuration of dynamic heap allocation libraries.
 
-% This is basically former background section
-Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~% todo: \cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
-Previous work looks at environmental factors such as the size of Unix environment variables, which can affect memory by offsetting the initial position of stack.
-Because of the complexity of modern hardware, it is often impossible to predict exactly how things such as environment variables ultimately affect program execution.
-Without an accurate understanding of the underlying memory interactions, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
-Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods.
-% Cite is basically the same as above, all three are about the same -- difficult to split.
+Our results show that the cost of aliasing can be significant, exemplified by a simple program with more than $2x$ speedup based in heap address alignment alone.
+We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing, by favoring page alignment for large allocations.
+With an understanding of how false dependencies in the CPU can impact performance, measurement bias caused by it can to some extent be predicted and even accounted for in software.
+We show how techniques like manual padding of allocations and dynamic detection of aliasing conditions can be used to improve performance. 
+Accounting for address aliasing can give significant speedups, and our findings should motivate to exploit this in architecture-specific optimization.
 
-Our approach is instead to try to understand some of the underlying reasons to why bias can occur, in an attempt to utilize this knowledge for potential optimization.
-% Could have a sentence with cite of Blind Optimization.
-Usually, the explanation will be that different memory layouts interact with various hardware buffers, hitting ``performance'' cliffs for certain worst or best cases.
-Certain mechanisms such as cache conflicts or branch misprediction are already well understood, thus we are looking for the more esoteric processor features.
+The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section.
+Then section 3 introduces ``4K aliasing'', providing necessary background for interpreting the later results.
+Section 4 presents an analysis of how the size of environment variables causes bias.
+In Section 5 we provide a similar analysis for bias to heap address alignment.
+Section 6 concludes and discusses directions for future work.
 
-Our starting point is a small program that was first presented by Mytkowicz et.
-al, exhibiting significant bias from environment size without any clear answer as to why in the original paper~\cite{Mytkowicz:2009:WrongData}.
-Collecting about 200 events using hardware performance counters, we find strong correlation with resource stalls from \emph{address aliasing}.
-Aliasing is an artifact of a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline.
-In short, the CPU uses a heuristic for determining whether memory loads are dependent on previous stores, comparing only the last 12 virtual address bits.
-False dependencies incur performance penalties, and we show how this can explain many common cases of measurement bias.
-
-In addition to environment variables, we look at dynamic heap allocation libraries as an external factor that can trigger (or amplify) bias resulting from address aliasing.
-We find that most libraries tend to produce worst case behavior by default with respect to aliasing, caused by favoring page alignment for large allocations.
-The cost of address aliasing can be significant, exemplified by a simple program with more than $2x$ speedup based on heap address alignment alone.
-Fortunately, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
-For some applications it might even make sense for programmers to actively take aliasing into account for architecture specific optimization, as neither compilers nor heap allocators seem to handle it well.
-
-\subsection{Methodology}
+\section{Methodology}
 In order to understand biased behavior it is important to be able to do precise measurements.
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{ACM-Reference-Format-Journals}
 \bibliography{references}

### 2014-05-24 paper.tex
Rewrite abstract

@@
 \begin{abstract} % Should be 150 -- 200 words max
-Seemingly innocious properties of the environment, such as the contents of system environment variables, can affect the performance of computer programs.
-This effect is known as \emph{measurement bias}, and has been shown to be significant, but unpredictable and difficult to deal with.
+Understanding program performance and obtaining accure measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
+Predicting or accounting for measurment bias is challenging, as the interaction with low level hardware mechanisms is complex.  
+In this paper we show how \emph{address aliasing} can often be the underlying mechanism resulting in measurement bias on modern Intel microarchitectures.
+Our approach is to use hardware performance counter to analyze performance data over controlled series of execution contexts.
+We show how address aliasing can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.
 
-In this article we identify one of the underlying mechanics resulting in measurement bias on modern Intel microarchitectures.
-Using hardware performance counters, we identify that bias is ultimately caused by \emph{address aliasing}.
-When issuing load operations, the CPU is only comparing the last 12 address bits to previous stores, sometimes resulting in performance penalties from \emph{false} dependencies.
-
-In addition to environment variable size, we show how bias can be triggered from variations in dynamically linked \emph{heap allocation} libraries.
-The impact of unfavorable memory address allocations can be significant, and we demonstrate up to $2x$ speedup for a real world application.
-Furthermore, the allocation pattern of common libraries such as glibc tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
-An understanding of address aliasing is necessary in order to achieve optimal performance, and we outline how to programmatically optimize for it.
+The perfomance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
+It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
+While the aliasing mechanism itself is not new, we provide the connection to measurement bias and show how to use this knowledge not only to account for bias but also explicitly optimize for it.
+Our results should provide motivation for more architecture specific optimization for example in heap allocator heuristics.
 \end{abstract}
@@
 Others treat the existence bias as a potential optimization problem, using search over variant spaces to find optimal environments~\cite{Knights:2009:BlindOpt}.
-Our approach is analyze isolated programs in detail using \emph{performance counters}, in an attempt to identify which underlying hardware mechanisms are causing bias.
+Our approach is to analyze isolated programs in detail using \emph{performance counters}, in an attempt to identify which underlying hardware mechanisms are causing bias.
 
@@
 The CPU uses a heuristic for determining whether loads are dependent on previous stores, comparing only the last 12 virtual address bits.
-False dependencies can happen for aliased memory accesses, incurring performance penalties.
+False dependencies can happen for aliased memory accesses, and incur performance penalties.
 We look at how address aliasing can trigger bias from two external factors; size of system environment variables and configuration of dynamic heap allocation libraries.

### 2014-05-24 paper.tex
Some changes to methodology and misc

@@
 
-To alliviate the impact of measurement bias in performance analysis, researchers have proposed to use causal analysis techniques and randomization of execution contexts~\cite{Mytkowicz:2008:OE&MB}.
+To alleviate the impact of measurement bias in performance analysis, researchers have proposed to use causal analysis techniques and randomization of execution contexts~\cite{Mytkowicz:2008:OE&MB}.
 Others treat the existence bias as a potential optimization problem, using search over variant spaces to find optimal environments~\cite{Knights:2009:BlindOpt}.
@@
 The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section.
-Then section 3 introduces ``4K aliasing'', providing necessary background for interpreting the later results.
-Section 4 presents an analysis of how the size of environment variables causes bias.
-In Section 5 we provide a similar analysis for bias to heap address alignment.
-Section 6 concludes and discusses directions for future work.
+Then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
+Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias.
+In Section \ref{sec:heap} we provide a similar analysis for bias to heap address alignment.
+Section \ref{sec:conclusions} concludes and discusses directions for future work.
 
-\section{Methodology}
-In order to understand biased behavior it is important to be able to do precise measurements.
-Besides measurement bias in itself, one must also consider \emph{observer effect}.
-Any kind of instrumentation added to the code, such as a counter variable for each invocation of a function, has the potential to produce misleading results~\cite{Mytkowicz:2008:OE&MB}.
-Instead, we use instrumentation provided directly by the hardware, in the form of performance counters. 
-Modern processors have a set of model specific registers (MSRs) that can be configured to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
-Recent Intel architectures have several hundred available events, which can give a very detailed view of what happens inside the CPU.
+\section{Experimental Methodology}
+\label{sec:methodology}
+In order to understand biased behavior it is important to be able to do precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
+This can be achieved by using performance counters, instrumentation support in hardware that can be used to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
+Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.
+Performance counters are supported in the Linux kernel, accessed via a tool called \emph{perf}.
+We use the perf-stat command in all experiments, which accepts raw event codes listed in the reference manual~\cite{Volume3B}.
 
-Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
-We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Volume3B}.
 A small Python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
-The same script simulates changes to environment size, which is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
+Only a small set of events are collected at a time, to ensure events are actually counted continuously and not sampled by multiplexing between a limited set of counter registers.
+Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
 Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
-Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
-Results are averaged over multiple runs to reduce potential random error, supplying the \texttt{-r} option to perf-stat. % Not median, good question. man perf-stat says average + std.dev.
+% Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
+Results are averaged over multiple runs to reduce potential random error, supplying the -r option to perf-stat.
 
@@
 \section{4K Address Aliasing}
+\label{sec:aliasing}
 Modern processors are \emph{superscalar}, and achieves parallelism by issuing multiple instructions simultaneously and out of order.
@@
 \section{Bias from Environment Size}
+\label{sec:environment}
 Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack.
@@
 \section{Bias from Heap Allocation}
+\label{sec:heap}
 Address aliasing can be caused by conflicting pairs of load and store operations to any part of memory.
@@
 \section{Conclusions}
+\label{sec:conclusions}
 In this article we have looked at how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 

### 2014-05-24 paper.tex
Various changes

@@
 By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
-All benchmarks are done on a machine under minimal load, and \emph{frequency scaling} is disabled to keep the CPU's clock speed fixed.
+All experiments are done on a machine under minimal load, and \emph{frequency scaling} is disabled to keep the CPU's clock speed fixed.
 Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads.
@@
 One of the issues that can limit throughput is dependencies between a load and previous stores.
-According to Intel, typical software consists of about 38~\% memory access instructions, with about two thirds of them being loads.
-To increase parallelism, modern Intel architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}. 
+% According to Intel, typical software consists of about 38~\% memory access instructions, with about two thirds of them being loads.
+To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}. 
 More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
@@
 % Generic intro to memory layout (with focus on stack/env interaction)
+% todo: Consider removing?
 Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
-The stack is normally placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables.
-There is still some level of stack alignment happening, independent of environment offsets. 
-Normally, the stack is aligned to 16 bytes, which will be enforced by the compiler. % [ref]
+The stack is placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables and program arguments.
+After some variable amount of offset, the stack is normally aligned to a 16 byte boundary, which will be enforced by the compiler. % [ref]
 Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
@@
 This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data. 
+% Note that there is no clear relationship between environment size and stack location with ASLR enabled. % todo: check again
+% However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
 
-Note that there is no clear relationship between environment size and stack location with ASLR enabled.
-However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
 % In addition to affecting stack, ASLR also randomizes location of dynamically linked files (memory mapped), and initial position of heap.
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \centering
@@
 
+\subsection{Microkernel Analysis}
+To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!''~\cite{Mytkowicz:2009:WrongData}, reproduced here in Listing~\ref{lst:loopkernel}.
+This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straightforward to analyze.
+Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
+
 % Cite does not work in acm style
@@
 \end{lstlisting}
-
-\subsection{Microkernel Analysis}
-To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by T.
-Mytkowicz et.
-al.~\cite{Mytkowicz:2009:WrongData}, reproduced here in Listing~\ref{lst:loopkernel}.
-This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straightforward to analyze.
-Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
 
@@
 % See analysis for results showing this does not alias
-\begin{lstlisting}[float=b, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
+\begin{lstlisting}[float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
 #define ALIAS(a, b) \

### 2014-05-25 paper.tex
Formatting for tables, figures and listings

@@
 \usepackage{filecontents}
-\usepackage{booktabs}
+\usepackage{booktabs} % Only used for horizontal lines
 \usepackage{tikz}
 \usepackage{listings}
-\usepackage{caption}
+%\usepackage{caption}
 \usepackage{mathtools}
@@
 
-\section{Bias from Environment Size}
-\label{sec:environment}
-Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack.
-Environment variables and program arguments are allocated in the stack section of virtual memory, before the first call frame.
-Changing environment variables will therefore offset the addresses of stack, and consequently all stack allocated variables.
-
-% Generic intro to memory layout (with focus on stack/env interaction)
-% todo: Consider removing?
-Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
-The stack is placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables and program arguments.
-After some variable amount of offset, the stack is normally aligned to a 16 byte boundary, which will be enforced by the compiler. % [ref]
-Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
-Any one of these contexts could trigger aliasing issues between stack and some other area of memory remaining unaffected by environment size.
-This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data. 
-% Note that there is no clear relationship between environment size and stack location with ASLR enabled. % todo: check again
-% However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
-
-% In addition to affecting stack, ASLR also randomizes location of dynamically linked files (memory mapped), and initial position of heap.
-% Moving to the lower end of the address space, we find the text, data and bss sections, which contains program code and statically allocated data.
-% The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.
-
-\begin{figure}[t]
+\begin{narrowfig}{0.5\textwidth}[t]
   \centering
-  \caption{Organization of virtual memory for a 64-bit process.}
-  \label{fig:virtualmemory}
   \begin{tikzpicture}[font=\footnotesize]
@@
   \end{tikzpicture}
-\end{figure}
+  \caption{Relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
+  The stack is placed close to the top, but offset by environment variables and program arguments. 
+  Program code and statically allocated data is allocated at the lower end of the address space.
+  The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.}
+  % \footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}
+  \label{fig:virtualmemory}
+\end{narrowfig}
+
+\section{Bias from Environment Size}
+\label{sec:environment}
+Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack.
+Environment variables and program arguments are allocated in the stack section of virtual memory, before the first call frame.
+Changing environment variables will therefore offset the addresses of stack, and consequently all stack allocated variables.
+
+% Generic intro to memory layout (with focus on stack/env interaction)
+% todo: Consider removing?
+
+% Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
+% The stack is placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables and program arguments.
+% After some variable amount of offset, the stack is normally aligned to a 16 byte boundary, which will be enforced by the compiler. % [ref]
+% Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
+% Any one of these contexts could trigger aliasing issues between stack and some other area of memory remaining unaffected by environment size.
+% This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data.
+
+% Note that there is no clear relationship between environment size and stack location with ASLR enabled. % todo: check again
+% However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
+
+% In addition to affecting stack, ASLR also randomizes location of dynamically linked files (memory mapped), and initial position of heap.
+% Moving to the lower end of the address space, we find the text, data and bss sections, which contains program code and statically allocated data.
+% The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.
 
 \subsection{Microkernel Analysis}
-To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!''~\cite{Mytkowicz:2009:WrongData}, reproduced here in Listing~\ref{lst:loopkernel}.
+To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!''~\cite{Mytkowicz:2009:WrongData}, reproduced below.
 This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straightforward to analyze.
@@
 % Cite does not work in acm style
-% Micro-kernel presented by Mytkowicz et al~\cite{Mytkowicz:2009:WrongData}, 
-\begin{lstlisting}[float=h, language=C, caption={Micro-kernel succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
+%\begin{figure}
+  % \lstset{%
+  %   basicstyle=\ttfamily\small\bfseries,
+  %   frame=tb,
+  %   xleftmargin=0.2\textwidth,
+  %   xrightmargin=0.2\textwidth
+  % }
+  % [xleftmargin=0.2\textwidth,xrightmargin=0.2\textwidth]
+  \begin{lstlisting}
     static int i, j, k;
@@
 
+  % Can have caption with figure environment
+  %\caption{Micro-kernel succeptible to bias from variations to system environment variables.\label{lst:loopkernel}}
+%\end{figure}
+
 % Large plot of cycle count for two 4K stack sections.
@@
 \begin{figure*}[t]
-  \caption{Bias from environment size for micro-kernel in Listing~\ref{lst:loopkernel}.
-Measured average of 10 cycle count samples for 512 different environments.
-Spikes show aliasing case, occurring once for each 4K period}
-  \label{fig:envbias}
   \begin{tikzpicture}
@@
   \end{tikzpicture}
+  \caption{Bias from environment size for microkernel. Measured average of 10 cycle count samples for 512 different environments. Spikes show aliasing case, occurring once for each 4K period}
+  \label{fig:envbias}
 \end{figure*}
 
+Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
+We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
+A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
+The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
+
+There are clearly two worst cases, indicated by significant spikes at the 3184 and 7280 byte offsets.
+We sampled an extensive set of performance counters in addition to cycle count for each execution context.
+By looking at the \emph{median} value of each counter compared to the extreme cases, we narrowed down the performance events that behaved similarly to cycle count.
+Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases of 3184 and 7280 bytes, compared to the median over all environments.
+
 % Detailed comparison between median and worst cases.
-\begin{table}[b]
-  \centering
-  \caption{Events with significant correlation to cycle count}
-  \label{tab:loopcorrelation}
-  %\renewcommand{\tabcolsep}{1pt} % Less padding between cells
-  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
+\begin{table}
+  \tbl{Events with significant correlation to cycle count\label{tab:loopcorrelation}}{
     \pgfplotstabletypeset[
-    font=\footnotesize,
       int detect, % Output whole numbers for counter values
@@
     ]{bin/micro-kernel-comparison.csv}
+  }
 \end{table}
-
-Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
-We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
-A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
-The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
-
-There are clearly two worst cases, indicated by significant spikes at the 3184 and 7280 byte offsets.
-We sampled an extensive set of performance counters in addition to cycle count for each execution context.
-By looking at the \emph{median} value of each counter compared to the extreme cases, we narrowed down the performance events that behaved similarly to cycle count.
-Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases of 3184 and 7280 bytes, compared to the median over all environments.
 
@@
 \lstinputlisting[
+  basicstyle=\ttfamily\footnotesize,
   linerange={22-38}, 
@@
 Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
-Shown in Listing \ref{lst:loopfixed} is a proof of concept of how alias-free code could be generated in this particular case.
+Shown in Figure \ref{lst:loopfixed} is a proof of concept of how alias-free code could be generated in this particular case.
 If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
@@
 % See analysis for results showing this does not alias
-\begin{lstlisting}[float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
+\begin{figure}
+  % [float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
+  \begin{lstlisting}
   #define ALIAS(a, b) \
@@
   \end{lstlisting}
+  \caption{Dynamically detect aliasing case, and avoid by pushing another stack frame.}
+  \label{lst:loopfixed}
+\end{figure}
 
@@
 A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
-\begin{itemize}
+\begin{longitem}
   \item The ``regular'' heap area in virtual memory is used for smaller allocations. 
@@
   One drawback of this is that it requires the memory to be zero initialized, and whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
-\end{itemize}
+\end{longitem}
 The heap section starts at a relatively low address (close to the data section), growing upwards.
@@
 This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
-As an example of this, consider a naive implementation of linear correlation shown in Listing \ref{lst:conv} (for simplicity, endpoints are not handled).
+As an example of this, consider a naive implementation of linear correlation shown in Figure \ref{lst:conv} (for simplicity, endpoints are not handled).
 The function containins lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
@@
 
+\begin{figure}[t]
+  \centering
   \lstinputlisting[
-  float=t, 
+    %float=t, 
     language=C, 
-  caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.},
-  label={lst:conv},
-  frame=lines
+    %caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.},
+    %label={lst:conv},
+    %frame=lines
   ]{bin/convolution-kernel.c}
+  \caption{Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}
+  \label{lst:conv}
+\end{figure}
 
-% Use mmap directly instead maybe?
 By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
@@
   \centering
-  \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Listing~\ref{lst:conv}.
-Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
-Input size $n=2^{20}$.}
-  \label{fig:conv-default}
   \begin{tikzpicture}
@@
   \end{tikzpicture}
+  \label{fig:conv-default}
+  \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}.
+Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
+Input size $n=2^{20}$.}
 \end{figure*}
@@
 % Table with detailed performance data, use estimated for O2 because it does not need as many columns (!)
-\begin{table*}[t]
-  \centering
-  \caption{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}
+\begin{table}[t]
+  %\centering
   \label{tab:convstats}
-  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
+  \renewcommand{\tabcolsep}{3pt} % Less padding between cells
+  \tbl{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}{
     \pgfplotstabletypeset[
@@
     ]{bin/conv-default-o2.estimate.csv}
-\end{table*}
+  }
+\end{table}
 
@@
   \centering
-  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}
   \label{tab:mallocompare}
   \renewcommand{\tabcolsep}{4pt} % Less padding between cells
+  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
     \pgfplotstabletypeset[
@@
     ]{bin/malloc-comparison.csv}
+  }
 \end{table*}
@@
 
-\paragraph{Manually adjust memory}
+\paragraph{Manually adjust address offsets}
 In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.

### 2014-05-25 paper.tex
Fixes and cleanup in microkernel section

@@
   \caption{Relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory.
-  The stack is placed close to the top, but offset by environment variables and program arguments. 
-  Program code and statically allocated data is allocated at the lower end of the address space.
-  The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.}
-  % \footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}
+  Initial addresses of stack, heap and memory mapped files are often randomized for security reasons.
+  Addresses of code and statically allocated data are allocated at compile time by the linker, and can be determined by inspecting the executable.}
   \label{fig:virtualmemory}
@@
 \label{sec:environment}
-Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack.
-Environment variables and program arguments are allocated in the stack section of virtual memory, before the first call frame.
+Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the position and alignment of stack.
+As indicated in Figure \ref{fig:virtualmemory}, environment variables and program arguments are allocated in the stack section of virtual memory close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, before the first call frame.
 Changing environment variables will therefore offset the addresses of stack, and consequently all stack allocated variables.
-
-% Generic intro to memory layout (with focus on stack/env interaction)
-% todo: Consider removing?
-
-% Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
-% The stack is placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables and program arguments.
-% After some variable amount of offset, the stack is normally aligned to a 16 byte boundary, which will be enforced by the compiler. % [ref]
-% Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
-% Any one of these contexts could trigger aliasing issues between stack and some other area of memory remaining unaffected by environment size.
-% This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data.
-
-% Note that there is no clear relationship between environment size and stack location with ASLR enabled. % todo: check again
-% However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
-
-% In addition to affecting stack, ASLR also randomizes location of dynamically linked files (memory mapped), and initial position of heap.
-% Moving to the lower end of the address space, we find the text, data and bss sections, which contains program code and statically allocated data.
-% The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.
+After some variable amount of offset, the stack is normally aligned to a 16 byte boundary, which will be enforced by the compiler.
+Within a span of 4096 bytes there are thus 256 possible initial stack addresses, each representing a different execution context with respect to address aliasing.\footnote{Note that there is no clear relationship between environment size and stack location with ASLR enabled.
+However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.}
 
@@
 
-% Cite does not work in acm style
 %\begin{figure}
@@
 By looking at the \emph{median} value of each counter compared to the extreme cases, we narrowed down the performance events that behaved similarly to cycle count.
-Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases of 3184 and 7280 bytes, compared to the median over all environments.
+Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases, compared to the median over all environments.
 
@@
   }
+  \begin{tabnote}
+    \Note{Note:}{Performance events that are obviously not indicative of any causal relationship are omitted, for example bus-cycles which will naturally vary with total cycle count.}
+  \end{tabnote}
 \end{table}
 
-Notice the most extreme change from median to worst case is clearly the number of alias events.
+We see the most extreme change from median to worst case is clearly the number of alias events.
 If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
 The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
-% Some of the other counters are less interesting; for example bus cycles, which will naturally vary with the total cycle count.
-% Performance events that are obviously not indicative of any causal relationship are omitted.
-
 On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270,000 to about 135,000.
@@
 
-\subsection{The Offending Variables}
-Our performance counter data clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
+Performance counter data clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
 For that we need to know the addresses of each variable at runtime.
@@
 Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
-In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR. 
+In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR or program arguments. 
 Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.

### 2014-05-25 paper.tex
Reduced bloat in allocator comparison table
Plus other changes

@@
 \lstinputlisting[
-  basicstyle=\ttfamily\footnotesize,
+  basicstyle=\ttfamily\small,
   linerange={22-38}, 
@@
 A heap allocator is responsible for managing dynamic memory, and ultimately assigns the actual addresses of heap allocated variables at runtime.
-Heap allocation routines such as \texttt{malloc} and \texttt{free} are typically dynamically linked, for example as part of glibc, or some other library.
+Heap allocation routines such as \texttt{malloc} and \texttt{free} are typically dynamically linked, for example as part of glibc.
 The particular library used therefore constitutes an important part of the execution context, as linking to a different library, or a library with some alternative configuration, can impact heap addresses at runtime.
 
-\subsection{\texttt{malloc} will Alias by Default}
+\subsection{Why \texttt{malloc} will Alias by Default}
 Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
 A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
-\begin{longitem}
+\begin{itemize}
   \item The ``regular'' heap area in virtual memory is used for smaller allocations. 
   Programs are initially given a relatively small heap, which the allocator can increase by calling \texttt{brk} or \texttt{sbrk} system calls.
-
   \item For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
   One drawback of this is that it requires the memory to be zero initialized, and whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
-\end{longitem}
+\end{itemize}
 The heap section starts at a relatively low address (close to the data section), growing upwards.
@@
 However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address end with 0x010.}
+The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
 This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.
@@
 This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
-As an example of this, consider a naive implementation of linear correlation shown in Figure \ref{lst:conv} (for simplicity, endpoints are not handled).
+As an example of this, consider a naive implementation of linear correlation shown in Figure \ref{lst:conv}.
 The function containins lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
@@
   \lstinputlisting[
-    %float=t, 
     language=C, 
-    %caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.},
-    %label={lst:conv},
     %frame=lines
   ]{bin/convolution-kernel.c}
-  \caption{Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}
+  \caption{Naive implementation of convolution, ignoring endpoints for simplicity. This program is highly sensitive to aliasing between input and output arrays.}
   \label{lst:conv}
@@
 
-\begin{table*}[t]
-  \centering
-  \label{tab:mallocompare}
-  \renewcommand{\tabcolsep}{4pt} % Less padding between cells
-  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
-    \pgfplotstabletypeset[
-      font=\footnotesize,
-      col sep=comma,
-      string type,
-      columns={Allocation, 64, 5120, 65536, 1048576, 4194304},
-      column type=r,
-      columns/Allocation/.style={
-        string type, 
-        column type=l
-      },
-      every head row/.style={
-        output empty row,
-        before row={\toprule
-           & 64 B & 5,120 B & 65,536 B & 1,048,576 B & 4,194,304 B \\
-          },
-        after row=\midrule,
-      },
-      every odd row/.style={after row=\midrule},
-      every last row/.style={after row=\bottomrule}
-    ]{bin/malloc-comparison.csv}
-  }
-\end{table*}
-
 Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
@@
 We do not care about performance, only whether the actual memory addresses returned alias or not.
+
+% Removed some data here, too much info
+\begin{table*}[t]
+  \centering
+  \label{tab:mallocompare}
+  %\renewcommand{\tabcolsep}{4pt} % Less padding between cells
+  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
+    \pgfplotstabletypeset[
+      font=\footnotesize,
+      col sep=comma,
+      string type,
+      columns={Allocation, 64, 5120, 1048576},
+      column type=r,
+      columns/Allocation/.style={
+        string type, 
+        column type=l
+      },
+      every head row/.style={
+        output empty row,
+        before row={\toprule
+           & 64 B & 5,120 B & 1,048,576 B \\
+          },
+        after row=\midrule,
+      },
+      every odd row/.style={after row=\midrule},
+      every last row/.style={after row=\bottomrule}
+    ]{bin/malloc-comparison.csv}
+  }
+\end{table*}
 

### 2014-05-25 paper.tex
Remove booktabs style from tables

@@
 \usepackage{filecontents}
-\usepackage{booktabs} % Only used for horizontal lines
+%\usepackage{booktabs} % Only used for horizontal lines
 \usepackage{tikz}
@@
         column type=l,
+        column type/.add={|}{},
         postproc cell content/.append code={
@@
         output empty row,
-        before row={\toprule
+        before row={\hline
           Performance counter & Median & Spike 1 & Spike 2 \\
         },
-        after row=\midrule,
+        after row=\hline\hline
       },
-      every last row/.style={after row=\bottomrule}
+      every last row/.style={after row=\hline},
+      every last column/.style={column type/.add={}{|}}
     ]{bin/micro-kernel-comparison.csv}
@@
 \begin{table}[t]
-  %\centering
-  \label{tab:convstats}
-  \renewcommand{\tabcolsep}{3pt} % Less padding between cells
-  \tbl{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}{
+  %\renewcommand{\tabcolsep}{3pt} % Less padding between cells
+  \tbl{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.\label{tab:convstats}}{
     \pgfplotstabletypeset[
-      font=\footnotesize,
+      %font=\footnotesize,
       int detect, % Output whole numbers for counter values
@@
         output empty row,
-        before row={\toprule
+        before row={\hline
           Performance counter & $r$ & 0 & 2 & 4 & 8 \\
         },
-        after row=\midrule,
+        after row=\hline\hline
       },
-      every last row/.style={after row=\bottomrule}
+      every last row/.style={after row=\hline},
+      every last column/.style={column type/.add={}{|}}
     ]{bin/conv-default-o2.estimate.csv}
@@
 
-% Removed some data here, too much info
-\begin{table*}[t]
+\begin{table}[t]
   \centering
@@
     \pgfplotstabletypeset[
-      font=\footnotesize,
       col sep=comma,
@@
         string type, 
-        column type=l
+        column type=l,
+        column type/.add={|}{}
       },
@@
         output empty row,
-        before row={\toprule
+        before row={\hline
            & 64 B & 5,120 B & 1,048,576 B \\
           },
-        after row=\midrule,
+        after row=\hline\hline,
       },
-      every odd row/.style={after row=\midrule},
-      every last row/.style={after row=\bottomrule}
+      every last row/.style={after row=\hline},
+      every last column/.style={column type/.add={}{|}},
+      every odd row/.style={after row=\hline},
     ]{bin/malloc-comparison.csv}
   }
-\end{table*}
+\end{table}
 

### 2014-05-25 paper.tex
Fix width of heap table

@@
 \begin{table}[t]
-  %\renewcommand{\tabcolsep}{3pt} % Less padding between cells
-  \tbl{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.\label{tab:convstats}}{
+  \renewcommand{\tabcolsep}{3pt} % Less padding between cells
+  \tbl{Relevant performance counters and correlation ($r$) with cycle count for optimization O2. Estimated cost accounting for constant overhead.\label{tab:convstats}}{
     \pgfplotstabletypeset[
-      %font=\footnotesize,
       int detect, % Output whole numbers for counter values

### 2014-05-25 paper.tex
Misc. changes and improvements

@@
 Predicting or accounting for measurment bias is challenging, as the interaction with low level hardware mechanisms is complex.  
-In this paper we show how \emph{address aliasing} can often be the underlying mechanism resulting in measurement bias on modern Intel microarchitectures.
+In this paper we show how \emph{address aliasing} can be the underlying mechanism resulting in measurement bias on modern Intel microarchitectures.
 Our approach is to use hardware performance counter to analyze performance data over controlled series of execution contexts.
@@
 While the aliasing mechanism itself is not new, we provide the connection to measurement bias and show how to use this knowledge not only to account for bias but also explicitly optimize for it.
-Our results should provide motivation for more architecture specific optimization for example in heap allocator heuristics.
+Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics.
 \end{abstract}
@@
 
-The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section.
-Then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
+The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section, then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
 Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias.
-In Section \ref{sec:heap} we provide a similar analysis for bias to heap address alignment.
-Section \ref{sec:conclusions} concludes and discusses directions for future work.
+In Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
+Section \ref{sec:conclusions} includes a summary and conclusion.
 
@@
 \begin{description}
-  \item[LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS]
-  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{OptimizationManual} % todo: where
+  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}] ``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.'' Intel Optimization Manual~\citeyear[B.3.4.4]{OptimizationManual}
 \end{description}
@@
 \subsection{Microkernel Analysis}
-To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!''~\cite{Mytkowicz:2009:WrongData}, reproduced below.
+To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, reproduced below.
 This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straightforward to analyze.
@@
 The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
-This phenomenon is only observed for address offsets close to zero, thus we only show the first 32 data points.
+This phenomenon is only observed for address offsets close to zero, thus we only show the first 20 data points.
 If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
@@
 The numbers are for the level 2 optimization case, which was chosen because the behavior was less chaotic and thus more easily represented by only a few numbers.
-However, the performance data looks fairly similar between the two data sets, with a couple of events that seem to stand out.
+However, the performance data looks fairly similar between the two data sets, with a couple of events that seem to stand out:
 \begin{itemize}
@@
     \item Changes to the number of micro-ops executed for certain ports.
-    \item Offcore requests. (++) % todo!
 \end{itemize}
@@
 
-It is also worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
+It is worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
 The L1 hit rate remains stable across all offsets, and only a neglible number of memory loads actually misses L1.
+On the other hand, we see a fairly strong correlation between cycle count and outstanding offcore requests. 
+These events count the number of outstanding loads to memory outside the processor core each cycle.
+Since we do not see any significant number of loads missing L1, the correlation to outstanding loads is probably a result of stalling and more cycles executed.
 
@@
 Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
-
 We do not attempt to pinpoint exactly which assembly instructions might alias, but on a high level we can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
@@
 \begin{table}[t]
-  \centering
   \label{tab:mallocompare}
-  %\renewcommand{\tabcolsep}{4pt} % Less padding between cells
   \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
@@
   }
+  \begin{tabnote}
+    \Note{Note:}{Memory mapped addresses starting with the 0x2aa...
+prefix is an artifact of using \emph{make} to generate results.
+Executing the test program from \emph{bash} directly result in 0x7fff...
+prefixes.
+    The difference is not important to our discussion.}
+  \end{tabnote}
 \end{table}
@@
 
-\subsection{Ways to Deal with Address Aliasing}
+\subsection{Ways to Deal with Heap Address Aliasing}
 Measurement bias from address aliasing can be significant in real applications, with potentially large performance penalties.
@@
 \label{sec:conclusions}
-In this article we have looked at how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
+In this paper we have shown how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
 The effect is caused by the way speculative and out-of-order memory operations are handled by the CPU, only considering the last 12 address bits to resolve conflicts between load and store operations.
 
-By analyzing an example originally presented by Mytkowicz et al.
-\cite{Mytkowicz:2009:WrongData}, we determined that collisions between stack variables and static data resulted in aliasing for certain stack positions.
-In this case, the external trigger of bias is variations in environment size, which in turn offset the virtual addresses of stack allocated variables.
+In general, any change to virtual memory layout of data can potentially introduce bias effects from address aliasing.
+Analyzing an example with bias to enviornment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
+Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
+We show how aliasing can have significant performance impact in practice for algorithms operating on heap allocated buffers, analyzing and example with more than 50 \% performance variation between different memory layouts.
+Bacause typical heap allocators will return aliased pointers for large allocations, worst case performance is likely to be the default.
 
-In general, any change to virtual memory layout of data can potentially introduce bias effects from address aliasing. 
-We present a code example with extreme sensitivity to heap address alignment, with more than 50 \% performance variation between different memory layouts.
-We find that typical heap allocators return page-aligned memory by default, which is the worst case with respect to aliasing for many algorithms.
-Despite recommendations from Intel, we are not aware of any allocators that specifically addresses this issue. 
-We show how manual padding of variables, and alternative alias-free code paths, can be used to avoid aliasing at runtime.
-This means that an understanding of 4K address aliasing is sometimes needed to achieve optimal performance.
+We show how techniques like padding of variables and alternative alias-free code paths can be used to avoid aliasing at runtime.
+For heap address aliasing especially, we find that manual intervention is sometimes required in order to achieve optimal performance.
+Our results should inspire more clever optimizations and heuristics taking address aliasing effects into account, as the potential performance improvements can be substantial.
 
-\section{Acknowledgements}
+% Acknowledgements
+\begin{acks}
 This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurment bias on the Intel Ivy Bridge architecture~\cite{MasterThesis}.
+\end{acks}
 
+% Bibliography
 \bibliographystyle{ACM-Reference-Format-Journals}
@@
 
+% History dates
+\received{}{}{}
+
 \end{document}

### 2014-05-25 paper.tex
Put borders around code listings

@@
 \begin{figure}
-  % [float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
-  \begin{lstlisting}
+  \begin{lstlisting}[frame=single, xleftmargin=.1\textwidth, xrightmargin=.1\textwidth]
 #define ALIAS(a, b) \
@@
     language=C,
-    %frame=lines
+    frame=single,
+    xleftmargin=.1\textwidth,
+    xrightmargin=.1\textwidth
   ]{bin/convolution-kernel.c}

### 2014-05-25 paper.tex
Added related work section

@@
 
+\section{Related Work}
+\label{sec:related}
+Measurement bias and observer effect in performance analysis has been studied in detail by Mytkowicz et al~\shortcite{Mytkowicz:2008:OE&MB,Mytkowicz:2008:Easy}.
+The authors introduce environment variables and link ordering as examples of external bias triggers, showing how standardised {\small SPECint} benchmarks suffer from measurement bias when trying to evaluate the effectiveness of O3 over O2.
+Strategies for detecting bias through causal analysis and randomization of experimental setups is introduced.
+The microkernel example we analyzed is taken from a followup paper by the same authors, presenting similar results for an older Core 2 processor~\cite{Mytkowicz:2009:WrongData}.
+Assuming the root causes of measurement bias were the same, our results indicate that address aliasing issues is probably relevant on several previous generations of Intel architectures as well.
+
+Techniques for ``Blind'' optimization using search over variant spaces, is introduced by \citeN{Knights:2009:BlindOpt}.
+Their space of variants consist of different position and alignments for each function and global variable, in principle exploring the different configurations possible by altering link order.
+The authors of MAO present a framework for extending compilers to provide very low level microarchitectural assembly optimization~\cite{Hundt:2011:MAO}.
+Different techniques are used, including rules, pattern matching, and random insertion of nop-instructions.
+
 \section{Conclusions}

### 2014-05-26 paper.tex
Rewrite malloc subsection slightly

@@
   \begin{tikzpicture}[font=\footnotesize]
-
     % See page 453 in pgf manual
@@
       };
-
     \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.seven split west) 
@@
         { Program code and static data } ;
-
     \node[right] at (m.north east)
       { \scriptsize{0x7fff'ffffffff} } ;
-
     \node[right] at (m.south east)
       { \scriptsize{0x400000} } ;
-
   \end{tikzpicture}
@@
 Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
-A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
-\begin{itemize}
-  \item The ``regular'' heap area in virtual memory is used for smaller allocations. 
-  Programs are initially given a relatively small heap, which the allocator can increase by calling \texttt{brk} or \texttt{sbrk} system calls.
-  \item For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
-  One drawback of this is that it requires the memory to be zero initialized, and whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
-\end{itemize}
-The heap section starts at a relatively low address (close to the data section), growing upwards.
-Memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack.
+Depending on the particular request and allocator used, the returned value will either point to the ``regular'' heap, or to a memory mapped area (see Figure \ref{fig:virtualmemory}).
+Different strategies for how to allocate memory will often depend on the \emph{size} of the request.
+
+\begin{enumerate}
+  \item The heap is marked by a brake point representing the end of uninitialized data in virtual memory, and more space is requested by the \texttt{brk} or \texttt{sbrk} system calls. 
+  Allocators typically use this area for smaller allocations.
+  \item Anonymous memory mappings, i.e. buffers not backed by a file, can be created through the \texttt{mmap} system call.
+  This mechanism is typically used for large allocations.
+  %For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
+\end{enumerate}
+% One drawback of using memory mapping is that it requires the buffer to be zero initialized, so which mechanism is used can be based on some cost heuristic employed by the allocator.
+% A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
+% The normal heap is often used for smaller requests, while memory mapping is used for large buffers.
+
+The heap section starts at a relatively low address right above static code and data.
+Memory mapped chunks are placed towards the upper end of the virtual address space, closer to stack.
 Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the actual pointer values returned.
-While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger -- for example 0x7f0318a8f010 or 0x7f03105d2010.
+While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
 This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.

### 2014-05-26 paper.tex
Merge malloc/allocator sections

@@
 
-\subsection{Why \texttt{malloc} will Alias by Default}
+\subsection{Most Allocators Alias by Default}
 Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
@@
   This mechanism is typically used for large allocations.
-  %For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
 \end{enumerate}
-% One drawback of using memory mapping is that it requires the buffer to be zero initialized, so which mechanism is used can be based on some cost heuristic employed by the allocator.
-% A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
-% The normal heap is often used for smaller requests, while memory mapping is used for large buffers.
 
@@
 Memory mapped chunks are placed towards the upper end of the virtual address space, closer to stack.
-Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the actual pointer values returned.
+Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the pointer values returned:
 While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
@@
 However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
+The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
 This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.
+
+Table \ref{tab:mallocompare} illustrates how different dynamic allocators tend to produce aliasing conditions for large requests\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
+The addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} are observed for different size parameters, and equal three digit address suffix indicate an aliasing pair.
+In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
+\begin{itemize}
+  \item Thread-Caching Malloc (tcmalloc) by Google~\cite{TCMalloc}.
+  \item jemalloc, originally developed for FreeBSD~\cite{JEMalloc}.
+  \item Hoard~\cite{Berger:2000:Hoard}.
+\end{itemize}
+All of the above focuses heavily on performance in multithreaded environments.
+Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
+Here we only look at behavior for a single thread, and whether the addresses returned alias or not.
+\begin{table}[t]
+  \label{tab:mallocompare}
+  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
+    \pgfplotstabletypeset[
+      col sep=comma,
+      string type,
+      columns={Allocation, 64, 5120, 1048576},
+      column type=r,
+      columns/Allocation/.style={
+        string type, 
+        column type=l,
+        column type/.add={|}{}
+      },
+      every head row/.style={
+        output empty row,
+        before row={\hline
+           & 64 B & 5,120 B & 1,048,576 B \\
+          },
+        after row=\hline\hline,
+      },
+      every last row/.style={after row=\hline},
+      every last column/.style={column type/.add={}{|}},
+      every odd row/.style={after row=\hline},
+    ]{bin/malloc-comparison.csv}
+  }
+  \begin{tabnote}
+    \Note{Note:}{Memory mapped addresses starting with the 0x2aa...
+prefix is an artifact of using \emph{make} to generate results.
+Executing the test program from \emph{bash} directly result in 0x7fff...
+prefixes.
+    The difference is not important to our discussion.}
+  \end{tabnote}
+\end{table}
+
+For smaller allocations, we see that glibc and tcmalloc utilize the normal heap area -- returning numerically low addresses.
+Interestingly, jemalloc and Hoard appears to never use the heap, but allocate to memory mapped areas even for smaller requests.
+Conversely, tcmalloc seem manage only the heap.
+We see one example where one allocator yields aliasing buffers while another does not.
+Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
+Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.
+But even with randomization, addresses returned by \texttt{mmap} must still be page aligned.
+This means that addresses returned by allocators using this mechanism directly will \emph{always} alias, giving a deterministic execution context considering only the address suffix.
+From our limited experiment, this seems to be the case for glibc, jemalloc and Hoard.
 
@@
 As an example of this, consider a naive implementation of linear correlation shown in Figure \ref{lst:conv}.
-The function containins lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
+The function contains lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
 We will see that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.
@@
 To analyse performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
-This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the arguments before calling the function.
+This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
 Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
 Allocating and managing these buffers at startup takes a non-neglegible amount of work.
-The overhead can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
+The overhead can be masked by repeatedly invoking the convolution kernel after allocating and initializing all the inputs.
 \begin{lstlisting}
     for (i = 0; i < k; ++i)
-        convolution(n, input, output + offset);
+        conv(n, input, output + offset);
 \end{lstlisting}
 An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
-
 \[
@@
 \]
-
 Our results are with $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
@@
 
-\subsection{Behavior of other Allocators}
-As with the contents of environment variables, the exact behavior of dynamic memory allocation can typically not be known ahead of time.
-Heap allocation routines such as \texttt{malloc} and \texttt{free} are usually dynamically linked and resolved at runtime, which again represents a potential source of bias.
-For the code examples presented earlier, the dynamic linker will resolve \texttt{malloc} to some version of glibc, which is the default on our system.
-Variations between different versions of libc, or local configurations on parameters such as thresholds for when to use \texttt{mmap} over \texttt{brk}, could trigger aliasing cases.
-
-Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
-In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
-\begin{itemize}
-  \item Thread-Caching Malloc (tcmalloc) from Google. It claims to be time- and space efficient, and reduces lock contention for multithreaded applications~\cite{TCMalloc}.
-  \item jemalloc was originally developed for FreeBSD. It aims to be a scalable general purpose allocator in single or multithreaded environments~\cite{JEMalloc}.
-  \item Hoard is another project that focuses extensively on improving performance for multithreaded programs~\cite{Berger:2000:Hoard}.
-\end{itemize}
-The addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} are observed for different size parameters. 
-We do not care about performance, only whether the actual memory addresses returned alias or not.
-
-\begin{table}[t]
-  \label{tab:mallocompare}
-  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
-    \pgfplotstabletypeset[
-      col sep=comma,
-      string type,
-      columns={Allocation, 64, 5120, 1048576},
-      column type=r,
-      columns/Allocation/.style={
-        string type, 
-        column type=l,
-        column type/.add={|}{}
-      },
-      every head row/.style={
-        output empty row,
-        before row={\hline
-           & 64 B & 5,120 B & 1,048,576 B \\
-          },
-        after row=\hline\hline,
-      },
-      every last row/.style={after row=\hline},
-      every last column/.style={column type/.add={}{|}},
-      every odd row/.style={after row=\hline},
-    ]{bin/malloc-comparison.csv}
-  }
-  \begin{tabnote}
-    \Note{Note:}{Memory mapped addresses starting with the 0x2aa...
-prefix is an artifact of using \emph{make} to generate results.
-Executing the test program from \emph{bash} directly result in 0x7fff...
-prefixes.
-    The difference is not important to our discussion.}
-  \end{tabnote}
-\end{table}
-
-For smaller allocations, we see that glibc and tcmalloc uses the normal heap area -- returning numerically low addresses.
-Interestingly, jemalloc and Hoard appears to never use the normal heap, but allocate to memory mapped areas even for smaller sizes.
-We see one example where one allocator yields aliasing buffers while another does not.
-Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
-Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.
-
-%The purpose of this is to illustrate that allocators is yet another source of bias.
-Using heap allocated memory, address aliasing is an artifact of the allocator used.
-With different versions or configurations of memory allocators, any performance impact from aliasing can qualify as measurement bias.
-
-% Most malloc alternatives focuses heavily on performance in multithreaded environments.
-% Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
-
 \subsection{Ways to Deal with Heap Address Aliasing}

### 2014-05-26 paper.tex
Update aslr reference

@@
 Most importantly, this means keeping the memory address space under control.
-For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Shackham:2004:ASLR}.
-% Should maybe have more generic reference
+For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Pax:ASLR,Bhatkar:AddressObfuscation}.
 By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.

### 2014-05-27 paper.tex
Add related work reference
Remove comments
Big changes to the abstract.

@@
 \begin{abstract} % Should be 150 -- 200 words max
-Understanding program performance and obtaining accure measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
-Predicting or accounting for measurment bias is challenging, as the interaction with low level hardware mechanisms is complex.  
-In this paper we show how \emph{address aliasing} can be the underlying mechanism resulting in measurement bias on modern Intel microarchitectures.
-Our approach is to use hardware performance counter to analyze performance data over controlled series of execution contexts.
-We show how address aliasing can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.
+%Understanding program performance and obtaining accure measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
+%Predicting or accounting for measurment bias is challenging, as the interaction with low level hardware mechanisms is complex.  
+%In this paper we show how \emph{address aliasing} can be the underlying mechanism resulting in measurement bias on modern Intel microarchitectures.
+%Our approach is to use hardware performance counter to analyze performance data over controlled series of execution contexts.
+%We show how address aliasing can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.
 
-The perfomance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
+%The perfomance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
+%It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
+%While the aliasing mechanism itself is not new, we provide the connection to measurement bias and show how to use this knowledge not only to account for bias but also explicitly optimize for it.
+%Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics.
+Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
+It has been shown that there can be significant performance differences based seemingly irrelevant factors like user name length.
+In this paper we show an important underlying mechanism resulting in this type of measurement bias on modern Intel microarchitectures.
+
+Our approach is to search for correlations between a large set of hardware performance counters and performance figures from multiple environment contexts.
+We find that \emph{address aliasing} can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.
+
+This clear result not only makes accounting and avoiding measurement bias much simpler, but enables explicit optimizations for it.
 It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
-While the aliasing mechanism itself is not new, we provide the connection to measurement bias and show how to use this knowledge not only to account for bias but also explicitly optimize for it.
-Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics.
+The performance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
+Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics and compiler rules.
+
 \end{abstract}
@@
 In Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
+Related work is discussed in Section \ref{sec:related}.
 Section \ref{sec:conclusions} includes a summary and conclusion.
@@
 
-%\begin{figure}
-  % \lstset{%
-  %   basicstyle=\ttfamily\small\bfseries,
-  %   frame=tb,
-  %   xleftmargin=0.2\textwidth,
-  %   xrightmargin=0.2\textwidth
-  % }
-  % [xleftmargin=0.2\textwidth,xrightmargin=0.2\textwidth]
 \begin{lstlisting}
@@
 \end{lstlisting}
-
-  % Can have caption with figure environment
-  %\caption{Micro-kernel succeptible to bias from variations to system environment variables.\label{lst:loopkernel}}
-%\end{figure}
 

### 2014-05-27 paper.tex
Minor fixes to the abstract.

@@
 Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
-It has been shown that there can be significant performance differences based seemingly irrelevant factors like user name length.
+Recent research has shown that there can be significant performance differences based seemingly irrelevant factors like user name length.
 In this paper we show an important underlying mechanism resulting in this type of measurement bias on modern Intel microarchitectures.
-
 Our approach is to search for correlations between a large set of hardware performance counters and performance figures from multiple environment contexts.
@@
 
-This clear result not only makes accounting and avoiding measurement bias much simpler, but enables explicit optimizations for it.
-It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
-The performance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
+This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
+Further, we show that the allocation logic of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
+The performance impact of unfavorable memory layout combinations can be significant, with as much as $2x$ speedup in one of our examples.
 Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics and compiler rules.

### 2014-05-27 paper.tex
More fixes to the abstract. Fixed some typos elsewhere.

@@
 Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
-Recent research has shown that there can be significant performance differences based seemingly irrelevant factors like user name length.
-In this paper we show an important underlying mechanism resulting in this type of measurement bias on modern Intel microarchitectures.
+Recent research has shown that there can be significant performance differences based on seemingly irrelevant factors like user name length.
+In this paper we identify an important underlying mechanism that cause this type of measurement bias on modern Intel microarchitectures.
 Our approach is to search for correlations between a large set of hardware performance counters and performance figures from multiple environment contexts.
@@
 This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
+We demonstrate this by implementing runtime detection and correction of address aliasing, eliminating the bias.
 Further, we show that the allocation logic of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
 The performance impact of unfavorable memory layout combinations can be significant, with as much as $2x$ speedup in one of our examples.
-Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics and compiler rules.
+%Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics and compiler rules.
+In the longer term, our results can enable novel compiler and architectural optimization strategies.
 
@@
 Accurately measuring the performance of computer programs is important in order to evaluate new algorithms or compare different implementations.
-This process is complicated by \emph{measurment bias}, which can be a challenge for performance analysist and systems researchers.
+This process is complicated by \emph{measurement bias}, which can be a challenge for performance analysist and systems researchers.
 Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
@@
 By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
-To analyse performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
+To analyze performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
 This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
 Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
-Allocating and managing these buffers at startup takes a non-neglegible amount of work.
+Allocating and managing these buffers at startup takes a non-negligible amount of work.
 The overhead can be masked by repeatedly invoking the convolution kernel after allocating and initializing all the inputs.
@@
 It is worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
-The L1 hit rate remains stable across all offsets, and only a neglible number of memory loads actually misses L1.
+The L1 hit rate remains stable across all offsets, and only a negligible number of memory loads actually misses L1.
 On the other hand, we see a fairly strong correlation between cycle count and outstanding offcore requests. 
@@
 A potential solution could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
-This goes somewhat agains conventional wisdom that more alignment is better.
-The Intel optimization manual actually mentiones this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
+This goes somewhat against conventional wisdom that more alignment is better.
+The Intel optimization manual actually mentions this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
 However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
@@
 In general, any change to virtual memory layout of data can potentially introduce bias effects from address aliasing.
-Analyzing an example with bias to enviornment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
+Analyzing an example with bias to environment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
 Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
 We show how aliasing can have significant performance impact in practice for algorithms operating on heap allocated buffers, analyzing and example with more than 50 \% performance variation between different memory layouts.
-Bacause typical heap allocators will return aliased pointers for large allocations, worst case performance is likely to be the default.
+Because typical heap allocators will return aliased pointers for large allocations, worst case performance is likely to be the default.
 
@@
 \begin{acks}
-This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurment bias on the Intel Ivy Bridge architecture~\cite{MasterThesis}.
+This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel Ivy Bridge architecture~\cite{MasterThesis}.
 \end{acks}

### 2014-05-27 paper.tex
Improve malloc/heap section

@@
 Depending on the particular request and allocator used, the returned value will either point to the ``regular'' heap, or to a memory mapped area (see Figure \ref{fig:virtualmemory}).
-Different strategies for how to allocate memory will often depend on the \emph{size} of the request.
-
-\begin{enumerate}
-  \item The heap is marked by a brake point representing the end of uninitialized data in virtual memory, and more space is requested by the \texttt{brk} or \texttt{sbrk} system calls. 
-  Allocators typically use this area for smaller allocations.
-  \item Anonymous memory mappings, i.e. buffers not backed by a file, can be created through the \texttt{mmap} system call.
-  This mechanism is typically used for large allocations.
-\end{enumerate}
+\begin{itemize}
+  \item The \emph{heap} is marked by a break point representing the end of uninitialized data in virtual memory, and more space is requested by the \texttt{brk} or \texttt{sbrk} system calls.
+  %Allocators typically use this area for smaller allocations.
+  \item The \texttt{mmap} system call is used to map file descriptors to virtual memory. \emph{Anonymous} mappings, i.e. buffers not backed by a file, can be used for general purpose allocations.
+  %This mechanism is typically used for large allocations.
+\end{itemize}
+The allocator included as part of glibc uses both mechanisms, choosing which based on the \emph{size} of the request~\cite{GlibcManual}.
+Small sizes generally live in the heap, while long lived and large allocations tend to be memory mapped.
 
@@
 Memory mapped chunks are placed towards the upper end of the virtual address space, closer to stack.
-Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the pointer values returned:
-While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
+Whether a request is served by the heap or by memory mapping is therefore easy to determine just by looking at the pointer values returned:
+Addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, while pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
 This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
 However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
-This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.
+The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address end with 0x010.}
+This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following section.
 
-Table \ref{tab:mallocompare} illustrates how different dynamic allocators tend to produce aliasing conditions for large requests\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
-The addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} are observed for different size parameters, and equal three digit address suffix indicate an aliasing pair.
+Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers \footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
+We observe the addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} for different size parameters.
+Equal three digit address suffix indicate an aliasing pair.
 In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
-\begin{itemize}
+\begin{enumerate}
   \item Thread-Caching Malloc (tcmalloc) by Google~\cite{TCMalloc}.
@@
   \item Hoard~\cite{Berger:2000:Hoard}.
-\end{itemize}
-All of the above focuses heavily on performance in multithreaded environments.
+\end{enumerate}
+All of the above focus heavily on performance in multithreaded environments.
 Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
-Here we only look at behavior for a single thread, and whether the addresses returned alias or not.
+Here, we only look at behavior for a single thread, and whether the addresses returned alias or not.
 \begin{table}[t]
@@
 
-For smaller allocations, we see that glibc and tcmalloc utilize the normal heap area -- returning numerically low addresses.
+We see that glibc and tcmalloc utilize the normal heap area for smaller allocations, returning numerically low addresses. % Verified by looking at strace that tcmalloc actually uses brk
 Interestingly, jemalloc and Hoard appears to never use the heap, but allocate to memory mapped areas even for smaller requests.
 Conversely, tcmalloc seem manage only the heap.
-We see one example where one allocator yields aliasing buffers while another does not.
+There is one example where an allocator yields aliasing buffers while another does not.
 Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
 Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.
-But even with randomization, addresses returned by \texttt{mmap} must still be page aligned.
+But even with randomization, pointers returned by \texttt{mmap} must still be page aligned.
 This means that addresses returned by allocators using this mechanism directly will \emph{always} alias, giving a deterministic execution context considering only the address suffix.
@@
 This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
-As an example of this, consider a naive implementation of linear correlation shown in Figure \ref{lst:conv}.
+As an example of this, consider a na{\"i}ve implementation of linear correlation shown in Figure \ref{lst:conv}.
 The function contains lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
-We will see that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.
+The performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.
 
@@
 
-By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
-By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
+We use an input size of $n=2^{20}$ (4 MiB in memory for each array), which results in glibc's heap allocator always choosing \texttt{mmap} to serve requests.
+By default, even with address randomization enabled, both \texttt{input} and \texttt{output} will have start addresses with the same address suffix of 0x010.
 To analyze performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
 This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
-Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
+Controlling the offset parameter, we can create environments where the address suffixes are some number of \texttt{sizeof(float)} bytes apart.
 Allocating and managing these buffers at startup takes a non-negligible amount of work.
@@
 Our results are with $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
-For every run, performance counter measurements is also averaged over 10 samples, using the repeat mechanism of perf.
+For every run, performance counter measurements are also averaged over 10 samples, using the repeat mechanism of perf.
 
@@
 
-More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we have selected a subset that seem to correlate well with the change to cycle count.
+More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we have selected a subset that correlates well with the change to cycle count.
 The numbers are for the level 2 optimization case, which was chosen because the behavior was less chaotic and thus more easily represented by only a few numbers.
-However, the performance data looks fairly similar between the two data sets, with a couple of events that seem to stand out:
+However, the performance data looks fairly similar between the two data sets, with a couple of events that appear to stand out:
 \begin{itemize}

### 2014-05-27 paper.tex
Minor abstract change

@@
 Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
-Recent research has shown that there can be significant performance differences based on seemingly irrelevant factors like user name length.
+Recent research has shown that there can be significant performance differences based on seemingly irrelevant factors such as user name length.
 In this paper we identify an important underlying mechanism that cause this type of measurement bias on modern Intel microarchitectures.
@@
 %Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics and compiler rules.
-In the longer term, our results can enable novel compiler and architectural optimization strategies.
+In the longer term, our results can enable new architecture specific optimization strategies.
 

### 2014-05-29 paper.tex
Update analysis of env alias

@@
 The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
-On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270,000 to about 135,000.
+On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from around 270,000 to 135,000.
 The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
 Fewer stalls in the aliasing case could indicate less contention on the reservation station.
-This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports seem to have fewer operations going in parallel.
-Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stays in the execution pipeline for a shorter time.
-Our assumption is that this is a by-product of stalls generated from aliasing, causing micro-operations to be reissued.
+This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports have fewer operations going in parallel.
+Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stay in the execution pipeline for a shorter time.
+Our assumption is that this is caused by the CPU being more conservative in the aliasing case, issuing fewer operations (speculatively) in parallel.
+With fewer instructions per cycle, there is also less pressure on the reservation station.
 

### 2014-05-29 paper.tex
Fix parentesis error

@@
 #define ALIAS(a, b) \
-    (((long)&a) & 0xfff == ((long)&b) & 0xfff)
+    ((((long)&a) & 0xfff) == (((long)&b) & 0xfff))
 static int i, j, k;

### 2014-05-29 paper.tex
Minor changes

@@
 Changing environment variables will therefore offset the addresses of stack, and consequently all stack allocated variables.
-After some variable amount of offset, the stack is normally aligned to a 16 byte boundary, which will be enforced by the compiler.
-Within a span of 4096 bytes there are thus 256 possible initial stack addresses, each representing a different execution context with respect to address aliasing.\footnote{Note that there is no clear relationship between environment size and stack location with ASLR enabled.
-However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.}
+After this offset, the stack is normally aligned to a 16 byte boundary, enforced by the compiler.
+Within a span of 4096 bytes there are thus 256 possible initial stack addresses, each representing a different execution context with respect to address aliasing.
+Note that there is no clear relationship between environment size and stack location with ASLR enabled.
+However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
 
@@
 To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, reproduced below.
-This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straightforward to analyze.
+This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.
 Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
@@
 We sampled an extensive set of performance counters in addition to cycle count for each execution context.
-By looking at the \emph{median} value of each counter compared to the extreme cases, we narrowed down the performance events that behaved similarly to cycle count.
-Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases, compared to the median over all environments.
+To analyze the bias effects at these points, we calculate the \emph{median} value of each counter over all contexts, and compare that to the two extreme cases.
+Going through the list of events manually, we narrowed down to a set that seem to correlate well with cycle count.
+Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases, compared to the median values.
 
@@
   \begin{tabnote}
-    \Note{Note:}{Performance events that are obviously not indicative of any causal relationship are omitted, for example bus-cycles which will naturally vary with total cycle count.}
+    \Note{Note:}{Some correlating performance events are omitted, for example bus-cycles which will naturally vary with total cycle count.}
   \end{tabnote}
@@
 The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
-Virtual memory layout of static data is decided at compile time, and we can find the location of these variables by looking at the symbol table in the ELF executable.
+Memory layout of static data is decided at compile time, and we can find the location of these variables by looking at the symbol table in the ELF executable.
 We find the addresses to be \texttt{\&i} = 0x60103c, \texttt{\&j} = 0x601040, and \texttt{\&k} = 0x601044.\footnote{ELF symbol tables can be read using \texttt{readelf -s}}
@@
 Observing addresses of stack allocated data at runtime is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
-A small amount of assembly code was added to calculate the addresses of \texttt{g} and \texttt{inc}, outputting to stdout directly using the \texttt{syscall} instruction.
-Making no change to the stack allocation instructions and not affecting the addresses of static variables, the instrumented program have the exact same bias to environment size, free from observer effects.
-We find that the spike in cycle count occurs precisely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe{\underline{038} and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
-The assembly output of {\small GCC} for the inner loop is shown below, indicating the relevant load and store instructions.
-There is only one store to variable \texttt{i}, which will alias with loads of \texttt{inc}.
+Assembly code was injected to calculate and output the addresses of \texttt{g} and \texttt{inc}.
+We made no change to the stack allocation instructions, and the code offset did not affect the addresses of static variables.
+The same experiment was run again, showing that the modified program had the exact same bias to environment size.
+From the output at each spike, we found the addresses to be \texttt{\&g} = 0x7fffffffe038, \texttt{\&inc} = 0x7fffffffe03c, and \texttt{\&g} = 0x7fffffffd038, \texttt{\&inc} = 0x7fffffffd03c, respectively
+Notice the common suffix 0x03c between \texttt{inc} and \texttt{i}, which is the aliasing pair.
+
+% The assembly output of {\small GCC} for the inner loop is shown below, indicating the relevant load and store instructions.
+% There is only one store to variable \texttt{i}, which will alias with loads of \texttt{inc}.
 
 % Refer to relevant section of the generated assembly file (in gas syntax)
-\lstinputlisting[
-  basicstyle=\ttfamily\small,
-  linerange={22-38}, 
-  numbers=left,
-  firstnumber=22
-  ]{bin/micro-kernel-annotated.s}
+% \lstinputlisting[
+%   basicstyle=\ttfamily\small,
+%   linerange={22-38}, 
+%   numbers=left,
+%   firstnumber=22
+%   ]{bin/micro-kernel-annotated.s}
 
@@
 Static variables are fixed and covers 12 contiguous bytes (3 words), in our case the addresses end in 0x0, 0x4 and 0xc, leaving the 0x8 slot free.
-Automatic variables occupies 8 contiguous bytes on stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
+The two automatic variables \texttt{g} and \texttt{inc} occupies 8 contiguous bytes on stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
 In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.
+
 A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots.
-While this will give significantly more alias counts, it has little effect on the total number of cycles executed.
+While this will give significantly more alias counts, we found that it has little effect on the total number of cycles executed.
 
@@
 
-While this particular solution is clearly not something one would want to do in practice, it at least shows that compilers and programmers \emph{could} take measures to account for aliasing.
-This becomes much more enticing when the performance penalties are more significant, which we find is often the case for aliasing heap allocations.
+While this particular solution is clearly not something one would want to do in practice, it shows that compilers and programmers \emph{could} take measures to account for aliasing.
 
@@
   \item The \emph{heap} is marked by a break point representing the end of uninitialized data in virtual memory, and more space is requested by the \texttt{brk} or \texttt{sbrk} system calls.
-  %Allocators typically use this area for smaller allocations.
   \item The \texttt{mmap} system call is used to map file descriptors to virtual memory. \emph{Anonymous} mappings, i.e. buffers not backed by a file, can be used for general purpose allocations.
-  %This mechanism is typically used for large allocations.
 \end{itemize}
@@
 Addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, while pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
-This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
+This distinction is unimportant for application developers, as everything is conceptually the same ``heap''.
 However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
 The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address end with 0x010.}
-This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following section.
+This behavior is often the worst case for functions that operate on two or more independent buffers.
 
@@
 Conversely, tcmalloc seem manage only the heap.
-There is one example where an allocator yields aliasing buffers while another does not.
+We see one example where an allocator yields aliasing buffers while another does not.
 Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.

### 2014-05-29 paper.tex
Rephrasing in introduction

@@
 %Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics.
-Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
-Recent research has shown that there can be significant performance differences based on seemingly irrelevant factors such as user name length.
+
+%Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
+%Understanding and accounting for external measurement \emph{bias} is important in all science.
+Understanding program performance and obtaining accurate measurements is important in all scientific work.
+Any source of \emph{bias} by external factors should therefore be known and accounted for.
+However, recent research has shown that there can be significant performance differences based on seemingly irrelevant factors such as user name length.
 In this paper we identify an important underlying mechanism that cause this type of measurement bias on modern Intel microarchitectures.
@@
 Accurately measuring the performance of computer programs is important in order to evaluate new algorithms or compare different implementations.
-This process is complicated by \emph{measurement bias}, which can be a challenge for performance analysist and systems researchers.
+%This process is complicated by \emph{measurement bias}, which can be a challenge for performance analysist and systems researchers.
+A key challenge for performance analysist and systems researchers is handling \emph{measurement bias}.
 Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
-The reason this happens is because certain variables has the potential to offset memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
-Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to things like system environment variables ultimately affect program execution.
+% A likely reason this happens is because certain variables has the potential to offset memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
+% Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes interact.
+%to things like system environment variables ultimately affect program execution.
+The underlying explanation is likely that this changes memory addresses of code and/or data, which in turn interacts with various low level hardware buffers at runtime.
+Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.
 
+A literature study of 88 papers showed that the median speedup was within the margin of bias, possibly invalidating the claimed speedups~\cite{Mytkowicz:2008:OE&MB}. 
 To alleviate the impact of measurement bias in performance analysis, researchers have proposed to use causal analysis techniques and randomization of execution contexts~\cite{Mytkowicz:2008:OE&MB}.
-Others treat the existence bias as a potential optimization problem, using search over variant spaces to find optimal environments~\cite{Knights:2009:BlindOpt}.
-Our approach is to analyze isolated programs in detail using \emph{performance counters}, in an attempt to identify which underlying hardware mechanisms are causing bias.
+Others treat the interaction complexity as a potential optimization problem, using search over variant spaces to find optimal environments~\cite{Knights:2009:BlindOpt}.
+Our approach is to analyze isolated programs in detail using \emph{hardware performance measurements}, in an attempt to identify which underlying mechanisms are causing bias.
 
@@
 We show how techniques like manual padding of allocations and dynamic detection of aliasing conditions can be used to improve performance. 
-Accounting for address aliasing can give significant speedups, and our findings should motivate to exploit this in architecture-specific optimization.
+Understanding and handling address aliasing can give significant speedups, and our findings shows potential for further architecture-specific optimizations.
 
@@
 Performance counters are supported in the Linux kernel, accessed via a tool called \emph{perf}.
-We use the perf-stat command in all experiments, which accepts raw event codes listed in the reference manual~\cite{Volume3B}.
+This tool instructs the kernel to enable the hardware Performance Monitoring Unit (PMU) for a different program, before it is executed.
+Using this technique avoids modifications to the executed program, unlike tools like PAPI [LARS cite: PAPI: A Portable Interface to Hardware Performance Counters].
+We use the perf-stat command in all experiments, which accepts raw PMU event codes listed in the reference manual~\cite{Volume3B}.
 
-A small Python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
-Only a small set of events are collected at a time, to ensure events are actually counted continuously and not sampled by multiplexing between a limited set of counter registers.
+Reliability of PMU measurements have been evaluated by Weaver and Sally in [Can hardware performance counters be trusted? 2008].
+The potential issues they find are either handled explicitly or are irrelevant to our setup.
+%They find intra-machine variations of up to 1.07\%,
+
+A small Python script is used to iterate over an exhaustive set of all known counters, which amounts to about 200 on our architecture.
+Only a small set of events are collected at a time, to ensure events are actually counted continuously.
+We do not use multiplexing between a limited set of counter registers, as it will introduce too many variations [Lars cite: Toward Accurate Performance Evaluation using Hardware Counters].
 Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
@@
 % Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
-Results are averaged over multiple runs to reduce potential random error, supplying the -r option to perf-stat.
+Results are also averaged over multiple runs to reduce potential random error, using built-in perf rerun functionality.
 
@@
 By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
-All experiments are done on a machine under minimal load, and \emph{frequency scaling} is disabled to keep the CPU's clock speed fixed.
+All experiments are done on a dedicated machine under minimal load, and \emph{frequency scaling} is disabled to keep the CPU's clock speed fixed.
 Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads.

### 2014-05-29 paper.tex
Abstract. Multiple changes to introduction.

@@
 Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
-% A likely reason this happens is because certain variables has the potential to offset memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
-% Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes interact.
+A likely reason this happens is because certain variables has the potential to offset memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
+Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes interact.
 %to things like system environment variables ultimately affect program execution.
-The underlying explanation is likely that this changes memory addresses of code and/or data, which in turn interacts with various low level hardware buffers at runtime.
-Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.
 

### 2014-05-29 paper.tex
Update figure with y axis legends

--- 2014-05-29 paper.tex
@@
 \pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
-\begin{figure*}[t]
+\begin{figure}[t]
   \centering
@@
         xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
+        ylabel=Event count,
         cycle list name=black white,
@@
   \label{fig:conv-default}
-  \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}.
+  \caption{Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}.
 Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
-Input size $n=2^{20}$.}
-\end{figure*}
+Showing optimization levels O2 and O3, input size $n=2^{20}$.}
+\end{figure}
 
@@
 The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
-This phenomenon is only observed for address offsets close to zero, thus we only show the first 20 data points.
+This phenomenon is only observed for address delta close to zero, thus we only show the first 20 data points.
 If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
@@
 More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we have selected a subset that correlates well with the change to cycle count.
-The numbers are for the level 2 optimization case, which was chosen because the behavior was less chaotic and thus more easily represented by only a few numbers.
+The numbers are for the level 2 optimization case, which we choose to study in more detail.
+ % which was chosen because the behavior was less chaotic and thus more easily represented by only a few numbers.
 However, the performance data looks fairly similar between the two data sets, with a couple of events that appear to stand out:

### 2014-05-29 paper.tex
Fix references in introduction

@@
 
-The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section, then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
-Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias.
-In Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
+The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section.
+Then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
+Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
 Related work is discussed in Section \ref{sec:related}.
@@
 This tool instructs the kernel to enable the hardware Performance Monitoring Unit (PMU) for a different program, before it is executed.
-Using this technique avoids modifications to the executed program, unlike tools like PAPI [LARS cite: PAPI: A Portable Interface to Hardware Performance Counters].
+Using this technique avoids modifications to the executed program, unlike tools like PAPI~\cite{PAPI:PortableInterface}.
 We use the perf-stat command in all experiments, which accepts raw PMU event codes listed in the reference manual~\cite{Volume3B}.
 
-Reliability of PMU measurements have been evaluated by Weaver and Sally in [Can hardware performance counters be trusted? 2008].
+Reliability of PMU measurements have been evaluated by \citeN{Weaver:2008:PCTrusted}.
 The potential issues they find are either handled explicitly or are irrelevant to our setup.
@@
 Only a small set of events are collected at a time, to ensure events are actually counted continuously.
-We do not use multiplexing between a limited set of counter registers, as it will introduce too many variations [Lars cite: Toward Accurate Performance Evaluation using Hardware Counters].
+We do not use multiplexing between a limited set of counter registers, as it will introduce too many variations~\cite{Wiplove:ImproveEstimation}.
+% Paper is rename of "Toward Accurate Performance Evaluation using Hardware Counters".
 Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
@@
 
-\section{4K Address Aliasing}
-\label{sec:aliasing}
-Modern processors are \emph{superscalar}, and achieves parallelism by issuing multiple instructions simultaneously and out of order.
-One of the issues that can limit throughput is dependencies between a load and previous stores.
-% According to Intel, typical software consists of about 38~\% memory access instructions, with about two thirds of them being loads.
-To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}. 
-More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
-Loads are therefore issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
-The prediction is later verified, replaying any instructions that was wrongly assumed to have no dependencies.
-Similarly, if the load and store locations are the same, the value can be \emph{forwarded} from the store before it retires.
-
-While optimizations such as these are good on the average case, there are corner cases. 
-In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
-A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
-Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
-The number of times this happens can be counted by the following performance counter:
-\begin{description}
-  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}] ``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.'' Intel Optimization Manual~\citeyear[B.3.4.4]{OptimizationManual}
-\end{description}
-
-As address aliasing depends on the memory addresses of loads and stores, environmental factors that affects memory has the potential to induce aliasing conditions.
-In the following sections, we show how performance penalties from address aliasing can be the root cause of measurement bias.
-
+% Virtual memory/randomization illustration
 \begin{narrowfig}{0.5\textwidth}[t]
@@
 \end{narrowfig}
+
+\section{4K Address Aliasing}
+\label{sec:aliasing}
+Modern processors are \emph{superscalar}, and achieves parallelism by issuing multiple instructions simultaneously and out of order.
+One of the issues that can limit throughput is dependencies between a load and previous stores.
+% According to Intel, typical software consists of about 38~\% memory access instructions, with about two thirds of them being loads.
+To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}. 
+More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
+Loads are therefore issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
+The prediction is later verified, replaying any instructions that was wrongly assumed to have no dependencies.
+Similarly, if the load and store locations are the same, the value can be \emph{forwarded} from the store before it retires.
+
+While optimizations such as these are good on the average case, there are corner cases. 
+In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
+A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
+Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
+The number of times this happens can be counted by the following performance counter:
+\begin{description}
+  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}] ``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.'' Intel Optimization Manual~\citeyear[B.3.4.4]{OptimizationManual}
+\end{description}
+
+As address aliasing depends on the memory addresses of loads and stores, environmental factors that affects memory has the potential to induce aliasing conditions.
+In the following sections, we show how performance penalties from address aliasing can be the root cause of measurement bias.
 

### 2014-05-29 paper.tex
Improve second half based on feedback

@@
 This means that addresses returned by allocators using this mechanism directly will \emph{always} alias, giving a deterministic execution context considering only the address suffix.
-From our limited experiment, this seems to be the case for glibc, jemalloc and Hoard.
+From our limited analysis, this seems to be the case for glibc, jemalloc and Hoard.
 
@@
 Many functions operate in a ``sliding window'' fashion; reading and writing to different buffers in some loop construction.
-This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
-As an example of this, consider a na{\"i}ve implementation of linear correlation shown in Figure \ref{lst:conv}.
-The function contains lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
+This type of access pattern is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
+As an example of this, consider a simple implementation of convolution shown in Figure \ref{lst:conv}.
 The performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.
@@
   ]{bin/convolution-kernel.c}
-  \caption{Naive implementation of convolution, ignoring endpoints for simplicity. This program is highly sensitive to aliasing between input and output arrays.}
+  \caption{Basic implementation of convolution with a fixed kernel, ignoring endpoints for simplicity. This program is highly sensitive to aliasing between input and output arrays.}
   \label{lst:conv}
@@
 By default, even with address randomization enabled, both \texttt{input} and \texttt{output} will have start addresses with the same address suffix of 0x010.
-To analyze performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
+To analyze performance for different addresses, we manually insert padding to offset the start address of one buffer.
 This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
-Controlling the offset parameter, we can create environments where the address suffixes are some number of \texttt{sizeof(float)} bytes apart.
+Controlling the offset parameter, we can create environments where the address suffixes are any desired number of \texttt{sizeof(float)} bytes apart.
 Allocating and managing these buffers at startup takes a non-negligible amount of work.
 The overhead can be masked by repeatedly invoking the convolution kernel after allocating and initializing all the inputs.
+Repeated invocation can also even out performance by warming up cache.
 \begin{lstlisting}
@@
 \end{lstlisting}
-An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
+An estimate of the actual cost a single invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
 \[
@@
 Our results are with $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
-For every run, performance counter measurements are also averaged over 10 samples, using the repeat mechanism of perf.
+In addition, performance counter measurements are averaged over 10 samples using the repeat mechanism of perf.
 
@@
     \begin{axis}[
-        title=\texttt{cc -O2},
+        title=Optimization -O2,
         font=\footnotesize,
@@
     \begin{axis}[
-        title=\texttt{cc -O3},
+        title=Optimization -O3,
         font=\footnotesize,
@@
 Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
-Showing optimization levels O2 and O3, input size $n=2^{20}$.}
+Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
 \end{figure}
@@
 The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
-This phenomenon is only observed for address delta close to zero, thus we only show the first 20 data points.
+This phenomenon is only observed for address deltas close to zero, thus we only show the first 20 data points.
 If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
@@
 \subsection{Ways to Deal with Heap Address Aliasing}
-Measurement bias from address aliasing can be significant in real applications, with potentially large performance penalties.
+Performance penalties caused by address aliasing can be significant in real applications, creating bias towards certain memory layouts.
 However, with an understanding of the underlying mechanism that causes bias, the effects can to some extent be predicted and accounted for in software.
 The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
+We identify some mitigation or optimization techniques that can be used:
 
@@
 \end{lstlisting}
-With this updated function prototype, the number of alias events is reduced by about 10 million on optimization level O2 for the default alignment, with a corresponding improvement in cycle count.
-% File backing this in results
+With this updated function prototype, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% on optimization level O2.
+Alias events is reduced by about 10 million for the default alignment, with a corresponding improvement in cycle count. % File backing this in results
 
 \paragraph{Use a special purpose allocator}
-We found that heap allocators are prone to generate pairwise aliasing buffers, the worst case for functions such as the convolution example.
+The Intel optimization manual mentions this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
+However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
+We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for functions such as the convolution example.
+%his goes somewhat against conventional wisdom that more alignment is better.
+
 A potential solution could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
-This goes somewhat against conventional wisdom that more alignment is better.
-The Intel optimization manual actually mentions this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
-However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
+The allocator could also accept hints to which buffers are going to be accessed in parallel, for example as an optional argument to a custom version of \texttt{malloc}.
 
@@
 \begin{lstlisting}[breaklines=true]
-    mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
+    mmap(NULL, (n + d), PROT_READ|PROT_WRITE, MAP_PRIVATE|MAP_ANONYMOUS, -1, 0) + d;
 \end{lstlisting}
@@
 \label{sec:conclusions}
-In this paper we have shown how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
-The effect is caused by the way speculative and out-of-order memory operations are handled by the CPU, only considering the last 12 address bits to resolve conflicts between load and store operations.
+We have shown how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
+The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between load and store operations.
 
-In general, any change to virtual memory layout of data can potentially introduce bias effects from address aliasing.
+In general, any change to memory layout of data can potentially introduce bias effects from address aliasing.
 Analyzing an example with bias to environment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
 Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
-We show how aliasing can have significant performance impact in practice for algorithms operating on heap allocated buffers, analyzing and example with more than 50 \% performance variation between different memory layouts.
-Because typical heap allocators will return aliased pointers for large allocations, worst case performance is likely to be the default.
+We show how aliasing can have significant performance impact in practice for algorithms operating on heap allocated buffers, analyzing an example with more than $2x$ performance variation between different memory layouts.
+Heap memory is controlled by allocators, and we show that most libraries tend to favor page alignment for larger requests.
+This means that pointers to structures such as arrays and matrices are likely to alias by default, which can be worst case input for many algorithms.
 
 We show how techniques like padding of variables and alternative alias-free code paths can be used to avoid aliasing at runtime.
-For heap address aliasing especially, we find that manual intervention is sometimes required in order to achieve optimal performance.
+For heap address aliasing especially, we find that manual intervention can be required in order to achieve optimal performance.
 Our results should inspire more clever optimizations and heuristics taking address aliasing effects into account, as the potential performance improvements can be substantial.

### 2014-05-30 paper.tex
Improved the discussion both for the microkernel and heap exsample.

@@
 However, recent research has shown that there can be significant performance differences based on seemingly irrelevant factors such as user name length.
-In this paper we identify an important underlying mechanism that cause this type of measurement bias on modern Intel microarchitectures.
+In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
 Our approach is to search for correlations between a large set of hardware performance counters and performance figures from multiple environment contexts.
@@
 \begin{bottomstuff}
+New Paper, Not an Extension of a Conference Paper.
+Authors' addresses: Lars Kirkholt Melhus, Rune Erlend Jensen, Computer and Information Science Department, Norwegian University of Science and Technology, 
+Sem S{\ae}lands vei 9, Gl{\o}shaugen, 7491 Trondheim, Norway; email: runeerle@idi.ntnu.no
+%LARS: din addresse er ok (NTNU)
 \end{bottomstuff}
@@
 \maketitle
+
+%\footnote{New Paper, Not an Extension of a Conference Paper.}
 
@@
 \subsection{Microkernel Analysis}
+\label{sec:microkernel}
 To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, reproduced below.
@@
 This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports have fewer operations going in parallel.
-Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stay in the execution pipeline for a shorter time.
-Our assumption is that this is caused by the CPU being more conservative in the aliasing case, issuing fewer operations (speculatively) in parallel.
-With fewer instructions per cycle, there is also less pressure on the reservation station.
+Note that the number of micro-ops \emph{retired} overall does not change.%, but for some reason the micro-ops stay in the execution pipeline for a shorter time.
+This match with the fact that the memory disambiguation system do not handle address aliasing, limiting the potential for out of order execution~\cite[Page 2-20]{OptimizationManual}.
+We assume that the tiny kernel size immediately prevents instructions from executing out of order, leading to the lower counts.
+%Our assumption is that this is caused by the CPU being more conservative in the aliasing case, issuing fewer operations (speculatively) in parallel.
+%With fewer instructions per cycle, there is also less pressure on the reservation station.
+
+The \texttt{cycles_ldm_pending} and \texttt{stalls_l2_pending} are interesting as they are related to the memory and cache system.
+However, this program use only 2 cache-lines except for the initial startup.
+Using the L1 hit and miss counts also revealed that the L1 load is 99.3\% and store hit rate is ~100\%.
+Thus, these events do not indicate a cache hit rate issue, and is probably linked to how the CPU handles address alias.
 
@@
 The overhead can be masked by repeatedly invoking the convolution kernel after allocating and initializing all the inputs.
-Repeated invocation can also even out performance by warming up cache.
+Repeated invocation will also even out performance by warming up the cache.
 \begin{lstlisting}
@@
 As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branch micro-ops are being re-issued.
+Unlike the microkernel in \ref{sec:microkernel} the execution and RS stall counts drop with together with aliasing events.
+Speculation is possible in this kernel as there are many independent store addresses.
+False aliasing limits this speculation however, and makes the CPU incorrectly discard executed instructions.
+This leads to the inverted behavior compared to the microkernel.
 
 It is worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
+Again the \texttt{cycles_ldm_pending} and \texttt{stalls_l2_pending} behavior is the same as in the microkernel in \ref{sec:microkernel}.
 The L1 hit rate remains stable across all offsets, and only a negligible number of memory loads actually misses L1.

### 2014-05-30 paper.tex
Lots of fixes all over.

@@
 By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
-All experiments are done on a dedicated machine under minimal load, and \emph{frequency scaling} is disabled to keep the CPU's clock speed fixed.
+All experiments are done on a dedicated machine under minimal load, with \emph{frequency scaling} disabled to keep the CPU's clock speed fixed.
 Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads.
@@
 While optimizations such as these are good on the average case, there are corner cases. 
-In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
+In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
 A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
@@
 Changing environment variables will therefore offset the addresses of stack, and consequently all stack allocated variables.
-After this offset, the stack is normally aligned to a 16 byte boundary, enforced by the compiler.
+After this offset, the stack is normally realigned to a 16 byte boundary, enforced by the compiler.
 Within a span of 4096 bytes there are thus 256 possible initial stack addresses, each representing a different execution context with respect to address aliasing. 
@@
 To analyze the bias effects at these points, we calculate the \emph{median} value of each counter over all contexts, and compare that to the two extreme cases.
-Going through the list of events manually, we narrowed down to a set that seem to correlate well with cycle count.
-Table \ref{tab:loopcorrelation} shows numerical counter values for worst cases, compared to the median values.
+Going through the list of events manually, we selected the set that had the highest correlation with cycle count.
+Table \ref{tab:loopcorrelation} shows numerical counter values for these cases, compared to the median values.
 
@@
 
-We see the most extreme change from median to worst case is clearly the number of alias events. 
+We see the most extreme change from median to worst case in the number of alias events. 
 If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
@@
 A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots.
-While this will give significantly more alias counts, we found that it has little effect on the total number of cycles executed.
+While this will give significantly more alias counts, we found that it had little effect on the total number of cycles executed.
 
@@
 
+% TODO: Lars: make this a separate section or move it up a bit?
+This type of bias can occur in a number of memory configurations.
+Here the aliasing was caused by interactions between the stack and static variables.
+The stack and heap can interact the same way, depending on how the malloc function behaves.
+Since the stack is used for local variables complex code is likely contain many potential conflicts.
+A evaluation of how malloc behaves is presented in \ref{sec:heap}.
+
 \subsection{Avoiding Aliasing}
@@
 Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
-Shown in Figure \ref{lst:loopfixed} is a proof of concept of how alias-free code could be generated in this particular case.
+A proof of concept of how alias-free code could be generated in this particular case is shown in Figure \ref{lst:loopfixed}.
 If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
@@
 
-While this particular solution is clearly not something one would want to do in practice, it shows that compilers and programmers \emph{could} take measures to account for aliasing.
+While this particular solution is impractical, it shows that compilers and programmers \emph{could} take measures to account for aliasing.
 
@@
 Heap allocation routines such as \texttt{malloc} and \texttt{free} are typically dynamically linked, for example as part of glibc.
+If this assignment is independent of the initial stack offset all heap memory accesses can be biased.
 The particular library used therefore constitutes an important part of the execution context, as linking to a different library, or a library with some alternative configuration, can impact heap addresses at runtime.
@@
 Conversely, tcmalloc seem manage only the heap.
-We see one example where an allocator yields aliasing buffers while another does not.
+We also find an example where one allocator yields aliasing buffers while another does not.
 Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
 Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.
-But even with randomization, pointers returned by \texttt{mmap} must still be page aligned.
+But even with ASLR, pointers returned by \texttt{mmap} must still be page aligned.
 This means that addresses returned by allocators using this mechanism directly will \emph{always} alias, giving a deterministic execution context considering only the address suffix.
@@
 \end{lstlisting}
-An estimate of the actual cost a single invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
+An estimate of the actual cost a single invocation can be calculated by averaging over a number of repeated function calls, after subtracting the overhead of invoking it the first time.
 \[
@@
 
-Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
+Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between the heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
 The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
@@
 The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
-This phenomenon is only observed for address deltas close to zero, thus we only show the first 20 data points.
+This phenomenon is only observed for address separations close to zero, thus we only show the first 20 data points.
 If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
 
-More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we have selected a subset that correlates well with the change to cycle count.
+More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we show the subset that correlates best with the cycle count.
 The numbers are for the level 2 optimization case, which we choose to study in more detail.
@@
 Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
-We do not attempt to pinpoint exactly which assembly instructions might alias, but on a high level we can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
+We do not attempt to pinpoint exactly which assembly instructions might alias, but we assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
 By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
@@
 Performance penalties caused by address aliasing can be significant in real applications, creating bias towards certain memory layouts.
-However, with an understanding of the underlying mechanism that causes bias, the effects can to some extent be predicted and accounted for in software.
+However, with an understanding of the underlying mechanism that causes bias, the effects can be predicted and to some extent eliminated in software.
 The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
@@
 However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
+Similarly none offset their allocations based on the initial stack frame.
 We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for functions such as the convolution example.
@@
 We have shown how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
-The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between load and store operations.
+The effect is caused by how speculative and out-of-order memory operations are handled by the Intel ``Haswell'' CPU, which only considers the last 12 address bits to resolve conflicts between load and store operations.
 
@@
 Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
-We show how aliasing can have significant performance impact in practice for algorithms operating on heap allocated buffers, analyzing an example with more than $2x$ performance variation between different memory layouts.
+We show how aliasing can have significant performance impact for algorithms operating on heap allocated buffers, analyzing an example with more than $2x$ performance variation between different memory layouts.
 Heap memory is controlled by allocators, and we show that most libraries tend to favor page alignment for larger requests.
-This means that pointers to structures such as arrays and matrices are likely to alias by default, which can be worst case input for many algorithms.
+This means structures such as vectors and matrices are likely to alias by default, which can be worst case scenarios for many algorithms.
 

### 2014-05-30 paper.tex
Add new reference and improvements

@@
 Authors' addresses: Lars Kirkholt Melhus, Rune Erlend Jensen, Computer and Information Science Department, Norwegian University of Science and Technology, 
-Sem S{\ae}lands vei 9, Gl{\o}shaugen, 7491 Trondheim, Norway; email: runeerle@idi.ntnu.no
-%LARS: din addresse er ok (NTNU)
+Sem S{\ae}lands vei 9, Gl{\o}shaugen, 7491 Trondheim, Norway; email: larskirk@alumni.ntnu.no, runeerle@idi.ntnu.no
 \end{bottomstuff}
@@
 \maketitle
-
-%\footnote{New Paper, Not an Extension of a Conference Paper.}
 
@@
 
-The \texttt{cycles_ldm_pending} and \texttt{stalls_l2_pending} are interesting as they are related to the memory and cache system.
-However, this program use only 2 cache-lines except for the initial startup.
+The cycles\_ldm\_pending and stalls\_l2\_pending are interesting, as they are related to the memory and cache system.
+However, this program use only 2 cache-lines, except for the initial startup.
 Using the L1 hit and miss counts also revealed that the L1 load is 99.3\% and store hit rate is ~100\%.
@@
 In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.
-
 A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots.
@@
 The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.
-
-% TODO: Lars: make this a separate section or move it up a bit?
 This type of bias can occur in a number of memory configurations.
-Here the aliasing was caused by interactions between the stack and static variables.
-The stack and heap can interact the same way, depending on how the malloc function behaves.
-Since the stack is used for local variables complex code is likely contain many potential conflicts.
-A evaluation of how malloc behaves is presented in \ref{sec:heap}.
+Here, the aliasing was caused by interactions between the stack and static variables, but we can imagine similar conflicts between stack and heap, depending also on how dynamic memory allocators behave.
+Because the stack is used for local variables, complex code is likely to contain many potential conflicts with other areas of memory.
 
@@
 The numbers are for the level 2 optimization case, which we choose to study in more detail.
- % which was chosen because the behavior was less chaotic and thus more easily represented by only a few numbers.
 However, the performance data looks fairly similar between the two data sets, with a couple of events that appear to stand out:
@@
 Interestingly, it looks like small address offsets incur a massive increase in operations issued on port 0 for the O2 case.
-On Haswell, this port handles various ALU operations, and branching together with port 6.~\cite[Figure 2.1]{OptimizationManual}
-As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branch micro-ops are being re-issued.
-Unlike the microkernel in \ref{sec:microkernel} the execution and RS stall counts drop with together with aliasing events.
-Speculation is possible in this kernel as there are many independent store addresses.
-False aliasing limits this speculation however, and makes the CPU incorrectly discard executed instructions.
-This leads to the inverted behavior compared to the microkernel.
+On Haswell, this port handles various ALU operations, and branching together with port 6~\cite[Figure 2.1]{OptimizationManual}.
+As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branches are being re-issued.
+Unlike the microkernel in section \ref{sec:microkernel}, the micro-ops executed and RS stall counts drop together with aliasing events.
+Speculation is possible in this kernel, as there are many independent store addresses.
+False dependencies will limit the amount of speculation, making the CPU incorrectly discard executed instructions.
+This can explain the inverted behavior compared to the microkernel, where the store addresses were fixed between loop iterations.
 
-It is worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
-Again the \texttt{cycles_ldm_pending} and \texttt{stalls_l2_pending} behavior is the same as in the microkernel in \ref{sec:microkernel}.
-The L1 hit rate remains stable across all offsets, and only a negligible number of memory loads actually misses L1.
+It is worth noticing that most cache related metrics do \emph{not} stand out in this experiment.
+For cycles\_ldm\_pending and stalls\_l2\_pending, we see similar behavior to the microkernel analysis in section \ref{sec:microkernel}.
+These events can be explained by pipeline stalls waiting for aliasing store operations to be retired.
+The L1 hit rate remains stable across all offsets, and only a negligible number of memory accesses actually miss L1.
 On the other hand, we see a fairly strong correlation between cycle count and outstanding offcore requests. 
 These events count the number of outstanding loads to memory outside the processor core each cycle.
-Since we do not see any significant number of loads missing L1, the correlation to outstanding loads is probably a result of stalling and more cycles executed.
+Since we do not see any significant number of L1 misses, the correlation to outstanding loads is probably a result of stalling and more cycles executed.
 
-Overall it seems reasonable to conclude that the resource stalling is causing delayed execution, ultimately generated by false dependencies from address aliasing.
+Overall, it seems reasonable to conclude that the resource stalling is causing the slowdown, ultimately generated by false dependencies from address aliasing.
 Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
-We do not attempt to pinpoint exactly which assembly instructions might alias, but we assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
+We do not attempt to pinpoint exactly which assembly instructions might alias, but one can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
 By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
@@
 
-Techniques for ``Blind'' optimization using search over variant spaces, is introduced by \citeN{Knights:2009:BlindOpt}.
-Their space of variants consist of different position and alignments for each function and global variable, in principle exploring the different configurations possible by altering link order.
-The authors of MAO present a framework for extending compilers to provide very low level microarchitectural assembly optimization~\cite{Hundt:2011:MAO}.
-Different techniques are used, including rules, pattern matching, and random insertion of nop-instructions.
+Many different approaches have been explored in order to take performance counter data and processor intrinsics into account for optimization.
+\citeN{Knights:2009:BlindOpt} introduce ``blind'' optimization, treating the underlying hardware as a black box and use search over \emph{variant spaces}.
+Their space of variants consist of different position and alignment for each function and global variable, in principle exploring the different configurations possible by altering link order.
+Others have used machine learning to classify performance counter data, recognizing patterns or \emph{pathologies} in measurements to recognize optimization opportunities~\cite{Yoo:ADD}.
+The authors of MAO present a framework for extending compilers to provide very low level microarchitectural assembly optimization, using rules, pattern matching, and random insertion of nop-instructions to discover optimal assembly outputs~\cite{Hundt:2011:MAO}.
 

### 2014-05-30 paper.tex
Delete comments and improve various formulations

@@
 \usepackage{filecontents}
-%\usepackage{booktabs} % Only used for horizontal lines
 \usepackage{tikz}
 \usepackage{listings}
-%\usepackage{caption}
 \usepackage{mathtools}
@@
 \begin{abstract} % Should be 150 -- 200 words max
-%Understanding program performance and obtaining accure measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
-%Predicting or accounting for measurment bias is challenging, as the interaction with low level hardware mechanisms is complex.  
-%In this paper we show how \emph{address aliasing} can be the underlying mechanism resulting in measurement bias on modern Intel microarchitectures.
-%Our approach is to use hardware performance counter to analyze performance data over controlled series of execution contexts.
-%We show how address aliasing can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.
-
-%The perfomance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
-%It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
-%While the aliasing mechanism itself is not new, we provide the connection to measurement bias and show how to use this knowledge not only to account for bias but also explicitly optimize for it.
-%Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics.
-
-%Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
-%Understanding and accounting for external measurement \emph{bias} is important in all science.
 Understanding program performance and obtaining accurate measurements is important in all scientific work.
@@
 The performance impact of unfavorable memory layout combinations can be significant, with as much as $2x$ speedup in one of our examples.
-%Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics and compiler rules.
 In the longer term, our results can enable new architecture specific optimization strategies.
-
 \end{abstract}
@@
 Accurately measuring the performance of computer programs is important in order to evaluate new algorithms or compare different implementations.
-%This process is complicated by \emph{measurement bias}, which can be a challenge for performance analysist and systems researchers.
 A key challenge for performance analysist and systems researchers is handling \emph{measurement bias}.
 Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
-% A likely reason this happens is because certain variables has the potential to offset memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
-% Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes interact.
-%to things like system environment variables ultimately affect program execution.
-The underlying explanation is likely that this changes memory addresses of code and/or data, which in turn interacts with various low level hardware buffers at runtime.
+The underlying explanation is likely that this changes memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
 Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.
@@
 
-Our results show that the cost of aliasing can be significant, exemplified by a simple program with more than $2x$ speedup based in heap address alignment alone.
+Our results show that the cost of aliasing can be significant, exemplified by a simple program with up to $2x$ speedup based in heap address alignment alone.
 We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing, by favoring page alignment for large allocations.
@@
   \end{tikzpicture}
-  \caption{Relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory.
+  \caption{Memory execution context, assuming a 64-bit process running on a Linux system.
+  %Relative positions of some important sections of memory at runtime
   Initial addresses of stack, heap and memory mapped files are often randomized for security reasons.
+  The stack is also offset by environment variables and program arguments.
   Addresses of code and statically allocated data are allocated at compile time by the linker, and can be determined by inspecting the executable.}
@@
 
-Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers \footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
+Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers.\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
 We observe the addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} for different size parameters.
@@
 \end{itemize}
+
 Interestingly, it looks like small address offsets incur a massive increase in operations issued on port 0 for the O2 case.
@@
 Speculation is possible in this kernel, as there are many independent store addresses.
-False dependencies will limit the amount of speculation, making the CPU incorrectly discard executed instructions.
+However, false dependencies will limit the amount of speculation, making the CPU incorrectly discard executed instructions.
 This can explain the inverted behavior compared to the microkernel, where the store addresses were fixed between loop iterations.
 
-It is worth noticing that most cache related metrics do \emph{not} stand out in this experiment.
+It is worth noting that most cache related metrics do \emph{not} stand out in this experiment.
 For cycles\_ldm\_pending and stalls\_l2\_pending, we see similar behavior to the microkernel analysis in section \ref{sec:microkernel}.
 These events can be explained by pipeline stalls waiting for aliasing store operations to be retired.
-The L1 hit rate remains stable across all offsets, and only a negligible number of memory accesses actually miss L1.
+The L1 hit rate also remains stable across all offsets, and only a negligible number of memory accesses actually miss L1.
 On the other hand, we see a fairly strong correlation between cycle count and outstanding offcore requests. 
@@
 Overall, it seems reasonable to conclude that the resource stalling is causing the slowdown, ultimately generated by false dependencies from address aliasing.
-Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
-We do not attempt to pinpoint exactly which assembly instructions might alias, but one can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
+We do not attempt to pinpoint exactly which accesses generates these conflicts, but at a high level the CPU falsely assumes dependencies between \texttt{input[i]} and \texttt{output[i]}.
 By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
 This is a speedup on top of already aggressive compiler optimization.
+% Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
+% We do not attempt to pinpoint exactly which assembly instructions might alias, but one can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
 
@@
 However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
-Similarly none offset their allocations based on the initial stack frame.
+%Similarly none offset their allocations based on the initial stack frame.
 We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for functions such as the convolution example.
-%his goes somewhat against conventional wisdom that more alignment is better.
 
@@
 \begin{lstlisting}[breaklines=true]
-    mmap(NULL, (n + d), PROT_READ|PROT_WRITE, MAP_PRIVATE|MAP_ANONYMOUS, -1, 0) + d;
+    mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
 \end{lstlisting}
@@
 \label{sec:conclusions}
-We have shown how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
+We have shown how address aliasing can affect program performance under different memory contexts, and how it can explain certain cases of measurement bias. 
 The effect is caused by how speculative and out-of-order memory operations are handled by the Intel ``Haswell'' CPU, which only considers the last 12 address bits to resolve conflicts between load and store operations.
@@
 Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
-We show how aliasing can have significant performance impact for algorithms operating on heap allocated buffers, analyzing an example with more than $2x$ performance variation between different memory layouts.
-Heap memory is controlled by allocators, and we show that most libraries tend to favor page alignment for larger requests.
-This means structures such as vectors and matrices are likely to alias by default, which can be worst case scenarios for many algorithms.
+Dynamically allocated data is controlled by heap allocators, which we introduce as another source of bias.
+Comparing four different libraries, we show that most implementations tend to favor page alignment for larger requests.
+This means that structures such as vectors and matrices are likely to alias by default, which can be worst case scenarios for many algorithms.
+%Address aliasing can have significant performance impact for algorithms operating on heap allocated buffers, which 
+We analyze an example with up to $2x$ variation in cycle count between different heap alignments, showing that address aliasing can have a significant performance impact in practice.
 

### 2014-05-30 paper.tex
Fix spellcheck errors

@@
 Accurately measuring the performance of computer programs is important in order to evaluate new algorithms or compare different implementations.
-A key challenge for performance analysist and systems researchers is handling \emph{measurement bias}.
+A key challenge for performance analysts and systems researchers is handling \emph{measurement bias}.
 Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
@@
 Measurement bias and observer effect in performance analysis has been studied in detail by Mytkowicz et al~\shortcite{Mytkowicz:2008:OE&MB,Mytkowicz:2008:Easy}.
-The authors introduce environment variables and link ordering as examples of external bias triggers, showing how standardised {\small SPECint} benchmarks suffer from measurement bias when trying to evaluate the effectiveness of O3 over O2.
+The authors introduce environment variables and link ordering as examples of external bias triggers, showing how standardized {\small SPECint} benchmarks suffer from measurement bias when trying to evaluate the effectiveness of O3 over O2.
 Strategies for detecting bias through causal analysis and randomization of experimental setups is introduced.

### 2014-05-31 paper.tex
Update abstract

@@
 \begin{abstract} % Should be 150 -- 200 words max
-Understanding program performance and obtaining accurate measurements is important in all scientific work.
-Any source of \emph{bias} by external factors should therefore be known and accounted for.
-However, recent research has shown that there can be significant performance differences based on seemingly irrelevant factors such as user name length.
+% Understanding program performance and obtaining accurate measurements is important in all scientific work.
+% Any source of \emph{bias} by external factors should therefore be known and accounted for.
+% However, recent research has shown that there can be significant performance differences based on seemingly irrelevant factors such as user name length.
+Understanding program performance through accurate measurements is important in performance analysis, but experiments can be \emph{biased} by external factors.
+Seemingly irrelevant variables, such as the length of your user name, has been shown to potentially impact performance.
 In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
-Our approach is to search for correlations between a large set of hardware performance counters and performance figures from multiple environment contexts.
+Our approach is to search for correlations between a large set of hardware performance counters from a controlled series of execution contexts.
 We find that \emph{address aliasing} can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.
 
-This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
-We demonstrate this by implementing runtime detection and correction of address aliasing, eliminating the bias.
-Further, we show that the allocation logic of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
-The performance impact of unfavorable memory layout combinations can be significant, with as much as $2x$ speedup in one of our examples.
-In the longer term, our results can enable new architecture specific optimization strategies.
+This result not only makes accounting for, and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
+We demonstrate this by implementing runtime detection and correction of aliasing conditions.
+%Further, we show that the allocation logic of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
+For heap allocators, we show that common libraries tend to give worst case alignment by default, favoring page alignment for large allocations.
+The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in one of our examples.
+Our results show there is a need for new architecture specific optimization strategies to account for this phenomenon.
 \end{abstract}
@@
 \section{Introduction}
-Accurately measuring the performance of computer programs is important in order to evaluate new algorithms or compare different implementations.
+Accurately measuring the performance of computer programs is important in order to evaluate new algorithms, or compare different implementations.
 A key challenge for performance analysts and systems researchers is handling \emph{measurement bias}.
@@
 
-A literature study of 88 papers showed that the median speedup was within the margin of bias, possibly invalidating the claimed speedups~\cite{Mytkowicz:2008:OE&MB}. 
+A literature study of 88 papers by \citeN{Mytkowicz:2008:OE&MB} showed that the median reported speedup was within the margin of bias, possibly invalidating the claimed speedups. 
 To alleviate the impact of measurement bias in performance analysis, researchers have proposed to use causal analysis techniques and randomization of execution contexts~\cite{Mytkowicz:2008:OE&MB}.

### 2014-05-31 paper.tex
Update abstract

@@
 \begin{abstract} % Should be 150 -- 200 words max
-% Understanding program performance and obtaining accurate measurements is important in all scientific work.
-% Any source of \emph{bias} by external factors should therefore be known and accounted for.
-% However, recent research has shown that there can be significant performance differences based on seemingly irrelevant factors such as user name length.
-Understanding program performance through accurate measurements is important in performance analysis, but experiments can be \emph{biased} by external factors.
-Seemingly irrelevant variables, such as the length of your user name, has been shown to potentially impact performance.
+Understanding program behavior and obtaining accurate measurements is important in performance analysis.
+However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
+Seemingly irrelevant variables, such as the length of your user name, can impact program performance.
 In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
-Our approach is to search for correlations between a large set of hardware performance counters from a controlled series of execution contexts.
+Our approach is to search for correlations over multiple environment configurations, using a large set of hardware performance counter measurements.
 We find that \emph{address aliasing} can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.
 
-This result not only makes accounting for, and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
+This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
 We demonstrate this by implementing runtime detection and correction of aliasing conditions.
-%Further, we show that the allocation logic of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
-For heap allocators, we show that common libraries tend to give worst case alignment by default, favoring page alignment for large allocations.
+For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
 The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in one of our examples.
-Our results show there is a need for new architecture specific optimization strategies to account for this phenomenon.
+Our results can enable new architecture specific optimization strategies to account for this phenomenon.
 \end{abstract}

### 2014-05-31 paper.tex
Add context on environment variables

@@
 \label{sec:environment}
-Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the position and alignment of stack.
-As indicated in Figure \ref{fig:virtualmemory}, environment variables and program arguments are allocated in the stack section of virtual memory close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, before the first call frame.
+System environment variables contain things like the name of the logged in user, home directory, shell state, and various other settings.
+As a source of bias, it is not the environment variables themselves that are important, but rather the effect their total \emph{size} has on the position and alignment of stack.
+As indicated in Figure \ref{fig:virtualmemory}, environment variables and program arguments are allocated in the stack section of virtual memory, close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory} and before the first call frame.
 Changing environment variables will therefore offset the addresses of stack, and consequently all stack allocated variables.
@@
 Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between stack and static data.
-The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.
+The program is \emph{biased} towards environment sizes that avoids this specific stack alignment, which in principle could be triggered just by changing to a different user name.
 This type of bias can occur in a number of memory configurations.

### 2014-05-31 paper.tex
Remove unnecessary wording

@@
 Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between stack and static data.
-The program is \emph{biased} towards environment sizes that avoids this specific stack alignment, which in principle could be triggered just by changing to a different user name.
+The program is \emph{biased} towards environment sizes that avoids this specific stack alignment, which in principle could be triggered just by changing user name.
 This type of bias can occur in a number of memory configurations.

### 2014-08-07 util/lperf.py
Added the new/correct cmask\`s to the LDM events. Added a hack to rerun all samples that are taken when the cycle count is higher than the 75\`th percentile \(in the first complete run\).

@@ -125,7 +125,12 @@ haswell = [Counter(code if mnemonic == "" else mnemonic, name) for code, mnemoni
     ("r08a2:u", "", "RESOURCE_STALLS.SB"),
     ("r10a2:u", "", "RESOURCE_STALLS.ROB"),
     ("r01a3:u", "", "CYCLE_ACTIVITY.CYCLES_L2_PENDING"),
-    ("r02a3:u", "", "CYCLE_ACTIVITY.CYCLES_LDM_PENDING"),
+    ("r02a3:u", "", "CYCLE_ACTIVITY.CYCLES_LDM_PENDING_NO_CMASK"),
+    ("r20002a3:u", "", "CYCLE_ACTIVITY.CYCLES_LDM_PENDING_CMASK_2"),
+    ("r06a3:u", "", "CYCLE_ACTIVITY.STALLS_LDM_PENDING_NO_CMASK"),
+    ("r20006a3:u", "", "CYCLE_ACTIVITY.STALLS_LDM_PENDING_CMASK_2"),
+    ("r40006a3:u", "", "CYCLE_ACTIVITY.STALLS_LDM_PENDING_CMASK_4"),
+    ("r60006a3:u", "", "CYCLE_ACTIVITY.STALLS_LDM_PENDING_CMASK_6"),
     ("r05a3:u", "", "CYCLE_ACTIVITY.STALLS_L2_PENDING"),
     ("r08a3:u", "", "CYCLE_ACTIVITY.CYCLES_L1D_PENDING"),
     ("r01a8:u", "", "LSD.UOPS"),
@@ -246,27 +251,87 @@ def benchmark(args):
     counters = args.events
     data = { counter: {'count': [0] * args.iterations } for counter in args.events }
 
+    def perfctr(s):
+        counter = filter(lambda (ctr): ctr.mnemonic == s.lower(), haswell)
+        if not counter:
+            raise argparse.ArgumentTypeError("Unrecognized performance counter " + s)
+        return counter[0]
+    
     # Measure all counters under n different environments
     for x in range(args.iterations):
         argument = "" if not args.enumerate else str(x)
         env = {'X': '0' * (args.env_offset + x*args.env_increment)}
-
+		
+        # Store the data for each run separately
+        data_local = { counter: {'count': [0] * len(counters) } for counter in args.events }
+        cycles = perfctr('cycles:u')
+        command_cache = []
+        
         # Sample at most 4 events at a time because of register limitations
         for i in range(0, len(counters), 4):
[…]

### 2014-08-09 util/lperf.py
Revert "Added the new/correct cmask (...)"
This reverts commit 2c8c09c95e7b5207c6a8bd5f0ecfc9dcb9eb2535.

@@ -125,12 +125,7 @@ haswell = [Counter(code if mnemonic == "" else mnemonic, name) for code, mnemoni
     ("r08a2:u", "", "RESOURCE_STALLS.SB"),
     ("r10a2:u", "", "RESOURCE_STALLS.ROB"),
     ("r01a3:u", "", "CYCLE_ACTIVITY.CYCLES_L2_PENDING"),
-    ("r02a3:u", "", "CYCLE_ACTIVITY.CYCLES_LDM_PENDING_NO_CMASK"),
-    ("r20002a3:u", "", "CYCLE_ACTIVITY.CYCLES_LDM_PENDING_CMASK_2"),
-    ("r06a3:u", "", "CYCLE_ACTIVITY.STALLS_LDM_PENDING_NO_CMASK"),
-    ("r20006a3:u", "", "CYCLE_ACTIVITY.STALLS_LDM_PENDING_CMASK_2"),
-    ("r40006a3:u", "", "CYCLE_ACTIVITY.STALLS_LDM_PENDING_CMASK_4"),
-    ("r60006a3:u", "", "CYCLE_ACTIVITY.STALLS_LDM_PENDING_CMASK_6"),
+    ("r02a3:u", "", "CYCLE_ACTIVITY.CYCLES_LDM_PENDING"),
     ("r05a3:u", "", "CYCLE_ACTIVITY.STALLS_L2_PENDING"),
     ("r08a3:u", "", "CYCLE_ACTIVITY.CYCLES_L1D_PENDING"),
     ("r01a8:u", "", "LSD.UOPS"),
@@ -251,87 +246,27 @@ def benchmark(args):
     counters = args.events
     data = { counter: {'count': [0] * args.iterations } for counter in args.events }
 
-    def perfctr(s):
-        counter = filter(lambda (ctr): ctr.mnemonic == s.lower(), haswell)
-        if not counter:
-            raise argparse.ArgumentTypeError("Unrecognized performance counter " + s)
-        return counter[0]
-    
     # Measure all counters under n different environments
     for x in range(args.iterations):
         argument = "" if not args.enumerate else str(x)
         env = {'X': '0' * (args.env_offset + x*args.env_increment)}
-		
-        # Store the data for each run separately
-        data_local = { counter: {'count': [0] * len(counters) } for counter in args.events }
-        cycles = perfctr('cycles:u')
-        command_cache = []
-        
+
         # Sample at most 4 events at a time because of register limitations
         for i in range(0, len(counters), 4):
[…]

### 2014-08-09 util/lperf-experimental.py
Separate experimental script
Want to keep lperf stable if possible

@@ -1,0 +1,350 @@
+#!/usr/bin/env python 
+import subprocess, argparse, sys, numpy, datetime
+from collections import namedtuple
+from scipy import stats
+
+# Used for experimental purposes, not producing data in the paper.
+
+
+Counter = namedtuple('Counter', ['mnemonic', 'name'])
+
+# See performance-counters.ods for full reference. Only support user-mode flag
+# selected, could be made more generic.
+haswell = [Counter(code if mnemonic == "" else mnemonic, name) for code, mnemonic, name in [
+    ("r0203:u", "", "LD_BLOCKS.STORE_FORWARD"),
+    ("r0803:u", "", "LD_BLOCKS.NO_SR"),
+    ("r0105:u", "", "MISALIGN_MEM_REF.LOADS"),
+    ("r0205:u", "", "MISALIGN_MEM_REF.STORES"),
+    ("r0107:u", "", "LD_BLOCKS_PARTIAL.ADDRESS_ALIAS"),
+    ("r0108:u", "", "DTLB_LOAD_MISSES.MISS_CAUSES_A_WALK"),
+    ("r0208:u", "", "DTLB_LOAD_MISSES.WALK_COMPLETED_4K"),
+    ("r0408:u", "", "DTLB_LOAD_MISSES.WALK_COMPLETED_2M_4M"),
+    ("r0e08:u", "", "DTLB_LOAD_MISSES.WALK_COMPLETED"),
+    ("r1008:u", "", "DTLB_LOAD_MISSES.WALK_DURATION"),
+    ("r2008:u", "", "DTLB_LOAD_MISSES.STLB_HIT_4K"),
+    ("r4008:u", "", "DTLB_LOAD_MISSES.STLB_HIT_2M"),
+    ("r6008:u", "", "DTLB_LOAD_MISSES.STLB_HIT"),
+    ("r8008:u", "", "DTLB_LOAD_MISSES.PDE_CACHE_MISS"),
+    ("r030d:u", "", "INT_MISC.RECOVERY_CYCLES"),
+    ("r010e:u", "", "UOPS_ISSUED.ANY"),
[…]

### 2014-08-09 util/lperf.py
Apply CMASK to cache counters
As described in manual, also adding L1D pending which was missing
from the list.

@@ -124,10 +124,11 @@ haswell = [Counter(code if mnemonic == "" else mnemonic, name) for code, mnemoni
     ("r04a2:u", "", "RESOURCE_STALLS.RS"),
     ("r08a2:u", "", "RESOURCE_STALLS.SB"),
     ("r10a2:u", "", "RESOURCE_STALLS.ROB"),
-    ("r01a3:u", "", "CYCLE_ACTIVITY.CYCLES_L2_PENDING"),
-    ("r02a3:u", "", "CYCLE_ACTIVITY.CYCLES_LDM_PENDING"),
-    ("r05a3:u", "", "CYCLE_ACTIVITY.STALLS_L2_PENDING"),
-    ("r08a3:u", "", "CYCLE_ACTIVITY.CYCLES_L1D_PENDING"),
+    ("r020001a3:u", "", "CYCLE_ACTIVITY.CYCLES_L2_PENDING"),
+    ("r020002a3:u", "", "CYCLE_ACTIVITY.CYCLES_LDM_PENDING"),
+    ("r05a3:u",     "", "CYCLE_ACTIVITY.STALLS_L2_PENDING"),
+    ("r080008a3:u", "", "CYCLE_ACTIVITY.CYCLES_L1D_PENDING"),
+    ("r0c000ca3:u", "", "CYCLE_ACTIVITY.STALLS_L1D_PENDING"),
     ("r01a8:u", "", "LSD.UOPS"),
     ("r01ae:u", "", "ITLB.ITLB_FLUSH"),
     ("r01b0:u", "", "OFFCORE_REQUESTS.DEMAND_DATA_RD"),

### 2014-08-10 util/lperf.py
Add Core2 events
Rewrite lperf to support multiple cpu architectures

@@ -1,238 +1,25 @@
 #!/usr/bin/env python 
-import subprocess, argparse, sys, numpy, datetime
-from collections import namedtuple
-from scipy import stats
-
-Counter = namedtuple('Counter', ['mnemonic', 'name'])
-
-# See performance-counters.ods for full reference. Only support user-mode flag
-# selected, could be made more generic.
-haswell = [Counter(code if mnemonic == "" else mnemonic, name) for code, mnemonic, name in [
-    ("r0203:u", "", "LD_BLOCKS.STORE_FORWARD"),
-    ("r0803:u", "", "LD_BLOCKS.NO_SR"),
-    ("r0105:u", "", "MISALIGN_MEM_REF.LOADS"),
-    ("r0205:u", "", "MISALIGN_MEM_REF.STORES"),
-    ("r0107:u", "", "LD_BLOCKS_PARTIAL.ADDRESS_ALIAS"),
-    ("r0108:u", "", "DTLB_LOAD_MISSES.MISS_CAUSES_A_WALK"),
-    ("r0208:u", "", "DTLB_LOAD_MISSES.WALK_COMPLETED_4K"),
-    ("r0408:u", "", "DTLB_LOAD_MISSES.WALK_COMPLETED_2M_4M"),
-    ("r0e08:u", "", "DTLB_LOAD_MISSES.WALK_COMPLETED"),
-    ("r1008:u", "", "DTLB_LOAD_MISSES.WALK_DURATION"),
-    ("r2008:u", "", "DTLB_LOAD_MISSES.STLB_HIT_4K"),
-    ("r4008:u", "", "DTLB_LOAD_MISSES.STLB_HIT_2M"),
-    ("r6008:u", "", "DTLB_LOAD_MISSES.STLB_HIT"),
-    ("r8008:u", "", "DTLB_LOAD_MISSES.PDE_CACHE_MISS"),
-    ("r030d:u", "", "INT_MISC.RECOVERY_CYCLES"),
-    ("r010e:u", "", "UOPS_ISSUED.ANY"),
-    ("r100e:u", "", "UOPS_ISSUED.FLAGS_MERGE"),
-    ("r200e:u", "", "UOPS_ISSUED.SLOW_LEA"),
[…]

### 2014-08-10 util/perfevents.py

@@ -1,0 +1,422 @@
+import subprocess
+
+class Event:
+    def __init__(self, event, umask, name, cmask = 0x00, alias = None):
+        self.event = event
+        self.umask = umask
+        self.cmask = cmask
+        self.alias = alias
+        self.name = name
+
+    def mnemonic(self):
+        return self.perfcode() if self.alias == None else "%s:u" % self.alias
+
+    def perfcode(self):
+        register = (self.cmask << 24) | (self.umask << 8) | self.event
+        return "r%x:u" % register
+
+# Try to determine the CPU we are running on
+model_name = subprocess.check_output("grep 'model name' /proc/cpuinfo", shell=True)
+
+haswell = []
+core2 = []
+
+events = \
+    core2 if "Intel(R) Core(TM)2" in model_name else \
+    haswell if "Intel(R) Core(TM) i7-4" in model_name else \
+    []
+
+# See performance-counters.ods for full reference.
+haswell += [
+    Event(0x03, 0x02, name = "LD_BLOCKS.STORE_FORWARD"),
+    Event(0x03, 0x08, name = "LD_BLOCKS.NO_SR"),
+    Event(0x05, 0x01, name = "MISALIGN_MEM_REF.LOADS"),
+    Event(0x05, 0x02, name = "MISALIGN_MEM_REF.STORES"),
+    Event(0x07, 0x01, name = "LD_BLOCKS_PARTIAL.ADDRESS_ALIAS"),
+    Event(0x08, 0x01, name = "DTLB_LOAD_MISSES.MISS_CAUSES_A_WALK"),
+    Event(0x08, 0x02, name = "DTLB_LOAD_MISSES.WALK_COMPLETED_4K"),
+    Event(0x08, 0x04, name = "DTLB_LOAD_MISSES.WALK_COMPLETED_2M_4M"),
[…]

### 2014-08-10 util/lperf.py
Handle event register limitation
Move grouping of batches sent to perf stat to event module, adding
flag to events for Core2 Duo that only work for register 0 or 1

@@ -1,6 +1,6 @@
 #!/usr/bin/env python 
 import subprocess, argparse, sys
-from perfevents import events
+from perfevents import events, sample_events
 
 def parseargs():
     def perfctr(s):
@@ -39,10 +39,8 @@ def benchmark(args):
         argument = "" if not args.enumerate else str(x)
         env = {'X': '0' * (args.env_offset + x*args.env_increment)}
 
-        # Sample at most 4 events at a time because of register limitations
-        for i in range(0, len(counters), 4):
-            current = counters[i:i + 4]
-
+        # Sample subset because of register limitations
+        for current in sample_events():
             prfevnt = ','.join(map(lambda (c): c.mnemonic(), current))
             command = ' '.join(["perf stat -r", str(args.repeat), '-x"," -e', prfevnt, args.program, argument])
             process = subprocess.Popen(command, stdout=subprocess.PIPE, stderr=subprocess.PIPE, env=env, shell=True)

### 2014-08-10 util/perfevents.py

@@ -1,30 +1,57 @@
 import subprocess
 
 class Event:
-    def __init__(self, event, umask, name, cmask = 0x00, alias = None):
+    def __init__(self, event, umask, name, cmask = 0x00, alias = None, msr = 0xf):
         self.event = event
         self.umask = umask
         self.cmask = cmask
         self.alias = alias
         self.name = name
+        self.msr = msr
 
     def mnemonic(self):
         return self.perfcode() if self.alias == None else "%s:u" % self.alias
 
     def perfcode(self):
-        register = (self.cmask << 24) | (self.umask << 8) | self.event
-        return "r%x:u" % register
+        code = (self.cmask << 24) | (self.umask << 8) | self.event
+        return ("r%04x:u" if self.cmask == 0x0 else "r%08x:u") % code
 
 # Try to determine the CPU we are running on
 model_name = subprocess.check_output("grep 'model name' /proc/cpuinfo", shell=True)
 
 haswell = []
 core2 = []
+registers = 0
 
-events = \
-    core2 if "Intel(R) Core(TM)2" in model_name else \
-    haswell if "Intel(R) Core(TM) i7-4" in model_name else \
-    []
+# configuration based on detected architecture
+if "Intel(R) Core(TM)2" in model_name:
+    events = core2
+    registers = 2
+elif "Intel(R) Core(TM) i7-4" in model_name:
+    events = haswell
+    registers = 4
+
+# group events appropriate for a single perf invocation
+def sample_events():
+    batch = []
+    for e in events:
+        if e.msr < 0xf:
+            if len(batch) == 0:
+                batch += [e]
+                yield batch
+            else: 
+                yield batch
+                batch = [e]
+                yield batch
+                batch = []
+        else:
[…]

### 2014-08-11 util/lperf.py
Fix sampling subset of events

@@ -1,10 +1,10 @@
 #!/usr/bin/env python 
 import subprocess, argparse, sys
-from perfevents import events, sample_events
+from perfevents import events, sample
 
 def parseargs():
     def perfctr(s):
-        counter = filter(lambda (ctr): ctr.perfcode() == s.lower() or ctr.mnemonic() == s.lower(), events)
+        counter = filter(lambda (ctr): (ctr.perfcode() == s.lower()) or (ctr.mnemonic() == s.lower()), events)
         if not counter:
             raise argparse.ArgumentTypeError("Unrecognized event " + s)
         return counter[0]
@@ -40,7 +40,7 @@ def benchmark(args):
         env = {'X': '0' * (args.env_offset + x*args.env_increment)}
 
         # Sample subset because of register limitations
-        for current in sample_events():
+        for current in sample(args.events):
             prfevnt = ','.join(map(lambda (c): c.mnemonic(), current))
             command = ' '.join(["perf stat -r", str(args.repeat), '-x"," -e', prfevnt, args.program, argument])
             process = subprocess.Popen(command, stdout=subprocess.PIPE, stderr=subprocess.PIPE, env=env, shell=True)

### 2014-08-11 util/perfevents.py

@@ -16,6 +16,12 @@ class Event:
         code = (self.cmask << 24) | (self.umask << 8) | self.event
         return ("r%04x:u" if self.cmask == 0x0 else "r%08x:u") % code
 
+    def __hash__(self):
+        return hash((self.event, self.umask, self.cmask))
+
+    def __eq__(self, other):
+        return (self.event, self.umask, self.cmask) == (other.event, other.umask, other.cmask)
+
 # Try to determine the CPU we are running on
 model_name = subprocess.check_output("grep 'model name' /proc/cpuinfo", shell=True)
 
@@ -31,10 +37,9 @@ elif "Intel(R) Core(TM) i7-4" in model_name:
     events = haswell
     registers = 4
 
-# group events appropriate for a single perf invocation
-def sample_events():
+def sample(event_list):
     batch = []
-    for e in events:
+    for e in event_list:
         if e.msr < 0xf:
             if len(batch) == 0:
                 batch += [e]

### 2014-08-12 analysis/blas/gemv-grid.c
Experiment with different blas packages
Different patterns for libblas, libatlas and libopenblas, but
not entirelly clear correlation with cycle count

@@ -4,16 +4,16 @@
 #include <sys/mman.h>
 
 // M rows x N cols
-#define M 8192
-#define N 8192
+#define M 1024
+#define N 1024
 
 #define ALIGNED_ALLOC(n) \
     ((char*) mmap(NULL, n, PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0))
 
 int main(int argc, char **argv) {
-    int dx = argc > 1 ? atoi(argv[1]) : 0;
-    int dy = argc > 2 ? atoi(argv[2]) : 0;
-    int n  = argc > 3 ? atoi(argv[3]) : 1;
+    int n  = argc > 1 ? atoi(argv[1]) : 1;
+    int dx = argc > 2 ? 16 * atoi(argv[2]) : 0;
+    int dy = argc > 3 ? 16 * atoi(argv[3]) : 0;
     int i;
 
     // Matrix-vector multiplication, y = alpha*Ax + beta*y

### 2014-08-12 paper.tex
Add missing punctuation
Minor changes
Add core2 results graphic in introduction
Might be a good idea

@@
 
+\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
+
+\begin{figure}[t]
+\begin{minipage}[t]{0.40\textwidth}
+  \vspace{3mm}
+  \begin{lstlisting}[frame=tb] %[frame=single]
+  static int i, j, k;
+  int main() {
+      int g = 0, inc = 1;
+      for (; g < 65536; g++) {
+          i += inc;
+          j += inc;
+          k += inc;
+      }
+      return 0;
+  }
+  \end{lstlisting}
+\end{minipage}
+%\hfill
+\begin{minipage}[t]{0.55\textwidth}
+  \begin{flushleft}
+    %\fbox{
+    \begin{tikzpicture}
+      \begin{axis}[
+          width = 7.8cm,
+          height = 4.5cm,
+          font = \footnotesize,
+          xlabel=Bytes added to environment,
+          ylabel=Cycles,
+          domain = 0:4096,
+          xtick = {0,1024,...,4096},
+          xmin = 0,
+          xmax = 4096,
+          cycle list name = exotic
+        ]
+        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \microkernelcoretwotable ;
+      \end{axis}
+    \end{tikzpicture}
+    %}
+  \end{flushleft}
+\end{minipage}
+\label{fig:motivation}
+\caption{Microkernel first presented in~\cite{Mytkowicz:2009:WrongData}, showing bias to environment size}
+\end{figure}
+
 A literature study of 88 papers by \citeN{Mytkowicz:2008:OE&MB} showed that the median reported speedup was within the margin of bias, possibly invalidating the claimed speedups. 
@@
 All experiments are done on a dedicated machine under minimal load, with \emph{frequency scaling} disabled to keep the CPU's clock speed fixed.
-Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads.
+Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads. Also some PMU events require HT to be turned off, or are inaccurate with HT.
 
@@
   \end{tikzpicture}
-  \caption{Bias from environment size for microkernel. Measured average of 10 cycle count samples for 512 different environments. Spikes show aliasing case, occurring once for each 4K period}
+  \caption{Bias from environment size for microkernel. Measured average of 10 cycle count samples for 512 different environments. Spikes show aliasing case, occurring once for each 4K period.}
   \label{fig:envbias}

### 2014-08-13 paper.tex
Add table describing test machines

@@
 Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.
+In Figure~\ref{fig:motivation} we have reproduced a result originally presented by \citeN{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program depends on environment variables.
+This example provides the basis for our analysis.
 
 \pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
-
 \begin{figure}[t]
@@
       \begin{axis}[
+          title = Core2 Duo,
           width = 7.8cm,
@@
 \label{fig:motivation}
-\caption{Microkernel first presented in~\cite{Mytkowicz:2009:WrongData}, showing bias to environment size}
+\caption{Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size}
 \end{figure}
@@
 
-Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus on the Intel ``Haswell'' microarchitecture specifically.
-Our setup consists of a 4th generation Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K processor, running 64-bit Ubuntu 14.04 LTS on kernel version 3.13.0-24-generic.
-Code samples are compiled using the {\small GCC} toolchain, version 4.8.2-19ubuntu1.
+Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture specifically.
+We later reproduced similar results on the previous ``Sandy Bridge'' CPU, as well as an older Core2 Duo machine.
+Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.
+
+\begin{table}
+  \tbl{Experimental setups used\label{tab:machines}}{
+  \begin{tabular}{|l|l|}
+    \hline
+Core2 Duo     & Intel\textsuperscript{\textregistered{}} Core\texttrademark{}2 Duo P8600 (laptop) \\
+              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
+Sandy Bridge  & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-3???? \\
+              & Ubuntu ?? \\ \hline
+Haswell       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K \\
+              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
+  \end{tabular}}
+\end{table}
 
@@
 Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
-The number of times this happens can be counted by the following performance counter:
+On recent architectures such as Haswell and Ivy Bridge, the number of times this happens can be counted by the following performance counter:
 \begin{description}
   \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}] ``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.'' Intel Optimization Manual~\citeyear[B.3.4.4]{OptimizationManual}
+\end{description}
+Older architectures such as Core2 does not have the same event available.
+However, the effect is still described in the manual for these processors, and covered by the following event:
+\begin{description}
+  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE}] ``(...) Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.'' ~\citeN[Table 19-17]{Volume3B}
 \end{description}

### 2014-08-14 paper.tex
Clean up description of alias counters

@@
 Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
-On recent architectures such as Haswell and Ivy Bridge, the number of times this happens can be counted by the following performance counter:
-\begin{description}
-  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}] ``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.'' Intel Optimization Manual~\citeyear[B.3.4.4]{OptimizationManual}
-\end{description}
-Older architectures such as Core2 does not have the same event available.
-However, the effect is still described in the manual for these processors, and covered by the following event:
-\begin{description}
-  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE}] ``(...) Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.'' ~\citeN[Table 19-17]{Volume3B}
-\end{description}
+The number of times this happens can be counted by the following performance counter:
+\begin{describe}{{paddin}}
+  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} Intel Optimization Manual~\citeyear[B.3.4.4]{OptimizationManual}
+\end{describe}
+This event is listed in the manual for microarchitectures going back to Nehalem i7, including Ivy Bridge and Haswell CPUs~\cite{Volume3B}.
+Older architectures such as Core 2 does not have this particular event listed, however 4K aliasing is covered by a more general event:
+\begin{describe}{{paddin}}
+  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''}~\citeN[Table 19-17]{Volume3B}
+\end{describe}
 

### 2014-08-14 paper.tex
Improve coverage of multiple architectures
Various edits and additions to describe Core in addition to Haswell.
Fix most syntax errors from review.

@@
   \vspace{3mm}
-  \begin{lstlisting}[frame=tb] %[frame=single]
+  \begin{lstlisting}[frame=tb]
   static int i, j, k;
@@
       \begin{axis}[
-          title = Core2 Duo,
+          title = Core,
           width = 7.8cm,
@@
 \label{fig:motivation}
-\caption{Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size}
+\caption{Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size.
+Note the clear spike around 3200 bytes, where the number of cycles executed nearly doubles.
+Measuring performance in this configuration would give a biased result.}
 \end{figure}
@@
     \hline
-Core2 Duo     & Intel\textsuperscript{\textregistered{}} Core\texttrademark{}2 Duo P8600 (laptop) \\
+Core          & Intel\textsuperscript{\textregistered{}} Core\texttrademark{}2 Duo P8600 (laptop) \\
               & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
@@
 Loads are therefore issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
-The prediction is later verified, replaying any instructions that was wrongly assumed to have no dependencies.
+The prediction is later verified, replaying any instructions that were wrongly assumed to have no dependencies.
 Similarly, if the load and store locations are the same, the value can be \emph{forwarded} from the store before it retires.
@@
 A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
-Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
-The number of times this happens can be counted by the following performance counter:
-\begin{describe}{{paddin}}
-  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} Intel Optimization Manual~\citeyear[B.3.4.4]{OptimizationManual}
+Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
+The number of times this happens can be counted by the following event:
+\begin{describe}{{pad}}
+  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
 \end{describe}
@@
 Older architectures such as Core 2 does not have this particular event listed, however 4K aliasing is covered by a more general event:
-\begin{describe}{{paddin}}
-  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''}~\citeN[Table 19-17]{Volume3B}
+\begin{describe}{{pad}}
+  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
 \end{describe}
@@
 System environment variables contain things like the name of the logged in user, home directory, shell state, and various other settings.
-As a source of bias, it is not the environment variables themselves that are important, but rather the effect their total \emph{size} has on the position and alignment of stack.
-As indicated in Figure \ref{fig:virtualmemory}, environment variables and program arguments are allocated in the stack section of virtual memory, close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory} and before the first call frame.
-Changing environment variables will therefore offset the addresses of stack, and consequently all stack allocated variables.
+As a source of bias, it is not the environment variables themselves that are important, but rather the effect their total \emph{size} has on the position and alignment of the stack.
+As indicated in Figure \ref{fig:virtualmemory}, environment variables and program arguments are allocated in the stack section of virtual memory, close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory.} and before the first call frame.
+Changing environment variables will therefore offset the location of the stack, and consequently all stack allocated variables.
 After this offset, the stack is normally realigned to a 16 byte boundary, enforced by the compiler.
@@
 Note that there is no clear relationship between environment size and stack location with ASLR enabled.
-However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
+However, there will still be as many execution contexts with respect to aliasing (considering the stack only), making any occurrences of measurement bias indeed random.
 
@@
 \label{sec:microkernel}
-To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, reproduced below.
+To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, a small C program listed in Figure~\ref{fig:motivation}.
 This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.
 Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
-
-\begin{lstlisting}
-    static int i, j, k;
-    int main() {
-        int g = 0, inc = 1;
-        for (; g < 65536; g++) {
-            i += inc;
-            j += inc;
-            k += inc; 
-        }
-        return 0;
-    }
-\end{lstlisting}
 
@@
     \begin{axis}[
+        title = Haswell,
         width = \textwidth,
@@
 
+Measuring the performance over different environment sizes for our Haswell setup reveals a similar pattern to the one shown in Figure~\ref{fig:motivation} for Core~2.
 Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
 We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
-A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
+A finer sampling is not necessary, because the stack is by default aligned to 16 bytes.
 The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
@@
 This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports have fewer operations going in parallel.
-Note that the number of micro-ops \emph{retired} overall does not change.%, but for some reason the micro-ops stay in the execution pipeline for a shorter time.
-This match with the fact that the memory disambiguation system do not handle address aliasing, limiting the potential for out of order execution~\cite[Page 2-20]{OptimizationManual}.
+Note that the number of micro-ops \emph{retired} overall does not change. %, but for some reason the micro-ops stay in the execution pipeline for a shorter time.
+Bacause the memory disambiguation system does not handle address aliasing, the potential for out of order execution is limited~\cite[Page 2-20]{OptimizationManual}.
 We assume that the tiny kernel size immediately prevents instructions from executing out of order, leading to the lower counts.
@@
 Static variables are fixed and covers 12 contiguous bytes (3 words), in our case the addresses end in 0x0, 0x4 and 0xc, leaving the 0x8 slot free.
-The two automatic variables \texttt{g} and \texttt{inc} occupies 8 contiguous bytes on stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
+The two automatic variables \texttt{g} and \texttt{inc} occupies 8 contiguous bytes on the stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
 In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.
@@
 In conclusion, we identified that address aliasing is the root cause of measurement bias from environment size for this program.
-Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between stack and static data.
+Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between the stack and static data.
 The program is \emph{biased} towards environment sizes that avoids this specific stack alignment, which in principle could be triggered just by changing user name.
 This type of bias can occur in a number of memory configurations.
-Here, the aliasing was caused by interactions between the stack and static variables, but we can imagine similar conflicts between stack and heap, depending also on how dynamic memory allocators behave.
+Here, the aliasing was caused by interactions between the stack and static variables, but we can imagine similar conflicts between the stack and the heap, depending also on how dynamic memory allocators behave.
 Because the stack is used for local variables, complex code is likely to contain many potential conflicts with other areas of memory.
-
-\subsection{Avoiding Aliasing}
-Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
-In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR or program arguments. 
-Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
-A proof of concept of how alias-free code could be generated in this particular case is shown in Figure \ref{lst:loopfixed}.
-If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
-Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.
 
@@
 
+\subsection{Avoiding Aliasing}
+Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
+In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR or program arguments. 
+Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
+A proof of concept of how alias-free code could be generated in this particular case is shown in Figure \ref{lst:loopfixed}.
+If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
+Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.
+
 While this particular solution is impractical, it shows that compilers and programmers \emph{could} take measures to account for aliasing.
+
+\subsection{Other Architectures}
+While we show only the analysis for Haswell here, the same approach gave similar results for all CPUs we tested.
+The biggest practical difference occur for Core 2, where a different set of performance events were used, in particular load\_blocks.overlap_store indicating alias.
+Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}.
+We were able to verify that partial address overlap was indeed occuring synchronously with this event spiking, and we draw the conclusion that address aliasing must be the underlying explanation.
+Based on our results, and existing documentation on available performance events, it is reasonable to assume that the address alias issue is present in (at least) Intel microarchitectures Core, Nehalem, Sandy Bridge and Haswell.
+
+\subsection{Hyper-Threading}
+With Hyper-Threading we see smalls spikes like Core2, likely cache bank issues. % todo: fill in something here
 
@@
 Address aliasing can be caused by conflicting pairs of load and store operations to any part of memory.
-In the previous sections we saw collisions between static data and stack, observing bias from external conditions that affected addresses of automatic variables.
+In the previous sections we saw collisions between static data and the stack, observing bias from external conditions that affected addresses of automatic variables.
 Most dynamic memory is allocated on the \emph{heap}, which is managed by an \emph{allocator}.
@@
 The heap section starts at a relatively low address right above static code and data.
-Memory mapped chunks are placed towards the upper end of the virtual address space, closer to stack.
+Memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack.
 Whether a request is served by the heap or by memory mapping is therefore easy to determine just by looking at the pointer values returned:
@@
 However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address end with 0x010.}
+The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
 This behavior is often the worst case for functions that operate on two or more independent buffers.
@@
 We see that glibc and tcmalloc utilize the normal heap area for smaller allocations, returning numerically low addresses. % Verified by looking at strace that tcmalloc actually uses brk
-Interestingly, jemalloc and Hoard appears to never use the heap, but allocate to memory mapped areas even for smaller requests.
+Interestingly, jemalloc and Hoard appear to never use the heap, but allocate to memory mapped areas even for smaller requests.
 Conversely, tcmalloc seem manage only the heap.
@@
     \begin{axis}[
-        title=Optimization -O2,
+        title=Haswell -O2,
         font=\footnotesize,
@@
     \begin{axis}[
-        title=Optimization -O3,
+        title=Haswell -O3,
         font=\footnotesize,
@@
   \end{tikzpicture}
-  \label{fig:conv-default}
-  \caption{Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}.
+  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}.
 Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
@@
 We have shown how address aliasing can affect program performance under different memory contexts, and how it can explain certain cases of measurement bias. 
-The effect is caused by how speculative and out-of-order memory operations are handled by the Intel ``Haswell'' CPU, which only considers the last 12 address bits to resolve conflicts between load and store operations.
+The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between load and store operations.
+This phenomenon is likely present on most consumer hardware based on x86 Intel platforms today, ranging from Core to Haswell microarchitectures.
 

### 2014-08-15 paper.tex
Update haswell results for microkernel
Includes the updated LDM and L2 event cmask

@@
 % Large plot of cycle count for two 4K stack sections.
-\pgfplotstableread{bin/micro-kernel-cycles.dat}{\stackoffsettable}
+\pgfplotstableread{bin/microkernel-cycles-haswell.dat}{\stackoffsettable}
 \begin{figure*}[t]
@@
       every last column/.style={column type/.add={}{|}}
-    ]{bin/micro-kernel-comparison.csv}
+    ]{bin/microkernel-comparison-haswell.csv}
   }
@@
 The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
-On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from around 270,000 to 135,000.
+On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from around 272,000 to 136,000.
 The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.

### 2014-08-15 paper.tex
Clarify explanation on execution port vacancy
Hopefully easier to follow, comment from one of the reviewers that
the previous explanation was unclear.

@@
 This event is listed in the manual for microarchitectures going back to Nehalem i7, including Ivy Bridge and Haswell CPUs~\cite{Volume3B}.
-Older architectures such as Core 2 does not have this particular event listed, however 4K aliasing is covered by a more general event:
+Older architectures such as Core 2 do not have this particular event listed, however 4K aliasing is covered by a more general event:
 \begin{describe}{{pad}}
@@
 This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports have fewer operations going in parallel.
-Note that the number of micro-ops \emph{retired} overall does not change. %, but for some reason the micro-ops stay in the execution pipeline for a shorter time.
-Bacause the memory disambiguation system does not handle address aliasing, the potential for out of order execution is limited~\cite[Page 2-20]{OptimizationManual}.
-We assume that the tiny kernel size immediately prevents instructions from executing out of order, leading to the lower counts.
-%Our assumption is that this is caused by the CPU being more conservative in the aliasing case, issuing fewer operations (speculatively) in parallel.
-%With fewer instructions per cycle, there is also less pressure on the reservation station.
+Note that the number of micro-ops \emph{retired} overall does not change.
+The memory disambiguation system automatically assumes dependency between a load and a previous store that are aliased, which can limit the potential for issuing many instructions simultaneously and out of order~\cite[Page 2-20]{OptimizationManual}.
+In our case, lower occupation of execution ports and less pressure on the reservation station can be explained by the CPU not being able to issue more operations while waiting for an aliased memory access to be resolved.
 

### 2014-08-15 paper.tex
Reformulate some sentences

@@
 \subsection{Avoiding Aliasing}
-Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
-In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR or program arguments. 
-Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
+Addresses of automatic variables can not be determined statically, because the location of the stack at runtime is generally unknown.
+In addition to being offset by environment variables, the initial stack address can also be perturbed by for example program arguments or address randomization.
+Although one can not easily know if a collision is going to happen for a given environment a priori, it is possible to change the program to account for possible alias conditions.
 A proof of concept of how alias-free code could be generated in this particular case is shown in Figure \ref{lst:loopfixed}.
@@
 We have shown how address aliasing can affect program performance under different memory contexts, and how it can explain certain cases of measurement bias. 
-The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between load and store operations.
-This phenomenon is likely present on most consumer hardware based on x86 Intel platforms today, ranging from Core to Haswell microarchitectures.
+The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
+This phenomenon is likely present on most consumer hardware based on x86 Intel platforms today, ranging from ``Core'' to ``Haswell'' microarchitectures.
 
@@
 \begin{acks}
-This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel Ivy Bridge architecture~\cite{MasterThesis}.
+This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' architecture~\cite{MasterThesis}.
 \end{acks}

### 2014-08-15 paper.tex
Update convolution results for haswell
Including measurements with new LDM CMASK

@@
 
-\pgfplotstableread{bin/conv-default-o2.estimate.dat}{\convtabletwo}
-\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
+\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
+\pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
 \begin{figure}[t]
@@
       ]
-      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwo ;
-      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwo ;
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
       \addlegendentry{Cycles} ;
@@
       ]
-      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethree ;
-      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethree ;
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethreehaswell ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethreehaswell ;
       \addlegendentry{Cycles} ;
@@
       every last column/.style={column type/.add={}{|}}
-    ]{bin/conv-default-o2.estimate.csv}
+    ]{bin/conv-default-o2-haswell.estimate.csv}
   }

### 2014-08-20 analysis/blas/blas.tex
Keep old atlas results and notes

\section{Optimizing Alignment for Matrix Vector Multiplication}
To illustrate how real world applications can be impacted by aliasing, we look at low level numerical applications.
In this section we show that even highly optimized numerical libraries suffer performance degradation from aliasing.
Basic Linear Algebra Subroutines (BLAS), is the de facto standard API for high performance linear algebra routines.
The functionality is divided into three categories:

\begin{description}
  \item{Level~1} Scalar and vector operations, such as dot product and vector addition.
  \item{Level~2} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
  \item{Level~3} Matrix-matrix operations, including the widely applied gemm routine for general matrix-matrix multiplication. 
\end{description}

Functions operating on vectors, from BLAS Level 1 or 2, intuitively appears the most likely to have potential for aliasing.
Many highly optimized implementations of BLAS exists, ATLAS being a widely used and open source alternative. One of the key features of ATLAS is that it uses automatic tuning to optimize for cache efficiency [Whaley:2000:ATLAS].

\pgfplotstableread{bin/libblas.dat}{\libblastable}
\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
\pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=libblas,
        font=\footnotesize,
        xlabel=Address offset of vector $y$,
        %ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.5, % Make it fit to text width side by side
        legend style={anchor=north east}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
      %\addlegendentry{Cycles} ;
      %\addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libatlas,
        font=\footnotesize,
        xlabel=Address offset of vector $y$,
        %ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.5,
        % skip coords between index={20}{32}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
      %\addlegendentry{Cycles} ;
      %\addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libopenblas,
        font=\footnotesize,
        xlabel=Address offset of vector $y$,
        %ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.5,
        % skip coords between index={20}{32}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libopenblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libopenblastable ;
      %\addlegendentry{Cycles} ;
      %\addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \label{fig:conv-default}
  \caption{Wat is dis}
\end{figure}

\section{Optimizing Alignment in BLAS}

To illustrate how real world applications can be impacted by aliasing, we look at low level numerical applications.
In this section we show that even highly optimized numerical libraries suffer performance degradation from aliasing.
Basic Linear Algebra Subroutines (BLAS), is the de facto standard API for high performance linear algebra routines.
The functionality is divided into three categories:

\begin{description}
  \item{Level~1} Scalar and vector operations, such as dot product and vector addition.
  \item{Level~2} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
  \item{Level~3} Matrix-matrix operations, including the widely applied gemm routine for general matrix-matrix multiplication. 
\end{description}

Functions operating on vectors, from BLAS Level 1 or 2, intuitively appears the most likely to have potential for aliasing.
Many highly optimized implementations of BLAS exists, ATLAS being a widely used and open source alternative. One of the key features of ATLAS is that it uses automatic tuning to optimize for cache efficiency [Whaley:2000:ATLAS].

We use the currently latest stable version 3.10.1 of ATLAS, built as a shared library from source.
The automatic tuning happens during the build process; A series of test programs are run to determine cache edges and other properties of the hardware, which in turn affects the resulting binary.

Carefully monitoring cache metrics is particularly important in this case study, as BLAS performance heavily relies on cache efficiency.
Relevant performance statistics for cache hit ratios was considered to rule out cache as the cause of any bias effects.
In addition, we disable hardware prefetching in BIOS to further minimize cache impacts. 
Note however that the same conclusions can be drawn independently of these measures.

\subsection{Aliasing in cblas\_daxpy}
Found aliasing here as well. Two spikes, not sure why. Might want to test other BLAS libraries. Include or leave out? Too much handwaving and too little analysis maybe. Depends on the target paper length as well?

\subsection{Matrix vector multiplication}

Consider matrix-vector multiplication of the form $\boldsymbol{y} = A\boldsymbol{x} + \boldsymbol{b}$.
Let A be of size $M \times N$, where M is the number of rows.

$
\left[\begin{array}{ccccc}
a_{0,0} & a_{0,1} &  &  & a_{0,N}\\
a_{1,0}\\
 &  &  & \ddots\\
a_{M,0} &  &  &  & a_{M,N}
\end{array}\right]\left[\begin{array}{c}
x_{0}\\
x_{1}\\
\vdots\\
\\
x_{N}
\end{array}\right]=\left[\begin{array}{c}
y_{0}\\
y_{1}\\
\vdots\\
y_{M}
\end{array}\right]
$

General matrix vector multiplication is implemented by the level 2 gemv routine, computing $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$
Here, $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix. 
We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
[There are three memory buffers involved, and one might suspect that implementations are succeptible to aliasing between of $\boldsymbol{A}$, $\boldsymbol{x}$ or $\boldsymbol{y}$.]
For the following discussion we look at a particular configuration where $A$ is of size $64\times64$.

In order to accurately measure any effects of aliasing, we need to isolate the actual function call from heap allocation and initialization code.
This is done by running the program twice for each memory configuration, one where cblas\_dgemv is executed $K = 101$ times in a tight loop.
An approximation of the cost for a single function invocation can be expressed as 
$$
t_{\text{estimate}}=\frac{t_{K=101}-t_{K=1}}{100}
$$
where t represents some metric, such as the number of cycles.
Subtracting the $K = 1$ run removes the constant overhead from the $K = 101$ run, and dividing by $100$ averages the values from the remaining iterations.
Other iteration counts could have been used as well.

Figure \ref{fig:heatmap} shows estimated number of alias events for every relevant memory configuration possible for cblas\_dgemv with matrix size 64 $\times$ 64. 
While matrix A is kept fixed at address suffix 0x000, every reasonable variation of (\&x, \&y) is measured.
Each increment represents an address offset of 16 bytes, which is the same granularity one would get from the default heap allocator.
In a 4K address space, there are thus $4096/16 = 256$ possible address suffixes for $\boldsymbol{x}$ and $\boldsymbol{y}$ independently, resulting in a grid of $256^2$ values.

\begin{figure}[h]
  \caption{Estimated aliasing cost of invoking cblas\_dgemv with a matrix size $64 \times 64$ for different memory layouts. Axis show address offsets times 0x10 for vectors $\boldsymbol{x}$ and $\boldsymbol{y}$. Matrix $A$ has fixed to address suffix 0x000.}
  \label{fig:heatmap}
  \includegraphics[width=\columnwidth]{resources/heatmap.eps}
\end{figure}

It appears that aliasing is mostly dependent on the location of $\boldsymbol{y}$, in this case with 8 distinct worst case regions.
This is consistent with false dependencies between $\boldsymbol{y}$ and specific rows in A. 
The first bad region is where $\&y ~= 0$, matching the first row of A.

\begin{table*}[t]
  \caption{Estimated cost of invoking cblas\_dgemv with a matrix size $8192 \times 8192$}
  \label{tab:atlas8k}
  \small
  \centering
  \begin{tabular}{l r r r}
    \toprule
      & (0x010, 0x010, 0x020) & (0x010, 0x010, 0x0b0) & (0x010, 0x140, 0x150) \\
    \midrule
    \perfctr{cpu_clk_unhalted.thread_p} & 162,926,895 & 152,151,263 & 144,583,160 \\
    \perfctr{ld_blocs_partial.address_alias} & \textbf{68,408,790} & \textbf{5,154,981} & \textbf{2,919} \\

    \perfctr{mem_uops_retired.all.loads} & 100,701,777 & 100,702,138 & 100,701,752 \\
    \perfctr{mem_load_uops_retired.hit_lfb} & 31,628,539 & 32,495,287 & 29,853,850 \\
    \perfctr{mem_load_uops_retired.l1_hit} & 68,309,447 & 65,625,881 & 69,387,368 \\
    \perfctr{mem_load_uops_retired.l2_hit} & 711,973 & 1,540,956 & 864,720 \\
    \perfctr{mem_load_uops_retired.l3_hit} & 154,345 & 651,985 & 363,331 \\
    \perfctr{mem_load_uops_retired.l3_miss} & 106,313 & 350,449 & 295,235 \\
    \perfctr{mem_load_uops_retired.all_stores} & 33,562,692 & 33,562,692 & 33,562,692 \\

    \perfctr{load_hit_pre.hw_pf} & 42,914,191 & 47,315,935 & 59,211,320 \\
    \bottomrule
  \end{tabular}
\end{table*}

Notes: 
 - More BR\_INST.COND.TAKEN taken in case (a), but number of retired instructions are the same. Suggests replays of instructions.
 - Inverse HW prefetch numbers from masters

\paragraph{Prefetching}
Measurements repeated with prefetching disabled, giving similar results.

Estimated performance counter statistics for each of the three heap address configurations are shown in Table [?tab:gemv-estimate].
In addition to cycle count and alias events, a number of relevant metrics related to cache activity are also included.
Notice that almost all load micro-ops are served by either the line fill buffer or L1 cache in all cases.
Only a small amount of loads come from L2, and almost none from L3.
The hit rate for L1 actually decreases somewhat (considering LFB as well) with better execution time.
This could be explained by less time for prefetchers to feed the L1 cache with data.
The hardware prefetch counter indicates more hits in cases a and b.
Again, we find that cache efficiency does not explain the performance cliffs we observe.

Our results show that address aliasing between matrix and vector heap buffers can significantly impact performance of dgemv in ATLAS.
Variations in heap addresses alone can give a speedup of more than 30 \%.

\subsection{Dealing with Aliasing}
For the particular case we investigated, a good heuristic is to align heap segments “far apart” within the 4 KiB area of 12-bit suffixes.
More specifically, it appears that address suffixes of A and $\boldsymbol{y}$ are the most important to separate.

As discussed in section [x], aliasing cases like these can be accounted for in software using padding techniques.
A possible runtime solution to adjust heap addresses can be realized as follows:

\begin{itemize}
  \item Allocate some extra space for one of the vectors when calling malloc, for instance sizeof(double) * (M + 0x100) for $\boldsymbol{y}$.

  \item Check the returned pointers for potential alias, i.e. the difference between \&A and \&y. Offset using pointer arithmetic into the array with extra padding at the end, i.e. y += 0x100. 
\end{itemize}

Another option is to explicitly account for “worst cases” in the implementation of routines that are vulnerable to aliasing.
Addresses can be explicitly checked for potential conflicts in cblas\_dgemv, and if possible branch to code that will not suffer from aliasing.

### 2014-08-20 analysis/blas/gemv-mmap.c
Add results for simplified blas benchmark

@@ -50,6 +50,6 @@ int main(int argc, char **argv) {
             y, 1
         );
 
-    printf("A:%p x:%p y:%p\n", A, x, y);
+    // printf("A:%p x:%p y:%p\n", A, x, y);
     return 0;
 }

### 2014-08-21 analysis/perf-env/printenv.c
Improve environment introspection program
Use base pointer to indicate stack position, and add program
arguments to output.

@@ -5,13 +5,13 @@
 ///     <empty>
 ///
 /// $ env -i perf stat ./printenv
-///	    PWD=/home/lars/bias-paper/code/perf-env
+///     PWD=/home/lars/bias-paper/code/perf-env
 ///     SHLVL=0
 ///     PERF_BUILDID_DIR=.debug
 ///     PATH=/usr/libexec/perf-core:/usr/lib/linux-tools/3.11.0-15-generic:/usr/local/bin:/usr/bin:/bin
 ///
 /// $ perf stat ./printenv
-///	    PWD=/home/lars/bias-paper/code/perf-env
+///     PWD=/home/lars/bias-paper/code/perf-env
 ///     SHLVL=0
 ///     PERF_BUILDID_DIR=.debug
 ///     PATH=/usr/libexec/perf-core:/usr/lib/linux-tools/3.11.0-15-generic:/usr/local/bin:/usr/bin:/bin
@@ -20,20 +20,27 @@
 #include <stdio.h>
 #include <string.h>
 
-int main() {
-	char file[64], command[128];
-	sprintf(file, "/proc/%d/environ", getpid());
+int main(int argc, char *argv[]) {
+    long i = 0, *addr;
+    char file[64], command[128];
 
-	strcpy(command, "cat ");
-	strcat(command, file);
-	system(command);
+    for (; i < argc; ++i) 
+        puts(argv[i]);
+    puts("");
 
-	putchar('\n');
+    sprintf(file, "/proc/%d/environ", getpid());
 
-	strcat(command, " | wc -c");
-	system(command);
+    strcpy(command, "cat ");
+    strcat(command, file);
+    system(command);
 
-	printf("stack: %p\n", (void*)file);
+    puts("");
 
-	return 0;
+    strcat(command, " | wc -c");
+    system(command);
+
+    asm("movq %%rbp, %0" : "=r"(addr));
+    printf("%%rbp: %p\n", (void*)&addr);
+
+    return 0;
 }

### 2014-08-21 analysis/perf-env/printenv.py

@@ -15,5 +15,5 @@ import subprocess
 # PERF_BUILDID_DIR=.debug
 # PATH=/usr/libexec/perf-core:/usr/lib/linux-tools/3.11.0-15-generic:/usr/local/bin:/usr/bin:/bin
 
-p = subprocess.Popen("perf stat -e cycles:u ./printenv", env={}, shell=True)
+p = subprocess.Popen("perf stat -e cycles:u bin/printenv", env={}, shell=True)
 p.wait()

### 2014-08-23 paper.tex
Comment on other archs and ht
Explain additional bias effect for Core, and include comment on
hyper-threading. Various smaller rephrasing and fixes.

@@
 Our approach is to search for correlations over multiple environment configurations, using a large set of hardware performance counter measurements.
-We find that \emph{address aliasing} can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.
+We find that \emph{address aliasing} can explain bias from two external factors; environment variable size, and characteristics of dynamically linked heap allocators.
 
@@
 \caption{Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size.
-Note the clear spike around 3200 bytes, where the number of cycles executed nearly doubles.
+Note the clear spike at around 3200 bytes, where the number of cycles executed nearly doubles.
 Measuring performance in this configuration would give a biased result.}
@@
 All experiments are done on a dedicated machine under minimal load, with \emph{frequency scaling} disabled to keep the CPU's clock speed fixed.
-Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads. Also some PMU events require HT to be turned off, or are inaccurate with HT.
+Finally, we disable \emph{Hyper-threading} (HT), which reduces the possibility for resource contention between threads.
+Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.
 
@@
 This event is listed in the manual for microarchitectures going back to Nehalem i7, including Ivy Bridge and Haswell CPUs~\cite{Volume3B}.
-Older architectures such as Core 2 do not have this particular event listed, however 4K aliasing is covered by a more general event:
+Older architectures such as Core do not have this particular event listed, however 4K aliasing is covered by a more general event:
 \begin{describe}{{pad}}
@@
 \label{sec:microkernel}
-To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, a small C program listed in Figure~\ref{fig:motivation}.
+To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:motivation}.
 This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.
 Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
+The following outlines our data and analysis of this program conducted on the ``Haswell'' configuration.
+Similar experiments were performed on each of the other experimental setups as well, all arriving at the same conclusions.
 
@@
 
-Measuring the performance over different environment sizes for our Haswell setup reveals a similar pattern to the one shown in Figure~\ref{fig:motivation} for Core~2.
 Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
@@
 \begin{table}
-  \tbl{Events with significant correlation to cycle count\label{tab:loopcorrelation}}{
+  \tbl{Events with significant correlation to cycle count.\label{tab:loopcorrelation}}{
     \pgfplotstabletypeset[
@@
 The memory disambiguation system automatically assumes dependency between a load and a previous store that are aliased, which can limit the potential for issuing many instructions simultaneously and out of order~\cite[Page 2-20]{OptimizationManual}.
-In our case, lower occupation of execution ports and less pressure on the reservation station can be explained by the CPU not being able to issue more operations while waiting for an aliased memory access to be resolved.
+In our case, lower occupation of execution ports, and less pressure on the reservation station, can be explained by the CPU not being able to issue more operations while waiting for an aliased memory access to be resolved.
 
@@
 While we show only the analysis for Haswell here, the same approach gave similar results for all CPUs we tested.
-The biggest practical difference occur for Core 2, where a different set of performance events were used, in particular load\_blocks.overlap_store indicating alias.
+The biggest practical difference occur for Core 2, where a significantly different set of performance events were used.
+In particular; load\_blocks.overlap_store indicating alias.
 Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}.
-We were able to verify that partial address overlap was indeed occuring synchronously with this event spiking, and we draw the conclusion that address aliasing must be the underlying explanation.
+We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and we draw the conclusion that address aliasing must be the underlying explanation.
+However, this only explains part of what is going on in the Core setup. 
+As shown in Figure~\ref{fig:motivation}, there seem to be yet another bias effect visible at every fourth environment increment.
+These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
+Our assumption is that this is a different phenomenon, not related to 4K aliasing.
+We were only able to observe the effect on this particular architecture, and did not investigate the root cause further.
+
 Based on our results, and existing documentation on available performance events, it is reasonable to assume that the address alias issue is present in (at least) Intel microarchitectures Core, Nehalem, Sandy Bridge and Haswell.
@@
 \subsection{Hyper-Threading}
-With Hyper-Threading we see smalls spikes like Core2, likely cache bank issues. % todo: fill in something here
+As described in Section~\ref{sec:methodology}, disabling support for Hyper-Threading on applicable architectures was done in order to avoid any potential interference while collecting data.
+However, this begs the question if doing so might affect the existence of aliasing itself.
+To verify this was not the case, we ran the experiment again on the Haswell setup with Hyper-Threading enabled.
+Results for cycles executed and alias events were virtually identical, indicating that configuration of HTT either way is not impactful to our conclusions.
 
@@
 
-Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers.\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
-We observe the addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} for different size parameters.
-Equal three digit address suffix indicate an aliasing pair.
-In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
-\begin{enumerate}
-  \item Thread-Caching Malloc (tcmalloc) by Google~\cite{TCMalloc}.
-  \item jemalloc, originally developed for FreeBSD~\cite{JEMalloc}.
-  \item Hoard~\cite{Berger:2000:Hoard}.
-\end{enumerate}
-All of the above focus heavily on performance in multithreaded environments.
-Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
-Here, we only look at behavior for a single thread, and whether the addresses returned alias or not.
 \begin{table}[t]
-  \label{tab:mallocompare}
-  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
+  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.\label{tab:mallocompare}}{
     \pgfplotstabletypeset[
@@
 \end{table}
+
+Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers.\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
+We observe the addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} for different size parameters.
+Equal three digit address suffix indicate an aliasing pair.
+In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
+\begin{enumerate}
+  \item Thread-Caching Malloc (tcmalloc) by Google~\cite{TCMalloc}.
+  \item jemalloc, originally developed for FreeBSD~\cite{JEMalloc}.
+  \item Hoard~\cite{Berger:2000:Hoard}.
+\end{enumerate}
+All of the above focus heavily on performance in multithreaded environments.
+Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
+Here, we only look at behavior for a single thread, and whether the addresses returned alias or not.
 

### 2014-08-26 paper.tex
Updated the setup table.

@@
 Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture specifically.
-We later reproduced similar results on the previous ``Sandy Bridge'' CPU, as well as an older Core2 Duo machine.
-Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.
+We later reproduced similar results on the previous ``Nehalem'', ``Sandy Bridge'' and ``Ivy Bridge'' CPUs, as well as an older Core2 Duo machine.
+Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper. The compiler used is GCC 4.8.2-19ubuntu1 and the same binary is tested on all systems.
 
@@
 Core          & Intel\textsuperscript{\textregistered{}} Core\texttrademark{}2 Duo P8600 (laptop) \\
-              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
-Sandy Bridge  & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-3???? \\
-              & Ubuntu ?? \\ \hline
+              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0) \\ \hline
+Nehalem       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-950 \\
+              & 64-bit Ubuntu 12.04 LTS (Linux 3.2.0) \\ \hline
+Ivy Bridge    & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i5-3470 \\
+              & 64-bit Ubuntu 12.04 LTS (Linux 3.8.0) \\ \hline
+Sandy Bridge  & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-2670QM (laptop) \\
+              & 64-bit Ubuntu 11.04 (Linux 3.0.0) \\ \hline
 Haswell       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K \\
-              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
+              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0) \\ \hline
   \end{tabular}}
@@
 Most importantly, this means keeping the memory address space under control.
-For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Pax:ASLR,Bhatkar:AddressObfuscation}.
+For security reasons, addresses of the stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Pax:ASLR,Bhatkar:AddressObfuscation}.
 By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.

### 2014-08-26 paper.tex
Rephrase some sentences and cleanup table

@@
 However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
-Seemingly irrelevant variables, such as the length of your user name, can impact program performance.
+Seemingly irrelevant properties---such as the length of your user name---can impact program performance.
 In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
-Our approach is to search for correlations over multiple environment configurations, using a large set of hardware performance counter measurements.
-We find that \emph{address aliasing} can explain bias from two external factors; environment variable size, and characteristics of dynamically linked heap allocators.
+Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
+We find that \emph{address aliasing} can explain bias from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
 
@@
 Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.
-In Figure~\ref{fig:motivation} we have reproduced a result originally presented by \citeN{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program depends on environment variables.
-This example provides the basis for our analysis.
 
@@
 \caption{Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size.
-Note the clear spike at around 3200 bytes, where the number of cycles executed nearly doubles.
+Notice the clear spike at around 3200 bytes, where the number of cycles executed nearly doubles.
 Measuring performance in this configuration would give a biased result.}
@@
 A literature study of 88 papers by \citeN{Mytkowicz:2008:OE&MB} showed that the median reported speedup was within the margin of bias, possibly invalidating the claimed speedups. 
-To alleviate the impact of measurement bias in performance analysis, researchers have proposed to use causal analysis techniques and randomization of execution contexts~\cite{Mytkowicz:2008:OE&MB}.
+To alleviate the impact of measurement bias in performance analysis, researchers have proposed techniques like causal analysis, and randomization of execution contexts~\cite{Mytkowicz:2008:OE&MB}.
 Others treat the interaction complexity as a potential optimization problem, using search over variant spaces to find optimal environments~\cite{Knights:2009:BlindOpt}.
-Our approach is to analyze isolated programs in detail using \emph{hardware performance measurements}, in an attempt to identify which underlying mechanisms are causing bias.
+Our approach is to analyze isolated programs in detail using hardware performance counters, in an attempt to identify which underlying mechanisms are causing bias.
 
-In this paper we show how some instances of measurement bias can be explained by \emph{address aliasing}, an artifact of a presumably Intel specific optimization on the out-of-order execution pipeline.
-The CPU uses a heuristic for determining whether loads are dependent on previous stores, comparing only the last 12 virtual address bits.
-False dependencies can happen for aliased memory accesses, and incur performance penalties.
+In Figure~\ref{fig:motivation}, we have reproduced a result originally presented by \citeN{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program depends on environment variables.
+This example provides the basis for our analysis.
+We show how some instances of measurement bias, such as the above, can be explained by \emph{address aliasing}, an artifact of a presumably Intel specific optimization on the out-of-order execution pipeline.
+The CPU uses a heuristic for determining whether loads are dependent on previous stores, comparing only the 12 least significant virtual address bits.
+Aliased memory accesses can produce false dependencies, and incur performance penalties.
 We look at how address aliasing can trigger bias from two external factors; size of system environment variables and configuration of dynamic heap allocation libraries.
@@
 Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.
-Performance counters are supported in the Linux kernel, accessed via a tool called \emph{perf}.
-This tool instructs the kernel to enable the hardware Performance Monitoring Unit (PMU) for a different program, before it is executed.
+Performance counters are supported in the Linux kernel, and can be accessed via a tool called \emph{perf}.
+This utility instructs the kernel to enable the processor's Performance Monitoring Unit (PMU), before executing a specified program.
 Using this technique avoids modifications to the executed program, unlike tools like PAPI~\cite{PAPI:PortableInterface}.
@@
 Only a small set of events are collected at a time, to ensure events are actually counted continuously.
-We do not use multiplexing between a limited set of counter registers, as it will introduce too many variations~\cite{Wiplove:ImproveEstimation}.
+We do not use multiplexing between a limited set of counter registers, as it will introduce significant variations~\cite{Wiplove:ImproveEstimation}.
 % Paper is rename of "Toward Accurate Performance Evaluation using Hardware Counters".
@@
 
-Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture specifically.
-We later reproduced similar results on the previous ``Nehalem'', ``Sandy Bridge'' and ``Ivy Bridge'' CPUs, as well as an older Core2 Duo machine.
-Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper. The compiler used is GCC 4.8.2-19ubuntu1 and the same binary is tested on all systems.
+Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture.
+We later reproduced similar results for the ``Core'', ``Nehalem'', and ``Ivy Bridge'' microarchitectures as well.
+Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper. 
+% The compiler used is GCC 4.8.2-19ubuntu1, and the same binaries are tested on all systems.
 
@@
 Core          & Intel\textsuperscript{\textregistered{}} Core\texttrademark{}2 Duo P8600 (laptop) \\
-              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0) \\ \hline
+              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
 Nehalem       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-950 \\
-              & 64-bit Ubuntu 12.04 LTS (Linux 3.2.0) \\ \hline
+              & 64-bit Ubuntu 12.04 LTS (Linux 3.2.0), GCC 4.8.2-19ubuntu1 \\ \hline
+% Sandy Bridge  & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-2670QM (laptop) \\
+%               & 64-bit Ubuntu 11.04 (Linux 3.0.0) \\ \hline
 Ivy Bridge    & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i5-3470 \\
-              & 64-bit Ubuntu 12.04 LTS (Linux 3.8.0) \\ \hline
-Sandy Bridge  & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-2670QM (laptop) \\
-              & 64-bit Ubuntu 11.04 (Linux 3.0.0) \\ \hline
+              & 64-bit Ubuntu 12.04 LTS (Linux 3.8.0), GCC 4.8.2-19ubuntu1 \\ \hline
 Haswell       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K \\
-              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0) \\ \hline
+              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
   \end{tabular}}
@@
 This event is listed in the manual for microarchitectures going back to Nehalem i7, including Ivy Bridge and Haswell CPUs~\cite{Volume3B}.
-Older architectures such as Core do not have this particular event listed, however 4K aliasing is covered by a more general event:
+Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
 \begin{describe}{{pad}}
@@
 
-As address aliasing depends on the memory addresses of loads and stores, environmental factors that affects memory has the potential to induce aliasing conditions.
+As address aliasing depends on the memory addresses of loads and stores, any environmental factor that affects memory layout has the potential to induce aliasing conditions.
 In the following sections, we show how performance penalties from address aliasing can be the root cause of measurement bias.
@@
 The following outlines our data and analysis of this program conducted on the ``Haswell'' configuration.
-Similar experiments were performed on each of the other experimental setups as well, all arriving at the same conclusions.
+Similar experiments were conducted on the other experimental setups listed in Table~\ref{tab:machines} as well, all arriving at the same conclusion.
 
@@
 The biggest practical difference occur for Core 2, where a significantly different set of performance events were used.
-In particular; load\_blocks.overlap_store indicating alias.
-Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}.
-We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and we draw the conclusion that address aliasing must be the underlying explanation.
+In particular; \emph{load\_blocks.overlap_store} indicating alias.
+Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}, where the authors also used a Core 2 processor.
+We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and draw the conclusion that address aliasing must be the underlying explanation.
 However, this only explains part of what is going on in the Core setup. 
@@
 
-Based on our results, and existing documentation on available performance events, it is reasonable to assume that the address alias issue is present in (at least) Intel microarchitectures Core, Nehalem, Sandy Bridge and Haswell.
+Based on our results, and existing documentation on available performance events, it is reasonable to assume that the address alias issue exists for a wide range of Intel microarchitectures, going as far back as the ``Core'' platform.
 

### 2014-08-28 paper.tex
Add plots for additional architectures

@@
 \usepackage{tikz}
+\usepackage{subfigure}
 \usepackage{listings}
@@
 
+% EXPERIMENTAL!
+\pgfplotstableread{bin/microkernel-core.dat}{\tblmicrokernelcore}
+\pgfplotstableread{bin/microkernel-nehalem.dat}{\tblmicrokernelnehalem}
+\pgfplotstableread{bin/microkernel-ivybridge.dat}{\tblmicrokernelivybridge}
+\pgfplotstableread{bin/microkernel-haswell-ht.dat}{\tblmicrokernelhaswellht}
+\begin{figure}[t]
+  \begin{tabular}{lr}
+    \subfigure[Core]
+    {\begin{tikzpicture}
+      \begin{axis}[
+          width=\textwidth/2.0, % Make it fit to text width side by side
+          height = 4cm,
+          font = \footnotesize,
+          domain = 0:4096,
+          xtick = {0,1024,...,4096},
+          xmin = 0,
+          xmax = 4096,
+          cycle list name = exotic
+        ]
+        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelcore ;
+      \end{axis}
+      \label{subfig:microkernel-core}
+    \end{tikzpicture}} &
+    \subfigure[Nehalem]
+    {\begin{tikzpicture}
+      \begin{axis}[
+          width=\textwidth/2.0, % Make it fit to text width side by side
+          height = 4cm,
+          font = \footnotesize,
+          domain = 0:4096,
+          xtick = {0,1024,...,4096},
+          xmin = 0,
+          xmax = 4096,
+          cycle list name = exotic
+        ]
+        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelnehalem ;
+      \end{axis}
+      \label{subfig:microkernel-nehalem}
+    \end{tikzpicture}} \\
+    \subfigure[Ivy Bridge]
+    {\begin{tikzpicture}
+      \begin{axis}[
+          width=\textwidth/2.0, % Make it fit to text width side by side
+          height = 4cm,
+          font = \footnotesize,
+          domain = 0:4096,
+          xtick = {0,1024,...,4096},
+          xmin = 0,
+          xmax = 4096,
+          cycle list name = exotic
+        ]
+        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelivybridge ;
+      \end{axis}
+      \label{subfig:microkernel-ivybridge}
+    \end{tikzpicture}} &
+    \subfigure[Haswell (HT)]
+    {\begin{tikzpicture}
+      \begin{axis}[
+          width=\textwidth/2.0, % Make it fit to text width side by side
+          height = 4cm,
+          font = \footnotesize,
+          domain = 0:4096,
+          xtick = {0,1024,...,4096},
+          xmin = 0,
+          xmax = 4096,
+          cycle list name = exotic
+        ]
+        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelhaswellht ;
+      \end{axis}
+      \label{subfig:microkernel-haswell-ht}
+    \end{tikzpicture}}
+  \end{tabular}
+  \caption{\label{fig:microkernel-all-archs}Microkernel experiment on different architectures. Horizontal axis represents bytes added to environment, and vertical axis is number of cycles executed.}
+\end{figure}
+
 \subsection{Other Architectures}
 While we show only the analysis for Haswell here, the same approach gave similar results for all CPUs we tested.
-The biggest practical difference occur for Core 2, where a significantly different set of performance events were used.
-In particular; \emph{load\_blocks.overlap_store} indicating alias.
+Figure~\ref{fig:microkernel-all-archs} shows cycle count plots for Core, Nehalem and Ivy Bridge as well, revealing very similar patterns.
+
+There were some practical differences in obtaining these results, in particular slightly different sets of performance events supported on each architecture.
+For the Core setup we use \emph{load\_blocks.overlap\_store} instead of \emph{load\_blocks\_partial.address\_alias} to indicate alias.
 Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}, where the authors also used a Core 2 processor.
@@
 However, this only explains part of what is going on in the Core setup. 
-As shown in Figure~\ref{fig:motivation}, there seem to be yet another bias effect visible at every fourth environment increment.
+As shown in Figure~\ref{subfig:microkernel-core}, there seem to be yet another bias effect visible at every fourth environment increment.
 These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
@@
 To verify this was not the case, we ran the experiment again on the Haswell setup with Hyper-Threading enabled.
-Results for cycles executed and alias events were virtually identical, indicating that configuration of HTT either way is not impactful to our conclusions.
+The raw cycle count results are shown in Figure~\ref{subfig:microkernel-haswell-ht}.
+During this experiment we pinned the process to a particular core using \texttt{taskset}, and ran \texttt{cat /dev/zero > /dev/null} to create some background noise on the other (shared) core.
+While the results are more noisy, we see the same trend with respect to cycles executed and alias events spiking at a particular point.
+This indicates that configuration of Hyper-Threading either way is not impactful to our conclusions, address aliasing is still present.
 

### 2014-08-30 paper.tex
Rewrite other architectures
Remove duplicate Core chart, cleanup and rewrite various paragraphs

--- 2014-08-26 paper.tex
@@
 
-\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
+%\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
+\pgfplotstableread{bin/microkernel-core.dat}{\tblmicrokernelcore}
 \begin{figure}[t]
@@
 \end{minipage}
-%\hfill
 \begin{minipage}[t]{0.55\textwidth}
   \begin{flushleft}
-    %\fbox{
     \begin{tikzpicture}
@@
         ]
-        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \microkernelcoretwotable ;
+        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelcore ;
       \end{axis}
     \end{tikzpicture}
-    %}
   \end{flushleft}
@@
 We show how techniques like manual padding of allocations and dynamic detection of aliasing conditions can be used to improve performance. 
-Understanding and handling address aliasing can give significant speedups, and our findings shows potential for further architecture-specific optimizations.
+Being able to handle address aliasing can give significant speedups, and our findings show potential for further architecture-specific optimizations.
 
@@
 Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.
-% The compiler used is GCC 4.8.2-19ubuntu1, and the same binaries are tested on all systems.
 
 \begin{table}
-  \tbl{Experimental setups used\label{tab:machines}}{
+  \tbl{Experimental setups used.\label{tab:machines}}{
   \begin{tabular}{|l|l|}
@@
       draw, anchor=center, text width=2cm
-    ] (m)
-      {
+    ] (m) {
         \nodepart{one}
@@
   \caption{Memory execution context, assuming a 64-bit process running on a Linux system.
-  %Relative positions of some important sections of memory at runtime
   Initial addresses of stack, heap and memory mapped files are often randomized for security reasons.
@@
 One of the issues that can limit throughput is dependencies between a load and previous stores.
-% According to Intel, typical software consists of about 38~\% memory access instructions, with about two thirds of them being loads.
 To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}. 
@@
 \end{describe}
-This event is listed in the manual for microarchitectures going back to Nehalem i7, including Ivy Bridge and Haswell CPUs~\cite{Volume3B}.
+This event is listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}.
 Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
@@
 
-Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
-We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
+Performance counter measurements of cycle counts over 512 different environment sizes are shown in Figure~\ref{fig:envbias}.
+Every 16 byte increment of environment size is measured, covering two 4K periods of initial stack addresses.
 A finer sampling is not necessary, because the stack is by default aligned to 16 bytes.
-The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
+The microkernel is compiled with {\small{GCC}} using no optimization; any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
 
@@
 
-% See analysis for results showing this does not alias
+% See /analysis for results showing this does not alias
 \begin{figure}
@@
   \end{lstlisting}
-  \caption{Dynamically detect aliasing case, and avoid by pushing another stack frame.}
+  \caption{Modified microkernel that can dynamically detect aliasing case, and avoid it by pushing another stack frame.}
   \label{lst:loopfixed}
@@
 
-While this particular solution is impractical, it shows that compilers and programmers \emph{could} take measures to account for aliasing.
+Even though this particular solution is impractical, it shows that compilers and programmers \emph{could} take measures to account for aliasing.
+
+\pgfplotstableread{bin/microkernel-nehalem.dat}{\tblmicrokernelnehalem}
+\pgfplotstableread{bin/microkernel-ivybridge.dat}{\tblmicrokernelivybridge}
+\begin{figure}[t]
+  \begin{tikzpicture}
+    \begin{axis}[
+        title = Nehalem (HT),
+        width = \textwidth/2.0,
+        height = 4.5cm,
+        xlabel = Bytes added to environment,
+        ylabel = Cycles,
+        font = \footnotesize,
+        domain = 0:4096,
+        xtick = {0,1024,...,4096},
+        xmin = 0,
+        xmax = 4096,
+        cycle list name = exotic
+      ]
+      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelnehalem ;
+    \end{axis}
+  \end{tikzpicture}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title = Ivy Bridge,
+        width = \textwidth/2.0,
+        height = 4.5cm,
+        xlabel = Bytes added to environment,
+        ylabel = Cycles,
+        font = \footnotesize,
+        domain = 0:4096,
+        xtick = {0,1024,...,4096},
+        xmin = 0,
+        xmax = 4096,
+        cycle list name = exotic
+      ]
+      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelivybridge ;
+    \end{axis}
+  \end{tikzpicture}
+  \caption{\label{fig:microkernel-all-archs}Microkernel experiment on the ``Nehalem'' and ``Ivy Bridge'' microarchitectures.}
+\end{figure}
 
 \subsection{Other Architectures}
-While we show only the analysis for Haswell here, the same approach gave similar results for all CPUs we tested.
-The biggest practical difference occur for Core 2, where a significantly different set of performance events were used.
-In particular; \emph{load\_blocks.overlap_store} indicating alias.
-Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}, where the authors also used a Core 2 processor.
+While we show only the analysis for ``Haswell'' here, the same approach gave similar results for all the CPUs we tested.
+Figure~\ref{fig:microkernel-all-archs} shows additional cycle count plots for ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
+The experiment was also done for the older ``Core'' setup, shown in Figure~\ref{fig:motivation}.
+On this architecture, the \emph{load\_blocks.overlap\_store} event is used to indicate address aliasing.
+Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}, where the authors used a similar setup.
 We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and draw the conclusion that address aliasing must be the underlying explanation.
-However, this only explains part of what is going on in the Core setup. 
-As shown in Figure~\ref{fig:motivation}, there seem to be yet another bias effect visible at every fourth environment increment.
+Still, this only explains part of what is going on in Figure~\ref{fig:motivation}, as there seem to be yet another bias effect at every fourth environment increment.
 These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
@@
 
-Based on our results, and existing documentation on available performance events, it is reasonable to assume that the address alias issue exists for a wide range of Intel microarchitectures, going as far back as the ``Core'' platform.
+Results for the ``Nehalem'' and ``Ivy Bridge'' architectures are more clear-cut, with large spikes only at the aliasing environment configuration.
+It is clear that the address alias issue exist for a wide range of Intel microarchitectures.
 
-\subsection{Hyper-Threading}
-As described in Section~\ref{sec:methodology}, disabling support for Hyper-Threading on applicable architectures was done in order to avoid any potential interference while collecting data.
-However, this begs the question if doing so might affect the existence of aliasing itself.
-To verify this was not the case, we ran the experiment again on the Haswell setup with Hyper-Threading enabled.
-Results for cycles executed and alias events were virtually identical, indicating that configuration of HTT either way is not impactful to our conclusions.
+\paragraph{Note on Hyper-Threading}
+As described in Section~\ref{sec:methodology}, Hyper-Threading was disabled by default in order to avoid any potential interference while collecting data.
+However, we found that configuration of Hyper-Threading either way did not change the alias behavior.
+As an example of this, we show the results for ``Nehalem'' with HT enabled in Figure~\ref{fig:microkernel-all-archs}.
 
@@
 
+Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between the heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
+The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
+Numbers on the x axis represent the amount of offset, measured in number of \texttt{sizeof(float)} bytes.
+An offset of zero is the default behavior for this program when using \texttt{malloc} and moderately large inputs.
+Both for optimization level 2 and 3, the default alignment is close to worst case performance.
+The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
+This phenomenon is only observed for address separations close to zero, thus we only show the first 20 data points.
+If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
+
+More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we show the subset that correlates best with the cycle count.
+The numbers are for the level 2 optimization case, which we choose to study in more detail.
+However, the performance data looks fairly similar between the two data sets, with a couple of events that appear to stand out:
+\begin{itemize}
+    \item A high number of resource stalls for the default alignment, which is reduced substantially together with increasing offsets. 
+    \item A high number of cycles with memory loads pending, indicating that the pipeline is stalled waiting for load operations to resolve.
+    \item Changes to the number of micro-ops executed for certain ports.
+\end{itemize}
+
 \pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
@@
 
-Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between the heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
-The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
-Numbers on the x axis represent the amount of offset, measured in number of \texttt{sizeof(float)} bytes.
-An offset of zero is the default behavior for this program when using \texttt{malloc} and moderately large inputs.
-Both for optimization level 2 and 3, the default alignment is close to worst case performance.
-The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
-This phenomenon is only observed for address separations close to zero, thus we only show the first 20 data points.
-If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
-
-More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we show the subset that correlates best with the cycle count.
-The numbers are for the level 2 optimization case, which we choose to study in more detail.
-However, the performance data looks fairly similar between the two data sets, with a couple of events that appear to stand out:
-\begin{itemize}
-    \item A high number of resource stalls for the default alignment, which is reduced substantially together with increasing offsets. 
-    \item A high number of cycles with memory loads pending, indicating that the pipeline is stalled waiting for load operations to resolve.
-    \item Changes to the number of micro-ops executed for certain ports.
-\end{itemize}
-
 Interestingly, it looks like small address offsets incur a massive increase in operations issued on port 0 for the O2 case.
-On Haswell, this port handles various ALU operations, and branching together with port 6~\cite[Figure 2.1]{OptimizationManual}.
+On ``Haswell'', this port handles various ALU operations, and branching together with port 6~\cite[Figure 2.1]{OptimizationManual}.
 As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branches are being re-issued.
@@
 Strategies for detecting bias through causal analysis and randomization of experimental setups is introduced.
-The microkernel example we analyzed is taken from a followup paper by the same authors, presenting similar results for an older Core 2 processor~\cite{Mytkowicz:2009:WrongData}.
-Assuming the root causes of measurement bias were the same, our results indicate that address aliasing issues is probably relevant on several previous generations of Intel architectures as well.
+The microkernel example we analyzed is taken from a followup paper by the same authors, where it is used as an example of how environment size can cause bias~\cite{Mytkowicz:2009:WrongData}.
+%Assuming the root causes of measurement bias were the same, our results indicate that address aliasing issues is probably relevant on several previous generations of Intel architectures as well.
 
@@
 \label{sec:conclusions}
-We have shown how address aliasing can affect program performance under different memory contexts, and how it can explain certain cases of measurement bias. 
+We have shown how address aliasing affects program performance under different memory contexts, and how it can explain certain cases of measurement bias.
 The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
-This phenomenon is likely present on most consumer hardware based on x86 Intel platforms today, ranging from ``Core'' to ``Haswell'' microarchitectures.
+This phenomenon is likely present on most consumer hardware based on Intel platforms today, ranging from ``Core'' to ``Haswell'' microarchitectures.
 
-In general, any change to memory layout of data can potentially introduce bias effects from address aliasing.
+In general, any change to how data is laid out in memory layout can potentially introduce bias effects from address aliasing.
 Analyzing an example with bias to environment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
@@
 \begin{acks}
-This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' architecture~\cite{MasterThesis}.
+This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
 \end{acks}

### 2014-08-30 paper.tex
Clarifying some sentences
Based on reviewer feedback

@@
 
-We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe, and in general follow best practices on gathering data~\cite{Mytkowicz:2009:WrongData}.
+Best practices for controlling the execution context were applied, ensuring that we are not affected by any unwanted bias~\cite{Mytkowicz:2009:WrongData}.
 Most importantly, this means keeping the memory address space under control.
@@
 Using the L1 hit and miss counts also revealed that the L1 load is 99.3\% and store hit rate is ~100\%.
-Thus, these events do not indicate a cache hit rate issue, and is probably linked to how the CPU handles address alias.
+These events do not indicate a cache hit rate issue, but are probably due to having to wait for aliased writes to be retired and committed to cache.
 
@@
 \subsection{Ways to Deal with Heap Address Aliasing}
-Performance penalties caused by address aliasing can be significant in real applications, creating bias towards certain memory layouts.
+Performance penalties caused by address aliasing can be significant, creating bias towards certain memory layouts.
 However, with an understanding of the underlying mechanism that causes bias, the effects can be predicted and to some extent eliminated in software.
@@
 Their space of variants consist of different position and alignment for each function and global variable, in principle exploring the different configurations possible by altering link order.
-Others have used machine learning to classify performance counter data, recognizing patterns or \emph{pathologies} in measurements to recognize optimization opportunities~\cite{Yoo:ADD}.
+Others have used machine learning to classify performance counter data, recognizing patterns or \emph{pathologies} in measurements to identify optimization opportunities~\cite{Yoo:ADD}.
 The authors of MAO present a framework for extending compilers to provide very low level microarchitectural assembly optimization, using rules, pattern matching, and random insertion of nop-instructions to discover optimal assembly outputs~\cite{Hundt:2011:MAO}.

### 2014-08-31 paper.tex
Reintroduce blas
Might as well

@@
         font=\footnotesize,
-        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
+        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
         ylabel=Event count,
@@
 
+\section{Aliasing in Real Applications}
+To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
+Basic Linear Algebra Subroutines (BLAS), is the de facto standard API for high performance linear algebra routines.
+The functionality is divided into three categories:
+\begin{describe}{{pad}}
+  \item[\emph{Level 1}.] Scalar and vector operations, such as dot product and vector addition.
+  \item[\emph{Level 2}.] Matrix-vector operations, such as gemv for general matrix-vector multiplication.
+  \item[\emph{Level 3}.] Matrix-matrix operations, including the widely applied gemm routine for general matrix-matrix multiplication.
+\end{describe}
+Procedures operating on vectors, from level 1 or 2, intuitively appears the most likely to have potential for aliasing.
+
+We look at three different libraries available from the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}.
+We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$
+In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
+We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
+Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
+In code, the function is invoked with the following parameters:
+\begin{lstlisting}
+cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
+\end{lstlisting}
+
+Each of the structures \texttt{A}, \texttt{x} and \texttt{y} represent different locations in memory, and their addresses is part of the execution context.
+The \texttt{mmap} method outlined in the previous section is used to control the location of each of these buffers on the heap.
+Out of trial and error, it turns out the memory address offset of vector $y$ compared to matrix $A$ is the most important with respect to the number of alias events.
+For the following experiment, we set $A = I$ (identity matrix) and $x = [1, 2, ... N]$, with parameters $M = N = 8192$.
+Address suffix of $A$ and $x$ is fixed to 0x000, while performance events are sampled for increasing address offset of $y$, starting at a baseline of 0x000.
+The experiment was run on the ``Haswell'' setup, with Hyper-Threading enabled.
+In an attempt to isolate the cost of the \texttt{cblas_dgemv} invocation, we use the estimation approach described in Section~\ref{sec:heap} to subtract the constant overhead from heap allocation and initialization.
+
+\pgfplotstableread{bin/libblas.dat}{\libblastable}
+\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
+\pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
+\begin{figure}[t]
+  \centering
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=libblas,
+        font=\footnotesize,
+        xlabel=Address offset of vector $y$,
+        ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/2.7, % Make it fit to text width side by side
+        legend style={anchor=north east}
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
+    \end{axis}
+  \end{tikzpicture}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=libatlas,
+        font=\footnotesize,
+        xlabel=Address offset of vector $y$,
+        ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/2.7
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
+    \end{axis}
+  \end{tikzpicture}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=libopenblas,
+        font=\footnotesize,
+        xlabel=Address offset of vector $y$,
+        ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/2.7
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libopenblastable ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libopenblastable ;
+    \end{axis}
+  \end{tikzpicture}
+  \label{fig:blas}
+  \caption{Performance of \texttt{dgemv} for different {\small{BLAS}} packages, with varying address offset between matrix $A$ and vector $y$.}
+\end{figure}
+
+Cycles executed and alias events for each library is shown in Figure~\ref{fig:blas}.
+Each library behaves differently, but in all cases we identify a significant amount of alias. 
+The diagrams can be understood the same way as for the convolution example; partial address overlap occur when the address suffix delta of $A$ and $y$ is close to zero.
+Libblas generates the most distinct pattern, with a relatively clear spike for deltas between 1 and 6 \texttt{double} elements.
+% Need to run experiment again
+Correlation between cycles executed and address aliasing events were calculated to 95\% for libblas, 78\% for libatlas and 44\% for libopenblas.
+
+We do not go into a more detailed analysis of the raw performance data here, but note that we see many of the previously discussed events stand out; such as execution port occupancy, resource stalls, branch instructions executed, offcore requests, and ldm pending.
+Especially for the libblas case, results look quite similar to what we saw for the convolution kernel.
+
+Our results show that address aliasing can indeed occur in real, performance critical applications.
+The amount of \emph{bias} this creates with respect to number of cycles executed is not as clear in all cases, but at least for the libblas case we are pretty confident in claiming that aliasing is indeed causing the performance degradation.
+
 \section{Related Work}
@@
 For heap address aliasing especially, we find that manual intervention can be required in order to achieve optimal performance.
+Finally, we look at different linear algebra libraries and find examples of alias issues in matrix-vector multiplication routines.
 Our results should inspire more clever optimizations and heuristics taking address aliasing effects into account, as the potential performance improvements can be substantial.

### 2014-08-31 paper.tex
Revamp blas discussion
Add legends to plot, fix references, update abstract, and many
other smaller fixes

@@
 The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in one of our examples.
+Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
 Our results can enable new architecture specific optimization strategies to account for this phenomenon.
@@
 % Free to choose keywords, but must be in alphabetical order
-\keywords{4K address aliasing, heap allocators, measurement bias, memory disambiguation}
+\keywords{4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation}
 
@@
 To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
-Basic Linear Algebra Subroutines (BLAS), is the de facto standard API for high performance linear algebra routines.
+Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
 The functionality is divided into three categories:
@@
   \item[\emph{Level 2}.] Matrix-vector operations, such as gemv for general matrix-vector multiplication.
-  \item[\emph{Level 3}.] Matrix-matrix operations, including the widely applied gemm routine for general matrix-matrix multiplication.
+  \item[\emph{Level 3}.] Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
 \end{describe}
@@
 
-We look at three different libraries available from the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}.
-We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$
+We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
+We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
 In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
@@
 The \texttt{mmap} method outlined in the previous section is used to control the location of each of these buffers on the heap.
-Out of trial and error, it turns out the memory address offset of vector $y$ compared to matrix $A$ is the most important with respect to the number of alias events.
-For the following experiment, we set $A = I$ (identity matrix) and $x = [1, 2, ... N]$, with parameters $M = N = 8192$.
-Address suffix of $A$ and $x$ is fixed to 0x000, while performance events are sampled for increasing address offset of $y$, starting at a baseline of 0x000.
+From our experiments, we found that the memory address offset of vector $\boldsymbol{y}$ compared to matrix $A$ is the most important with respect to the number of alias events.
+For the results presented here, we set $A = I$ (identity matrix) and $\boldsymbol{x} = [0, 1, 2, ..., N-1]$, with parameters $M = N = 1024$.
+Address suffix of $A$ and $\boldsymbol{x}$ is fixed to 0x000, while performance events are sampled for increasing address offset of $\boldsymbol{y}$, starting at a baseline of 0x000.
 The experiment was run on the ``Haswell'' setup, with Hyper-Threading enabled.
-In an attempt to isolate the cost of the \texttt{cblas_dgemv} invocation, we use the estimation approach described in Section~\ref{sec:heap} to subtract the constant overhead from heap allocation and initialization.
+In an attempt to isolate the cost of a single \texttt{cblas_dgemv} invocation, we use the estimation approach described in Section~\ref{sec:heap} to subtract the constant overhead from heap allocation and initialization.
 
@@
         title=libblas,
+        height=5.0cm,
         font=\footnotesize,
-        xlabel=Address offset of vector $y$,
+        xlabel=Address offset of $\boldsymbol{y}$,
         ylabel=Event count,
         cycle list name=black white,
-        width=\textwidth/2.7, % Make it fit to text width side by side
-        legend style={anchor=north east}
+        width=\textwidth/2.8, % Make it fit to text width side by side
+        legend style={at={(0.4,0.55)},anchor=west,draw=none}
       ]
@@
       \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
     \end{axis}
@@
         title=libatlas,
+        height=5.0cm,
         font=\footnotesize,
-        xlabel=Address offset of vector $y$,
+        xlabel=Address offset of $\boldsymbol{y}$,
         ylabel=Event count,
         cycle list name=black white,
-        width=\textwidth/2.7
+        width=\textwidth/2.8,
+        legend style={at={(0.4,0.55)},anchor=west,draw=none}
       ]
@@
       \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
     \end{axis}
@@
         title=libopenblas,
+        height=5.0cm,
         font=\footnotesize,
-        xlabel=Address offset of vector $y$,
+        xlabel=Address offset of $\boldsymbol{y}$,
         ylabel=Event count,
         cycle list name=black white,
-        width=\textwidth/2.7
+        width=\textwidth/2.8,
+        legend style={at={(0.4,0.55)},anchor=west,draw=none}
       ]
@@
       \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libopenblastable ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
     \end{axis}
   \end{tikzpicture}
-  \label{fig:blas}
-  \caption{Performance of \texttt{dgemv} for different {\small{BLAS}} packages, with varying address offset between matrix $A$ and vector $y$.}
+  \caption{\label{fig:blas}Performance of \texttt{cblas\_dgemv} for different BLAS packages, measuring cycles executed and address alias events for varying relative address offset between matrix $A$ and vector $\boldsymbol{y}$.
+Each sample point represents an increment of 16 bytes, or 0x10.}
 \end{figure}
@@
 Each library behaves differently, but in all cases we identify a significant amount of alias. 
-The diagrams can be understood the same way as for the convolution example; partial address overlap occur when the address suffix delta of $A$ and $y$ is close to zero.
-Libblas generates the most distinct pattern, with a relatively clear spike for deltas between 1 and 6 \texttt{double} elements.
-% Need to run experiment again
-Correlation between cycles executed and address aliasing events were calculated to 95\% for libblas, 78\% for libatlas and 44\% for libopenblas.
+The diagrams can be understood the same way as for the convolution example; partial address overlap occur when the address suffix delta of $A$ and $\boldsymbol{y}$ is close to zero.
+Libblas generates the most distinct pattern, with a relatively clear spike for address deltas between 0x10 and 0x60.
+% NB: Taken from analysis directory, update manually on new results
+Correlation between cycles executed and address aliasing events were calculated to 95\% for libblas, 81\% for libatlas and 45\% for libopenblas.
 
@@
 Our results show that address aliasing can indeed occur in real, performance critical applications.
-The amount of \emph{bias} this creates with respect to number of cycles executed is not as clear in all cases, but at least for the libblas case we are pretty confident in claiming that aliasing is indeed causing the performance degradation.
+The amount of \emph{bias} this creates with respect to number of cycles executed is not as clear in all cases, but at least for libblas we are confident in claiming that aliasing is indeed causing a performance degradation.
 
@@
 Strategies for detecting bias through causal analysis and randomization of experimental setups is introduced.
-The microkernel example we analyzed is taken from a followup paper by the same authors, where it is used as an example of how environment size can cause bias~\cite{Mytkowicz:2009:WrongData}.
-%Assuming the root causes of measurement bias were the same, our results indicate that address aliasing issues is probably relevant on several previous generations of Intel architectures as well.
+The microkernel we analyzed is taken from a followup paper by the same authors, where it is used as an example of how environment size can cause bias~\cite{Mytkowicz:2009:WrongData}.
 
@@
 %Address aliasing can have significant performance impact for algorithms operating on heap allocated buffers, which 
-We analyze an example with up to $2x$ variation in cycle count between different heap alignments, showing that address aliasing can have a significant performance impact in practice.
+We analyze an example with up to $2x$ variation in cycle count between different heap alignments, showing that address aliasing can have a significant performance impact.
+We are able to identify instances of alias issues in matrix-vector multiplication routines for different BLAS libraries, proving that this phenomenon also appear in real, performance critical code.
 
@@
 For heap address aliasing especially, we find that manual intervention can be required in order to achieve optimal performance.
-Finally, we look at different linear algebra libraries and find examples of alias issues in matrix-vector multiplication routines.
 Our results should inspire more clever optimizations and heuristics taking address aliasing effects into account, as the potential performance improvements can be substantial.

### 2014-08-31 paper.tex
Reference blas section in outline

@@
 Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
-Related work is discussed in Section \ref{sec:related}.
-Section \ref{sec:conclusions} includes a summary and conclusion.
+Finally, in Section \ref{sec:blas} we look at alias in BLAS libraries.
+Related work is discussed in Section \ref{sec:related}, before we summarize and conclude in Section \ref{sec:conclusions}.
 
@@
 \section{Aliasing in Real Applications}
+\label{sec:blas}
 To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.

### 2014-11-17 paper.tex
Fixed some typos.

@@
 \label{sec:aliasing}
-Modern processors are \emph{superscalar}, and achieves parallelism by issuing multiple instructions simultaneously and out of order.
+Modern processors are \emph{superscalar}, and achieve parallelism by issuing multiple instructions simultaneously and out of order.
 One of the issues that can limit throughput is dependencies between a load and previous stores.
@@
 Because the stack is aligned to 16 bytes, or 4 \emph{words}, there are a couple of different scenarios that could have been observed here.
-Static variables are fixed and covers 12 contiguous bytes (3 words), in our case the addresses end in 0x0, 0x4 and 0xc, leaving the 0x8 slot free.
+Static variables are fixed and cover 12 contiguous bytes (3 words), in our case the addresses end in 0x0, 0x4 and 0xc, leaving the 0x8 slot free.
 The two automatic variables \texttt{g} and \texttt{inc} occupies 8 contiguous bytes on the stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
@@
 Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between the stack and static data.
-The program is \emph{biased} towards environment sizes that avoids this specific stack alignment, which in principle could be triggered just by changing user name.
+The program is \emph{biased} towards environment sizes that avoid this specific stack alignment, which in principle could be triggered just by changing user name.
 This type of bias can occur in a number of memory configurations.
@@
 The Intel optimization manual mentions this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
-However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.
+However, to our knowledge there are no commonly used allocators that specifically try to mitigate aliasing in heap allocated memory.
 %Similarly none offset their allocations based on the initial stack frame.

### 2014-11-19 paper.tex
Changed to ACM Proceedings Format

@@
 % Typeset for Transactions on Architecture and Code Optimization
-\documentclass[prodmode,acmtaco]{acmsmall}
+% \documentclass[prodmode,acmtaco]{acmsmall}
+\documentclass{acm_proc_article-sp}
 \usepackage{graphicx}
@@
 
+\begin{document}
+
 %\acmVolume{V}
@@
 % Page headers
-\markboth{L. K. Melhus and R. E. Jensen}{Measurement Bias from Address Aliasing}
+%\markboth{L. K. Melhus and R. E. Jensen}{Measurement Bias from Address Aliasing}
 
 \title{Measurement Bias from Address Aliasing}
+
+\subtitle{[Extended Abstract]
+\titlenote{Just a test}}
+\numberofauthors{2}
+
+% names here
 \author{
-LARS KIRKHOLT MELHUS and RUNE ERLEND JENSEN
-  \affil{Norwegian University of Science and Technology}}
+% You can go ahead and credit any number of authors here,
+% e.g. one 'row of three' or two rows (consisting of one row of three
+% and a second row of one, two or three).
+%
+% The command \alignauthor (no curly braces needed) should
+% precede each author name, affiliation/snail-mail address and
+% e-mail address. Additionally, tag each line of
+% affiliation/address with \affaddr, and tag the
+% e-mail address with \email.
+%
+% 1st. author
+\alignauthor
+LARS KIRKHOLT MELHUS\\
+       \affaddr{Norwegian University of Science and Technology}\\
+       \affaddr{fix 1932 Wallamaloo Lane}\\
+       \affaddr{fix Wallamaloo, New Zealand}\\
+       \email{fix@corporation.com}
+% 2nd. author
+\alignauthor
+RUNE ERLEND JENSEN\\
+       \affaddr{Norwegian University of Science and Technology}\\
+       \affaddr{fix P.O. Box 1212}\\
+       \affaddr{fix Dublin, Ohio 43017-6221}\\
+       \email{runeerle@idi.ntnu.no}
+}
+
+\date{19 November 2014}
+%LARS KIRKHOLT MELHUS and RUNE ERLEND JENSEN
+%  \affil{Norwegian University of Science and Technology}
+
+\maketitle
 
@@
 % Minimum is author, year and title
-\acmformat{Lars K. Melhus and Rune E. Jensen. 2014. Measurement Bias from Address Aliasing}
-
-\begin{document}
+%\acmformat{Lars K. Melhus and Rune E. Jensen. 2014. Measurement Bias from Address Aliasing}
 
 % Acknowledgement of support, author(s) address(es)
-\begin{bottomstuff}
-New Paper, Not an Extension of a Conference Paper.
-Authors' addresses: Lars Kirkholt Melhus, Rune Erlend Jensen, Computer and Information Science Department, Norwegian University of Science and Technology, 
-Sem S{\ae}lands vei 9, Gl{\o}shaugen, 7491 Trondheim, Norway; email: larskirk@alumni.ntnu.no, runeerle@idi.ntnu.no
-\end{bottomstuff}
-
-\maketitle
+%\begin{bottomstuff}
+%New Paper, Not an Extension of a Conference Paper.
+%Authors' addresses: Lars Kirkholt Melhus, Rune Erlend Jensen, Computer and Information Science Department, Norwegian University of Science and Technology, 
+%Sem S{\ae}lands vei 9, Gl{\o}shaugen, 7491 Trondheim, Norway; email: larskirk@alumni.ntnu.no, runeerle@idi.ntnu.no
+%\end{bottomstuff}
 
@@
 \begin{table}
-  \tbl{Experimental setups used.\label{tab:machines}}{
+  \caption{Experimental setups used.\label{tab:machines}}{
   \begin{tabular}{|l|l|}
@@
 % Virtual memory/randomization illustration
-\begin{narrowfig}{0.5\textwidth}[t]
+\begin{figure}
   \centering
@@
   \label{fig:virtualmemory}
-\end{narrowfig}
+\end{figure}
 
@@
 The number of times this happens can be counted by the following event:
-\begin{describe}{{pad}}
+%\begin{describe}{{pad}}
+\begin{list}{{}}
   \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
-\end{describe}
+%\end{describe}
+\end{list}
 This event is listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}.
 Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
-\begin{describe}{{pad}}
+%\begin{describe}{{pad}}
+\begin{list}{{}}
   \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
-\end{describe}
+\end{list}
+%\end{describe}
 
@@
 \begin{table}
-  \tbl{Events with significant correlation to cycle count.\label{tab:loopcorrelation}}{
+  \caption{Events with significant correlation to cycle count.\label{tab:loopcorrelation}}{
     \pgfplotstabletypeset[
@@
   }
-  \begin{tabnote}
-    \Note{Note:}{Some correlating performance events are omitted, for example bus-cycles which will naturally vary with total cycle count.}
-  \end{tabnote}
+% TODO: make this a real note
+    {[Note:] Some correlating performance events are omitted, for example bus-cycles which will naturally vary with total cycle count.}
 \end{table}
@@
 \begin{table}[t]
-  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.\label{tab:mallocompare}}{
+  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.\label{tab:mallocompare}}{
     \pgfplotstabletypeset[
@@
   }
-  \begin{tabnote}
-    \Note{Note:}{Memory mapped addresses starting with the 0x2aa...
+  %\begin{tabnote}
+  %  \Note
+{Note:}{Memory mapped addresses starting with the 0x2aa...
 prefix is an artifact of using \emph{make} to generate results.
@@
     The difference is not important to our discussion.}
-  \end{tabnote}
+  %\end{tabnote}
 \end{table}
@@
 % Table with detailed performance data, use estimated for O2 because it does not need as many columns (!)
-\begin{table}[t]
+\begin{table*}[t]
   \renewcommand{\tabcolsep}{3pt} % Less padding between cells
-  \tbl{Relevant performance counters and correlation ($r$) with cycle count for optimization O2. Estimated cost accounting for constant overhead.\label{tab:convstats}}{
+  \caption{Relevant performance counters and correlation ($r$) with cycle count for optimization O2. Estimated cost accounting for constant overhead.\label{tab:convstats}}{
     \pgfplotstabletypeset[
@@
   }
-\end{table}
+\end{table*}
 
@@
 The functionality is divided into three categories:
-\begin{describe}{{pad}}
+%\begin{describe}{{pad}}
+\begin{list}{{}}
   \item[\emph{Level 1}.] Scalar and vector operations, such as dot product and vector addition.
@@
   \item[\emph{Level 3}.] Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
-\end{describe}
+\end{list}
+%\end{describe}
 Procedures operating on vectors, from level 1 or 2, intuitively appears the most likely to have potential for aliasing.
@@
 % Acknowledgements
-\begin{acks}
+%\begin{acks}
+\section{Acknowledgments}
 This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
-\end{acks}
+%\end{acks}
 
 % Bibliography
-\bibliographystyle{ACM-Reference-Format-Journals}
+\bibliographystyle{abbrv}
 \bibliography{references}
@@
 % History dates
-\received{}{}{}
+%\received{}{}{}
 

### 2014-11-19 paper.tex
Improved table and figure layout.

@@
 Core          & Intel\textsuperscript{\textregistered{}} Core\texttrademark{}2 Duo P8600 (laptop) \\
-              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
+              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0),\\
+              & GCC 4.8.2-19ubuntu1 \\ \hline
 Nehalem       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-950 \\
-              & 64-bit Ubuntu 12.04 LTS (Linux 3.2.0), GCC 4.8.2-19ubuntu1 \\ \hline
+              & 64-bit Ubuntu 12.04 LTS (Linux 3.2.0),\\
+              & GCC 4.8.2-19ubuntu1 \\ \hline
 % Sandy Bridge  & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-2670QM (laptop) \\
@@
 Ivy Bridge    & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i5-3470 \\
-              & 64-bit Ubuntu 12.04 LTS (Linux 3.8.0), GCC 4.8.2-19ubuntu1 \\ \hline
+              & 64-bit Ubuntu 12.04 LTS (Linux 3.8.0),\\
+              & GCC 4.8.2-19ubuntu1 \\ \hline
 Haswell       & Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K \\
-              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0), GCC 4.8.2-19ubuntu1 \\ \hline
+              & 64-bit Ubuntu 14.04 LTS (Linux 3.13.0),\\
+              & GCC 4.8.2-19ubuntu1 \\ \hline
   \end{tabular}}
@@
 \begin{figure}
-  \begin{lstlisting}[frame=single, xleftmargin=.1\textwidth, xrightmargin=.1\textwidth]
+  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
 #define ALIAS(a, b) \
@@
     frame=single,
-    xleftmargin=.1\textwidth,
-    xrightmargin=.1\textwidth
+    xleftmargin=.01\textwidth,
+    xrightmargin=.01\textwidth
   ]{bin/convolution-kernel.c}
@@
 \pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
-\begin{figure}[t]
+\begin{figure}
   \centering
@@
 In code, the function is invoked with the following parameters:
-\begin{lstlisting}
+\begin{lstlisting}[breaklines=true]
 cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
@@
 \pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
-\begin{figure}[t]
+\begin{figure}
   \centering

### 2014-11-19 paper.tex
Split motivation into separate figures

@@
 
-%\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
-\pgfplotstableread{bin/microkernel-core.dat}{\tblmicrokernelcore}
 \begin{figure}[t]
-\begin{minipage}[t]{0.40\textwidth}
-  \vspace{3mm}
-  \begin{lstlisting}[frame=tb]
+  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
 static int i, j, k;
@@
   \end{lstlisting}
-\end{minipage}
-\begin{minipage}[t]{0.55\textwidth}
-  \begin{flushleft}
+  \caption{\label{fig:microkernel}Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size.}
+\end{figure}
+
+%\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
+\pgfplotstableread{bin/microkernel-core.dat}{\tblmicrokernelcore}
+\begin{figure}[t]
   \begin{tikzpicture}
@@
   \end{tikzpicture}
-  \end{flushleft}
-\end{minipage}
-\label{fig:motivation}
-\caption{Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size.
+  \caption{\label{fig:motivation}Performance counter data running the micro-kernel from Figure~\ref{fig:microkernel} under different environment sizes.
 Notice the clear spike at around 3200 bytes, where the number of cycles executed nearly doubles.
@@
 \label{sec:microkernel}
-To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:motivation}.
+To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:microkernel}.
 This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.

### 2014-11-19 paper.tex
Replace all citeN with cite
Only plain cite is supported by new template

@@
   \end{lstlisting}
-  \caption{\label{fig:microkernel}Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size.}
+  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
 \end{figure}
@@
 
-A literature study of 88 papers by \citeN{Mytkowicz:2008:OE&MB} showed that the median reported speedup was within the margin of bias, possibly invalidating the claimed speedups. 
+A literature study of 88 papers by Mytkowicz et al. showed that the median reported speedup was within the margin of bias, possibly invalidating the claimed speedups~\cite{Mytkowicz:2008:OE&MB}.
 To alleviate the impact of measurement bias in performance analysis, researchers have proposed techniques like causal analysis, and randomization of execution contexts~\cite{Mytkowicz:2008:OE&MB}.
@@
 
-In Figure~\ref{fig:motivation}, we have reproduced a result originally presented by \citeN{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program depends on environment variables.
+In Figure~\ref{fig:motivation}, we have reproduced a result originally presented in a paper by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program depends on environment variables.
 This example provides the basis for our analysis.
@@
 The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section.
-Then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
-Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
-Finally, in Section \ref{sec:blas} we look at alias in BLAS libraries.
-Related work is discussed in Section \ref{sec:related}, before we summarize and conclude in Section \ref{sec:conclusions}.
+Then section~\ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
+Section~\ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section~\ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
+Finally, in Section~\ref{sec:blas} we look at alias in BLAS libraries.
+Related work is discussed in Section~\ref{sec:related}, before we summarize and conclude in Section~\ref{sec:conclusions}.
 
@@
 
-Reliability of PMU measurements have been evaluated by \citeN{Weaver:2008:PCTrusted}.
+Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
 The potential issues they find are either handled explicitly or are irrelevant to our setup.
@@
 \label{sec:microkernel}
-To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:microkernel}.
+To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by Mytkowicz et al.~\cite{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:microkernel}.
 This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.
@@
 On this architecture, the \emph{load\_blocks.overlap\_store} event is used to indicate address aliasing.
-Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}, where the authors used a similar setup.
+Notably, this event was also found to be significant in the original paper, where the authors used a similar setup~\cite{Mytkowicz:2009:WrongData}.
 We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and draw the conclusion that address aliasing must be the underlying explanation.
@@
 \label{sec:related}
-Measurement bias and observer effect in performance analysis has been studied in detail by Mytkowicz et al~\shortcite{Mytkowicz:2008:OE&MB,Mytkowicz:2008:Easy}.
+Measurement bias and observer effect in performance analysis has been studied in detail by Mytkowicz et al.~\cite{Mytkowicz:2008:OE&MB,Mytkowicz:2008:Easy}.
 The authors introduce environment variables and link ordering as examples of external bias triggers, showing how standardized {\small SPECint} benchmarks suffer from measurement bias when trying to evaluate the effectiveness of O3 over O2.
@@
 Many different approaches have been explored in order to take performance counter data and processor intrinsics into account for optimization.
-\citeN{Knights:2009:BlindOpt} introduce ``blind'' optimization, treating the underlying hardware as a black box and use search over \emph{variant spaces}.
+Knights et al. introduce ``blind'' optimization, treating the underlying hardware as a black box and use search over \emph{variant spaces}~\cite{Knights:2009:BlindOpt}.
 Their space of variants consist of different position and alignment for each function and global variable, in principle exploring the different configurations possible by altering link order.

### 2014-11-19 paper.tex
Remove spike 2 column from correlation table
Trim width to fit in one column, compensate by adding a sentence
explaining that the spikes were identical.

@@
 To analyze the bias effects at these points, we calculate the \emph{median} value of each counter over all contexts, and compare that to the two extreme cases.
-Going through the list of events manually, we selected the set that had the highest correlation with cycle count.
-Table \ref{tab:loopcorrelation} shows numerical counter values for these cases, compared to the median values.
+Going through the list of events manually, we selected the set that had the highest correlation with cycle count\footnote{Some correlating performance events are omitted; for example bus-cycles, which will naturally vary with total cycle count.}.
+Table \ref{tab:loopcorrelation} shows numerical counter values for the first spike, compared to the median values.
+Event counts on the second spike are virtually identical to the first, and omitted for brevity.
 
@@
 \begin{table}
-  \caption{Events with significant correlation to cycle count.\label{tab:loopcorrelation}}{
+  \caption{Events with significant correlation to cycle count.\label{tab:loopcorrelation}}
   \pgfplotstabletypeset[
@@
     col sep=comma,
-      columns={Performance counter, Median, [index]3, [index]4},
+    columns={Performance counter, Median, [index]3}, %, [index]4
     column type=r,
@@
       before row={\hline
-          Performance counter & Median & Spike 1 & Spike 2 \\
+        Performance counter & Median & Spike \\ % & Spike 2
       },
@@
   ]{bin/microkernel-comparison-haswell.csv}
-  }
-% TODO: make this a real note
-    {[Note:] Some correlating performance events are omitted, for example bus-cycles which will naturally vary with total cycle count.}
 \end{table}

### 2014-11-19 paper.tex
Remove column from allocator table
Done in order to fit in a single column, without removing too
much useful information. Change result file to yield slightly
larger first column, which fits better with the layout.

@@
 \begin{table}[t]
-  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.\label{tab:mallocompare}}{
+  \centering
+  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.\label{tab:mallocompare}}
   \pgfplotstabletypeset[
@@
     string type,
-      columns={Allocation, 64, 5120, 1048576},
+    columns={Allocation, 5120, 1048576},
     column type=r,
@@
       before row={\hline
-           & 64 B & 5,120 B & 1,048,576 B \\
+         & 5,120 B & 1,048,576 B \\
         },
@@
   ]{bin/malloc-comparison.csv}
-  }
-  %\begin{tabnote}
-  %  \Note
-{Note:}{Memory mapped addresses starting with the 0x2aa...
+\end{table}
+
+Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers.\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}\footnote{Memory mapped addresses starting with the 0x2aa...
 prefix is an artifact of using \emph{make} to generate results.
@@
 prefixes.
-    The difference is not important to our discussion.}
-  %\end{tabnote}
-\end{table}
-
-Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers.\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
+This difference is not important to our discussion.}
 We observe the addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} for different size parameters.

### 2014-11-19 paper.tex
Center and adjust large table
Use description environment instead of list

@@
 Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
-The number of times this happens can be counted by the following event:
-%\begin{describe}{{pad}}
-\begin{list}{{}}
+The number of times this happens can be counted by the following event:\vspace{-2mm}
+\begin{description}
   \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
-%\end{describe}
-\end{list}
+\end{description}
+\vspace{-2mm}
 This event is listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}.
-Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
-%\begin{describe}{{pad}}
-\begin{list}{{}}
+Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:\vspace{-2mm}
+\begin{description}
   \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
-\end{list}
-%\end{describe}
+\end{description}\vspace{-2mm}
 
@@
 \begin{table*}[t]
-  \renewcommand{\tabcolsep}{3pt} % Less padding between cells
+  \centering
   \caption{Relevant performance counters and correlation ($r$) with cycle count for optimization O2. Estimated cost accounting for constant overhead.\label{tab:convstats}}{
@@
 The functionality is divided into three categories:
-%\begin{describe}{{pad}}
-\begin{list}{{}}
-  \item[\emph{Level 1}.] Scalar and vector operations, such as dot product and vector addition.
-  \item[\emph{Level 2}.] Matrix-vector operations, such as gemv for general matrix-vector multiplication.
-  \item[\emph{Level 3}.] Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
-\end{list}
-%\end{describe}
+\begin{description}
+  \item[Level 1] Scalar and vector operations, such as dot product and vector addition.
+  \item[Level 2] Matrix-vector operations, such as gemv for general matrix-vector multiplication.
+  \item[Level 3] Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
+\end{description}
 Procedures operating on vectors, from level 1 or 2, intuitively appears the most likely to have potential for aliasing.

### 2014-11-19 paper.tex
Adjust figure size and positions

@@
         title = Core,
-        width = 7.8cm,
+        width = \textwidth/2.1,
         height = 4.5cm,
@@
         title = Nehalem (HT),
-        width = \textwidth/2.0,
+        width = \textwidth/2.1,
         height = 4.5cm,
@@
         title = Ivy Bridge,
-        width = \textwidth/2.0,
+        width = \textwidth/2.1,
         height = 4.5cm,
@@
 
+\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
+\pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
+\begin{figure*}[t]
+  \centering
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=Haswell -O2,
+        font=\footnotesize,
+        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
+        ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/2.05, % Make it fit to text width side by side
+        height=6.0cm,
+        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+  \hspace{0.5cm}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=Haswell -O3,
+        font=\footnotesize,
+        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
+        ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/2.05,
+        height=6.0cm,
+        skip coords between index={20}{32}
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethreehaswell ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethreehaswell ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+  \hspace{0.5cm}
+  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}.
+Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
+Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
+\end{figure*}
+
 Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between the heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
@@
 \end{itemize}
-
-\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
-\pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \begin{axis}[
-        title=Haswell -O2,
-        font=\footnotesize,
-        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
-        ylabel=Event count,
-        cycle list name=black white,
-        width=\textwidth/1.9, % Make it fit to text width side by side
-        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
-      ]
-      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
-      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
-      \addlegendentry{Cycles} ;
-      \addlegendentry{Alias} ;
-    \end{axis}
-  \end{tikzpicture}
-  \begin{tikzpicture}
-    \begin{axis}[
-        title=Haswell -O3,
-        font=\footnotesize,
-        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
-        ylabel=Event count,
-        cycle list name=black white,
-        width=\textwidth/1.9,
-        skip coords between index={20}{32}
-      ]
-      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethreehaswell ;
-      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethreehaswell ;
-      \addlegendentry{Cycles} ;
-      \addlegendentry{Alias} ;
-    \end{axis}
-  \end{tikzpicture}
-  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}.
-Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
-Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
-\end{figure}
 
@@
 \pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
-\begin{figure}
+\begin{figure*}[t]
   \centering
@@
 Each sample point represents an increment of 16 bytes, or 0x10.}
-\end{figure}
+\end{figure*}
 

### 2014-11-20 paper.tex
Use normal font size for code listings
Adjust borders around program listings.

@@
 
-\lstset{language=C, basicstyle=\ttfamily\small, columns=flexible}
+\lstset{language=C, basicstyle=\ttfamily, columns=flexible}
 
@@
 \begin{figure}
-  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
+  \begin{lstlisting}[frame=single, xleftmargin=.005\textwidth, xrightmargin=.005\textwidth]
 #define ALIAS(a, b) \
@@
     frame=single,
-    xleftmargin=.01\textwidth,
-    xrightmargin=.01\textwidth
+    xleftmargin=.00\textwidth,
+    xrightmargin=.00\textwidth
   ]{bin/convolution-kernel.c}

### 2014-11-20 paper.tex
Updated author data. Changed to utf8.
Update to ACM latex Housekeeping standard
Removed sub-headline.
Swaped first figure
Aligned number presision in graph

@@
-% Typeset for Transactions on Architecture and Code Optimization
-% \documentclass[prodmode,acmtaco]{acmsmall}
+% paper.tex
+% ACM sigmetric format
+% Author: L. K. Melhus and R. E. Jensen
 \documentclass{acm_proc_article-sp}
@@
 \usepackage{lmodern} % Much clearer text
+\usepackage[utf8]{inputenc}
 
@@
 
+\conferenceinfo{SIGMETRICS}{2015 Portland, Oregon USA}
+\CopyrightYear{2015}
 %\acmVolume{V}
@@
 
-\subtitle{[Extended Abstract]
-\titlenote{Just a test}}
 \numberofauthors{2}
@@
 \alignauthor
-LARS KIRKHOLT MELHUS\\
+Lars Kirkholt Melhus\\
        \affaddr{Norwegian University of Science and Technology}\\
-       \affaddr{fix 1932 Wallamaloo Lane}\\
-       \affaddr{fix Wallamaloo, New Zealand}\\
-       \email{fix@corporation.com}
+       \affaddr{Sem Sælands vei 9}\\
+       \affaddr{7491 Trondheim, Norway}\\
+       \email{lars.kirkholt.melhus@gmail.com}
 % 2nd. author
 \alignauthor
-RUNE ERLEND JENSEN\\
+Rune Erlend Jensen\\
        \affaddr{Norwegian University of Science and Technology}\\
-       \affaddr{fix P.O. Box 1212}\\
-       \affaddr{fix Dublin, Ohio 43017-6221}\\
+       \affaddr{Sem Sælands vei 9}\\
+       \affaddr{7491 Trondheim, Norway}\\
        \email{runeerle@idi.ntnu.no}
@@
 \date{19 November 2014}
-%LARS KIRKHOLT MELHUS and RUNE ERLEND JENSEN
-%  \affil{Norwegian University of Science and Technology}
 
@@
 \category{C.4}{Performance of Systems}{Measurement techniques; Performance attributes}
+\category{D.3.4}{Processors}{Optimization; Run-time environments}
 
@@
 \terms{Measurement, Performance}
-
 % Free to choose keywords, but must be in alphabetical order
-\keywords{4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation}
+\keywords{4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}
 
@@
 Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.
-
-\begin{figure}[t]
-  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
-static int i, j, k;
-int main() {
-    int g = 0, inc = 1;
-    for (; g < 65536; g++) {
-        i += inc;
-        j += inc;
-        k += inc;
-    }
-    return 0;
-}
-  \end{lstlisting}
-  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
-\end{figure}
 
@@
 Measuring performance in this configuration would give a biased result.}
+\end{figure}
+
+\begin{figure}[t]
+  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
+static int i, j, k;
+int main() {
+    int g = 0, inc = 1;
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    return 0;
+}
+  \end{lstlisting}
+  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
 \end{figure}
@@
         title = Nehalem (HT),
+        y tick label style={
+        /pgf/number format/.cd,
+            fixed,
+            fixed zerofill,
+            precision=1,
+        /tikz/.cd
+        },
         width = \textwidth/2.1,
@@
         title = Ivy Bridge,
+        y tick label style={
+        /pgf/number format/.cd,
+            fixed,
+            fixed zerofill,
+            precision=1,
+        /tikz/.cd
+        },
         width = \textwidth/2.1,

### 2014-11-20 paper.tex
Made it fully Anonymous

@@
 % 1st. author
+\iffalse
 \alignauthor
@@
        \email{runeerle@idi.ntnu.no}
+\fi
+\alignauthor
+Anonymous\\
+       \affaddr{Anonymous University}\\
+       \affaddr{Anonymous}\\
+       \affaddr{Anonymous, Anonymous}\\
+       \email{anonymous@anonymous.com}
+% 2nd. author
+\alignauthor
+Anonymous\\
+       \affaddr{Anonymous University }\\
+       \affaddr{Anonymous}\\
+       \affaddr{Anonymous, Anonymous}\\
+       \email{anonymous@anonymous.com}
 }
@@
 \section{Acknowledgments}
+\iffalse
 This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
+\fi
+This work is ... under the ..., .
+ under the supervision of ..., looking at sources of measurement bias ... ~\cite{MasterThesis}.
 %\end{acks}

### 2014-11-20 paper.tex
Added PDF meta data
Replace paragraph with subsubsection
Paragraph did not render very well, and subsubsection is the correct
next level in the hierarchy

@@
 \usepackage[utf8]{inputenc}
+\usepackage{hyperref} % For PDF metadata
 
@@
 \begin{document}
+
+\hypersetup
+{
+%    pdfauthor={L. K. Melhus and R. E. Jensen},
+    pdfauthor={Anonymous and Anonymous},
+    pdfsubject={Identification of an underlying mechanism that causes measurement bias},
+    pdftitle={Measurement Bias from Address Aliasing},
+    pdfkeywords={4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}
+}
 
@@
 
-\paragraph{Note on Hyper-Threading}
+\subsubsection{Note on Hyper-Threading}
 As described in Section~\ref{sec:methodology}, Hyper-Threading was disabled by default in order to avoid any potential interference while collecting data.
@@
 
-\paragraph{Mark buffers with \texttt{restrict}}
+\subsubsection{Mark buffers with \texttt{restrict}}
 In our convolution kernel implementation, the compiler has to account for the fact that input and output pointers might alias, or that the buffers partially overlap.
@@
 
-\paragraph{Use a special purpose allocator}
+\subsubsection{Use a special purpose allocator}
 The Intel optimization manual mentions this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
@@
 
-\paragraph{Manually adjust address offsets}
+\subsubsection{Manually adjust address offsets}
 In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.

### 2014-11-20 paper.tex
Indent code snippet
Should be the same for all inline code
Make lists more compact

@@
 \usepackage{hyperref} % For PDF metadata
+\usepackage{enumitem} % Controlling list styling
 
@@
 \lstset{language=C, basicstyle=\ttfamily, columns=flexible}
+
+\setlist{topsep=0pt}
 
@@
 Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
-The number of times this happens can be counted by the following event:\vspace{-2mm}
+The number of times this happens can be counted by the following event:
 \begin{description}
@@
 \end{description}
-\vspace{-2mm}
 This event is listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}.
-Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:\vspace{-2mm}
+Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
 \begin{description}
   \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
-\end{description}\vspace{-2mm}
-
+\end{description}
 As address aliasing depends on the memory addresses of loads and stores, any environmental factor that affects memory layout has the potential to induce aliasing conditions.

### 2014-11-24 paper.tex
Reduce whitespace below listings
Improved layout of some section headlines.

@@
 
-\lstset{language=C, basicstyle=\ttfamily, columns=flexible}
+\lstset{language=C, basicstyle=\ttfamily, columns=flexible, belowskip=0pt}
 
@@
 
-In Figure~\ref{fig:motivation}, we have reproduced a result originally presented in a paper by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program depends on environment variables.
+In Figure~\ref{fig:motivation}, we have reproduced a result originally presented in a paper by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program (Figure~\ref{fig:microkernel}) depends on environment variables.
 This example provides the basis for our analysis.
@@
 
+\newpage
 \subsection{Avoiding Aliasing}
@@
 
+\newpage
 \section{Aliasing in Real Applications}
@@
 \pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
-\begin{figure*}[t]
+\begin{figure*}
   \centering
@@
 
+\newpage
 \section{Conclusions}

### 2016-01-07 paper.tex
Update for iWAPT2016. Basic changes for compiling only.

@@
 % Author: L. K. Melhus and R. E. Jensen
-\documentclass{acm_proc_article-sp}
+%\documentclass{acm_proc_article-sp}
+\documentclass[10pt, conference, compsocconf]{IEEEtran}
 \usepackage{graphicx}
@@
 \usepackage{hyperref} % For PDF metadata
-\usepackage{enumitem} % Controlling list styling
+%\usepackage{enumitem} % Controlling list styling
 
@@
 
-\setlist{topsep=0pt}
+%\setlist{topsep=0pt}
 
@@
 
-\conferenceinfo{SIGMETRICS}{2015 Portland, Oregon USA}
-\CopyrightYear{2015}
+%\conferenceinfo{SIGMETRICS}{2015 Portland, Oregon USA}
+%\CopyrightYear{2015}
 %\acmVolume{V}
@@
 
-\numberofauthors{2}
-
+%\numberofauthors{2}
+
+\iffalse
 % names here
@@
 % 1st. author
-\iffalse
+
 \alignauthor
@@
        \email{runeerle@idi.ntnu.no}
-\fi
+
 \alignauthor
@@
 }
+\fi
+
+\author{\IEEEauthorblockN{Authors Name/s per 1st Affiliation (Author)}
+\IEEEauthorblockA{line 1 (of Affiliation): dept. name of organization\\
+line 2: name of organization, acronyms acceptable\\
+line 3: City, Country\\
+line 4: Email: name@xyz.com}
+\and
+\IEEEauthorblockN{Authors Name/s per 2nd Affiliation (Author)}
+\IEEEauthorblockA{line 1 (of Affiliation): dept. name of organization\\
+line 2: name of organization, acronyms acceptable\\
+line 3: City, Country\\
+line 4: Email: name@xyz.com}
+}
 
@@
 % Choose from classification tree http://www.acm.org/about/class/ccs98-html
-\category{C.4}{Performance of Systems}{Measurement techniques; Performance attributes}
-\category{D.3.4}{Processors}{Optimization; Run-time environments}
+%\category{C.4}{Performance of Systems}{Measurement techniques; Performance attributes}
+%\category{D.3.4}{Processors}{Optimization; Run-time environments}
 
 % Terms come from a fixed list of predefined terms
-\terms{Measurement, Performance}
+%\terms{Measurement, Performance}
 % Free to choose keywords, but must be in alphabetical order
-\keywords{4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}
+\IEEEkeywords{4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}
 
@@
 
-\newpage
+%\newpage
 \subsection{Avoiding Aliasing}
@@
 
-\newpage
+%\newpage
 \section{Aliasing in Real Applications}
@@
 
-\newpage
+%\newpage
 \section{Conclusions}

### 2016-01-07 analysis/heap-alias/conv.c
Compacted the code a bit for the paper.

@@ -1,12 +1,10 @@
-static float k[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
-
-void conv(int n, const float *input, float *output)
+static float k[5]= {0.1, 0.25, 0.3, 0.25, 0.1};
+void conv(int n, const float *in, float *out)
 {
     int i, j;
-    for (i = 2; i < n - 2; ++i)
-    {
-        output[i] = 0;
+    for (i = 2; i < n - 2; ++i) {
+        out[i] = 0;
         for (j = 0; j < 5; ++j)
-            output[i] += input[i-2+j] * k[j];
+            out[i] += in[i-2+j] * k[j];
     }
 }

### 2016-01-07 paper.tex
Fixed overlapped text.

@@
 \begin{description}
-  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
+  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \hfill \\ \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
 \end{description}
@@
 \begin{description}
-  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
+  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \hfill \\ \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
 \end{description}

### 2016-01-07 analysis/heap-alias/conv.c
Compacted the code a bit more for the paper.

@@ -1,10 +1,8 @@
 static float k[5]= {0.1, 0.25, 0.3, 0.25, 0.1};
-void conv(int n, const float *in, float *out)
-{
-    int i, j;
-    for (i = 2; i < n - 2; ++i) {
+void conv(int n, const float *in, float *out) {
+    for (int i = 2; i < n - 2; ++i) {
         out[i] = 0;
-        for (j = 0; j < 5; ++j)
+        for (int j = 0; j < 5; ++j)
             out[i] += in[i-2+j] * k[j];
     }
 }

### 2016-01-07 paper.tex
Fixed IEEE keyword issue.
Reduced the spacing of caption text.

@@
 %\usepackage{enumitem} % Controlling list styling
+\usepackage[font=small,skip=0pt]{caption}
+%http://tex.stackexchange.com/questions/94016/how-to-reduce-space-between-image-and-its-caption
 
@@
 % Free to choose keywords, but must be in alphabetical order
-\IEEEkeywords{4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}
+\begin{IEEEkeywords}
+4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters
+\end{IEEEkeywords}
 
@@
 \begin{lstlisting}[breaklines=true]
-    mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
+    mmap(NULL, (n+d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) +d;
 \end{lstlisting}

### 2016-01-07 paper.tex
Tuned the spacing of figures/tables.
Fixed overlap in BLAS item list.

@@
 %\usepackage{enumitem} % Controlling list styling
-\usepackage[font=small,skip=0pt]{caption}
+%\usepackage[font=small,skip=0pt]{caption}
 %http://tex.stackexchange.com/questions/94016/how-to-reduce-space-between-image-and-its-caption
+\usepackage[font=small,belowskip=-10pt,aboveskip=0pt]{caption}
+
+%\setlength{\intextsep}{10pt plus 2pt minus 2pt}
+
+%\setlength{\belowcaptionskip}{-5pt}
+% too much?
+
+%\captionsetup[table]{aboveskip=-25pt}
+
+%\setlength{\dblfloatsep}{-15pt}
+
+%\addtolength{\belowcaptionskip}{15pt}
 
@@
 \begin{description}
-  \item[Level 1] Scalar and vector operations, such as dot product and vector addition.
-  \item[Level 2] Matrix-vector operations, such as gemv for general matrix-vector multiplication.
-  \item[Level 3] Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
+  \item[Level 1] \hspace{0.1cm} Scalar and vector operations, such as dot product and vector addition.
+  \item[Level 2] \hspace{0.1cm} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
+  \item[Level 3] \hspace{0.1cm} Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
 \end{description}

### 2016-01-07 paper.tex
Make code compacting change consistent

@@
 Overall, it seems reasonable to conclude that the resource stalling is causing the slowdown, ultimately generated by false dependencies from address aliasing.
-We do not attempt to pinpoint exactly which accesses generates these conflicts, but at a high level the CPU falsely assumes dependencies between \texttt{input[i]} and \texttt{output[i]}.
+We do not attempt to pinpoint exactly which accesses generates these conflicts, but at a high level the CPU falsely assumes dependencies between \texttt{in[i]} and \texttt{out[i]}.
 By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
@@
 \begin{lstlisting}[breaklines=true]
-    void conv(int n, const float * restrict input, float * restrict output);
+    void conv(int n, const float * restrict in, float * restrict out);
 \end{lstlisting}

### 2016-01-07 analysis/heap-alias/conv.c

@@ -1,8 +1,9 @@
 static float k[5]= {0.1, 0.25, 0.3, 0.25, 0.1};
 void conv(int n, const float *in, float *out) {
-    for (int i = 2; i < n - 2; ++i) {
+    int i, j;
+    for (i = 2; i < n - 2; ++i) {
         out[i] = 0;
-        for (int j = 0; j < 5; ++j)
+        for (j = 0; j < 5; ++j)
             out[i] += in[i-2+j] * k[j];
     }
 }

### 2016-01-07 paper.tex
Comment out bloated sections
Length limit is now 10 pages.

@@
 
-Because the stack is aligned to 16 bytes, or 4 \emph{words}, there are a couple of different scenarios that could have been observed here.
-Static variables are fixed and cover 12 contiguous bytes (3 words), in our case the addresses end in 0x0, 0x4 and 0xc, leaving the 0x8 slot free.
-The two automatic variables \texttt{g} and \texttt{inc} occupies 8 contiguous bytes on the stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
-In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.
-A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots.
-While this will give significantly more alias counts, we found that it had little effect on the total number of cycles executed.
-
-In conclusion, we identified that address aliasing is the root cause of measurement bias from environment size for this program.
-Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between the stack and static data.
-The program is \emph{biased} towards environment sizes that avoid this specific stack alignment, which in principle could be triggered just by changing user name.
-This type of bias can occur in a number of memory configurations.
-Here, the aliasing was caused by interactions between the stack and static variables, but we can imagine similar conflicts between the stack and the heap, depending also on how dynamic memory allocators behave.
-Because the stack is used for local variables, complex code is likely to contain many potential conflicts with other areas of memory.
+% Because the stack is aligned to 16 bytes, or 4 \emph{words}, there are a couple of different scenarios that could have been observed here.
+% Static variables are fixed and cover 12 contiguous bytes (3 words), in our case the addresses end in 0x0, 0x4 and 0xc, leaving the 0x8 slot free.
+% The two automatic variables \texttt{g} and \texttt{inc} occupies 8 contiguous bytes on the stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots.
+% In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.
+% A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots.
+% While this will give significantly more alias counts, we found that it had little effect on the total number of cycles executed.
+
+% In conclusion, we identified that address aliasing is the root cause of measurement bias from environment size for this program.
+% Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between the stack and static data.
+% The program is \emph{biased} towards environment sizes that avoid this specific stack alignment, which in principle could be triggered just by changing user name.
+% This type of bias can occur in a number of memory configurations.
+% Here, the aliasing was caused by interactions between the stack and static variables, but we can imagine similar conflicts between the stack and the heap, depending also on how dynamic memory allocators behave.
+% Because the stack is used for local variables, complex code is likely to contain many potential conflicts with other areas of memory.
 
@@
 
-\subsubsection{Note on Hyper-Threading}
-As described in Section~\ref{sec:methodology}, Hyper-Threading was disabled by default in order to avoid any potential interference while collecting data.
-However, we found that configuration of Hyper-Threading either way did not change the alias behavior.
-As an example of this, we show the results for ``Nehalem'' with HT enabled in Figure~\ref{fig:microkernel-all-archs}.
+% \subsubsection{Note on Hyper-Threading}
+% As described in Section~\ref{sec:methodology}, Hyper-Threading was disabled by default in order to avoid any potential interference while collecting data.
+% However, we found that configuration of Hyper-Threading either way did not change the alias behavior.
+% As an example of this, we show the results for ``Nehalem'' with HT enabled in Figure~\ref{fig:microkernel-all-archs}.
 
@@
 The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer does not alias with any other, allowing for more efficient code generation with fewer memory accesses.
-\begin{lstlisting}[breaklines=true]
-    void conv(int n, const float * restrict in, float * restrict out);
-\end{lstlisting}
-With this updated function prototype, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% on optimization level O2.
+% \begin{lstlisting}[breaklines=true]
+%     void conv(int n, const float * restrict in, float * restrict out);
+% \end{lstlisting}
+Applied to our example, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% on optimization level O2.
 Alias events is reduced by about 10 million for the default alignment, with a corresponding improvement in cycle count. % File backing this in results
@@
 In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.
-This can be achieved by exploiting the fact the \texttt{mmap} is in fact guaranteed to be placed at a page boundary, and use that directly instead of \texttt{malloc}.
-The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.
-\begin{lstlisting}[breaklines=true]
-    mmap(NULL, (n+d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) +d;
-\end{lstlisting}
-The same pointer difference must of course by subtracted before unmapping the memory.
+This can be achieved by exploiting the fact the \texttt{mmap} is in fact guaranteed to be placed at a page boundary, and use that directly instead of \texttt{malloc} to create specific alignments.
+% The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.
+% \begin{lstlisting}[breaklines=true]
+%     mmap(NULL, (n+d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) +d;
+% \end{lstlisting}
+% The same pointer difference must of course by subtracted before unmapping the memory.
 

### 2016-01-09 paper.tex
Decrease listing font size
Reduce size of abstract

@@
 
-\lstset{language=C, basicstyle=\ttfamily, columns=flexible, belowskip=0pt}
+\lstset{language=C, basicstyle=\ttfamily\small, columns=flexible, belowskip=0pt}
 
@@
 
-\author{\IEEEauthorblockN{Authors Name/s per 1st Affiliation (Author)}
-\IEEEauthorblockA{line 1 (of Affiliation): dept. name of organization\\
-line 2: name of organization, acronyms acceptable\\
-line 3: City, Country\\
-line 4: Email: name@xyz.com}
-\and
-\IEEEauthorblockN{Authors Name/s per 2nd Affiliation (Author)}
-\IEEEauthorblockA{line 1 (of Affiliation): dept. name of organization\\
-line 2: name of organization, acronyms acceptable\\
-line 3: City, Country\\
-line 4: Email: name@xyz.com}
+\author{\IEEEauthorblockN{Lars Kirkholt Melhus, Rune Erlend Jensen}
+\IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
+NTNU, Norway\\
+lars.kirkholt.melhus@gmail.com, runeerle@idi.ntnu.no}
 }
@@
 \begin{abstract} % Should be 150 -- 200 words max
-Understanding program behavior and obtaining accurate measurements is important in performance analysis.
-However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
-Seemingly irrelevant properties---such as the length of your user name---can impact program performance.
-In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
-Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
-We find that \emph{address aliasing} can explain bias from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
-
-This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
-We demonstrate this by implementing runtime detection and correction of aliasing conditions.
-For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
-The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in one of our examples.
-Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
+%Understanding program behavior and obtaining accurate measurements is important in performance analysis.
+%However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
+%Seemingly irrelevant properties---such as the length of your user name---can impact program performance.
+In this paper, we identify \emph{address aliasing} as an important underlying mechanism causing measurement bias on modern Intel microarchitectures.
+By analyzing hardware performance counters, we show how bias can occur from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
+%Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
+%We find that \emph{address aliasing} can explain bias from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
+
+%This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
+We demonstrate ways to deal with this type of bias, through runtime detection and correction of aliasing conditions.
+This is especially important for heap allocations, as most allocators tend to produce worst case alignment by default; favoring page alignment.
+%For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
+The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in already optimized code.
+%Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
 Our results can enable new architecture specific optimization strategies to account for this phenomenon.
@@
 Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
-The following outlines our data and analysis of this program conducted on the ``Haswell'' configuration.
-Similar experiments were conducted on the other experimental setups listed in Table~\ref{tab:machines} as well, all arriving at the same conclusion.
+The following data and analysis of this program is from the ``Haswell'' configuration.
+Similar experiments were conducted on the other experimental setups listed in Table~\ref{tab:machines}, all arriving at the same conclusion.
 
@@
 If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
-Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.
+Calling the function recursively will effectively allocate a new set of variables further down the stack, and the alias condition is avoided.
 
@@
 These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
-Our assumption is that this is a different phenomenon, not related to 4K aliasing.
-We were only able to observe the effect on this particular architecture, and did not investigate the root cause further.
+%Our assumption is that this is a different phenomenon, not related to 4K aliasing.
+We were only able to observe the effect on this particular architecture, and did not investigate the root cause.% is related to 4K aliasing.
 
@@
 The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
-This phenomenon is only observed for address separations close to zero, thus we only show the first 20 data points.
+This effect is only observed for address separations close to zero, thus we only show the first 20 data points.
 If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
@@
 By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
-This is a speedup on top of already aggressive compiler optimization.
+%This is a speedup on top of already aggressive compiler optimization.
 % Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.
@@
 The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
-We identify some mitigation or optimization techniques that can be used:
+We identify some relevant optimization techniques:
 
@@
 % \end{lstlisting}
-Applied to our example, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% on optimization level O2.
-Alias events is reduced by about 10 million for the default alignment, with a corresponding improvement in cycle count. % File backing this in results
+Applied to our example, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% and alias events by approximately 10 million on optimization level O2.
+%Alias events are reduced by about 10 million for the default alignment, with a corresponding improvement in cycle count. % File backing this in results
 
@@
 The diagrams can be understood the same way as for the convolution example; partial address overlap occur when the address suffix delta of $A$ and $\boldsymbol{y}$ is close to zero.
-Libblas generates the most distinct pattern, with a relatively clear spike for address deltas between 0x10 and 0x60.
+Libblas generates the most distinct pattern, with a clear spike for address deltas between 0x10 and 0x60.
 % NB: Taken from analysis directory, update manually on new results
-Correlation between cycles executed and address aliasing events were calculated to 95\% for libblas, 81\% for libatlas and 45\% for libopenblas.
+Correlation between cycles executed and address aliasing events were 95\% for libblas, 81\% for libatlas and 45\% for libopenblas.
 
@@
 We have shown how address aliasing affects program performance under different memory contexts, and how it can explain certain cases of measurement bias.
-The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
-This phenomenon is likely present on most consumer hardware based on Intel platforms today, ranging from ``Core'' to ``Haswell'' microarchitectures.
+%The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
+Speculative and out-of-order memory operations handling only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
+This effect is observed on Intel platforms, ranging from ``Core'' to ``Haswell'' microarchitectures.
 
@@
 We analyze an example with up to $2x$ variation in cycle count between different heap alignments, showing that address aliasing can have a significant performance impact.
-We are able to identify instances of alias issues in matrix-vector multiplication routines for different BLAS libraries, proving that this phenomenon also appear in real, performance critical code.
+We are able to identify instances of alias issues in matrix-vector multiplication routines for different BLAS libraries, showing that this effect also appears in real, performance critical code.
 

### 2016-01-09 paper.tex
Shortend the end.

@@
 Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
-Our results can enable new architecture specific optimization strategies to account for this phenomenon.
+Our results can enable new architecture specific optimization strategies to account for this effect.
 \end{abstract}

### 2016-01-09 paper.tex
Trim down introduction

--- 2016-01-09 paper.tex
@@
 
-Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture.
-We later reproduced similar results for the ``Core'', ``Nehalem'', and ``Ivy Bridge'' microarchitectures as well.
-Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.
+%Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture.
+%We later reproduced similar results for the ``Core'', ``Nehalem'', and ``Ivy Bridge'' microarchitectures as well.
+%Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.
 
@@
 By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
-All experiments are done on a dedicated machine under minimal load, with \emph{frequency scaling} disabled to keep the CPU's clock speed fixed.
-Finally, we disable \emph{Hyper-threading} (HT), which reduces the possibility for resource contention between threads.
-Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.
+All experiments are done with the machine under minimal load, disabling \emph{frequency scaling} to keep the CPU's clock speed fixed, and \emph{Hyper-threading} (HT) to reduce the possibility for resource contention between threads.\footnote{Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.}
 

### 2016-01-09 paper.tex
Shortened the end.
Remove redundant stuff from address space caption

@@
   \end{tikzpicture}
-  \caption{Memory execution context, assuming a 64-bit process running on a Linux system.
-  Initial addresses of stack, heap and memory mapped files are often randomized for security reasons.
-  The stack is also offset by environment variables and program arguments.
-  Addresses of code and statically allocated data are allocated at compile time by the linker, and can be determined by inspecting the executable.}
+  \caption{Memory execution context, assuming a 64-bit process running on a Linux system.}
+  %Initial addresses of stack, heap and memory mapped files are often randomized for security reasons.
+  %The stack is also offset by environment variables and program arguments.
+  %Addresses of code and statically allocated data are allocated at compile time by the linker, and can be determined by inspecting the executable.}
   \label{fig:virtualmemory}

### 2016-01-09 paper.tex
Remove some redundant stuff in analysis

@@
 Within a span of 4096 bytes there are thus 256 possible initial stack addresses, each representing a different execution context with respect to address aliasing. 
-Note that there is no clear relationship between environment size and stack location with ASLR enabled.
-However, there will still be as many execution contexts with respect to aliasing (considering the stack only), making any occurrences of measurement bias indeed random.
+%Note that there is no clear relationship between environment size and stack location with ASLR enabled.
+%However, there will still be as many execution contexts with respect to aliasing (considering the stack only), making any occurrences of measurement bias indeed random.
 
@@
 Every 16 byte increment of environment size is measured, covering two 4K periods of initial stack addresses.
-A finer sampling is not necessary, because the stack is by default aligned to 16 bytes.
+%A finer sampling is not necessary, because the stack is by default aligned to 16 bytes.
 The microkernel is compiled with {\small{GCC}} using no optimization; any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
@@
 
-Performance counter data clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
-For that we need to know the addresses of each variable at runtime.
-The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
-Memory layout of static data is decided at compile time, and we can find the location of these variables by looking at the symbol table in the ELF executable.
-We find the addresses to be \texttt{\&i} = 0x60103c, \texttt{\&j} = 0x601040, and \texttt{\&k} = 0x601044.\footnote{ELF symbol tables can be read using \texttt{readelf -s}}
-
-Observing addresses of stack allocated data at runtime is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
-Assembly code was injected to calculate and output the addresses of \texttt{g} and \texttt{inc}.
-We made no change to the stack allocation instructions, and the code offset did not affect the addresses of static variables.
-The same experiment was run again, showing that the modified program had the exact same bias to environment size.
+Performance counter data clearly points to address aliasing as a plausible explanation, so the next step is to see exactly which memory accesses are aliasing.
+%For that we need to know the addresses of each variable at runtime.
+%The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
+Memory layout of static data is decided at compile time, and we find the addresses to be \texttt{\&i} = 0x60103c, \texttt{\&j} = 0x601040, and \texttt{\&k} = 0x601044 by reading the ELF symbol table.%\footnote{ELF symbol tables can be read using \texttt{readelf -s}}
+
+%Observing addresses of stack allocated data at runtime is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
+Assembly code was injected to calculate and output the runtime addresses of \texttt{g} and \texttt{inc}, carefully craftet to not introduce any observer effects.
+%We made no change to the stack allocation instructions, and the code offset did not affect the addresses of static variables.
+%The same experiment was run again, showing that the modified program had the exact same bias to environment size.
 From the output at each spike, we found the addresses to be \texttt{\&g} = 0x7fffffffe038, \texttt{\&inc} = 0x7fffffffe03c, and \texttt{\&g} = 0x7fffffffd038, \texttt{\&inc} = 0x7fffffffd03c, respectively

### 2016-01-09 paper.tex
Many small fixes.

--- 2016-01-09 paper.tex
@@
 \begin{abstract} % Should be 150 -- 200 words max
-%Understanding program behavior and obtaining accurate measurements is important in performance analysis.
-%However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
-%Seemingly irrelevant properties---such as the length of your user name---can impact program performance.
-In this paper, we identify \emph{address aliasing} as an important underlying mechanism causing measurement bias on modern Intel microarchitectures.
-By analyzing hardware performance counters, we show how bias can occur from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
-%Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
-%We find that \emph{address aliasing} can explain bias from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
-
-%This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
-We demonstrate ways to deal with this type of bias, through runtime detection and correction of aliasing conditions.
-This is especially important for heap allocations, as most allocators tend to produce worst case alignment by default; favoring page alignment.
-%For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
-The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in already optimized code.
-%Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
-Our results can enable new architecture specific optimization strategies to account for this phenomenon.
+Understanding program behavior and obtaining accurate measurements is important in performance analysis.
+However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
+Seemingly irrelevant properties---such as the length of your user name---can impact program performance.
+In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
+Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
+We find that \emph{address aliasing} can explain bias from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
+
+This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
+We demonstrate this by implementing runtime detection and correction of aliasing conditions.
+For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
+The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in one of our examples.
+Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
+Our results can enable new architecture specific optimization strategies to account for this effect.
 \end{abstract}
@@
 
-%Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture.
-%We later reproduced similar results for the ``Core'', ``Nehalem'', and ``Ivy Bridge'' microarchitectures as well.
-%Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.
+Because bias can be hardware dependent, and to keep the scope of this work manageable, we initially chose to focus our analysis on the 4th generation Intel ``Haswell'' microarchitecture.
+We later reproduced similar results for the ``Core'', ``Nehalem'', and ``Ivy Bridge'' microarchitectures as well.
+Each setup is described in Table~\ref{tab:machines}, and referenced throughout the paper.
 
@@
 By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
-All experiments are done with the machine under minimal load, disabling \emph{frequency scaling} to keep the CPU's clock speed fixed, and \emph{Hyper-threading} (HT) to reduce the possibility for resource contention between threads.\footnote{Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.}
+All experiments are done on a dedicated machine under minimal load, with \emph{frequency scaling} disabled to keep the CPU's clock speed fixed.
+Finally, we disable \emph{Hyper-threading} (HT), which reduces the possibility for resource contention between threads.
+Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.
 

### 2016-01-09 paper.tex
Move microkernel result out of introduction
With shorter article, no need to tease this in the beginning.

--- 2016-01-09 paper.tex
@@
 
-%\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
-\pgfplotstableread{bin/microkernel-core.dat}{\tblmicrokernelcore}
-\begin{figure}[t]
-  \begin{tikzpicture}
-    \begin{axis}[
-        title = Core,
-        width = \textwidth/2.1,
-        height = 4.5cm,
-        font = \footnotesize,
-        xlabel=Bytes added to environment,
-        ylabel=Cycles,
-        domain = 0:4096,
-        xtick = {0,1024,...,4096},
-        xmin = 0,
-        xmax = 4096,
-        cycle list name = exotic
-      ]
-      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelcore ;
-    \end{axis}
-  \end{tikzpicture}
-  \caption{\label{fig:motivation}Performance counter data running the micro-kernel from Figure~\ref{fig:microkernel} under different environment sizes.
-Notice the clear spike at around 3200 bytes, where the number of cycles executed nearly doubles.
-Measuring performance in this configuration would give a biased result.}
-\end{figure}
-
-\begin{figure}[t]
-  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
-static int i, j, k;
-int main() {
-    int g = 0, inc = 1;
-    for (; g < 65536; g++) {
-        i += inc;
-        j += inc;
-        k += inc;
-    }
-    return 0;
-}
-  \end{lstlisting}
-  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
-\end{figure}
-
 A literature study of 88 papers by Mytkowicz et al. showed that the median reported speedup was within the margin of bias, possibly invalidating the claimed speedups~\cite{Mytkowicz:2008:OE&MB}.
@@
 
-In Figure~\ref{fig:motivation}, we have reproduced a result originally presented in a paper by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program (Figure~\ref{fig:microkernel}) depends on environment variables.
-This example provides the basis for our analysis.
-We show how some instances of measurement bias, such as the above, can be explained by \emph{address aliasing}, an artifact of a presumably Intel specific optimization on the out-of-order execution pipeline.
+%In Figure~\ref{fig:motivation}, we have reproduced a result originally presented in a paper by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing how the number of cycles executed for a simple program (Figure~\ref{fig:microkernel}) depends on environment variables.
+%This example provides the basis for our analysis.
+We show how some instances of measurement bias can be explained by \emph{address aliasing}, an artifact of a presumably Intel specific optimization on the out-of-order execution pipeline.
 The CPU uses a heuristic for determining whether loads are dependent on previous stores, comparing only the 12 least significant virtual address bits.
@@
 
+% Original code presented my Mytkowicz et al.
+\begin{figure}[t]
+  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
+static int i, j, k;
+int main() {
+    int g = 0, inc = 1;
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    return 0;
+}
+  \end{lstlisting}
+  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
+\end{figure}
+
 % Large plot of cycle count for two 4K stack sections.
@@
 
+%\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
+\pgfplotstableread{bin/microkernel-core.dat}{\tblmicrokernelcore}
 \pgfplotstableread{bin/microkernel-nehalem.dat}{\tblmicrokernelnehalem}
@@
 \begin{figure}[t]
+  \begin{tikzpicture}
+    \begin{axis}[
+        title = Core,
+        width = \textwidth/2.1,
+        height = 4.5cm,
+        font = \footnotesize,
+        xlabel=Bytes added to environment,
+        ylabel=Cycles,
+        domain = 0:4096,
+        xtick = {0,1024,...,4096},
+        xmin = 0,
+        xmax = 4096,
+        cycle list name = exotic
+      ]
+      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelcore ;
+    \end{axis}
+  \end{tikzpicture}
+  %\caption{\label{fig:motivation}Performance counter data running the micro-kernel from Figure~\ref{fig:microkernel} under different environment sizes.
+Notice the clear spike at around 3200 bytes, where the number of cycles executed nearly doubles.
+Measuring performance in this configuration would give a biased result.}
   \begin{tikzpicture}
@@
 While we show only the analysis for ``Haswell'' here, the same approach gave similar results for all the CPUs we tested.
-Figure~\ref{fig:microkernel-all-archs} shows additional cycle count plots for ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
-The experiment was also done for the older ``Core'' setup, shown in Figure~\ref{fig:motivation}.
-On this architecture, the \emph{load\_blocks.overlap\_store} event is used to indicate address aliasing.
+Figure~\ref{fig:microkernel-all-archs} shows additional cycle count plots for ``Core'', ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.%\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
+%The experiment was also done for the older ``Core'' setup, shown in Figure~\ref{fig:motivation}.
+On the older ``Core'' architecture, \emph{load\_blocks.overlap\_store} is used to indicate address aliasing.
 Notably, this event was also found to be significant in the original paper, where the authors used a similar setup~\cite{Mytkowicz:2009:WrongData}.

### 2016-01-09 paper.tex
More small fixes.
Removed some redundant information.

@@
 Repeated invocation will also even out performance by warming up the cache.
-\begin{lstlisting}
-    for (i = 0; i < k; ++i)
-        conv(n, input, output + offset);
-\end{lstlisting}
+%\begin{lstlisting}
+%    for (i = 0; i < k; ++i)
+%        conv(n, input, output + offset);
+%\end{lstlisting}
+
 An estimate of the actual cost a single invocation can be calculated by averaging over a number of repeated function calls, after subtracting the overhead of invoking it the first time.
@@
 %Similarly none offset their allocations based on the initial stack frame.
-We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for functions such as the convolution example.
+We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for many functions.
 
@@
 Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
-In code, the function is invoked with the following parameters:
-\begin{lstlisting}[breaklines=true]
-    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
-\end{lstlisting}
+%In code, the function is invoked with the following parameters:
+%\begin{lstlisting}[breaklines=true]
+%    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
+%\end{lstlisting}
 

### 2016-01-09 paper.tex
Fixed headline.

@@
 
-\subsection{Most Allocators Alias by Default}
+\subsection{Memory Allocators and Alias}
 Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.

### 2016-01-09 paper.tex
Fixed figure ref.

@@
 While we show only the analysis for ``Haswell'' here, the same approach gave similar results for all the CPUs we tested.
-Figure~\ref{fig:microkernel-all-archs} shows additional cycle count plots for ``Core'', ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.%\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
+Figure~\ref{fig:microkernel-all-archs} shows cycle count plots for ``Core'', ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.%\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
 %The experiment was also done for the older ``Core'' setup, shown in Figure~\ref{fig:motivation}.
@@
 We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and draw the conclusion that address aliasing must be the underlying explanation.
-Still, this only explains part of what is going on in Figure~\ref{fig:motivation}, as there seem to be yet another bias effect at every fourth environment increment.
+Still, this only explains part of what is going on in Figure~\ref{fig:microkernel-all-archs}, as there seem to be yet another bias effect at every fourth environment increment.
 These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.

### 2016-01-09 paper.tex
Cut redundant text.

@@
 Our results show that the cost of aliasing can be significant, exemplified by a simple program with up to $2x$ speedup based in heap address alignment alone.
-We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing, by favoring page alignment for large allocations.
+We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing.%, by favoring page alignment for large allocations.
 With an understanding of how false dependencies in the CPU can impact performance, measurement bias caused by it can to some extent be predicted and even accounted for in software.
@@
 Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
-The potential issues they find are either handled explicitly or are irrelevant to our setup.
+The issues they find are either handled explicitly or are irrelevant to our setup.
 %They find intra-machine variations of up to 1.07\%,
@@
 % Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
-Results are also averaged over multiple runs to reduce potential random error, using built-in perf rerun functionality.
+Results are also averaged over multiple runs to reduce potential random error.%, using built-in perf rerun functionality.
 

### 2016-01-09 paper.tex
Compressed 4K aliasing intro.

@@
 One of the issues that can limit throughput is dependencies between a load and previous stores.
-To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}. 
-More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
-Loads are therefore issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
+To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory these operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess} as well.
+%More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
+Loads are then issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
 The prediction is later verified, replaying any instructions that were wrongly assumed to have no dependencies.

### 2016-01-09 paper.tex

@@
 Section~\ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section~\ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
-Finally, in Section~\ref{sec:blas} we look at alias in BLAS libraries.
+Section~\ref{sec:blas} looks at alias in BLAS libraries.
+Finally, in Section~\ref{seg:mitigation} we discuss some approaches to deal with aliasing effects.
 Related work is discussed in Section~\ref{sec:related}, before we summarize and conclude in Section~\ref{sec:conclusions}.
@@
 % Because the stack is used for local variables, complex code is likely to contain many potential conflicts with other areas of memory.
-
-% See /analysis for results showing this does not alias
-\begin{figure}
-  \begin{lstlisting}[frame=single, xleftmargin=.005\textwidth, xrightmargin=.005\textwidth]
-#define ALIAS(a, b) \
-    ((((long)&a) & 0xfff) == (((long)&b) & 0xfff))
-static int i, j, k;
-int main() {
-    int g = 0, inc = 1;
-    if (ALIAS(inc, i) || ALIAS(g, i))
-        return main();
-    for (; g < 65536; g++) {
-        i += inc;
-        j += inc;
-        k += inc;
-    }
-    return 0;
-}
-  \end{lstlisting}
-  \caption{Modified microkernel that can dynamically detect aliasing case, and avoid it by pushing another stack frame.}
-  \label{lst:loopfixed}
-\end{figure}
-
-%\newpage
-\subsection{Avoiding Aliasing}
-Addresses of automatic variables can not be determined statically, because the location of the stack at runtime is generally unknown.
-In addition to being offset by environment variables, the initial stack address can also be perturbed by for example program arguments or address randomization.
-Although one can not easily know if a collision is going to happen for a given environment a priori, it is possible to change the program to account for possible alias conditions.
-A proof of concept of how alias-free code could be generated in this particular case is shown in Figure \ref{lst:loopfixed}.
-If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
-Calling the function recursively will effectively allocate a new set of variables further down the stack, and the alias condition is avoided.
-
-Even though this particular solution is impractical, it shows that compilers and programmers \emph{could} take measures to account for aliasing.
 
@@
 % We do not attempt to pinpoint exactly which assembly instructions might alias, but one can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
-
-\subsection{Ways to Deal with Heap Address Aliasing}
-Performance penalties caused by address aliasing can be significant, creating bias towards certain memory layouts.
-However, with an understanding of the underlying mechanism that causes bias, the effects can be predicted and to some extent eliminated in software.
-The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
-We identify some relevant optimization techniques:
-
-\subsubsection{Mark buffers with \texttt{restrict}}
-In our convolution kernel implementation, the compiler has to account for the fact that input and output pointers might alias, or that the buffers partially overlap.
-This limits the extent generated code can keep data in registers without updating the values from memory, as a write to one buffer potentially could invalidate a cached value from the other.
-The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer does not alias with any other, allowing for more efficient code generation with fewer memory accesses.
-% \begin{lstlisting}[breaklines=true]
-%     void conv(int n, const float * restrict in, float * restrict out);
-% \end{lstlisting}
-Applied to our example, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% and alias events by approximately 10 million on optimization level O2.
-%Alias events are reduced by about 10 million for the default alignment, with a corresponding improvement in cycle count. % File backing this in results
-
-\subsubsection{Use a special purpose allocator}
-The Intel optimization manual mentions this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
-However, to our knowledge there are no commonly used allocators that specifically try to mitigate aliasing in heap allocated memory.
-%Similarly none offset their allocations based on the initial stack frame.
-We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for many functions.
-
-A potential solution could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
-The allocator could also accept hints to which buffers are going to be accessed in parallel, for example as an optional argument to a custom version of \texttt{malloc}.
-
-\subsubsection{Manually adjust address offsets}
-In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.
-This can be achieved by exploiting the fact the \texttt{mmap} is in fact guaranteed to be placed at a page boundary, and use that directly instead of \texttt{malloc} to create specific alignments.
-% The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.
-% \begin{lstlisting}[breaklines=true]
-%     mmap(NULL, (n+d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) +d;
-% \end{lstlisting}
-% The same pointer difference must of course by subtracted before unmapping the memory.
 
@@
 
+\section{Ways to Deal with Address Aliasing}
+\label{sec:mitigation}
+Performance penalties caused by address aliasing can be significant, creating bias towards certain memory layouts.
+However, with an understanding of the underlying mechanism that causes bias, the effects can be predicted and to some extent eliminated in software.
+The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
+We identify some relevant optimization techniques:
+
+\subsection{Mark buffers with \texttt{restrict}}
+In our convolution kernel implementation, the compiler has to account for the fact that input and output pointers might alias, or that the buffers partially overlap.
+This limits the extent generated code can keep data in registers without updating the values from memory, as a write to one buffer potentially could invalidate a cached value from the other.
+The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer does not alias with any other, allowing for more efficient code generation with fewer memory accesses.
+% \begin{lstlisting}[breaklines=true]
+%     void conv(int n, const float * restrict in, float * restrict out);
+% \end{lstlisting}
+Applied to our example, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% and alias events by approximately 10 million on optimization level O2.
+%Alias events are reduced by about 10 million for the default alignment, with a corresponding improvement in cycle count. % File backing this in results
+
+\subsection{Use a special purpose allocator}
+The Intel optimization manual mentions this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
+However, to our knowledge there are no commonly used allocators that specifically try to mitigate aliasing in heap allocated memory.
+%Similarly none offset their allocations based on the initial stack frame.
+We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for many functions.
+
+A potential solution could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
+The allocator could also accept hints to which buffers are going to be accessed in parallel, for example as an optional argument to a custom version of \texttt{malloc}.
+
+\subsection{Manually adjust address offsets}
+In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.
+This can be achieved by exploiting the fact the \texttt{mmap} is in fact guaranteed to be placed at a page boundary, and use that directly instead of \texttt{malloc} to create specific alignments.
+% The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.
+% \begin{lstlisting}[breaklines=true]
+%     mmap(NULL, (n+d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) +d;
+% \end{lstlisting}
+% The same pointer difference must of course by subtracted before unmapping the memory.
+
+% See /analysis for results showing this does not alias
+\begin{figure}
+  \begin{lstlisting}[frame=single, xleftmargin=.005\textwidth, xrightmargin=.005\textwidth]
+#define ALIAS(a, b) \
+    ((((long)&a) & 0xfff) == (((long)&b) & 0xfff))
+static int i, j, k;
+int main() {
+    int g = 0, inc = 1;
+    if (ALIAS(inc, i) || ALIAS(g, i))
+        return main();
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    return 0;
+}
+  \end{lstlisting}
+  \caption{Modified microkernel that can dynamically detect aliasing case, and avoid it by pushing another stack frame.}
+  \label{lst:loopfixed}
+\end{figure}
+
+% Addresses of automatic variables can not be determined statically, because the location of the stack at runtime is generally unknown.
+% In addition to being offset by environment variables, the initial stack address can also be perturbed by for example program arguments or address randomization.
+% Although one can not easily know if a collision is going to happen for a given environment a priori, it is possible to change the program to account for possible alias conditions.
+It is also possible to manipulate addresses of automatic (stack allocated) variables.
+A proof of concept of how alias-free code could be generated for the microkernel example is shown in Figure \ref{lst:loopfixed}.
+If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
+Calling the function recursively will effectively allocate a new set of variables further down the stack, and the alias condition is avoided.
+Even though this particular solution is impractical, it proves that compilers and programmers \emph{could} take measures to account for aliasing.
+
 \section{Related Work}

### 2016-01-09 paper.tex
Move alias mitigation to separate section

@@
 Our results show that the cost of aliasing can be significant, exemplified by a simple program with up to $2x$ speedup based in heap address alignment alone.
-We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing.%, by favoring page alignment for large allocations.
+We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing, by favoring page alignment for large allocations.
 With an understanding of how false dependencies in the CPU can impact performance, measurement bias caused by it can to some extent be predicted and even accounted for in software.
@@
 Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
-The issues they find are either handled explicitly or are irrelevant to our setup.
+The potential issues they find are either handled explicitly or are irrelevant to our setup.
 %They find intra-machine variations of up to 1.07\%,
@@
 % Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
-Results are also averaged over multiple runs to reduce potential random error.%, using built-in perf rerun functionality.
+Results are also averaged over multiple runs to reduce potential random error, using built-in perf rerun functionality.
 
@@
 One of the issues that can limit throughput is dependencies between a load and previous stores.
-To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory these operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess} as well.
-%More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
-Loads are then issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
+To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}. 
+More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
+Loads are therefore issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
 The prediction is later verified, replaying any instructions that were wrongly assumed to have no dependencies.

### 2016-01-09 paper.tex
Compressed the intro.

--- 2016-01-09 paper.tex
@@
 Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
-The underlying explanation is likely that this changes memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
+The underlying explanation is that this changes memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
 Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.

### 2016-01-09 paper.tex
Compressed the mmap's guarantee.

@@
 \subsection{Manually adjust address offsets}
-In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.
-This can be achieved by exploiting the fact the \texttt{mmap} is in fact guaranteed to be placed at a page boundary, and use that directly instead of \texttt{malloc} to create specific alignments.
+In some cases it might help to explicitly control the memory addresses used, for example by requiring fixed relative offset between input and output pointers.
+This can be achieved by exploiting \texttt{mmap's} guarantee to be page boundary aligned, and use that directly instead of \texttt{malloc} to create specific alignments.
 % The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.

### 2016-01-09 paper.tex
Compressed the ways to deal with..

@@
 % NB: Taken from analysis directory, update manually on new results
-Correlation between cycles executed and address aliasing events were 95\% for libblas, 81\% for libatlas and 45\% for libopenblas.
+Correlation between cycles and address aliasing were 95\% for libblas, 81\% for libatlas and 45\% for libopenblas.
 
@@
 \subsection{Use a special purpose allocator}
-The Intel optimization manual mentions this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
-However, to our knowledge there are no commonly used allocators that specifically try to mitigate aliasing in heap allocated memory.
+The Intel optimization manual mentions this% in \emph{User/Source Coding Rule 8}
+, suggesting that special purpose allocators could be used to avoid aliasing~\cite{OptimizationManual}.
+However, to our knowledge there are no allocators that specifically try to mitigate aliasing in heap allocated memory.
 %Similarly none offset their allocations based on the initial stack frame.

### 2016-01-09 paper.tex
Language: things->strings.

@@
 \label{sec:environment}
-System environment variables contain things like the name of the logged in user, home directory, shell state, and various other settings.
+System environment variables contain strings like the name of the logged in user, home directory, shell state, and various other settings.
 As a source of bias, it is not the environment variables themselves that are important, but rather the effect their total \emph{size} has on the position and alignment of the stack.

### 2016-01-09 paper.tex
Compressed/cleanup IV intro.

@@
 As a source of bias, it is not the environment variables themselves that are important, but rather the effect their total \emph{size} has on the position and alignment of the stack.
-As indicated in Figure \ref{fig:virtualmemory}, environment variables and program arguments are allocated in the stack section of virtual memory, close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory.} and before the first call frame.
-Changing environment variables will therefore offset the location of the stack, and consequently all stack allocated variables.
-After this offset, the stack is normally realigned to a 16 byte boundary, enforced by the compiler.
-Within a span of 4096 bytes there are thus 256 possible initial stack addresses, each representing a different execution context with respect to address aliasing. 
+Environment variables and program arguments are allocated in the stack section of virtual memory, close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory.} and before the first call frame.
+Figure \ref{fig:virtualmemory} shows this layout.
+Changing environment variables will offset the location of the stack, and consequently all stack allocated variables.
+After this offset, the stack is normally realigned to a 16 byte boundary by the compiler.
+Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context with respect to address aliasing. 
 %Note that there is no clear relationship between environment size and stack location with ASLR enabled.

### 2016-01-09 paper.tex
Compressed/cleanup Other Architectures.

@@
 Notably, this event was also found to be significant in the original paper, where the authors used a similar setup~\cite{Mytkowicz:2009:WrongData}.
-We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and draw the conclusion that address aliasing must be the underlying explanation.
+We were able to verify that partial address overlap was occurring at the same time this event is spiking, concluding that address aliasing must be the underlying explanation.
 Still, this only explains part of what is going on in Figure~\ref{fig:microkernel-all-archs}, as there seem to be yet another bias effect at every fourth environment increment.
@@
 
-Results for the ``Nehalem'' and ``Ivy Bridge'' architectures are more clear-cut, with large spikes only at the aliasing environment configuration.
-It is clear that the address alias issue exist for a wide range of Intel microarchitectures.
+Results for the ``Nehalem'' and ``Ivy Bridge'' architectures are clearer, with large spikes only at the aliasing environment configuration.
+It is clear that this issue exist for a wide range of Intel microarchitectures.
+%We believe that address alias issue exist for a wider range of Intel microarchitectures.
 

### 2016-01-09 paper.tex
Compressed malloc description-

@@
 \subsection{Memory Allocators and Alias}
-Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
+Acquiring dynamic memory at runtime is usually handled by the \texttt{malloc} function.
 Depending on the particular request and allocator used, the returned value will either point to the ``regular'' heap, or to a memory mapped area (see Figure \ref{fig:virtualmemory}).

### 2016-01-09 paper.tex
Compressed/cleanup Microkernel Analysis.

@@
 To analyze the bias effects at these points, we calculate the \emph{median} value of each counter over all contexts, and compare that to the two extreme cases.
-Going through the list of events manually, we selected the set that had the highest correlation with cycle count\footnote{Some correlating performance events are omitted; for example bus-cycles, which will naturally vary with total cycle count.}.
+Going through the list of events, we selected the set that had the highest correlation with cycle count\footnote{Some correlating performance events are omitted; for example bus-cycles, which will naturally vary with total cycle count.}.
 Table \ref{tab:loopcorrelation} shows numerical counter values for the first spike, compared to the median values.
-Event counts on the second spike are virtually identical to the first, and omitted for brevity.
+Event counts on the second spike are almost identical to the first, and omitted for brevity.
 
@@
 
-We see the most extreme change from median to worst case in the number of alias events. 
-If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
+We see a huge change from the median in the number of alias events. 
+%If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
+We found that there are near zero address aliasing events everywhere except exactly the points we observe bias. 
 The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
@@
 Fewer stalls in the aliasing case could indicate less contention on the reservation station.
-This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports have fewer operations going in parallel.
+This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the 8 execution ports have fewer operations going in parallel.
 Note that the number of micro-ops \emph{retired} overall does not change.
@@
 %Observing addresses of stack allocated data at runtime is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
-Assembly code was injected to calculate and output the runtime addresses of \texttt{g} and \texttt{inc}, carefully craftet to not introduce any observer effects.
+Carefully crafted assembly code was used to capture the runtime addresses of \texttt{g} and \texttt{inc}.
 %We made no change to the stack allocation instructions, and the code offset did not affect the addresses of static variables.

### 2016-01-09 paper.tex
Compressed BLAS intro.
Removed the BLAS list.

@@
 Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
+\iffalse
 The functionality is divided into three categories:
@@
 \end{description}
-Procedures operating on vectors, from level 1 or 2, intuitively appears the most likely to have potential for aliasing.
-
+Procedures operating on vectors, intuitively appears the most likely to have potential for aliasing.
+\fi
+Here the level 1 and 2 vector operations intuitively appears to be more likely to have potential for aliasing.
 We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.

### 2016-01-09 paper.tex
Compressed restrict section.

@@
 % \end{lstlisting}
-Applied to our example, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores executed by more than 80 \% and alias events by approximately 10 million on optimization level O2.
+Applied to our example, the compiler is able to move a store instruction out of the inner loop, reducing the number of stores by more than 80 \% and alias events by approximately 10 million on optimization level O2.
 %Alias events are reduced by about 10 million for the default alignment, with a corresponding improvement in cycle count. % File backing this in results

### 2016-01-09 paper.tex
Reduce bloat in mitigation section

@@
 \fi
-Here the level 1 and 2 vector operations intuitively appears to be more likely to have potential for aliasing.
+Here, the level 1 and 2 vector operations intuitively appears to be more likely to have potential for aliasing.
 We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
@@
 However, with an understanding of the underlying mechanism that causes bias, the effects can be predicted and to some extent eliminated in software.
-The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
+%The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
 We identify some relevant optimization techniques:
@@
 %Similarly none offset their allocations based on the initial stack frame.
-We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for many functions.
-
-A potential solution could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
+%We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for many functions.
+
+One approach could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
 The allocator could also accept hints to which buffers are going to be accessed in parallel, for example as an optional argument to a custom version of \texttt{malloc}.
@@
 In some cases it might help to explicitly control the memory addresses used, for example by requiring fixed relative offset between input and output pointers.
-This can be achieved by exploiting \texttt{mmap's} guarantee to be page boundary aligned, and use that directly instead of \texttt{malloc} to create specific alignments.
+For heap allocations, one can exploit that \texttt{mmap} is guaranteed to be page aligned, and use that directly to create specific alignments.
 % The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.
@@
 It is also possible to manipulate addresses of automatic (stack allocated) variables.
-A proof of concept of how alias-free code could be generated for the microkernel example is shown in Figure \ref{lst:loopfixed}.
+A proof of concept of how alias-free code could be generated for the program discussed in Section~\ref{sec:microkernel} is shown in Figure \ref{lst:loopfixed}.
 If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
@@
 For heap address aliasing especially, we find that manual intervention can be required in order to achieve optimal performance.
-Our results should inspire more clever optimizations and heuristics taking address aliasing effects into account, as the potential performance improvements can be substantial.
+Our results can inspire more clever optimizations and heuristics taking address aliasing effects into account, as the potential performance improvements can be substantial.
 

### 2016-01-09 paper.tex
Fix section reference in outline
Fix minor stuff

@@
 Section~\ref{sec:blas} looks at alias in BLAS libraries.
-Finally, in Section~\ref{seg:mitigation} we discuss some approaches to deal with aliasing effects.
+Finally, in Section~\ref{sec:mitigation} we discuss some approaches to deal with aliasing effects.
 Related work is discussed in Section~\ref{sec:related}, before we summarize and conclude in Section~\ref{sec:conclusions}.
@@
         width = \textwidth,
-        height = 6cm,
+        height = 5.5cm,
         font = \footnotesize,
@@
 %A finer sampling is not necessary, because the stack is by default aligned to 16 bytes.
-The microkernel is compiled with {\small{GCC}} using no optimization; any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.
+The code is compiled with {\small{GCC}} using no optimization; any optimization would likely disregard most of the function as redundant, reducing it to return zero immediately.
 
@@
 
-The cycles\_ldm\_pending and stalls\_l2\_pending are interesting, as they are related to the memory and cache system.
+The cycles\_ldm\_pending and stalls\_l2\_pending numbers are interesting, as they are related to the memory and cache system.
 However, this program use only 2 cache-lines, except for the initial startup.

### 2016-01-09 paper.tex
Remove some stuff in convolution discussion

@@
 This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
-Controlling the offset parameter, we can create environments where the address suffixes are any desired number of \texttt{sizeof(float)} bytes apart.
-Allocating and managing these buffers at startup takes a non-negligible amount of work.
-The overhead can be masked by repeatedly invoking the convolution kernel after allocating and initializing all the inputs.
+Controlling the offset parameter, we can create environments where the address suffixes are offset by any desired number of bytes.
+Allocating and managing these buffers at startup takes a non-negligible amount of work, but the overhead can be masked by repeatedly invoking the convolution kernel in a loop, after allocating and initializing all the inputs.
 Repeated invocation will also even out performance by warming up the cache.
@@
 %\end{lstlisting}
-
-An estimate of the actual cost a single invocation can be calculated by averaging over a number of repeated function calls, after subtracting the overhead of invoking it the first time.
+An estimate of the actual cost a single invocation can be calculated by the following formula:
+%An estimate of the actual cost a single invocation can be calculated by averaging over $ k $ repeated function calls, after subtracting time $ t_{1} $ which includes allocation overhead.
 \[
@@
 \]
-Our results are with $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
+We choose $k=11$, effectively getting the average over 10 iterations.
 In addition, performance counter measurements are averaged over 10 samples using the repeat mechanism of perf.
@@
 Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between the heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
-The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
-Numbers on the x axis represent the amount of offset, measured in number of \texttt{sizeof(float)} bytes.
+%The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
+%Numbers on the x axis represent the amount of offset, measured in number of \texttt{sizeof(float)} bytes.
 An offset of zero is the default behavior for this program when using \texttt{malloc} and moderately large inputs.
-Both for optimization level 2 and 3, the default alignment is close to worst case performance.
+For both for optimization levels 2 and 3, this is close to worst case performance.
 The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
-This effect is only observed for address separations close to zero, thus we only show the first 20 data points.
-If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
+%This effect is only observed for address separations close to zero, thus we only show the first 20 data points.
+%If extended to cover the full width of possible offsets within a 4K segment, we see that the performance is uniform everywhere else.
 

### 2016-01-09 paper.tex
Give some space to the poor captions
Shrink introduction

@@
 %\usepackage{enumitem} % Controlling list styling
-%\usepackage[font=small,skip=0pt]{caption}
 %http://tex.stackexchange.com/questions/94016/how-to-reduce-space-between-image-and-its-caption
-\usepackage[font=small,belowskip=-10pt,aboveskip=0pt]{caption}
+\usepackage[font=small]{caption}
 
@@
 \section{Introduction}
-Accurately measuring the performance of computer programs is important in order to evaluate new algorithms, or compare different implementations.
+%Accurately measuring the performance of computer programs is important in order to evaluate new algorithms, or compare different implementations.
 A key challenge for performance analysts and systems researchers is handling \emph{measurement bias}.
-Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
-The underlying explanation is that this changes memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
-Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to for example system environment variables ultimately affect program execution.
-
-A literature study of 88 papers by Mytkowicz et al. showed that the median reported speedup was within the margin of bias, possibly invalidating the claimed speedups~\cite{Mytkowicz:2008:OE&MB}.
+Seemingly innocious external factors---such as link ordering or contents of environment variables---have been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
+The underlying explanation is that these factors affect memory addresses of code or data at runtime, which in turn interacts with various low level hardware buffers.
+Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to things like environment variables ultimately affect program execution.
+
+%A literature study of 88 papers by Mytkowicz et al. showed that the median reported speedup was within the margin of bias, possibly invalidating the claimed speedups~\cite{Mytkowicz:2008:OE&MB}.
 To alleviate the impact of measurement bias in performance analysis, researchers have proposed techniques like causal analysis, and randomization of execution contexts~\cite{Mytkowicz:2008:OE&MB}.
@@
 Our results show that the cost of aliasing can be significant, exemplified by a simple program with up to $2x$ speedup based in heap address alignment alone.
-We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing.%, by favoring page alignment for large allocations.
+%We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing, by favoring page alignment for large allocations.
 With an understanding of how false dependencies in the CPU can impact performance, measurement bias caused by it can to some extent be predicted and even accounted for in software.
@@
 
-The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section.
-Then section~\ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
+The rest of this paper is structured as follows:
+Section~\ref{sec:methodology} starts by describing methodology and experimental setup.
+Then, section~\ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
 Section~\ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section~\ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
@@
 In order to understand biased behavior it is important to be able to do precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
-This can be achieved by using performance counters, instrumentation support in hardware that can be used to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
+This can be achieved by using performance counters; instrumentation support in hardware that can be used to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
 Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.

### 2016-01-09 paper.tex
Fix various smaller issues

@@
 
-Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
-The issues they find are either handled explicitly or are irrelevant to our setup.
+%Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
+%The issues they find are either handled explicitly or are irrelevant to our setup.
 %They find intra-machine variations of up to 1.07\%,
@@
 One of the issues that can limit throughput is dependencies between a load and previous stores.
-To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory these operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess} as well.
+To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}.
 %More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
-Loads are then issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
+Loads are issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
 The prediction is later verified, replaying any instructions that were wrongly assumed to have no dependencies.
@@
 In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
-A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
-Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
+%A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
+%Despite being independent, 
+In these cases, the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
 The number of times this happens can be counted by the following event:
@@
 
-We see a huge change from the median in the number of alias events. 
+We see a significant change from the median in the number of alias events. 
 %If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
@@
   \hspace{0.5cm}
-  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}.
-Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations.
-Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
+  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}, input size $n=2^{20}$.}
+  %Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
 \end{figure*}
@@
 \fi
-Here, the level 1 and 2 vector operations intuitively appears to be more likely to have potential for aliasing.
+%The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
 We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.

### 2016-01-09 paper.tex
Place blas graph on correct page
Reintroduce pmu comment and reference

@@
 We use the perf-stat command in all experiments, which accepts raw PMU event codes listed in the reference manual~\cite{Volume3B}.
-
-%Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
-%The issues they find are either handled explicitly or are irrelevant to our setup.
+Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
+The issues they find are either handled explicitly or are irrelevant to our setup.
 %They find intra-machine variations of up to 1.07\%,
@@
 % Paper is rename of "Toward Accurate Performance Evaluation using Hardware Counters".
-Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
+Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
 Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
@@
 Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
-\iffalse
-The functionality is divided into three categories:
-\begin{description}
-  \item[Level 1] \hspace{0.1cm} Scalar and vector operations, such as dot product and vector addition.
-  \item[Level 2] \hspace{0.1cm} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
-  \item[Level 3] \hspace{0.1cm} Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
-\end{description}
-Procedures operating on vectors, intuitively appears the most likely to have potential for aliasing.
-\fi
-%The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
-We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
-We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
-In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
-We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
-Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
-%In code, the function is invoked with the following parameters:
-%\begin{lstlisting}[breaklines=true]
-%    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
-%\end{lstlisting}
-
-Each of the structures \texttt{A}, \texttt{x} and \texttt{y} represent different locations in memory, and their addresses is part of the execution context.
-The \texttt{mmap} method outlined in the previous section is used to control the location of each of these buffers on the heap.
-From our experiments, we found that the memory address offset of vector $\boldsymbol{y}$ compared to matrix $A$ is the most important with respect to the number of alias events.
-For the results presented here, we set $A = I$ (identity matrix) and $\boldsymbol{x} = [0, 1, 2, ..., N-1]$, with parameters $M = N = 1024$.
-Address suffix of $A$ and $\boldsymbol{x}$ is fixed to 0x000, while performance events are sampled for increasing address offset of $\boldsymbol{y}$, starting at a baseline of 0x000.
-The experiment was run on the ``Haswell'' setup, with Hyper-Threading enabled.
-In an attempt to isolate the cost of a single \texttt{cblas_dgemv} invocation, we use the estimation approach described in Section~\ref{sec:heap} to subtract the constant overhead from heap allocation and initialization.
-
 \pgfplotstableread{bin/libblas.dat}{\libblastable}
@@
 
+\iffalse
+The functionality is divided into three categories:
+\begin{description}
+  \item[Level 1] \hspace{0.1cm} Scalar and vector operations, such as dot product and vector addition.
+  \item[Level 2] \hspace{0.1cm} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
+  \item[Level 3] \hspace{0.1cm} Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
+\end{description}
+Procedures operating on vectors, intuitively appears the most likely to have potential for aliasing.
+\fi
+%The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
+We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
+We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
+In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
+We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
+Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
+%In code, the function is invoked with the following parameters:
+%\begin{lstlisting}[breaklines=true]
+%    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
+%\end{lstlisting}
+
+Each of the structures \texttt{A}, \texttt{x} and \texttt{y} represent different locations in memory, and their addresses is part of the execution context.
+The \texttt{mmap} method outlined in the previous section is used to control the location of each of these buffers on the heap.
+From our experiments, we found that the memory address offset of vector $\boldsymbol{y}$ compared to matrix $A$ is the most important with respect to the number of alias events.
+For the results presented here, we set $A = I$ (identity matrix) and $\boldsymbol{x} = [0, 1, 2, ..., N-1]$, with parameters $M = N = 1024$.
+Address suffix of $A$ and $\boldsymbol{x}$ is fixed to 0x000, while performance events are sampled for increasing address offset of $\boldsymbol{y}$, starting at a baseline of 0x000.
+The experiment was run on the ``Haswell'' setup, with Hyper-Threading enabled.
+In an attempt to isolate the cost of a single \texttt{cblas_dgemv} invocation, we use the estimation approach described in Section~\ref{sec:heap} to subtract the constant overhead from heap allocation and initialization.
+
 Cycles executed and alias events for each library is shown in Figure~\ref{fig:blas}.

### 2016-01-10 paper.tex
Fix wording in some sections

@@
 
-We see a significant change from the median in the number of alias events. 
+The most distinctive change from the median is the number of alias events.
+This counter reports close to zero except exactly at the two spikes, where it reaches upwards of 320,000.
 %If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
-We found that there are near zero address aliasing events everywhere except exactly the points we observe bias. 
+%We found that there are near zero address aliasing events everywhere except exactly the points we observe bias. 
 The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
@@
 
-Performance counter data clearly points to address aliasing as a plausible explanation, so the next step is to see exactly which memory accesses are aliasing.
+Performance counter data clearly points to address aliasing as the underlying explanation, so the next step is to see exactly which memory accesses are aliasing.
 %For that we need to know the addresses of each variable at runtime.
 %The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
-Memory layout of static data is decided at compile time, and we find the addresses to be \texttt{\&i} = 0x60103c, \texttt{\&j} = 0x601040, and \texttt{\&k} = 0x601044 by reading the ELF symbol table.%\footnote{ELF symbol tables can be read using \texttt{readelf -s}}
-
+Memory layout of static data is decided at compile time, and we find the addresses to be \texttt{\&i} = 0x60103c, \texttt{\&j} = 0x601040, and \texttt{\&k} = 0x601044 by reading the ELF symbol table.
+%\footnote{ELF symbol tables can be read using \texttt{readelf -s}}
 %Observing addresses of stack allocated data at runtime is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
@@
   \end{tikzpicture}
-  \caption{\label{fig:microkernel-all-archs}Microkernel experiment on the ``Nehalem'' and ``Ivy Bridge'' microarchitectures.}
+  \caption{\label{fig:microkernel-all-archs}Microkernel experiment on the ``Core'', ``Nehalem'' and ``Ivy Bridge'' microarchitectures.}
 \end{figure}
@@
 While we show only the analysis for ``Haswell'' here, the same approach gave similar results for all the CPUs we tested.
-Figure~\ref{fig:microkernel-all-archs} shows cycle count plots for ``Core'', ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.%\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
+Figure~\ref{fig:microkernel-all-archs} shows cycle count plots for ``Core'', ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.
+%\footnote{For these results we also pinned the process to a single core during measurements using \texttt{taskset}, which helps remove some of the random noise.}
 %The experiment was also done for the older ``Core'' setup, shown in Figure~\ref{fig:motivation}.
@@
 %Our assumption is that this is a different phenomenon, not related to 4K aliasing.
-We were only able to observe the effect on this particular architecture, and did not investigate the root cause.% is related to 4K aliasing.
-
-Results for the ``Nehalem'' and ``Ivy Bridge'' architectures are clearer, with large spikes only at the aliasing environment configuration.
+We were only able to observe this effect on this particular architecture, and did not investigate further.% is related to 4K aliasing.
+
+Results for the ``Nehalem'' and ``Ivy Bridge'' architectures are more uniform, with large spikes only at the aliasing environment configuration.
 It is clear that this issue exist for a wide range of Intel microarchitectures.
@@
 Address aliasing can be caused by conflicting pairs of load and store operations to any part of memory.
-In the previous sections we saw collisions between static data and the stack, observing bias from external conditions that affected addresses of automatic variables.
+In the previous section, we saw collisions between static data and the stack, observing bias from external conditions that affected addresses of automatic variables.
 Most dynamic memory is allocated on the \emph{heap}, which is managed by an \emph{allocator}.
@@
 Heap allocation routines such as \texttt{malloc} and \texttt{free} are typically dynamically linked, for example as part of glibc.
-If this assignment is independent of the initial stack offset all heap memory accesses can be biased.
+%If this assignment is independent of the initial stack offset all heap memory accesses can be biased.
 The particular library used therefore constitutes an important part of the execution context, as linking to a different library, or a library with some alternative configuration, can impact heap addresses at runtime.
@@
 \subsection{Memory Allocators and Alias}
-Acquiring dynamic memory at runtime is usually handled by the \texttt{malloc} function.
-Depending on the particular request and allocator used, the returned value will either point to the ``regular'' heap, or to a memory mapped area (see Figure \ref{fig:virtualmemory}).
+Acquiring dynamic memory at runtime is usually handled by a variant of \texttt{malloc}.
+Depending on the particular request and allocator used, the returned value will either point to the ``regular'' heap, or to a memory mapped area.
 \begin{itemize}
@@
 The allocator included as part of glibc uses both mechanisms, choosing which based on the \emph{size} of the request~\cite{GlibcManual}.
-Small sizes generally live in the heap, while long lived and large allocations tend to be memory mapped.
-
-The heap section starts at a relatively low address right above static code and data.
-Memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack.
+Small sizes generally live in the heap, while larger allocations tend to be memory mapped.
+
+The heap section starts at a relatively low address right above static code and data, while memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack (see Figure \ref{fig:virtualmemory}).
 Whether a request is served by the heap or by memory mapping is therefore easy to determine just by looking at the pointer values returned:
@@
 However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
-The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
+The page size is usually 4096 bytes, meaning addresses returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
 This behavior is often the worst case for functions that operate on two or more independent buffers.
@@
   \centering
-  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.\label{tab:mallocompare}}
+  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers. Equal three digit suffix indicate an aliasing pair.\label{tab:mallocompare}}
   \pgfplotstabletypeset[
@@
 We observe the addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} for different size parameters.
-Equal three digit address suffix indicate an aliasing pair.
+%Equal three digit address suffix indicate an aliasing pair.
 In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
@@
 Controlling the offset parameter, we can create environments where the address suffixes are offset by any desired number of bytes.
-Allocating and managing these buffers at startup takes a non-negligible amount of work, but the overhead can be masked by repeatedly invoking the convolution kernel in a loop, after allocating and initializing all the inputs.
+Allocating and managing these buffers at startup takes a non-negligible amount of work, but the overhead can be masked by repeatedly invoking the convolution kernel in a loop.
 Repeated invocation will also even out performance by warming up the cache.
@@
 \]
-We choose $k=11$, effectively getting the average over 10 iterations.
+Time spent on a sigle invocation is subtracted from the estimate to mask initialization time.We choose $k=11$, effectively getting the average over 10 iterations.
 In addition, performance counter measurements are averaged over 10 samples using the repeat mechanism of perf.
@@
 For both for optimization levels 2 and 3, this is close to worst case performance.
-The differences in cycles executed is significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 for increasing relative offset.
+The differences in cycles executed is significant, with around $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 on increasing relative offset.
 %This effect is only observed for address separations close to zero, thus we only show the first 20 data points.
@@
 Each of the structures \texttt{A}, \texttt{x} and \texttt{y} represent different locations in memory, and their addresses is part of the execution context.
-The \texttt{mmap} method outlined in the previous section is used to control the location of each of these buffers on the heap.
+We use \texttt{mmap} with manually added offsets to control the location of each of these buffers on the heap.
 From our experiments, we found that the memory address offset of vector $\boldsymbol{y}$ compared to matrix $A$ is the most important with respect to the number of alias events.
@@
 Cycles executed and alias events for each library is shown in Figure~\ref{fig:blas}.
-Each library behaves differently, but in all cases we identify a significant amount of alias. 
+Each library behaves differently, but in all cases we identify a significant amount of aliasing. 
 The diagrams can be understood the same way as for the convolution example; partial address overlap occur when the address suffix delta of $A$ and $\boldsymbol{y}$ is close to zero.

### 2016-01-10 paper.tex
Changed the author layout and added anonymous option.

@@
 
-\author{\IEEEauthorblockN{Lars Kirkholt Melhus, Rune Erlend Jensen}
+\iftrue
+\author{\IEEEauthorblockN{Lars Kirkholt Melhus}
 \IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
 NTNU, Norway\\
-lars.kirkholt.melhus@gmail.com, runeerle@idi.ntnu.no}
-}
-
-\date{19 November 2014}
+lars.kirkholt.melhus@gmail.com}
+\and
+\IEEEauthorblockN{Rune Erlend Jensen}
+\IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
+NTNU, Norway\\
+runeerle@idi.ntnu.no}
+}
+\else
+\author{\IEEEauthorblockN{Anonymous}
+\IEEEauthorblockA{Organization\\
+University\\
+anonymous@mail.com}
+\and
+\IEEEauthorblockN{Anonymous}
+\IEEEauthorblockA{Organization\\
+University\\
+anonymous@mail.com}
+}
+
+\fi
+
+\date{10 January 2016}
 

### 2016-01-10 paper.tex
Lots of language improvements.

@@
 In this paper, we identify \emph{address aliasing} as an important underlying mechanism causing measurement bias on modern Intel microarchitectures.
-By analyzing hardware performance counters, we show how bias can occur from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
+By analyzing hardware performance counters, we show how bias arises from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
 %Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
@@
 We demonstrate ways to deal with this type of bias, through runtime detection and correction of aliasing conditions.
-This is especially important for heap allocations, as most allocators tend to produce worst case alignment by default; favoring page alignment.
+This is especially important for heap allocations, by default most allocators tend to produce worst case alignment, favoring page alignment.
 %For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
-The performance impact of unfavorable memory layouts can be significant, with as much as $2x$ speedup in already optimized code.
+The performance impact of unfavorable memory layouts can be significant, causing up to a $2x$ performance degradation in already optimized code.
 %Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
-Our results can enable new architecture specific optimization strategies to account for this phenomenon.
+Our results can enable new optimization strategies to account for this phenomenon.
 \end{abstract}
@@
 A key challenge for performance analysts and systems researchers is handling \emph{measurement bias}.
-Seemingly innocious external factors---such as link ordering or contents of environment variables---have been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
+Seemingly harmless external factors---such as link ordering or contents of environment variables---have been shown to have significant impact on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
 The underlying explanation is that these factors affect memory addresses of code or data at runtime, which in turn interacts with various low level hardware buffers.
-Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to things like environment variables ultimately affect program execution.
+Because of the complexity of modern processors, it is often hard or impossible to predict exactly how changes to things like environment variables ultimately affect program execution.
 
@@
 %This example provides the basis for our analysis.
-We show how some instances of measurement bias can be explained by \emph{address aliasing}, an artifact of a presumably Intel specific optimization on the out-of-order execution pipeline.
+We show how some instances of measurement bias can be explained by \emph{address aliasing}, an artifact of a known Intel specific optimization on the out-of-order execution pipeline.
 The CPU uses a heuristic for determining whether loads are dependent on previous stores, comparing only the 12 least significant virtual address bits.
@@
 %We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing, by favoring page alignment for large allocations.
-With an understanding of how false dependencies in the CPU can impact performance, measurement bias caused by it can to some extent be predicted and even accounted for in software.
+With an understanding of how false dependencies in the CPU can impact performance, measurement bias caused by it can to some extent be predicted, and even accounted for in software.
 We show how techniques like manual padding of allocations and dynamic detection of aliasing conditions can be used to improve performance.
-Being able to handle address aliasing can give significant speedups, and our findings show potential for further architecture-specific optimizations.
 
@@
 Section~\ref{sec:methodology} starts by describing methodology and experimental setup.
-Then, section~\ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
-Section~\ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section~\ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
+Section~\ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
+Section~\ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section~\ref{sec:heap} we go through a similar analysis for bias and heap address alignment.
 Section~\ref{sec:blas} looks at alias in BLAS libraries.
-Finally, in Section~\ref{sec:mitigation} we discuss some approaches to deal with aliasing effects.
+Finally, in Section~\ref{sec:mitigation} we discuss some approaches to handle aliasing effects.
 Related work is discussed in Section~\ref{sec:related}, before we summarize and conclude in Section~\ref{sec:conclusions}.
@@
 \label{sec:methodology}
-In order to understand biased behavior it is important to be able to do precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
-This can be achieved by using performance counters; instrumentation support in hardware that can be used to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
+In order to understand biased behavior it is important to be able to make precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
+This can be achieved by using performance counters; instrumentation support in hardware that can be used to count various events, such as cycles executed, branch misses, instructions fetched, and so on.
 Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.
@@
 
-A small Python script is used to iterate over an exhaustive set of all known counters, which amounts to about 200 on our architecture.
+To identify issues we use a small Python script to iterate over an exhaustive set of all known counters, which amounts to about 200 on our architecture.
 Only a small set of events are collected at a time, to ensure events are actually counted continuously.
@@
 By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
-All experiments are done with the machine under minimal load, disabling \emph{frequency scaling} to keep the CPU's clock speed fixed, and \emph{Hyper-threading} (HT) to reduce the possibility for resource contention between threads.\footnote{Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.}
+All experiments are conducted under minimal load, disabling \emph{frequency scaling} to keep the CPU's clock speed fixed, and \emph{Hyper-threading} (HT) to reduce the possibility for resource contention between threads.\footnote{Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.}
 
@@
 %More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
-Loads are issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
+Loads are issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is not yet retired.
 The prediction is later verified, replaying any instructions that were wrongly assumed to have no dependencies.
@@
 \begin{description}
-  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \hfill \\ \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
+  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] 
+  \hfill \\ \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} 
+  \cite[B.3.4.4]{OptimizationManual}
 \end{description}
@@
 \begin{description}
-  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \hfill \\ \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
+  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] 
+  \hfill \\ \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} 
+  \cite[Table 19-17]{Volume3B}
 \end{description}
@@
 Changing environment variables will offset the location of the stack, and consequently all stack allocated variables.
-After this offset, the stack is normally realigned to a 16 byte boundary by the compiler.
+After this offset, the stack is realigned to a 16 byte boundary on x86_64 Linux systems.
 Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context with respect to address aliasing. 
@@
 \label{sec:microkernel}
-To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by Mytkowicz et al.~\cite{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:microkernel}.
+To illustrate how address aliasing can cause bias, we revisit the example first presented by Mytkowicz \emph{et al.}~\cite{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:microkernel}.
+% in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' 
 This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.
 Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
-The following data and analysis of this program is from the ``Haswell'' configuration.
-Similar experiments were conducted on the other experimental setups listed in Table~\ref{tab:machines}, all arriving at the same conclusion.
+The following data and analysis of this program is from the ``Haswell'' configuration in Table~\ref{tab:machines}.
+Similar experiments were conducted on all other experimental setups listed, arriving at the same conclusion.
 
@@
 We sampled an extensive set of performance counters in addition to cycle count for each execution context.
-To analyze the bias effects at these points, we calculate the \emph{median} value of each counter over all contexts, and compare that to the two extreme cases.
+To analyze the bias effects at these points, we calculate the median value of each counter over all contexts, and compare that to the two extreme cases.
 Going through the list of events, we selected the set that had the highest correlation with cycle count\footnote{Some correlating performance events are omitted; for example bus-cycles, which will naturally vary with total cycle count.}.
@@
 The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
-Fewer stalls in the aliasing case could indicate less contention on the reservation station.
-This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the 8 execution ports have fewer operations going in parallel.
+We attribute fewer stalls in the aliasing case to less contention on the reservation station, 
+relating to the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the 8 execution ports perform fewer operations in parallel.
 Note that the number of micro-ops \emph{retired} overall does not change.
@@
 \subsection{Other Architectures}
-While we show only the analysis for ``Haswell'' here, the same approach gave similar results for all the CPUs we tested.
+While we show only the analysis for ``Haswell'', the same approach gave similar results for all the CPUs we tested.
 Figure~\ref{fig:microkernel-all-archs} shows cycle count plots for ``Core'', ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.
@@
 We were able to verify that partial address overlap was occurring at the same time this event is spiking, concluding that address aliasing must be the underlying explanation.
-Still, this only explains part of what is going on in Figure~\ref{fig:microkernel-all-archs}, as there seem to be yet another bias effect at every fourth environment increment.
+Still, this only explains part of what is shown in Figure~\ref{fig:microkernel-all-archs}, as there seems to be yet another bias effect at every fourth environment increment.
 These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
 %Our assumption is that this is a different phenomenon, not related to 4K aliasing.
-We were only able to observe this effect on this particular architecture, and did not investigate further.% is related to 4K aliasing.
+We were only able to observe this effect on this particular architecture, and did not investigate it further.% is related to 4K aliasing.
 
 Results for the ``Nehalem'' and ``Ivy Bridge'' architectures are more uniform, with large spikes only at the aliasing environment configuration.
-It is clear that this issue exist for a wide range of Intel microarchitectures.
+It is clear that this issue exists for a wide range of Intel microarchitectures.
 %We believe that address alias issue exist for a wider range of Intel microarchitectures.
@@
   \item The \emph{heap} is marked by a break point representing the end of uninitialized data in virtual memory, and more space is requested by the \texttt{brk} or \texttt{sbrk} system calls.
-  \item The \texttt{mmap} system call is used to map file descriptors to virtual memory. \emph{Anonymous} mappings, i.e. buffers not backed by a file, can be used for general purpose allocations.
+  \item The \texttt{mmap} system call is used to map file descriptors to virtual memory.
+\emph{Anonymous} mappings, \emph{i.e.} buffers not backed by a file, can be used for general purpose allocations.
 \end{itemize}
-The allocator included as part of glibc uses both mechanisms, choosing which based on the \emph{size} of the request~\cite{GlibcManual}.
-Small sizes generally live in the heap, while larger allocations tend to be memory mapped.
-
-The heap section starts at a relatively low address right above static code and data, while memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack (see Figure \ref{fig:virtualmemory}).
-Whether a request is served by the heap or by memory mapping is therefore easy to determine just by looking at the pointer values returned:
-Addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, while pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
+The allocator included as part of glibc uses both mechanisms, choosing which based on the size of the request~\cite{GlibcManual}.
+Small sizes are generally allocated on the heap, while larger allocations tend to be memory mapped.
+
+The heap section starts at a low address right above static code and data, while memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack (see Figure \ref{fig:virtualmemory}).
+Whether a request is served by the heap or by memory mapping can be determined by looking at the pointer values returned:
+%Addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, while pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
+Addresses in the regular heap are several orders of magnitude smaller than pointers returned by \texttt{mmap}.
 This distinction is unimportant for application developers, as everything is conceptually the same ``heap''.
@@
 
-Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers.\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}\footnote{Memory mapped addresses starting with the 0x2aa...
+Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers.\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.} \footnote{Memory mapped addresses starting with the 0x2aa...
 prefix is an artifact of using \emph{make} to generate results.
@@
 All of the above focus heavily on performance in multithreaded environments.
-Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
+Heap allocation is inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
 Here, we only look at behavior for a single thread, and whether the addresses returned alias or not.
@@
 Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
-Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.
-But even with ASLR, pointers returned by \texttt{mmap} must still be page aligned.
+Given that these results are deterministic (with ASLR disabled), it is possible to construct a program with significant bias towards one or the other allocator.
+Even with ASLR enabled, pointers returned by \texttt{mmap} must still be page aligned.
 This means that addresses returned by allocators using this mechanism directly will \emph{always} alias, giving a deterministic execution context considering only the address suffix.
@@
 \]
-Time spent on a sigle invocation is subtracted from the estimate to mask initialization time.We choose $k=11$, effectively getting the average over 10 iterations.
+Time spent on a single invocation is subtracted from the estimate to mask initialization time.We choose $k=11$, effectively getting the average over 10 iterations.
 In addition, performance counter measurements are averaged over 10 samples using the repeat mechanism of perf.
@@
 An offset of zero is the default behavior for this program when using \texttt{malloc} and moderately large inputs.
-For both for optimization levels 2 and 3, this is close to worst case performance.
+For both optimization levels 2 and 3, this is close to worst case performance.
 The differences in cycles executed is significant, with around $1.7x$ speedup for O2 and as much as $2x$ speedup for O3 on increasing relative offset.
@@
 The numbers are for the level 2 optimization case, which we choose to study in more detail.
-However, the performance data looks fairly similar between the two data sets, with a couple of events that appear to stand out:
+However, the performance data are similar between the two data sets, with a couple of events that stand out:
 \begin{itemize}
@@
 
-Interestingly, it looks like small address offsets incur a massive increase in operations issued on port 0 for the O2 case.
+Interestingly, small address offsets incur a substantial increase in operations issued on port 0 for the O2 case.
 On ``Haswell'', this port handles various ALU operations, and branching together with port 6~\cite[Figure 2.1]{OptimizationManual}.
-As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branches are being re-issued.
+As there are also variations in the number of branch instructions executed, it appears that these counters together show that certain branches are being re-issued.
 Unlike the microkernel in section \ref{sec:microkernel}, the micro-ops executed and RS stall counts drop together with aliasing events.
@@
 
-Overall, it seems reasonable to conclude that the resource stalling is causing the slowdown, ultimately generated by false dependencies from address aliasing.
-We do not attempt to pinpoint exactly which accesses generates these conflicts, but at a high level the CPU falsely assumes dependencies between \texttt{in[i]} and \texttt{out[i]}.
+Overall, we conclude that the resource stalling is causing the slowdown, ultimately generated by false dependencies from address aliasing.
+We do not attempt to pinpoint exactly which access generates these conflicts, but at a high level the CPU falsely assumes dependencies between \texttt{in[i]} and \texttt{out[i]}.
 By manually adjusting the address alignment of one of these buffers, cycle count can be reduced by as much as 50\%.
@@
 %The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
-We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
+We look at three different publicly available libraries; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
 We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
@@
 
-Each of the structures \texttt{A}, \texttt{x} and \texttt{y} represent different locations in memory, and their addresses is part of the execution context.
+Each of the structures \texttt{A}, \texttt{x} and \texttt{y} represent different locations in memory, and their addresses are part of the execution context.
 We use \texttt{mmap} with manually added offsets to control the location of each of these buffers on the heap.
@@
 We do not go into a more detailed analysis of the raw performance data here, but note that we see many of the previously discussed events stand out; such as execution port occupancy, resource stalls, branch instructions executed, offcore requests, and ldm pending.
-Especially for the libblas case, results look quite similar to what we saw for the convolution kernel.
-
-Our results show that address aliasing can indeed occur in real, performance critical applications.
+Especially for the libblas case, results are similar to what we saw for the convolution kernel.
+
+Our results show that address aliasing can indeed occur in real, compute bound applications.
 The amount of \emph{bias} this creates with respect to number of cycles executed is not as clear in all cases, but at least for libblas we are confident in claiming that aliasing is indeed causing a performance degradation.
 
-\section{Ways to Deal with Address Aliasing}
+\section{Address Aliasing Mitigation Techniques}
 \label{sec:mitigation}
 Performance penalties caused by address aliasing can be significant, creating bias towards certain memory layouts.
-However, with an understanding of the underlying mechanism that causes bias, the effects can be predicted and to some extent eliminated in software.
+However, with an understanding of the underlying mechanism that causes bias, the effects can be predicted, and to some extent eliminated in software.
 %The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
@@
 In our convolution kernel implementation, the compiler has to account for the fact that input and output pointers might alias, or that the buffers partially overlap.
-This limits the extent generated code can keep data in registers without updating the values from memory, as a write to one buffer potentially could invalidate a cached value from the other.
-The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer does not alias with any other, allowing for more efficient code generation with fewer memory accesses.
+This limits the extent to which generated code can keep data in registers without updating the values from memory, as a write to one buffer potentially could invalidate a cached value from the other.
+The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer do not alias with any other, allowing for more efficient code generation with fewer memory accesses.
 % \begin{lstlisting}[breaklines=true]

### 2016-01-10 paper.tex
Some minor language improvements.

@@
 Calling the function recursively will effectively allocate a new set of variables further down the stack, and the alias condition is avoided.
-Even though this particular solution is impractical, it proves that compilers and programmers \emph{could} take measures to account for aliasing.
+Even though this particular solution is impractical, it proves that compilers and programmers can take measures to account for aliasing.
 
@@
 Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
-Dynamically allocated data is controlled by heap allocators, which we introduce as another source of bias.
+Dynamically allocated data is controlled by heap allocators, which we find is another source of bias.
 Comparing four different libraries, we show that most implementations tend to favor page alignment for larger requests.

### 2016-01-10 paper.tex
emph på et al.

@@
   \end{lstlisting}
-  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz et al.\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
+  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz \emph{et al.}\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
 \end{figure}
@@
 \label{sec:related}
-Measurement bias and observer effect in performance analysis has been studied in detail by Mytkowicz et al.~\cite{Mytkowicz:2008:OE&MB,Mytkowicz:2008:Easy}.
+Measurement bias and observer effect in performance analysis has been studied in detail by Mytkowicz \emph{et al.}~\cite{Mytkowicz:2008:OE&MB,Mytkowicz:2008:Easy}.
 The authors introduce environment variables and link ordering as examples of external bias triggers, showing how standardized {\small SPECint} benchmarks suffer from measurement bias when trying to evaluate the effectiveness of O3 over O2.
@@
 Many different approaches have been explored in order to take performance counter data and processor intrinsics into account for optimization.
-Knights et al. introduce ``blind'' optimization, treating the underlying hardware as a black box and use search over \emph{variant spaces}~\cite{Knights:2009:BlindOpt}.
+Knights \emph{et al.} introduce ``blind'' optimization, treating the underlying hardware as a black box and use search over \emph{variant spaces}~\cite{Knights:2009:BlindOpt}.
 Their space of variants consist of different position and alignment for each function and global variable, in principle exploring the different configurations possible by altering link order.
@@
 Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
-Dynamically allocated data is controlled by heap allocators, which we find is another source of bias.
+Dynamically allocated data is controlled by heap allocators, which we show is another source of bias.
 Comparing four different libraries, we show that most implementations tend to favor page alignment for larger requests.

### 2016-01-10 paper.tex
Rephrase some sentences

@@
 We demonstrate ways to deal with this type of bias, through runtime detection and correction of aliasing conditions.
-This is especially important for heap allocations, by default most allocators tend to produce worst case alignment, favoring page alignment.
-%For heap allocators, we show that common implementations tend to give worst case alignment by default, favoring page alignment for large allocations.
-The performance impact of unfavorable memory layouts can be significant, causing up to a $2x$ performance degradation in already optimized code.
+%This is especially important for heap allocations, by default most allocators tend to produce worst case alignment, favoring page alignment.
+For heap allocators, we show that common implementations tend to give worst case alignment by default; favoring page alignment for large allocations.
+The performance impact of aliased memory accesses can be significant, causing up to a $2x$ performance degradation in already optimized code.
 %Finally, we show that even highly optimized BLAS libraries can be impacted by address aliasing.
@@
 %This example provides the basis for our analysis.
-We show how some instances of measurement bias can be explained by \emph{address aliasing}, an artifact of a known Intel specific optimization on the out-of-order execution pipeline.
+We show how some instances of measurement bias can be explained by \emph{address aliasing}, an artifact of a presumably Intel specific optimization on the out-of-order execution pipeline.
 The CPU uses a heuristic for determining whether loads are dependent on previous stores, comparing only the 12 least significant virtual address bits.
@@
 Changing environment variables will offset the location of the stack, and consequently all stack allocated variables.
-After this offset, the stack is realigned to a 16 byte boundary on x86_64 Linux systems.
-Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context with respect to address aliasing. 
+%After this offset, the stack is realigned to a 16 byte boundary on x86_64 Linux systems.
+Alignment is restricted to 16 byte boundaries accoring to the ABI, creating 256 possible initial addresses for every period of 4096 bytes.
+Each one representing a different execution context with respect to address aliasing.
 %Note that there is no clear relationship between environment size and stack location with ASLR enabled.
@@
 \]
-Time spent on a single invocation is subtracted from the estimate to mask initialization time.We choose $k=11$, effectively getting the average over 10 iterations.
+Time spent on a single invocation is subtracted from the estimate to mask initialization time.
+We choose $k=11$, effectively getting the average over 10 iterations.
 In addition, performance counter measurements are averaged over 10 samples using the repeat mechanism of perf.

### 2016-01-10 paper.tex
Revert stack alignment sentence

@@
 Changing environment variables will offset the location of the stack, and consequently all stack allocated variables.
-%After this offset, the stack is realigned to a 16 byte boundary on x86_64 Linux systems.
-Alignment is restricted to 16 byte boundaries accoring to the ABI, creating 256 possible initial addresses for every period of 4096 bytes.
-Each one representing a different execution context with respect to address aliasing.
+After this offset, the stack is aligned to a 16 byte boundary on x86_64 Linux systems.
+Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context with respect to address aliasing.
+%Alignment is restricted to 16 byte boundaries accoring to the ABI, creating 256 possible initial addresses for every period of 4096 bytes.
+%Each one representing a different execution context with respect to address aliasing.
 %Note that there is no clear relationship between environment size and stack location with ASLR enabled.

### 2016-02-28 paper.tex
Changed from the anonymous to the public version.

@@
 {
-%    pdfauthor={L. K. Melhus and R. E. Jensen},
-    pdfauthor={Anonymous and Anonymous},
+    pdfauthor={L. K. Melhus and R. E. Jensen},
+%    pdfauthor={Anonymous and Anonymous},
     pdfsubject={Identification of an underlying mechanism that causes measurement bias},
@@
 \section{Acknowledgments}
-\iffalse
+\iftrue
 This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
-\fi
+\else
 This work is ... under the ..., .
@@
 %\end{acks}
+\fi
 

### 2016-02-28 paper.tex
Address trivial spelling mistakes pointed out in review

@@
 In order to understand biased behavior it is important to be able to make precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
-This can be achieved by using performance counters; instrumentation support in hardware that can be used to count various events, such as cycles executed, branch misses, instructions fetched, and so on.
+This can be achieved by using performance counters; instrumentation support in hardware that can be used to count various events, such as cycles executed, branch prediction misses, instructions fetched, and so on.
 Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.
@@
 We use the perf-stat command in all experiments, which accepts raw PMU event codes listed in the reference manual~\cite{Volume3B}.
-Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
+Reliability of PMU measurements has been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
 The issues they find are either handled explicitly or are irrelevant to our setup.
@@
 % Paper is rename of "Toward Accurate Performance Evaluation using Hardware Counters".
-Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
+Controlled variations in environment size are achieved by setting a dummy environment variable to $n$ number of zero characters.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
 Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
@@
 Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
-The following data and analysis of this program is from the ``Haswell'' configuration in Table~\ref{tab:machines}.
+The following data and analysis of this program are from the ``Haswell'' configuration in Table~\ref{tab:machines}.
 Similar experiments were conducted on all other experimental setups listed, arriving at the same conclusion.
@@
 The cycles\_ldm\_pending and stalls\_l2\_pending numbers are interesting, as they are related to the memory and cache system.
-However, this program use only 2 cache-lines, except for the initial startup.
-Using the L1 hit and miss counts also revealed that the L1 load is 99.3\% and store hit rate is ~100\%.
+However, this program uses only 2 cache-lines, except for the initial startup.
+Using the L1 hit and miss counts also revealed that the L1 load hit rate is 99.3\% and store hit rate is {\raise.17ex\hbox{$\scriptstyle\mathtt{\sim}$}}100\%.
 These events do not indicate a cache hit rate issue, but are probably due to having to wait for aliased writes to be retired and committed to cache.
@@
 
-Cycles executed and alias events for each library is shown in Figure~\ref{fig:blas}.
+Cycles executed and alias events for each library are shown in Figure~\ref{fig:blas}.
 Each library behaves differently, but in all cases we identify a significant amount of aliasing. 

### 2016-02-28 paper.tex
Remove dead markup from old formatting

@@
 % paper.tex
-% ACM sigmetric format
 % Author: L. K. Melhus and R. E. Jensen
-%\documentclass{acm_proc_article-sp}
 \documentclass[10pt, conference, compsocconf]{IEEEtran}
@@
 \usepackage{hyperref} % For PDF metadata
-%\usepackage{enumitem} % Controlling list styling
-%http://tex.stackexchange.com/questions/94016/how-to-reduce-space-between-image-and-its-caption
 \usepackage[font=small]{caption}
-
-%\setlength{\intextsep}{10pt plus 2pt minus 2pt}
-
-%\setlength{\belowcaptionskip}{-5pt}
-% too much?
-
-%\captionsetup[table]{aboveskip=-25pt}
-
-%\setlength{\dblfloatsep}{-15pt}
-
-%\addtolength{\belowcaptionskip}{15pt}
 
@@
 \lstset{language=C, basicstyle=\ttfamily\small, columns=flexible, belowskip=0pt}
-
-%\setlist{topsep=0pt}
 
@@
     pdfauthor={L. K. Melhus and R. E. Jensen},
-%    pdfauthor={Anonymous and Anonymous},
     pdfsubject={Identification of an underlying mechanism that causes measurement bias},
@@
 
-%\conferenceinfo{SIGMETRICS}{2015 Portland, Oregon USA}
-%\CopyrightYear{2015}
-%\acmVolume{V}
-%\acmNumber{N}
-%\acmArticle{A}
-%\acmYear{YYYY}
-%\acmMonth{0}
-
-% Page headers
-%\markboth{L. K. Melhus and R. E. Jensen}{Measurement Bias from Address Aliasing}
-
 \title{Measurement Bias from Address Aliasing}
 
-%\numberofauthors{2}
-
-\iffalse
-% names here
-\author{
-% You can go ahead and credit any number of authors here,
-% e.g. one 'row of three' or two rows (consisting of one row of three
-% and a second row of one, two or three).
-%
-% The command \alignauthor (no curly braces needed) should
-% precede each author name, affiliation/snail-mail address and
-% e-mail address. Additionally, tag each line of
-% affiliation/address with \affaddr, and tag the
-% e-mail address with \email.
-%
-% 1st. author
-
-\alignauthor
-Lars Kirkholt Melhus\\
-       \affaddr{Norwegian University of Science and Technology}\\
-       \affaddr{Sem Sælands vei 9}\\
-       \affaddr{7491 Trondheim, Norway}\\
-       \email{lars.kirkholt.melhus@gmail.com}
-% 2nd. author
-\alignauthor
-Rune Erlend Jensen\\
-       \affaddr{Norwegian University of Science and Technology}\\
-       \affaddr{Sem Sælands vei 9}\\
-       \affaddr{7491 Trondheim, Norway}\\
-       \email{runeerle@idi.ntnu.no}
-
-\alignauthor
-Anonymous\\
-       \affaddr{Anonymous University}\\
-       \affaddr{Anonymous}\\
-       \affaddr{Anonymous, Anonymous}\\
-       \email{anonymous@anonymous.com}
-% 2nd. author
-\alignauthor
-Anonymous\\
-       \affaddr{Anonymous University }\\
-       \affaddr{Anonymous}\\
-       \affaddr{Anonymous, Anonymous}\\
-       \email{anonymous@anonymous.com}
-}
-\fi
-
-\iftrue
 \author{\IEEEauthorblockN{Lars Kirkholt Melhus}
@@
 }
-\else
-\author{\IEEEauthorblockN{Anonymous}
-\IEEEauthorblockA{Organization\\
-University\\
-anonymous@mail.com}
-\and
-\IEEEauthorblockN{Anonymous}
-\IEEEauthorblockA{Organization\\
-University\\
-anonymous@mail.com}
-}
-
-\fi
 
@@
 
-% Choose from classification tree http://www.acm.org/about/class/ccs98-html
-%\category{C.4}{Performance of Systems}{Measurement techniques; Performance attributes}
-%\category{D.3.4}{Processors}{Optimization; Run-time environments}
-
-% Terms come from a fixed list of predefined terms
-%\terms{Measurement, Performance}
 % Free to choose keywords, but must be in alphabetical order
@@
 \end{IEEEkeywords}
-
-% Minimum is author, year and title
-%\acmformat{Lars K. Melhus and Rune E. Jensen. 2014. Measurement Bias from Address Aliasing}
-
-% Acknowledgement of support, author(s) address(es)
-%\begin{bottomstuff}
-%New Paper, Not an Extension of a Conference Paper.
-%Authors' addresses: Lars Kirkholt Melhus, Rune Erlend Jensen, Computer and Information Science Department, Norwegian University of Science and Technology, 
-%Sem S{\ae}lands vei 9, Gl{\o}shaugen, 7491 Trondheim, Norway; email: larskirk@alumni.ntnu.no, runeerle@idi.ntnu.no
-%\end{bottomstuff}
 
@@
 
-%\newpage
 \section{Aliasing in Real Applications}
@@
 
-\iffalse
-The functionality is divided into three categories:
-\begin{description}
-  \item[Level 1] \hspace{0.1cm} Scalar and vector operations, such as dot product and vector addition.
-  \item[Level 2] \hspace{0.1cm} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
-  \item[Level 3] \hspace{0.1cm} Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
-\end{description}
-Procedures operating on vectors, intuitively appears the most likely to have potential for aliasing.
-\fi
 %The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
@@
 
-%\newpage
 \section{Conclusions}
@@
 
-% Acknowledgements
-%\begin{acks}
 \section{Acknowledgments}
-\iftrue
 This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
-\else
-This work is ... under the ..., .
- under the supervision of ..., looking at sources of measurement bias ... ~\cite{MasterThesis}.
-%\end{acks}
-\fi
 
@@
 
-% History dates
-%\received{}{}{}
-
 \end{document}

### 2016-03-01 paper.tex
Change formatting elements to conform to template
Remove unused elements from header. Also change section names to singular
Conclusion and Acknowledgement, to mimic bare_conf example.

@@
 
-\hypersetup
-{
-    pdfauthor={L. K. Melhus and R. E. Jensen},
-    pdfsubject={Identification of an underlying mechanism that causes measurement bias},
-    pdftitle={Measurement Bias from Address Aliasing},
-    pdfkeywords={4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}
-}
-
 \title{Measurement Bias from Address Aliasing}
@@
 }
-
-\date{10 January 2016}
 
@@
 \begin{IEEEkeywords}
-4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters
+4K address aliasing; BLAS; heap allocators; measurement bias; memory disambiguation; performance counters;
 \end{IEEEkeywords}
@@
 
-\section{Conclusions}
+\section{Conclusion}
 \label{sec:conclusions}
@@
 
-\section{Acknowledgments}
+% According to template, this is supposed to be \section*
+\section*{Acknowledgment}
 This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.

### 2016-03-03 paper.tex
Disable bookmarks
Not allowed by pdf validation.
Change bibliography style
Reformat references.bib to resemble IEEE example. Disable clickable hyperlinks
and cross-references, as they are discouraged for publication.
Display full name of university and use ntnu email

@@
 \usepackage{underscore}
-\usepackage{url}
+\usepackage{url} % Better line breaking of urls
 \usepackage{lmodern} % Much clearer text
 \usepackage[utf8]{inputenc}
-\usepackage{hyperref} % For PDF metadata
 \usepackage[font=small]{caption}
@@
 \author{\IEEEauthorblockN{Lars Kirkholt Melhus}
-\IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
-NTNU, Norway\\
-lars.kirkholt.melhus@gmail.com}
+\IEEEauthorblockA{Dept. of Computer and Information Science\\
+Norwegian University of Science and Technology\\
+Trondheim, Norway\\
+larskirk@alumni.ntnu.no}
 \and
 \IEEEauthorblockN{Rune Erlend Jensen}
-\IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
-NTNU, Norway\\
+\IEEEauthorblockA{Dept. of Computer and Information Science\\
+Norwegian University of Science and Technology\\
+Trondheim, Norway\\
 runeerle@idi.ntnu.no}
@@
 % Bibliography
-\bibliographystyle{abbrv}
+\bibliographystyle{IEEEtran}
 \bibliography{references}

### 2016-03-05 paper.tex
Fix wording after proofreading

@@
 In this paper, we identify \emph{address aliasing} as an important underlying mechanism causing measurement bias on modern Intel microarchitectures.
-By analyzing hardware performance counters, we show how bias arises from two external factors; size of environment variables, and characteristics of dynamically linked heap allocators.
+By analyzing hardware performance counters, we show how bias arises from two external factors: size of environment variables, and characteristics of dynamically linked heap allocators.
 %Our approach is to use a large set of hardware performance counter measurements to reveal the inner workings of the CPU, and search for correlations over a series of execution contexts.
@@
 Seemingly harmless external factors---such as link ordering or contents of environment variables---have been shown to have significant impact on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
-The underlying explanation is that these factors affect memory addresses of code or data at runtime, which in turn interacts with various low level hardware buffers.
-Because of the complexity of modern processors, it is often hard or impossible to predict exactly how changes to things like environment variables ultimately affect program execution.
+The underlying explanation is that these factors affect memory addresses of data at runtime, which in turn interacts with various low level hardware buffers.
+Because of the complexity of modern processors, it is often hard, or even impossible to predict exactly how changes to things like environment variables ultimately affect program execution.
 
@@
 Aliased memory accesses can produce false dependencies, and incur performance penalties.
-We look at how address aliasing can trigger bias from two external factors; size of system environment variables and configuration of dynamic heap allocation libraries.
+We look at how address aliasing can trigger bias from two external factors: size of system environment variables, and configuration of dynamic heap allocation libraries.
 
-Our results show that the cost of aliasing can be significant, exemplified by a simple program with up to $2x$ speedup based in heap address alignment alone.
+Our results show that the cost of aliasing can be significant, exemplified by a simple program with up to $2x$ speedup based on heap address alignment alone.
 %We also find that many heap allocation libraries tend to produce worst case behavior by default with respect to aliasing, by favoring page alignment for large allocations.
@@
 Section~\ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
-Section~\ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section~\ref{sec:heap} we go through a similar analysis for bias and heap address alignment.
+Section~\ref{sec:environment} presents an analysis of how the size of environment variables causes bias, and in Section~\ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
 Section~\ref{sec:blas} looks at alias in BLAS libraries.
@@
 \label{sec:methodology}
-In order to understand biased behavior it is important to be able to make precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
+In order to understand biased behavior, it is important to be able to make precise and detailed measurements without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
 This can be achieved by using performance counters; instrumentation support in hardware that can be used to count various events, such as cycles executed, branch prediction misses, instructions fetched, and so on.
-Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.
+Recent Intel architectures have several hundred available events, providing a detailed view of what is happening inside the CPU.
 Performance counters are supported in the Linux kernel, and can be accessed via a tool called \emph{perf}.
@@
 Reliability of PMU measurements has been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
-The issues they find are either handled explicitly or are irrelevant to our setup.
+The issues they find are either handled explicitly, or are irrelevant to our setup.
 %They find intra-machine variations of up to 1.07\%,
 
-To identify issues we use a small Python script to iterate over an exhaustive set of all known counters, which amounts to about 200 on our architecture.
+A small Python script is used to iterate over an exhaustive set of all known counters, which amounts to about 200 in our setups.
 Only a small set of events are collected at a time, to ensure events are actually counted continuously.
@@
 % Paper is rename of "Toward Accurate Performance Evaluation using Hardware Counters".
-Controlled variations in environment size are achieved by setting a dummy environment variable to $n$ number of zero characters.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
+Controlled variations in environment size are achieved by setting a dummy environment variable.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
+% to $n$ number of zero characters.
 Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
@@
 The following data and analysis of this program are from the ``Haswell'' configuration in Table~\ref{tab:machines}.
-Similar experiments were conducted on all other experimental setups listed, arriving at the same conclusion.
+%Similar experiments were conducted on all other experimental setups listed, arriving at the same conclusion.
 
@@
 %The same experiment was run again, showing that the modified program had the exact same bias to environment size.
-From the output at each spike, we found the addresses to be \texttt{\&g} = 0x7fffffffe038, \texttt{\&inc} = 0x7fffffffe03c, and \texttt{\&g} = 0x7fffffffd038, \texttt{\&inc} = 0x7fffffffd03c, respectively
+From the output at each spike, we found the addresses to be \texttt{\&g} = 0x7fffffffe038, \texttt{\&inc} = 0x7fffffffe03c, and \texttt{\&g} = 0x7fffffffd038, \texttt{\&inc} = 0x7fffffffd03c, respectively.
 Notice the common suffix 0x03c between \texttt{inc} and \texttt{i}, which is the aliasing pair.
@@
 \subsection{Other Architectures}
-While we show only the analysis for ``Haswell'', the same approach gave similar results for all the CPUs we tested.
+While we show the analysis only for ``Haswell'', the same approach gave similar results for all the CPUs we tested.
 Figure~\ref{fig:microkernel-all-archs} shows cycle count plots for ``Core'', ``Nehalem'' and ``Ivy Bridge'', revealing the same alias behavior.
@@
 The heap section starts at a low address right above static code and data, while memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack (see Figure \ref{fig:virtualmemory}).
-Whether a request is served by the heap or by memory mapping can be determined by looking at the pointer values returned:
+Whether a request is served by the heap or by memory mapping can be determined by looking at the pointer values returned;
 %Addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, while pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
-Addresses in the regular heap are several orders of magnitude smaller than pointers returned by \texttt{mmap}.
+addresses in the regular heap are several orders of magnitude smaller than pointers returned by \texttt{mmap}.
 This distinction is unimportant for application developers, as everything is conceptually the same ``heap''.
@@
 The page size is usually 4096 bytes, meaning addresses returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
-This behavior is often the worst case for functions that operate on two or more independent buffers.
+%This behavior is often the worst case for functions that operate on two or more independent buffers.
 
@@
 This difference is not important to our discussion.}
-We observe the addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} for different size parameters.
+We observe the addresses of two equally sized buffers allocated with \texttt{malloc}, for different size parameters.
 %Equal three digit address suffix indicate an aliasing pair.
@@
 Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
-Given that these results are deterministic (with ASLR disabled), it is possible to construct a program with significant bias towards one or the other allocator.
+Given that these results are deterministic (with ASLR disabled), it is possible to construct programs with significant bias towards one or the other allocator.
 Even with ASLR enabled, pointers returned by \texttt{mmap} must still be page aligned.
 This means that addresses returned by allocators using this mechanism directly will \emph{always} alias, giving a deterministic execution context considering only the address suffix.
-From our limited analysis, this seems to be the case for glibc, jemalloc and Hoard.
+%From our limited analysis, this seems to be the case for glibc, jemalloc and Hoard.
 
@@
 To analyze performance for different addresses, we manually insert padding to offset the start address of one buffer.
-This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
+%This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
 Controlling the offset parameter, we can create environments where the address suffixes are offset by any desired number of bytes.
@@
 %\end{lstlisting}
-An estimate of the actual cost a single invocation can be calculated by the following formula:
+An estimate of the actual cost of a single invocation can be calculated by the following formula:
 %An estimate of the actual cost a single invocation can be calculated by averaging over $ k $ repeated function calls, after subtracting time $ t_{1} $ which includes allocation overhead.
@@
 Each of the structures \texttt{A}, \texttt{x} and \texttt{y} represent different locations in memory, and their addresses are part of the execution context.
-We use \texttt{mmap} with manually added offsets to control the location of each of these buffers on the heap.
+We use \texttt{mmap} with manually added offsets to control the location of each of these buffers.
 From our experiments, we found that the memory address offset of vector $\boldsymbol{y}$ compared to matrix $A$ is the most important with respect to the number of alias events.
@@
 Cycles executed and alias events for each library are shown in Figure~\ref{fig:blas}.
-Each library behaves differently, but in all cases we identify a significant amount of aliasing. 
+Each implemetation behaves differently, but in all cases we identify a significant amount of aliasing.
 The diagrams can be understood the same way as for the convolution example; partial address overlap occur when the address suffix delta of $A$ and $\boldsymbol{y}$ is close to zero.
@@
 
-We do not go into a more detailed analysis of the raw performance data here, but note that we see many of the previously discussed events stand out; such as execution port occupancy, resource stalls, branch instructions executed, offcore requests, and ldm pending.
+We do not go into a more detailed analysis of the raw performance data here, but note that we see many of the previously discussed events stand out, such as execution port occupancy, resource stalls, branch instructions executed, offcore requests, and ldm pending.
 Especially for the libblas case, results are similar to what we saw for the convolution kernel.
 
-Our results show that address aliasing can indeed occur in real, compute bound applications.
+Our results show that address aliasing can indeed occur in real, and already highly optimized applications.
 The amount of \emph{bias} this creates with respect to number of cycles executed is not as clear in all cases, but at least for libblas we are confident in claiming that aliasing is indeed causing a performance degradation.
@@
 In our convolution kernel implementation, the compiler has to account for the fact that input and output pointers might alias, or that the buffers partially overlap.
-This limits the extent to which generated code can keep data in registers without updating the values from memory, as a write to one buffer potentially could invalidate a cached value from the other.
+This limits the extent to which generated code can keep data in registers without updating the values from memory, as a write to one buffer could potentially invalidate a cached value from the other.
 The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer do not alias with any other, allowing for more efficient code generation with fewer memory accesses.
@@
 \subsection{Use a special purpose allocator}
-The Intel optimization manual mentions this% in \emph{User/Source Coding Rule 8}
-, suggesting that special purpose allocators could be used to avoid aliasing~\cite{OptimizationManual}.
+The Intel optimization manual suggests that a special purpose memory allocator could be used to avoid aliasing~\cite{OptimizationManual}. % in User/Source Coding Rule 8
 However, to our knowledge there are no allocators that specifically try to mitigate aliasing in heap allocated memory.
@@
 %We show that heap allocators are prone to generate pairwise aliasing buffers, the worst case for many functions.
-
-One approach could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
+One approach to achieve this could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
 The allocator could also accept hints to which buffers are going to be accessed in parallel, for example as an optional argument to a custom version of \texttt{malloc}.
@@
 %The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
-Speculative and out-of-order memory operations handling only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
+False conflicts occur from speculatively executing memory operations out of order, when considering only the last 12 address bits to resolve conflicts between a load and previous store.
 This effect is observed on Intel platforms, ranging from ``Core'' to ``Haswell'' microarchitectures.
 
-In general, any change to how data is laid out in memory layout can potentially introduce bias effects from address aliasing.
+In general, any change to how data is laid out in memory can potentially introduce bias effects from address aliasing.
 Analyzing an example with bias to environment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
@@
 \section*{Acknowledgment}
-This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
+This work is forked from a Master's thesis project supervised by Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
 

### 2016-03-06 paper.tex
Adjust formatting to fit better

@@
 static int i, j, k;
+
 int main() {
@@
 By default, even with address randomization enabled, both \texttt{input} and \texttt{output} will have start addresses with the same address suffix of 0x010.
-To analyze performance for different addresses, we manually insert padding to offset the start address of one buffer.
-%This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
-Controlling the offset parameter, we can create environments where the address suffixes are offset by any desired number of bytes.
+To analyze performance for different addresses, we manually insert padding to control the address of one of the arrays.
+This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset the pointer past the aligned baseline.
+Controlling the offset parameter, we can create environments where the address suffixes differ by any desired number of bytes.
 Allocating and managing these buffers at startup takes a non-negligible amount of work, but the overhead can be masked by repeatedly invoking the convolution kernel in a loop.
@@
     ((((long)&a) & 0xfff) == (((long)&b) & 0xfff))
+
 static int i, j, k;
+
 int main() {

### 2016-03-06 analysis/heap-alias/conv.c

@@ -1,9 +1,11 @@
-static float k[5]= {0.1, 0.25, 0.3, 0.25, 0.1};
+static float k[5] = {0.1, 0.25, 0.3, 0.25, 0.1};
+
 void conv(int n, const float *in, float *out) {
     int i, j;
     for (i = 2; i < n - 2; ++i) {
         out[i] = 0;
-        for (j = 0; j < 5; ++j)
+        for (j = 0; j < 5; ++j) {
             out[i] += in[i-2+j] * k[j];
+        }
     }
 }

### 2016-03-07 paper.tex
Center all tables
Fixed some spelling errors.

@@
 \begin{table}
+  \centering
   \caption{Experimental setups used.\label{tab:machines}}{
@@
 \begin{table}
+  \centering
   \caption{Events with significant correlation to cycle count.\label{tab:loopcorrelation}}
@@
 To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
-Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
+Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance dense linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
 \pgfplotstableread{bin/libblas.dat}{\libblastable}
@@
 For the results presented here, we set $A = I$ (identity matrix) and $\boldsymbol{x} = [0, 1, 2, ..., N-1]$, with parameters $M = N = 1024$.
-Address suffix of $A$ and $\boldsymbol{x}$ is fixed to 0x000, while performance events are sampled for increasing address offset of $\boldsymbol{y}$, starting at a baseline of 0x000.
+Address suffixes of $A$ and $\boldsymbol{x}$ are fixed to 0x000, while performance events are sampled for increasing address offset of $\boldsymbol{y}$, starting at a baseline of 0x000.
 The experiment was run on the ``Haswell'' setup, with Hyper-Threading enabled.
@@
 This limits the extent to which generated code can keep data in registers without updating the values from memory, as a write to one buffer could potentially invalidate a cached value from the other.
-The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer do not alias with any other, allowing for more efficient code generation with fewer memory accesses.
+The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer do not alias with any others, allowing for more efficient code generation with fewer memory accesses.
 % \begin{lstlisting}[breaklines=true]

### 2016-03-07 paper.tex
Added several unused auto-tune formulations.

@@
 
-\subsection{Manually adjust address offsets}
+\subsection{Adjusting address offsets}
 In some cases it might help to explicitly control the memory addresses used, for example by requiring fixed relative offset between input and output pointers.
@@
 Even though this particular solution is impractical, it proves that compilers and programmers can take measures to account for aliasing.
+A complier can generate the same solution automatically by using data access relevant pointer identification
+%While this example uses C code to address the problem it is possible to perform the same directly by a compiler extension.
+%In general we need automated searches for both relevant pointer addresses, offset rules and good offsets.
+%For general code one need to use automated searches for both relevant pointer identification and offset generator rules.
+%By utilizing hardware performance counter information this issue can be checked
 
@@
 We are able to identify instances of alias issues in matrix-vector multiplication routines for different BLAS libraries, showing that this effect also appears in real, performance critical code.
+%We have shown a method of performance search with hardware performance counters that can identify unknown performance issues inside modern CPUs, and how to deal with the issue.
+%This shows that other issues with address sensitive memory interactions may also be detected and addressed automatically.
+%We have shown new possibilities in the area of auto-tuning, both how to detect general issues, and outlined how to deal with a specific one.
 
@@
 \section*{Acknowledgment}
-This work is forked from a Master's thesis project supervised by Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
+This work is derived from a Master's thesis project supervised by Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
 

### 2016-03-07 paper.tex
Expand explanation of mitigation techniques
Add some additional pointers to future work, and adjust formatting to
perfectly fit 10 pages.

@@
         width=\textwidth/2.05, % Make it fit to text width side by side
-        height=6.0cm,
+        height=5.6cm,
         skip coords between index={20}{32} % Limit level of detail to fit page width nicely
@@
         width=\textwidth/2.05,
-        height=6.0cm,
+        height=5.6cm,
         skip coords between index={20}{32}
@@
 %The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.
-We identify some relevant optimization techniques:
+In addition to controlling bias in experimental setups, avoiding alias issues is also necessary for achieving optimal performance.
+This opens up new avenues to explore, for example in making more intelligent architecture aware compiler optimizations, or use of automatic performance tuning to avoid suboptimal memory layouts.
+Some principles are also more directly applicable, and we identify the following relevant optimization techniques:
 
@@
 Calling the function recursively will effectively allocate a new set of variables further down the stack, and the alias condition is avoided.
-Even though this particular solution is impractical, it proves that compilers and programmers can take measures to account for aliasing.
-A complier can generate the same solution automatically by using data access relevant pointer identification
+
+Even though this particular solution is impractical, it proves the possibility for compilers, as well as programmers, to take measures to account for address aliasing.
+% A complier can generate the same solution automatically by using data access relevant pointer identification
 %While this example uses C code to address the problem it is possible to perform the same directly by a compiler extension.
@@
 Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
+
 Dynamically allocated data is controlled by heap allocators, which we show is another source of bias.
@@
 %Address aliasing can have significant performance impact for algorithms operating on heap allocated buffers, which
+
 We analyze an example with up to $2x$ variation in cycle count between different heap alignments, showing that address aliasing can have a significant performance impact.

### 2016-03-12 presentation.tex
Add skeleton for workshop presentation

\documentclass{beamer}

\title{Measurement Bias from Address Aliasing}
\subtitle{How your user name affects runtime performance}

\author[Lars Kirkholt Melhus, Rune Erlend Jensen]
{Lars Kirkholt Melhus \and Rune Erlend Jensen}
\institute{
  Dept. of Computer and Information Science\\
  Norwegian University of Science and Technology \\
  Trondheim, Norway
}

\date[2016] % (optional)
{International Workshop on Automatic Performance Tuning, 2016}

\subject{Computer Science}

% Allocated to each presentation: 30 mins. Probably 20 mins + questions,
% a couple of minutes per slide should mean around 10 slides at the most.

\begin{document}
  \frame{\titlepage}

  \begin{frame}
    \frametitle{Measurement Bias}

    % Show example code and plot from Mytkowitz.

    % Also an optimization problem (relevant for auto-tuning).

  \end{frame}

  \begin{frame}
    \frametitle{4K Address Aliasing}

    % Image from Master's (redo in Tikz probably) showing exact address
    % collisions. Update with addresses taken from paper.

    % Maybe show disassembly with read/write false conflicts.

  \end{frame}

  \begin{frame}
    \frametitle{How to Measure Aliasing}
    \framesubtitle{Using Hardware Performance Counters}

    % Some points on hardware perf counters.

    % Describe LD_BLOCKS_PARTIAL.ADDRESS_ALIAS
  \end{frame}

  \begin{frame}
    \frametitle{Cycle Correlation}

    % Correlation data for the microkernel (?)

  \end{frame}

  % After this, done with method for detecting alias, and answers question
  % from old paper. Now lets see what else can be discovered.

  \begin{frame}
    \frametitle{Bias in Heap Allocation}

    % Motivation for looking: Determines addresses.

    % Short survay, most use mmap for large allocations.

    % punchline: mmap always alias.

  \end{frame}

  \begin{frame}
    \frametitle{Optimization Resistant Code}

    % Convolution example code

    % Plots for O2 and O3

    % Manual offset code (?)

  \end{frame}

  \begin{frame}
    \frametitle{Aliasing in BLAS libraries}

    % Maybe mention FFTW results from Master's?

  \end{frame}

\end{document}

### 2016-05-15 presentation.tex
Write presentation

@@
 \documentclass{beamer}
+\usepackage{graphicx}
+\usepackage{listings}
+\usepackage{pgfplots}
+\usepackage{pgfplotstable}
+\usepackage{mathtools}
+\usepackage{tikz}
+
+\usetikzlibrary{patterns,shapes,positioning,calc}
+\setbeamertemplate{bibliography item}[text]
+\lstset{language=C, basicstyle=\ttfamily\scriptsize, columns=flexible, belowskip=0pt}
 
 \title{Measurement Bias from Address Aliasing}
-\subtitle{How your user name affects runtime performance}
 
@@
 \date[2016] % (optional)
-{International Workshop on Automatic Performance Tuning, 2016}
+{International Workshop on Automatic Performance Tuning, Chicago, May 2016}
 
@@
 \begin{document}
+
 \frame{\titlepage}
@@
 \begin{frame}
+
+\frametitle{Outline}
+
+\begin{itemize}
+  \item Identify address aliasing as an underlying explanation for measurement bias.
+  \item See how aliasing affects dynamic memory allocation.
+  \item Look at opportunities to optimize for aliasing.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile, allowframebreaks]
 \frametitle{Measurement Bias}
@@
 % Show example code and plot from Mytkowitz.
+
+\begin{itemize}
+  \item Runtime performance change from external factors
+  \begin{itemize}
+    \item Environment variables
+    \item Link ordering
+  \end{itemize}
+  \item Challenge for performance analysis
+\end{itemize}
+
+\begin{figure}
+  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth, aboveskip=-2pt]
+static int i, j, k;
+
+int main() {
+    int g = 0, inc = 1;
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    return 0;
+}
+  \end{lstlisting}
+  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz \emph{et al.}\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
+\end{figure}
+
+% Also an optimization problem (relevant for auto-tuning).
+
+Execute program under different environments, setting dummy variable to $K$ zero characters.
+
+\begin{lstlisting}[breaklines=true, belowskip=-10pt, basicstyle=\ttfamily\small]
+$ env -i X=`head -c K </dev/zero | tr '\0' '0'` ./loop
+\end{lstlisting}
+
+\pgfplotstableread{bin/microkernel-cycles-haswell.dat}{\stackoffsettable}
+\begin{figure}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title = Haswell,
+        width = \textwidth,
+        height = 4.7cm,
+        font = \footnotesize,
+        xlabel=Bytes added to environment,
+        ylabel=Cycles,
+        domain = 0:8192,
+        xtick = {0,1024,...,8192},
+        xmin = 0,
+        xmax = 8192,
+        cycle list name = exotic
+      ]
+      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
+    \end{axis}
+  \end{tikzpicture}
+  \caption{Measured cycle count samples for 512 different environments.}
+\end{figure}
 
@@
 
-    % Image from Master's (redo in Tikz probably) showing exact address
-    % collisions. Update with addresses taken from paper.
+\begin{columns}[T] % align columns
+\begin{column}{.50\textwidth}
 
-    % Maybe show disassembly with read/write false conflicts.
+\begin{itemize}
+  %\item Collisions between static and automatic variables, depending on alignment of stack.
+  \item Addresses of static i, j and k are fixed for all environments.
+  \item Addresses of g and inc are pushed down with size of environment variables.
+  \item Spike when address of inc is \emph{aliasing} with i in the last 12 bits.
+  \item Performance impact from \emph{false dependencies} in CPU out of order execution.
+\end{itemize}
+
+\end{column}
+\hfill
+\begin{column}{.50\textwidth}
+
+% Virtual memory/randomization illustration
+\begin{figure}
+  \begin{tikzpicture}[font=\footnotesize]
+    % See page 453 in pgf manual
+    \node [
+      rectangle split, rectangle split parts=5, 
+      rectangle split part fill={white, white, lightgray, white, lightgray},
+      draw, anchor=center, text width=2.5cm
+    ] (m) {
+      \nodepart{one}
+        \scriptsize{USER=lars;SHELL
+        =/bin/bash;X=0000
+        00000000000000...}
+      \nodepart{two}
+        .stack
+        \begin{flushright}
+        inc \\
+        g
+        \end{flushright}
+      \nodepart{four}
+        .bss
+        \begin{flushright}
+        k \\
+        j \\
+        i
+        \end{flushright}
+    };
+
+    \node[right, yshift=0cm] at (m.two east) {
+      \scriptsize{0x7fff'ffffe{\color{red}03c}} % inc
+    };
+    \node[right, yshift=-0.4cm] at (m.two east) {
+      \scriptsize{0x7fff'ffffe038} % g
+    };
+
+    \node[right, yshift=0cm] at (m.four east) {
+      \scriptsize{0x601044} % k
+    };
+    \node[right, yshift=-0.4cm] at (m.four east) {
+      \scriptsize{0x601040} % j
+    };
+    \node[right, yshift=-0.8cm] at (m.four east) {
+      \scriptsize{0x601{\color{red}03c}} % i
+    };
+
+    %\draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.seven split west) 
+    %  node [black, midway, xshift=-1.5cm, text width=2.0cm] 
+    %    { Program code and static data } ;
+
+    %\node[right] at (m.north east) { \scriptsize{0x7fff'ffffffff} } ;
+    %\node[right] at (m.south east) { \scriptsize{0x400000} } ;
+  \end{tikzpicture}
+  \caption{Addresses of each variable in the first spike.}
+\end{figure}
+
+% Note: Could have gotten more collisions if g also was colliding with something,
+%       but not noticeable effect on cycles executed.
+
+\end{column}
+\end{columns}
 
@@
 
-  \begin{frame}
-    \frametitle{How to Measure Aliasing}
-    \framesubtitle{Using Hardware Performance Counters}
+\begin{frame}[fragile]
+  \frametitle{How to Measure Address Aliasing}
 
   % Some points on hardware perf counters.
+  \begin{itemize}
+    \item Control execution environment (disable ASLR).
+    \item Instrument program using \emph{hardware performance counters}, sampling the following event:
+  \end{itemize}
+
+  \begin{description}
+    \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] 
+    \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} 
+    \cite[B.3.4.4]{OptimizationManual}
+  \end{description}
+  %This event is listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}.
+  %Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
+
+  \begin{itemize}
+    \item Older architectures, such as Core, have alias covered in a more general event:
+  \end{itemize}
+
+  \begin{description}
+    \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] 
+    \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} 
+    \cite[Table 19-17]{Volume3B}
+  \end{description}
 
   % Describe LD_BLOCKS_PARTIAL.ADDRESS_ALIAS
-  \end{frame}
-
-  \begin{frame}
-    \frametitle{Cycle Correlation}
-
-    % Correlation data for the microkernel (?)
-
 \end{frame}
@@
 
-  \begin{frame}
+\begin{frame}[fragile]
 \frametitle{Bias in Heap Allocation}
+
+\begin{itemize}
+  \item Alias conflicts can happen between any sections of memory.
+  \item While stack addresses depend on environment, heap addresses depend on allocators.
+\end{itemize}
+
+\begin{table}
+  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers. Same 12 bit suffix indicate an aliasing pair.}
+  \pgfplotstabletypeset[
+    font=\footnotesize,
+    col sep=comma,
+    string type,
+    columns={Allocation, 5120, 1048576},
+    column type=r,
+    columns/Allocation/.style={
+      string type, 
+      column type=l,
+      column type/.add={|}{}
+    },
+    every head row/.style={
+      output empty row,
+      before row={\hline
+         & 5,120 B & 1,048,576 B \\
+        },
+      after row=\hline\hline,
+    },
+    every last row/.style={after row=\hline},
+    every last column/.style={column type/.add={}{|}},
+    every odd row/.style={after row=\hline},
+  ]{bin/malloc-comparison.csv}
+\end{table}
 
@@
 
-  \begin{frame}
+\begin{frame}[fragile]
+  \frametitle{mmap}
+
+Most allocators tend to use anonymous memory mapping for large requests.
+
+\begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth, belowskip=1cm]
+$ man mmap
+
+(...) on Linux, the mapping will be created at a nearby page boundary.
+The address of the new mapping is returned as the result of the call.
+
+\end{lstlisting}
+
+Since page size is 4 KiB, such allocations will \emph{always} alias with each other.
+
+\end{frame}
+
+\begin{frame}[fragile, allowframebreaks]
   \frametitle{Optimization Resistant Code}
 
+% Image of sliding window with buffers in and out (Color variables?)
+
+Example program which is sensitive to aliasing between buffers, continuously generating false dependencies:
+
+\begin{tikzpicture}[
+  barstyle/.style={rectangle,fill=lightgray,draw,anchor=center,minimum width=9cm}
+]
+\node [barstyle] (in) at (0,1) { } ;
+\node [barstyle] (out) at (0,0) { } ;
+\node [rectangle,draw,minimum height=1.8cm] (bar) at (-2,0.5) { } ;
+
+\node[] (input label) [left=of in] { in } ;
+\node[] (output label) [left=of out] { out } ;
+
+\draw[->] (-2,0.5) -- (-1.5,0.5) node [right,align=center] {\scriptsize{Sliding window}} ;
+
+\end{tikzpicture}
+
+\begin{figure}[t]
 % Convolution example code
+  \lstinputlisting[
+    language=C,
+    frame=single,
+    xleftmargin=.00\textwidth,
+    xrightmargin=.00\textwidth,
+    aboveskip=-5pt
+  ]{bin/convolution-kernel.c}
+  \caption{Simplified implementation of convolution with a fixed kernel.}
+  \label{lst:conv}
+\end{figure}
 
-    % Plots for O2 and O3
+% \begin{itemize}
+%   \item Most heap allocators will \emph{always} align arrays on the same 12 bit suffix.
+%   \item Manually adjusting addresses necessary to get good performance.
+% \end{itemize}
 
-    % Manual offset code (?)
+Adjust address of output array by \texttt{d} bytes manually to avoid alising:
+
+%mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
+\begin{lstlisting}[belowskip=-4pt, basicstyle=\ttfamily\small]
+out = malloc((N + d) * sizeof(float));
+conv(N, in, out + d);
+\end{lstlisting}
+
+% The most awesome graph, O2 of convolution kernel
+\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
+\pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
+\begin{figure}[t]
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=Haswell -O2,
+        font=\footnotesize,
+        cycle list name=black white,
+        width=\textwidth/1.8,
+        height=4.7cm,
+        skip coords between index={18}{32} % Limit level of detail to fit page width nicely
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=Haswell -O3,
+        font=\footnotesize,
+        cycle list name=black white,
+        width=\textwidth/1.8,
+        height=4.7cm,
+        skip coords between index={18}{32}
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethreehaswell ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethreehaswell ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel, input size $N=2^{20}$.}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Aliasing in BLAS libraries}
+
+% Maybe mention FFTW results from Master's?
+
+Measured matrix-vector multiplication (cblas\_dgemv) with increasing offset between buffer addresses.
+Computing $\boldsymbol{y}=A\boldsymbol{x}$.
+
+\pgfplotstableread{bin/libblas.dat}{\libblastable}
+\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
+\begin{figure}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=libblas,
+        height=4.7cm,
+        font=\footnotesize,
+        xlabel=Address offset of $\boldsymbol{y}$,
+        %ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/1.8, % Make it fit to text width side by side
+        legend style={at={(0.4,0.55)},anchor=west,draw=none}
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=libatlas,
+        height=4.7cm,
+        font=\footnotesize,
+        xlabel=Address offset of $\boldsymbol{y}$,
+        %ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/1.8,
+        legend style={at={(0.4,0.55)},anchor=west,draw=none}
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+  \caption{\label{fig:blas}Cycles executed and address alias events for varying relative address offset between matrix $A$ and vector $\boldsymbol{y}$.}
+  % Each sample point represents an increment of 16 bytes, or 0x10.}
+\end{figure}
 
@@
 \begin{frame}
-    \frametitle{Aliasing in BLAS libraries}
+\frametitle{Optimizing for address aliasing}
 
-    % Maybe mention FFTW results from Master's?
+% Effects of address aliasing are real, these are some ideas for optimization:
+
+\begin{itemize}
+  \item Mark buffers with \texttt{restrict}.
+  \begin{itemize}
+    \item Can allow the optimizer to reduce number of memory accesses.
+  \end{itemize}
+  \item Use a special purpose allocator.
+  \begin{itemize}
+    \item Heuristics for avoiding pairwise aliasing buffers.
+    \item No known implementation with this goal in mind, to our knowledge.
+  \end{itemize}
+  \item Manually adjust address offset.
+  \begin{itemize}
+    \item Consider memory address alignment for performance tuning inner loops.
+  \end{itemize}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+
+\frametitle{References}
+
+\scriptsize{\bibliographystyle{IEEEtran}}
+\bibliography{references}
 

### 2016-05-15 presentation.tex
wip: Write first half of presentation

--- 2016-03-12 presentation.tex
@@
 \documentclass{beamer}
+\usepackage{graphicx}
+\usepackage{listings}
+\usepackage{pgfplots}
+\usepackage{pgfplotstable}
+\usepackage{mathtools}
+\usepackage{tikz}
+
+\usetikzlibrary{patterns,shapes,positioning,calc}
+\setbeamertemplate{bibliography item}[text]
+\lstset{language=C, basicstyle=\ttfamily\scriptsize, columns=flexible, belowskip=0pt, aboveskip=-1pt}
 
 \title{Measurement Bias from Address Aliasing}
-\subtitle{How your user name affects runtime performance}
 
@@
 \begin{frame}
+
+\frametitle{Outline}
+
+\begin{itemize}
+  \item What is measurement bias?
+  \item What is address aliasing?
+  \item How does this help?
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile, allowframebreaks]
 \frametitle{Measurement Bias}
@@
 % Show example code and plot from Mytkowitz.
+
+\begin{itemize}
+  \item Runtime performance change from external factors
+  \begin{itemize}
+    \item Environment variables
+    \item Link ordering
+  \end{itemize}
+  \item Challenge for performance analysis
+\end{itemize}
+
+\begin{figure}
+  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
+static int i, j, k;
+
+int main() {
+    int g = 0, inc = 1;
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    return 0;
+}
+  \end{lstlisting}
+  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz \emph{et al.}\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
+\end{figure}
+
+% Also an optimization problem (relevant for auto-tuning).
+
+What is the execution time of this function?
+
+\pgfplotstableread{bin/microkernel-cycles-haswell.dat}{\stackoffsettable}
+\begin{figure}
+  \begin{tikzpicture}
+    \begin{axis}[
+        width = \textwidth,
+        height = 5.0cm,
+        font = \footnotesize,
+        xlabel=Bytes added to environment,
+        ylabel=Cycles,
+        domain = 0:8192,
+        xtick = {0,1024,...,8192},
+        xmin = 0,
+        xmax = 8192,
+        cycle list name = exotic
+      ]
+      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
+    \end{axis}
+  \end{tikzpicture}
+  \caption{Measured average of 10 cycle count samples for 512 different environments; offseting stack by multiples of 16 bytes.}
+\end{figure}
 
@@
 
-    % Image from Master's (redo in Tikz probably) showing exact address
-    % collisions. Update with addresses taken from paper.
+\begin{columns}[T] % align columns
+\begin{column}{.48\textwidth}
 
-    % Maybe show disassembly with read/write false conflicts.
+\begin{itemize}
+  %\item Collisions between static and automatic variables, depending on alignment of stack.
+  \item Address of i, j and k are fixed for all environments.
+  \item Address of g and inc is pushed down with length of environment variables.
+  \item Spike when inc is at 0x7fffffffe03c + $ k \times $ 0x1000, \emph{aliasing} with i in the last 12 bits.
+  \item Performance impact from \emph{false dependencies} in CPU out of order execution.
+  \item Effect known as \emph{address aliasing}.
+\end{itemize}
+
+\end{column}
+\hfill
+\begin{column}{.48\textwidth}
+
+% Virtual memory/randomization illustration
+\begin{figure}
+  \begin{tikzpicture}[font=\footnotesize]
+    % See page 453 in pgf manual
+    \node [
+      rectangle split, rectangle split parts=5, 
+      rectangle split part fill={white, white, lightgray, white, lightgray},
+      draw, anchor=center, text width=2.5cm
+    ] (m) {
+        \nodepart{one}
+          \scriptsize{USER=lars;SHELL
+          =/bin/bash;PWD=
+          /home (...)}
+        \nodepart{two}
+          .stack
+          \begin{flushright}
+          inc \\
+          g
+          \end{flushright}
+        \nodepart{four}
+          .bss
+          \begin{flushright}
+          k \\
+          j \\
+          i
+          \end{flushright}
+      };
+
+      \node[right, yshift=0cm] at (m.two east) {
+        \scriptsize{0x7fff'ffffe{\color{red}03c}} % inc
+      };
+      \node[right, yshift=-0.4cm] at (m.two east) {
+        \scriptsize{0x7fff'ffffe038} % g
+      };
+
+      \node[right, yshift=0cm] at (m.four east) {
+        \scriptsize{0x601044} % k
+      };
+      \node[right, yshift=-0.4cm] at (m.four east) {
+        \scriptsize{0x601040} % j
+      };
+      \node[right, yshift=-0.8cm] at (m.four east) {
+        \scriptsize{0x601{\color{red}03c}} % i
+      };
+
+    %\draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.seven split west) 
+    %  node [black, midway, xshift=-1.5cm, text width=2.0cm] 
+    %    { Program code and static data } ;
+
+    %\node[right] at (m.north east) { \scriptsize{0x7fff'ffffffff} } ;
+    %\node[right] at (m.south east) { \scriptsize{0x400000} } ;
+  \end{tikzpicture}
+  \caption{4K address aliasing between stack and static data.}
+\end{figure}
+
+% Note: Could have gotten more collisions if g also was colliding with something,
+%       but not noticeable effect on cycles executed.
+
+\end{column}
+\end{columns}
 
@@
   % Some points on hardware perf counters.
+  \begin{itemize}
+    \item Using processor-specific instrumentation, we can \emph{count} various low-level events during execution.
+    \item Intel manual lists hundreds of events.
+  \end{itemize}
+
+  \begin{description}
+    \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] 
+    \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} 
+    \cite[B.3.4.4]{OptimizationManual}
+  \end{description}
+  %This event is listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}.
+  %Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
+  \begin{description}
+    \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] 
+    \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} 
+    \cite[Table 19-17]{Volume3B}
+  \end{description}
 
@@
 
+\begin{frame}
+
+\frametitle{References}
+
+\scriptsize{\bibliographystyle{IEEEtran}}
+\bibliography{references}
+
+\end{frame}
+
 \end{document}

### 2016-05-15 presentation.tex
Start on a picture of sliding window

@@
 \setbeamertemplate{bibliography item}[text]
-\lstset{language=C, basicstyle=\ttfamily\scriptsize, columns=flexible, belowskip=0pt, aboveskip=-1pt}
+\lstset{language=C, basicstyle=\ttfamily\scriptsize, columns=flexible, belowskip=0pt, aboveskip=0pt}
 
@@
 
+% More generic on methodology? Not necessary for presentation maybe..
+% Just to straight to the juice.
 \begin{frame}
-  \frametitle{How to Measure Aliasing}
+  \frametitle{How to Measure Address Aliasing}
   \framesubtitle{Using Hardware Performance Counters}
@@
   \begin{itemize}
+    \item Control execution environment (disable ASLR).
     \item Using processor-specific instrumentation, we can \emph{count} various low-level events during execution.
-    \item Intel manual lists hundreds of events.
+      \begin{itemize}
+        \item Cycles executed, branch prediction misses, instructions fetched, etc.
       \end{itemize}
+    %\item Intel manual lists hundreds of events.
+  \end{itemize}
+
+  Events covering 4K aliasing:
 
@@
 
-  \begin{frame}
-    \frametitle{Cycle Correlation}
-
-    % Correlation data for the microkernel (?)
-
-  \end{frame}
-
   % After this, done with method for detecting alias, and answers question
@@
     \frametitle{Bias in Heap Allocation}
+
+    \begin{itemize}
+      \item Alias conflicts can happen between any sections of memory.
+      \item While stack addresses depend on environment, heap addresses depend on allocators.
+    \end{itemize}
+
+    Looked at several allocators:
+
+    \begin{itemize}
+      \item ptmalloc (glibc)
+      \item tcmalloc (Google)
+      \item jemalloc
+      \item Hoard
+    \end{itemize}
 
@@
 
-  \begin{frame}
+\begin{frame}[fragile]
+  \frametitle{mmap}
+
+    Most allocators use anonymous memory mapping for large allocations.
+
+\begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
+$ man mmap
+
+(...) on Linux, the mapping will be created at a nearby page boundary.
+The address of the new mapping is returned as the result of the call.
+
+\end{lstlisting}
+
+\end{frame}
+
+\begin{frame}[fragile]
   \frametitle{Optimization Resistant Code}
 
+% Image of sliding window with buffers in and out (Color variables?)
+
+\begin{tikzpicture}[
+  barstyle/.style={rectangle,fill=lightgray,draw,anchor=center,minimum width=9cm}
+]
+\node [barstyle] (in) at (0,1) { } ;
+\node [barstyle] (out) at (0,0) { } ;
+\node [rectangle,draw,minimum height=1.8cm] (bar) at (-2,0.5) { } ;
+
+\node[] (input label) [left=of in] { in } ;
+\node[] (output label) [left=of out] { out } ;
+
+\draw[->] (-2,0.5) -- (-1.5,0.5) node [right,align=center] {\scriptsize{Sliding window}} ;
+
+\end{tikzpicture}
+
 % Convolution example code
+\begin{figure}[t]
+  \lstinputlisting[
+    language=C,
+    frame=single,
+    xleftmargin=.00\textwidth,
+    xrightmargin=.00\textwidth
+  ]{bin/convolution-kernel.c}
+  \caption{Simplified implementation of convolution with a fixed kernel. This program is highly sensitive to aliasing between input and output arrays, continuously generating false dependencies.}
+  \label{lst:conv}
+\end{figure}
 

### 2016-05-15 presentation.tex
Do some more shit

@@
 \setbeamertemplate{bibliography item}[text]
-\lstset{language=C, basicstyle=\ttfamily\scriptsize, columns=flexible, belowskip=0pt, aboveskip=0pt}
+\lstset{language=C, basicstyle=\ttfamily\scriptsize, columns=flexible, belowskip=0pt}
 
@@
 \begin{figure}
-  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
+  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth, aboveskip=-2pt]
 static int i, j, k;
@@
 
-\begin{frame}[fragile]
+\begin{frame}[fragile, allowframebreaks]
   \frametitle{Optimization Resistant Code}
@@
 
-    % Plots for O2 and O3
+% \begin{itemize}
+%   \item Most heap allocators will \emph{always} align arrays on the same 12 bit suffix.
+%   \item Manually adjusting addresses necessary to get good performance.
+% \end{itemize}
+
+Adjust address of output array by \texttt{d} bytes manually to avoid alising:
+
+%mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
+\begin{lstlisting}[belowskip=-4pt]
+out = malloc((N + d) * sizeof(float));
+...
+conv(N, in, out + d);
+\end{lstlisting}
+
+% The most awesome graph, O2 of convolution kernel
+\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
+\begin{figure}[t]
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=Haswell -O2,
+        font=\footnotesize,
+        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
+        ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/1.5,
+        height=4.7cm,
+        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel, input size $n=2^{20}$.}
+  %Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
+\end{figure}
 

### 2016-05-15 presentation.tex
Fill in shim of remaining slides

@@
 \begin{itemize}
-  \item What is measurement bias?
-  \item What is address aliasing?
-  \item How does this help?
+  \item Identify address aliasing as underlying explanation for measurement bias.
+  % \begin{itemize}
+  %   \item Looking at example from previous work
+  % \end{itemize}
+  \item See how aliasing affects dynamic memory allocation.
+  \item Identify opportunities to optimize for aliasing.
 \end{itemize}
@@
 
-What is the execution time of this function?
+Execute program under different environments, setting dummy variable to $K$ zero characters.
+
+\begin{lstlisting}[breaklines=true, belowskip=-10pt]
+$ env -i X=`head -c K </dev/zero | tr '\0' '0'` ./loop
+\end{lstlisting}
 
@@
 \begin{columns}[T] % align columns
-\begin{column}{.48\textwidth}
+\begin{column}{.50\textwidth}
 
@@
   %\item Collisions between static and automatic variables, depending on alignment of stack.
-  \item Address of i, j and k are fixed for all environments.
-  \item Address of g and inc is pushed down with length of environment variables.
-  \item Spike when inc is at 0x7fffffffe03c + $ k \times $ 0x1000, \emph{aliasing} with i in the last 12 bits.
+  \item Addresses of static i, j and k are fixed for all environments.
+  \item Addresses of g and inc are pushed down with size of environment variables.
+  \item Spike when address of inc \emph{aliasing} with i in the last 12 bits.
   \item Performance impact from \emph{false dependencies} in CPU out of order execution.
-  \item Effect known as \emph{address aliasing}.
 \end{itemize}
@@
 \hfill
-\begin{column}{.48\textwidth}
+\begin{column}{.50\textwidth}
 
@@
         \scriptsize{USER=lars;SHELL
-        =/bin/bash;PWD=
-        /home (...)}
+        =/bin/bash;X=0000
+        00000000000000...}
       \nodepart{two}
@@
   \end{tikzpicture}
-  \caption{4K address aliasing between stack and static data.}
+  \caption{Addresses of each variable in the first spike.}
 \end{figure}
@@
 
-  \begin{frame}
+\begin{frame}[fragile]
 \frametitle{Bias in Heap Allocation}
@@
 
-    Looked at several allocators:
-
-    \begin{itemize}
-      \item ptmalloc (glibc)
-      \item tcmalloc (Google)
-      \item jemalloc
-      \item Hoard
-    \end{itemize}
+\begin{table}
+  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers. Same 12 bit suffix indicate an aliasing pair.}
+  \pgfplotstabletypeset[
+    font=\footnotesize,
+    col sep=comma,
+    string type,
+    columns={Allocation, 5120, 1048576},
+    column type=r,
+    columns/Allocation/.style={
+      string type, 
+      column type=l,
+      column type/.add={|}{}
+    },
+    every head row/.style={
+      output empty row,
+      before row={\hline
+         & 5,120 B & 1,048,576 B \\
+        },
+      after row=\hline\hline,
+    },
+    every last row/.style={after row=\hline},
+    every last column/.style={column type/.add={}{|}},
+    every odd row/.style={after row=\hline},
+  ]{bin/malloc-comparison.csv}
+\end{table}
 
@@
 
-\begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
+\begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth, belowskip=1cm]
 $ man mmap
@@
 
+Since page size is usually 4096 byte, such allocations will \emph{always} alias with each other.
+
 \end{frame}
@@
 % Image of sliding window with buffers in and out (Color variables?)
+
+Example program which is sensitive to aliasing between buffers, continuously generating false dependencies:
 
@@
 
+\begin{figure}[t]
 % Convolution example code
-\begin{figure}[t]
   \lstinputlisting[
@@
     xleftmargin=.00\textwidth,
-    xrightmargin=.00\textwidth
+    xrightmargin=.00\textwidth,
+    aboveskip=-5pt
   ]{bin/convolution-kernel.c}
-  \caption{Simplified implementation of convolution with a fixed kernel. This program is highly sensitive to aliasing between input and output arrays, continuously generating false dependencies.}
+  \caption{Simplified implementation of convolution with a fixed kernel.}
   \label{lst:conv}
@@
   \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel, input size $n=2^{20}$.}
-  %Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
 \end{figure}
 
-    % Manual offset code (?)
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Aliasing in BLAS libraries}
+
+% Maybe mention FFTW results from Master's?
+
+Measured matrix-vector multiplication (cblas\_dgemv) with increasing offset between buffer addresses.
+
+\pgfplotstableread{bin/libblas.dat}{\libblastable}
+\begin{figure}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=libblas,
+        height=4.7cm,
+        font=\footnotesize,
+        xlabel=Address offset of $\boldsymbol{y}$,
+        ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/1.5, % Make it fit to text width side by side
+        legend style={at={(0.4,0.55)},anchor=west,draw=none}
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+\end{figure}
 
@@
 \begin{frame}
-    \frametitle{Aliasing in BLAS libraries}
+\frametitle{Optimizing for address alias}
 
-    % Maybe mention FFTW results from Master's?
+% Effects of address aliasing are real, these are some ideas for optimization:
+
+\begin{itemize}
+  \item Mark buffers with restrict
+  \item Use a special purpose allocator
+  \item Manually adjust address offset
+\end{itemize}
 

### 2016-05-15 presentation.tex
Some progress

@@
 
-Most allocators use anonymous memory mapping for large allocations.
+Most allocators tend to use anonymous memory mapping for large requests.
 
@@
   \end{tikzpicture}
-  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel, input size $n=2^{20}$.}
+  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel, input size $N=2^{20}$.}
 \end{figure}
@@
 Measured matrix-vector multiplication (cblas\_dgemv) with increasing offset between buffer addresses.
+Computing $\boldsymbol{y}=A\boldsymbol{x}$.
 
 \pgfplotstableread{bin/libblas.dat}{\libblastable}
+\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
 \begin{figure}
@@
         xlabel=Address offset of $\boldsymbol{y}$,
-        ylabel=Event count,
+        %ylabel=Event count,
         cycle list name=black white,
-        width=\textwidth/1.5, % Make it fit to text width side by side
+        width=\textwidth/1.8, % Make it fit to text width side by side
         legend style={at={(0.4,0.55)},anchor=west,draw=none}
@@
   \end{tikzpicture}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=libatlas,
+        height=4.7cm,
+        font=\footnotesize,
+        xlabel=Address offset of $\boldsymbol{y}$,
+        %ylabel=Event count,
+        cycle list name=black white,
+        width=\textwidth/1.8,
+        legend style={at={(0.4,0.55)},anchor=west,draw=none}
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+  \caption{\label{fig:blas}Cycles executed and address alias events for varying relative address offset between matrix $A$ and vector $\boldsymbol{y}$.}
+  % Each sample point represents an increment of 16 bytes, or 0x10.}
 \end{figure}
@@
 \begin{itemize}
-  \item Mark buffers with restrict
-  \item Use a special purpose allocator
-  \item Manually adjust address offset
+  \item Mark buffers with \texttt{restrict}.
+  \item Use a special purpose allocator.
+  \item Manually adjust address offset.
 \end{itemize}

### 2016-05-15 presentation.tex
Almost something to keep

@@
 \begin{itemize}
-  \item Identify address aliasing as underlying explanation for measurement bias.
-  % \begin{itemize}
-  %   \item Looking at example from previous work
-  % \end{itemize}
+  \item Identify address aliasing as an underlying explanation for measurement bias.
   \item See how aliasing affects dynamic memory allocation.
-  \item Identify opportunities to optimize for aliasing.
+  \item Look at opportunities to optimize for aliasing.
 \end{itemize}
@@
     \begin{axis}[
+        title = Haswell,
         width = \textwidth,
-        height = 5.0cm,
+        height = 4.7cm,
         font = \footnotesize,
@@
   \end{tikzpicture}
-  \caption{Measured average of 10 cycle count samples for 512 different environments; offseting stack by multiples of 16 bytes.}
+  \caption{Measured cycle count samples for 512 different environments.}
 \end{figure}
@@
 
-% More generic on methodology? Not necessary for presentation maybe..
-% Just to straight to the juice.
-\begin{frame}
+\begin{frame}[fragile]
   \frametitle{How to Measure Address Aliasing}
-  \framesubtitle{Using Hardware Performance Counters}
 
@@
     \item Control execution environment (disable ASLR).
-    \item Using processor-specific instrumentation, we can \emph{count} various low-level events during execution.
-      \begin{itemize}
-        \item Cycles executed, branch prediction misses, instructions fetched, etc.
-      \end{itemize}
+    \item Sample alias occurrence using \emph{hardware performance counters}.
+      % \begin{itemize}
+      %   \item Cycles executed, branch prediction misses, instructions fetched, etc.
+      % \end{itemize}
     %\item Intel manual lists hundreds of events.
   \end{itemize}
-
-  Events covering 4K aliasing:
 
@@
     \cite[B.3.4.4]{OptimizationManual}
+    Listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}
   \end{description}
@@
     \cite[Table 19-17]{Volume3B}
+    Covers 4K aliasing for older architectures, such as Core.
   \end{description}
@@
 
-Since page size is usually 4096 byte, such allocations will \emph{always} alias with each other.
+Since page size is 4 KiB, such allocations will \emph{always} alias with each other.
 
@@
 \begin{frame}
-\frametitle{Optimizing for address alias}
+\frametitle{Optimizing for address aliasing}
 
@@
   \item Mark buffers with \texttt{restrict}.
+  \begin{itemize}
+    \item Can allow the optimizer to reduce number of memory accesses.
+  \end{itemize}
   \item Use a special purpose allocator.
+  \begin{itemize}
+    \item Heuristics for avoiding pairwise aliasing buffers.
+    \item No known implementation with this goal in mind, to our knowledge.
+  \end{itemize}
   \item Manually adjust address offset.
+  \begin{itemize}
+    \item Consider memory address alignment for performance tuning inner loops.
+  \end{itemize}
 \end{itemize}

### 2016-05-25 presentation.tex
Improve presentation

@@
 
-\begin{frame}[fragile, allowframebreaks]
+\begin{frame}
 \frametitle{Measurement Bias}
@@
   \item Challenge for performance analysis
+  \item Effects can invalidate perceived speedup\cite{Mytkowicz:2009:WrongData}
 \end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Methodology}
+
+  % Some points on hardware perf counters.
+  \begin{itemize}
+    \item Control execution environment (disable ASLR).
+    \item Instrument program using \emph{hardware performance counters}
+    \item Python script to iterate over environments.
+  \end{itemize}
+
+%todo: Memory layout.
+
+\end{frame}
+
+\begin{frame}[fragile, allowframebreaks]
+\frametitle{Bias from Environment Size}
 
@@
 
-\begin{lstlisting}[breaklines=true, belowskip=-10pt]
+\begin{lstlisting}[breaklines=true, belowskip=-10pt, basicstyle=\ttfamily\small]
 $ env -i X=`head -c K </dev/zero | tr '\0' '0'` ./loop
@@
   \item Addresses of g and inc are pushed down with size of environment variables.
-  \item Spike when address of inc \emph{aliasing} with i in the last 12 bits.
+  \item Spike when address of inc is \emph{aliasing} with i in the last 12 bits.
   \item Performance impact from \emph{false dependencies} in CPU out of order execution.
@@
 
-  % Some points on hardware perf counters.
-  \begin{itemize}
-    \item Control execution environment (disable ASLR).
-    \item Sample alias occurrence using \emph{hardware performance counters}.
-      % \begin{itemize}
-      %   \item Cycles executed, branch prediction misses, instructions fetched, etc.
-      % \end{itemize}
-    %\item Intel manual lists hundreds of events.
-  \end{itemize}
-
   \begin{description}
@@
     \cite[B.3.4.4]{OptimizationManual}
-    Listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}
   \end{description}
@@
   %Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
+
+  \begin{itemize}
+    \item Older architectures, such as Core, have alias covered in a more general event:
+  \end{itemize}
+
   \begin{description}
@@
     \cite[Table 19-17]{Volume3B}
-    Covers 4K aliasing for older architectures, such as Core.
   \end{description}
@@
 %mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
-\begin{lstlisting}[belowskip=-4pt]
+\begin{lstlisting}[belowskip=-4pt, basicstyle=\ttfamily\small]
 out = malloc((N + d) * sizeof(float));
-...
 conv(N, in, out + d);
@@
 \pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
+\pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
 \begin{figure}[t]
@@
         font=\footnotesize,
-        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
-        ylabel=Event count,
-        cycle list name=black white,
-        width=\textwidth/1.5,
+        cycle list name=exotic,
+        width=\textwidth/1.8,
         height=4.7cm,
-        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
+        skip coords between index={18}{32} % Limit level of detail to fit page width nicely
       ]
@@
       \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
+      \addlegendentry{Cycles} ;
+      \addlegendentry{Alias} ;
+    \end{axis}
+  \end{tikzpicture}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title=Haswell -O3,
+        font=\footnotesize,
+        cycle list name=exotic,
+        width=\textwidth/1.8,
+        height=4.7cm,
+        skip coords between index={18}{32}
+      ]
+      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethreehaswell ;
+      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethreehaswell ;
       \addlegendentry{Cycles} ;
@@
         %ylabel=Event count,
-        cycle list name=black white,
+        cycle list name=exotic,
         width=\textwidth/1.8, % Make it fit to text width side by side
@@
         %ylabel=Event count,
-        cycle list name=black white,
+        cycle list name=exotic,
         width=\textwidth/1.8,
@@
     \item Consider memory address alignment for performance tuning inner loops.
+    \item Auto-tuning?
   \end{itemize}

### 2016-05-25 presentation.tex
Add table with counters for microkernel

@@
 \usepackage{pgfplotstable}
+\usepackage{filecontents}
 \usepackage{mathtools}
+\usepackage{underscore}
 \usepackage{tikz}
@@
 \lstset{language=C, basicstyle=\ttfamily\scriptsize, columns=flexible, belowskip=0pt}
+\newcommand{\perfctr}[1] {
+  {\lowercase{#1}}
+}
 
@@
 % Also an optimization problem (relevant for auto-tuning).
+
+\begin{table}
+  %\centering
+  \caption{Events with significant correlation to cycle count.\label{tab:loopcorrelation}}
+  \pgfplotstabletypeset[
+    int detect, % Output whole numbers for counter values
+    col sep=comma,
+    columns={Performance counter, Median, [index]3, [index]4},
+    column type=r,
+    font=\footnotesize,
+    columns/Performance counter/.style={
+      string type, 
+      column type=l,
+      column type/.add={|}{},
+      postproc cell content/.append code={
+        \pgfkeysalso{@cell content=\perfctr{##1}}
+      }
+    },
+    every head row/.style={
+      output empty row,
+      before row={\hline
+        Performance counter & Median & Spike 1 & Spike 2 \\
+      },
+      after row=\hline\hline
+    },
+    every last row/.style={after row=\hline},
+    every last column/.style={column type/.add={}{|}}
+  ]{bin/microkernel-comparison-haswell.csv}
+\end{table}
 

### 2016-05-25 presentation.tex
Clean up counter info

@@
 
-  \begin{itemize}
-    \item Older architectures, such as Core, have alias covered in a more general event:
-  \end{itemize}
+  Same or similar counters found in architectures dating back through Core 2.
 
-  \begin{description}
-    \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] 
-    \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} 
-    \cite[Table 19-17]{Volume3B}
-  \end{description}
+  % \begin{itemize}
+  %   \item Older architectures, such as Core, have alias covered in a more general event:
+  % \end{itemize}
+
+  % \begin{description}
+  %   \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] 
+  %   \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} 
+  %   \cite[Table 19-17]{Volume3B}
+  % \end{description}
 
@@
   \end{itemize}
-  \item Manually adjust address offset.
+  \item Explicitly adjust address offset.
   \begin{itemize}
     \item Consider memory address alignment for performance tuning inner loops.
-    \item Auto-tuning?
+    \item Buffer addresses as candidate for automatic performance tuning
   \end{itemize}

### 2016-05-25 presentation.tex
Add picture of memory address space

@@
   \item Look at opportunities to optimize for aliasing.
+  \item Up to 2x speedup.
 \end{itemize}
@@
 \begin{itemize}
-  \item Runtime performance change from external factors
+  \item Runtime performance change from external factors.
   \begin{itemize}
@@
   \end{itemize}
-  \item Challenge for performance analysis
-  \item Effects can invalidate perceived speedup\cite{Mytkowicz:2009:WrongData}
+  \item Challenge for performance analysis.
+  \item Effects can invalidate perceived speedup~\cite{Mytkowicz:2008:OE&MB}.
 \end{itemize}
@@
 
+\begin{columns}[T] % align columns
+\begin{column}{.40\textwidth}
+
   % Some points on hardware perf counters.
@@
     \item Control execution environment (disable ASLR).
-    \item Instrument program using \emph{hardware performance counters}
-    \item Python script to iterate over environments.
+    \item Instrument program using hardware performance counters.
+    \item Measure over range of environments.
   \end{itemize}
 
-%todo: Memory layout.
+\end{column}
+\begin{column}{.60\textwidth}
+
+\begin{figure}
+  \begin{tikzpicture}[font=\footnotesize]
+    % See page 453 in pgf manual
+    \node [
+      rectangle split, rectangle split parts=10, 
+      rectangle split part fill={white, white, lightgray, white, lightgray, white, lightgray, white},
+      draw, anchor=center, text width=1.8cm
+    ] (m) {
+        \nodepart{one}
+          env. vars
+        \nodepart{two}
+          stack
+        \nodepart{four}
+          mmap area
+        \nodepart{six}
+          heap
+        \nodepart{eight}
+          bss
+        \nodepart{nine}
+          data
+        \nodepart{ten}
+          text
+      };
+    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.seven split west) 
+      node [black, midway, xshift=-1.5cm, text width=2.0cm] 
+        { Program code and static data } ;
+    \node[right] at (m.north east)
+      { \scriptsize{0x7fff'ffffffff} } ;
+    \node[right] at (m.south east)
+      { \scriptsize{0x400000} } ;
+  \end{tikzpicture}
+  \caption{Memory execution context, assuming a 64-bit process running on a Linux system.}
+  \label{fig:virtualmemory}
+\end{figure}
+
+\end{column}
+\end{columns}
 

### 2016-05-25 presentation.tex
Reorder slides
Improve stuff

@@
 \setbeamertemplate{bibliography item}[text]
-\lstset{language=C, basicstyle=\ttfamily\scriptsize, columns=flexible, belowskip=0pt}
+\lstset{language=C, basicstyle=\ttfamily\footnotesize, columns=flexible, belowskip=0pt}
 \newcommand{\perfctr}[1] {
@@
 \date[2016] % (optional)
-{International Workshop on Automatic Performance Tuning, Chicago, May 2016}
+{International Workshop on Automatic Performance Tuning, Chicago, May 27. 2016}
 
@@
 \begin{frame}
-
-\frametitle{Outline}
-
-\begin{itemize}
-  \item Identify address aliasing as an underlying explanation for measurement bias.
-  \item See how aliasing affects dynamic memory allocation.
-  \item Look at opportunities to optimize for aliasing.
-  \item Up to 2x speedup.
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}
 \frametitle{Measurement Bias}
@@
 % Show example code and plot from Mytkowitz.
+
+What if changing your user name affected the execution time of your programs?
+%How changing user name can affect performance of executing programs.
+
+%Results being skewed from \"true\" value due to how measurements are performed.
+\vspace{1cm}
 
@@
   \end{itemize}
-  \item Challenge for performance analysis.
+  \item Challenge for performance analysis. Comparison of algorithms on different system configurations can give misleading results.
   \item Effects can invalidate perceived speedup~\cite{Mytkowicz:2008:OE&MB}.
 \end{itemize}
+
+\end{frame}
+
+\begin{frame}
+
+\frametitle{Outline}
+
+\begin{itemize}
+  \item Identify address aliasing as an underlying explanation for measurement bias.
+  \item How aliasing affects dynamic memory allocation.
+  \item Benchmark results.
+  \item Opportunities to optimize for aliasing.
+  %\item Up to 2x speedup.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile, allowframebreaks]
+\frametitle{Bias from Environment Size}
+
+\begin{figure}
+  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
+static int i, j, k;
+
+int main() {
+    int g = 0, inc = 1;
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    return 0;
+}
+  \end{lstlisting}
+  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz \emph{et al.}\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
+\end{figure}
+
+% Also an optimization problem (relevant for auto-tuning).
+\framebreak
+
+Execute program under different environments, setting dummy variable to $K$ zero characters.
+
+\begin{lstlisting}[breaklines=true, belowskip=-10pt, basicstyle=\ttfamily\small]
+$ env -i X=`head -c K </dev/zero | tr '\0' '0'` ./loop
+\end{lstlisting}
+
+\pgfplotstableread{bin/microkernel-cycles-haswell.dat}{\stackoffsettable}
+\begin{figure}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title = Haswell,
+        width = \textwidth,
+        height = 4.7cm,
+        font = \footnotesize,
+        xlabel=Bytes added to environment,
+        ylabel=Cycles,
+        domain = 0:8192,
+        xtick = {0,1024,...,8192},
+        xmin = 0,
+        xmax = 8192,
+        cycle list name = exotic
+      ]
+      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
+    \end{axis}
+  \end{tikzpicture}
+  \caption{Measured cycle count samples for 512 different environments.}
+\end{figure}
 
@@
 
-\begin{frame}[fragile, allowframebreaks]
-\frametitle{Bias from Environment Size}
-
-\begin{figure}
-  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth, aboveskip=-2pt]
-static int i, j, k;
-
-int main() {
-    int g = 0, inc = 1;
-    for (; g < 65536; g++) {
-        i += inc;
-        j += inc;
-        k += inc;
-    }
-    return 0;
-}
-  \end{lstlisting}
-  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz \emph{et al.}\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
-\end{figure}
-
-% Also an optimization problem (relevant for auto-tuning).
-
-Execute program under different environments, setting dummy variable to $K$ zero characters.
-
-\begin{lstlisting}[breaklines=true, belowskip=-10pt, basicstyle=\ttfamily\small]
-$ env -i X=`head -c K </dev/zero | tr '\0' '0'` ./loop
-\end{lstlisting}
-
-\pgfplotstableread{bin/microkernel-cycles-haswell.dat}{\stackoffsettable}
-\begin{figure}
-  \begin{tikzpicture}
-    \begin{axis}[
-        title = Haswell,
-        width = \textwidth,
-        height = 4.7cm,
-        font = \footnotesize,
-        xlabel=Bytes added to environment,
-        ylabel=Cycles,
-        domain = 0:8192,
-        xtick = {0,1024,...,8192},
-        xmin = 0,
-        xmax = 8192,
-        cycle list name = exotic
-      ]
-      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
-    \end{axis}
-  \end{tikzpicture}
-  \caption{Measured cycle count samples for 512 different environments.}
-\end{figure}
-
-% Also an optimization problem (relevant for auto-tuning).
-
+\begin{frame}[fragile]
+\frametitle{Results}
 \begin{table}
-  %\centering
   \caption{Events with significant correlation to cycle count.\label{tab:loopcorrelation}}
@@
 \end{table}
-
 \end{frame}
@@
 
-(...) on Linux, the mapping will be created at a nearby page boundary.
-The address of the new mapping is returned as the result of the call.
+(...) on Linux, the mapping will be created at a nearby page
+boundary. The address of the new mapping is returned as the
+result of the call.
 
@@
     xrightmargin=.00\textwidth,
-    aboveskip=-5pt
+    aboveskip=-5pt,
+    basicstyle=\ttfamily\scriptsize
   ]{bin/convolution-kernel.c}

### 2016-05-26 presentation.tex
Fix intro slides

@@
 
-What if changing your user name affected the execution time of your programs?
+What if changing your user name affected the execution time of programs?
 %How changing user name can affect performance of executing programs.
@@
   \item Challenge for performance analysis. Comparison of algorithms on different system configurations can give misleading results.
-  \item Effects can invalidate perceived speedup~\cite{Mytkowicz:2008:OE&MB}.
+  \item Effects can invalidate perceived speedup (Mytkowicz et al.~\cite{Mytkowicz:2008:OE&MB}).
 \end{itemize}
@@
 \end{itemize}
-
-\end{frame}
-
-\begin{frame}[fragile, allowframebreaks]
-\frametitle{Bias from Environment Size}
-
-\begin{figure}
-  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
-static int i, j, k;
-
-int main() {
-    int g = 0, inc = 1;
-    for (; g < 65536; g++) {
-        i += inc;
-        j += inc;
-        k += inc;
-    }
-    return 0;
-}
-  \end{lstlisting}
-  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz \emph{et al.}\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
-\end{figure}
-
-% Also an optimization problem (relevant for auto-tuning).
-\framebreak
-
-Execute program under different environments, setting dummy variable to $K$ zero characters.
-
-\begin{lstlisting}[breaklines=true, belowskip=-10pt, basicstyle=\ttfamily\small]
-$ env -i X=`head -c K </dev/zero | tr '\0' '0'` ./loop
-\end{lstlisting}
-
-\pgfplotstableread{bin/microkernel-cycles-haswell.dat}{\stackoffsettable}
-\begin{figure}
-  \begin{tikzpicture}
-    \begin{axis}[
-        title = Haswell,
-        width = \textwidth,
-        height = 4.7cm,
-        font = \footnotesize,
-        xlabel=Bytes added to environment,
-        ylabel=Cycles,
-        domain = 0:8192,
-        xtick = {0,1024,...,8192},
-        xmin = 0,
-        xmax = 8192,
-        cycle list name = exotic
-      ]
-      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
-    \end{axis}
-  \end{tikzpicture}
-  \caption{Measured cycle count samples for 512 different environments.}
-\end{figure}
 
@@
 
-\begin{frame}[fragile]
-\frametitle{Results}
+\begin{frame}[fragile, allowframebreaks]
+\frametitle{Bias from Environment Size}
+
+\begin{figure}
+  \begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
+static int i, j, k;
+
+int main() {
+    int g = 0, inc = 1;
+    for (; g < 65536; g++) {
+        i += inc;
+        j += inc;
+        k += inc;
+    }
+    return 0;
+}
+  \end{lstlisting}
+  \caption{\label{fig:microkernel}Microkernel first presented by Mytkowicz \emph{et al.}\cite{Mytkowicz:2009:WrongData}, showing bias to environment size.}
+\end{figure}
+
+% Also an optimization problem (relevant for auto-tuning).
+\framebreak
+
+Execute program under different environments, setting dummy variable to $K$ zero characters.
+
+\begin{lstlisting}[breaklines=true, belowskip=-10pt, basicstyle=\ttfamily\small]
+$ env -i X=`head -c K </dev/zero | tr '\0' '0'` ./loop
+\end{lstlisting}
+
+\pgfplotstableread{bin/microkernel-cycles-haswell.dat}{\stackoffsettable}
+\begin{figure}
+  \begin{tikzpicture}
+    \begin{axis}[
+        title = Haswell,
+        width = \textwidth,
+        height = 4.7cm,
+        font = \footnotesize,
+        xlabel=Bytes added to environment,
+        ylabel=Cycles,
+        domain = 0:8192,
+        xtick = {0,1024,...,8192},
+        xmin = 0,
+        xmax = 8192,
+        cycle list name=exotic
+      ]
+      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
+    \end{axis}
+  \end{tikzpicture}
+  \caption{Measured cycle count samples for 512 different environments.}
+\end{figure}
+
 \begin{table}

### 2016-05-26 presentation.tex
Move outline back in front
Highlight alias numbers

@@
 \begin{frame}
+
+\frametitle{Outline}
+
+\begin{itemize}
+  \item Background on measurement bias.
+  \item Address aliasing in Intel CPUs.
+  \item How aliasing affects dynamic memory allocation.
+  \item Benchmark results.
+  \item Opportunities to optimize by avoiding aliasing.
+  %\item Up to 2x speedup.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
 \frametitle{Measurement Bias}
@@
   \item Effects can invalidate perceived speedup (Mytkowicz et al.~\cite{Mytkowicz:2008:OE&MB}).
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}
-
-\frametitle{Outline}
-
-\begin{itemize}
-  \item Identify address aliasing as an underlying explanation for measurement bias.
-  \item How aliasing affects dynamic memory allocation.
-  \item Benchmark results.
-  \item Opportunities to optimize for aliasing.
-  %\item Up to 2x speedup.
 \end{itemize}
@@
 
+\pgfplotstableset{
+  highlightrow/.style={
+    postproc cell content/.append code={
+     \count0=\pgfplotstablerow
+      \advance\count0 by1
+      \ifnum\count0=#1
+        \pgfkeysalso{@cell content/.add={$\bf}{$}}
+      \fi
+    }
+  }
+}
+
 \begin{table}
@@
     col sep=comma,
+    highlightrow={1},
     columns={Performance counter, Median, [index]3, [index]4},
@@
 
-  Same or similar counters found in architectures dating back through Core 2.
+  Same or similar counters found in architectures dating back to Core 2.
 

### 2016-05-26 presentation.tex
Add slide with results from other architectures

@@
 \end{table}
+
 \end{frame}
@@
   % Describe LD_BLOCKS_PARTIAL.ADDRESS_ALIAS
+\end{frame}
+
+\begin{frame}[fragile]
+  \pgfplotstableread{bin/microkernel-core.dat}{\tblmicrokernelcore}
+  \pgfplotstableread{bin/microkernel-nehalem.dat}{\tblmicrokernelnehalem}
+  \pgfplotstableread{bin/microkernel-ivybridge.dat}{\tblmicrokernelivybridge}
+  \frametitle{Other architectures}
+  \begin{columns}[T] % align columns
+    \begin{column}{.50\textwidth}
+      \begin{tikzpicture}
+        \begin{axis}[
+            title = Core,
+            width = \textwidth/0.85,
+            height = 4.0cm,
+            font = \footnotesize,
+            domain = 0:4096,
+            xtick = {0,1024,...,4096},
+            xmin = 0,
+            xmax = 4096,
+            cycle list name = exotic
+          ]
+          \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelcore ;
+        \end{axis}
+      \end{tikzpicture}
+      \begin{tikzpicture}
+        \begin{axis}[
+            title = Nehalem (HT),
+            y tick label style={
+            /pgf/number format/.cd,
+                fixed,
+                fixed zerofill,
+                precision=1,
+            /tikz/.cd
+            },
+            width = \textwidth/0.85,
+            height = 4.0cm,
+            font = \footnotesize,
+            domain = 0:4096,
+            xtick = {0,1024,...,4096},
+            xmin = 0,
+            xmax = 4096,
+            cycle list name = exotic
+          ]
+          \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelnehalem ;
+        \end{axis}
+      \end{tikzpicture}
+    \end{column}
+    \begin{column}{.50\textwidth}
+      \begin{tikzpicture}
+        \begin{axis}[
+            title = Ivy Bridge,
+            y tick label style={
+            /pgf/number format/.cd,
+                fixed,
+                fixed zerofill,
+                precision=1,
+            /tikz/.cd
+            },
+            width = \textwidth/0.85,
+            height = 4.0cm,
+            font = \footnotesize,
+            domain = 0:4096,
+            xtick = {0,1024,...,4096},
+            xmin = 0,
+            xmax = 4096,
+            cycle list name = exotic
+          ]
+          \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \tblmicrokernelivybridge ;
+        \end{axis}
+      \end{tikzpicture}
+    \end{column}
+  \end{columns}
 \end{frame}
